%This TeX file is just to store the prompt used for LLM grading. When grading an LLM, one can just copy and paste the required prompts. Each prompt will be separated by dotted lines.

% When copy-pasting, make sure no other additional text is accidentally copied.

%No Markscheme LLM Grading Prompts

% Credit to R Mok et al for Prompt 
% All solutions come from 4o.tex file in LLM solutions

%--------------------------------------------------------------

\subsection{EM}

Electromagnetism Prompts

\subsubsection{EMQ1}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 8 marks. Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem: 

"Name the three types of magnetic materials. Explain what determines their response when brought near a bar magnet."

Mark Scheme:

"{Q1: (8 marks total)}

[1 mark] naming 3 types: paramagnetism, diamagnetism, ferromagnetism. 

[1 mark] mention differing magnetic susceptibility/magnetic dipole moment values which determine response 

[2 marks] paramagnetism, magnetic susceptibility positive, much less than 1, slight attraction to bar magnet (weaker dipole moment relative to ferromagnetic material)

[2 marks] diamagnetism, magnetic susceptibility negative, magnitude much less than 1, slight repulsion to bar magnet, dipole moment induced opposes field (weak dipole moment relative to ferromagnetic, paramagnetic material)

[2 marks] ferromagnetism, magnetic susceptibility much larger than 1, much stronger attraction to bar magnet relative to paramagnet. (large dipole moment relative to para and diamagnetic material)
"

Solution 1: 

"Magnetic materials are classified into three types: diamagnetic, paramagnetic, and ferromagnetic. The response of these materials when brought near a bar magnet depends on the alignment of their atomic magnetic moments and the external magnetic field.

\begin{itemize}
    \item \textbf{Diamagnetic materials:} These materials have no permanent magnetic moment, and when exposed to a magnetic field, they induce a weak, negative magnetization opposing the applied field. This results from the orbital motion of electrons that create small current loops opposing the magnetic field. Examples include copper, gold, and graphite.
    \item \textbf{Paramagnetic materials:} These materials possess weak, positive magnetization. They have unpaired electrons, and when placed in an external magnetic field, their atomic magnetic moments align with the field, reinforcing it. However, this alignment is weak and disappears once the external field is removed. Examples include aluminum, platinum, and certain metal ions.
    \item \textbf{Ferromagnetic materials:} These materials have a strong permanent magnetization due to the parallel alignment of atomic magnetic moments within regions called domains. When placed near a bar magnet, ferromagnetic materials experience a strong attraction and can retain their magnetization even after the external field is removed. Examples include iron, cobalt, and nickel.
\end{itemize}

The behavior of each material is governed by the interaction between the material's atomic magnetic moments and the external magnetic field. Diamagnetic materials oppose the field, paramagnetic materials align weakly with the field, and ferromagnetic materials strongly align and can become magnetized."

Solution 2: 

"Magnetic materials are categorized into diamagnetic, paramagnetic, and ferromagnetic types, based on their response to an external magnetic field.

\begin{itemize}
    \item \textbf{Diamagnetic materials:} These materials do not have any inherent magnetization. When exposed to a magnetic field, they induce a weak, negative magnetization, which opposes the applied field. This behavior is a result of the change in electron orbits induced by the magnetic field. Examples include bismuth and copper.
    \item \textbf{Paramagnetic materials:} These materials have unpaired electrons, and their magnetic moments are weakly aligned in the direction of the applied field. This alignment is temporary and disappears once the field is removed. Examples of paramagnetic materials include aluminum and platinum.
    \item \textbf{Ferromagnetic materials:} In these materials, the atomic magnetic moments are strongly aligned in the same direction within domains. They experience a strong attraction to a magnetic field and can retain their magnetization even in the absence of the external field. Iron, cobalt, and nickel are common examples.
\end{itemize}

The response of these materials to a bar magnet is determined by their magnetic properties. Diamagnetic materials will be repelled, paramagnetic materials will be weakly attracted, and ferromagnetic materials will experience a strong attraction."

Solution 3: 

"The three types of magnetic materials are diamagnetic, paramagnetic, and ferromagnetic. Each type exhibits distinct behaviors when placed in a magnetic field, such as that of a bar magnet.

\begin{itemize}
    \item \textbf{Diamagnetic materials:} These materials are characterized by the absence of a permanent magnetic moment. When subjected to an external magnetic field, they produce a weak magnetic field in the opposite direction, leading to a repulsive force. This occurs because the motion of electrons induces a small current that opposes the applied field. Examples include water and most non-metallic elements.
    \item \textbf{Paramagnetic materials:} These materials have unpaired electrons that create small magnetic moments. When exposed to an external magnetic field, these moments align in the direction of the field, resulting in a weak attraction. However, this effect is temporary, vanishing once the external field is removed. Examples include magnesium and titanium.
    \item \textbf{Ferromagnetic materials:} These materials have permanent magnetic moments due to the parallel alignment of atomic magnetic moments in domains. Upon exposure to a magnetic field, the domains align with the field, producing a strong attraction. Ferromagnetic materials can retain their magnetization even after the field is removed. Examples include iron, nickel, and cobalt.
\end{itemize}

Thus, the material's response to the bar magnet is determined by the nature of its atomic magnetic moments and how they interact with the external field. Diamagnetic materials are repelled, paramagnetic materials are weakly attracted, and ferromagnetic materials experience a strong attraction."

\subsubsection{EMQ2}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 5 marks. Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem: 

"Within the context of classical electromagnetism, briefly define what is 'free space'. In free space the electric field, \underline{E}, satisfies 3 partial differential equations, one for each of its components

\[
\nabla^{2}\underline{E} =  \epsilon_{0}\mu_{0}\frac{\partial^2 \underline{E}}{\partial t^{2}}
\]

Briefly describe the physical interpretation of this equation for showing how electric fields can propagate in space. Explain how this equation gives the speed of light c and state its value." 

Mark Scheme:

"{Q2: (5 marks total)}

[1 mark] free space - vacuum with no matter, free charge or free current.

[1 mark] The equation given defines a wave equation for electric fields. Solutions are plane waves propagating in space

[3 marks] Some mention that light is an EM wave or partly formed by electric field waves. Speed of wave propagation in the equation is equal to speed of light. The speed of light, c,  is equal to \(1/\sqrt{\epsilon_{0}\mu_{0}}\). \(c = 3 \times 10^8 \, ms^{-1}\).

"

Solution 1: 

"\begin{enumerate}
    \item 'Free space' refers to a vacuum or empty space, where there are no charges or currents present. In this context, the electromagnetic field propagates without any material medium or interference from other physical phenomena.
    
    \item The equation given, $\nabla^2 \mathbf{E} = \epsilon_0 \mu_0 \frac{\partial^2 \mathbf{E}}{\partial t^2}$, is the wave equation for the electric field $\mathbf{E}$ in free space. This equation indicates that the electric field behaves as a wave, propagating through space over time. The Laplacian operator $\nabla^2$ represents the spatial variation of the electric field, while the term $\frac{\partial^2 \mathbf{E}}{\partial t^2}$ represents the temporal variation. This shows that changes in the electric field at one point in space can cause the field to propagate outward in the form of a wave, demonstrating that electric fields can travel through space without requiring a material medium.

    \item The equation also reveals that the speed at which the electric field propagates, known as the speed of light $c$, is determined by the constants $\epsilon_0$ (the permittivity of free space) and $\mu_0$ (the permeability of free space). To find the speed of light, we rearrange the equation:
    \[
    c = \frac{1}{\sqrt{\epsilon_0 \mu_0}}.
    \]
    Substituting the known values of $\epsilon_0 = 8.854 \times 10^{-12} \, \text{F/m}$ and $\mu_0 = 4\pi \times 10^{-7} \, \text{N/A}^2$, we find:
    \[
    c \approx 3.00 \times 10^8 \, \text{m/s}.
    \]
\end{enumerate}"

Solution 2:

"In classical electromagnetism, 'free space' refers to a region in space devoid of matter, where the electric field $\mathbf{E}$ and magnetic field $\mathbf{B}$ are not influenced by any charges or currents. This region is characterized by the permittivity $\epsilon_0$ and permeability $\mu_0$, which are constants in vacuum.

The equation $\nabla^2 \mathbf{E} = \epsilon_0 \mu_0 \frac{\partial^2 \mathbf{E}}{\partial t^2}$ represents the wave equation for the electric field in free space. This equation indicates that the electric field propagates as a wave. The term $\nabla^2 \mathbf{E}$ is the spatial second derivative, and $\frac{\partial^2 \mathbf{E}}{\partial t^2}$ is the time second derivative, showing that changes in the electric field over time are related to its spatial changes.

This wave equation implies that electric fields can propagate through space without needing a medium. The speed of propagation is determined by the relationship between $\epsilon_0$ and $\mu_0$. Specifically, the wave speed $v$ is given by:
\[
v = \frac{1}{\sqrt{\epsilon_0 \mu_0}}
\]
In free space, this speed is the speed of light $c$. Using the known values of $\epsilon_0 = 8.854 \times 10^{-12} \, \mathrm{F/m}$ and $\mu_0 = 4\pi \times 10^{-7} \, \mathrm{H/m}$, we calculate:
\[
c = \frac{1}{\sqrt{(8.854 \times 10^{-12})(4\pi \times 10^{-7})}} \approx 3.00 \times 10^8 \, \mathrm{m/s}
\]
Thus, the electric field propagates through free space at the speed of light, which is approximately $3.00 \times 10^8 \, \mathrm{m/s}$."

Solution 3: 

"In classical electromagnetism, 'free space' refers to a region of space that is devoid of any charges or currents, meaning that it is not influenced by any external material properties such as conductivity or magnetization. In free space, the electric field, $\mathbf{E}$, is governed by Maxwell's equations, which describe how electric and magnetic fields propagate and interact with matter.

The equation provided,
\[
\nabla^2 \mathbf{E} = \epsilon_0 \mu_0 \frac{\partial^2 \mathbf{E}}{\partial t^2},
\]
is the wave equation for the electric field in free space. This form implies that the electric field satisfies the conditions for wave propagation. The operator $\nabla^2$ is the Laplacian, representing spatial variation, while the term $\frac{\partial^2 \mathbf{E}}{\partial t^2}$ describes the time variation. The constants $\epsilon_0$ and $\mu_0$ are the permittivity and permeability of free space, respectively.

This equation shows that changes in the electric field propagate through space as waves. The speed at which these waves travel can be derived from the equation. Comparing it to the standard wave equation of the form,
\[
\nabla^2 \psi = \frac{1}{v^2} \frac{\partial^2 \psi}{\partial t^2},
\]
we identify the wave speed $v$ as the speed of light, $c$. Therefore, the speed of light $c$ in free space is given by,
\[
c = \frac{1}{\sqrt{\epsilon_0 \mu_0}}.
\]
Using the known values of $\epsilon_0 = 8.854 \times 10^{-12} \, \text{F/m}$ and $\mu_0 = 4\pi \times 10^{-7} \, \text{N/A}^2$, we find that
\[
c \approx 3 \times 10^8 \, \text{m/s}.
\]
Thus, the electric field propagates as a wave at the speed of light, which is approximately $3 \times 10^8 \, \text{m/s}$."

\subsubsection{EMQ3}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 14 marks. Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Find the electric field a distance, \( d \), above one end of a straight line segment of length \( L \) that has a constant line charge density \( \lambda \) (as seen in the figure). In the limit of \( d \gg L \), what does the electric field reduce to and what is its physical interpretation?"

Mark Scheme:

"{Q3: (14 marks total)}

[1 mark] Use of coulombs law to integrate and find electric field

[3 marks] correct integral setup to evaluate x-component \( E_x \). \( dq = \lambda dx \)

\[ E_x = - \frac{1}{4\pi \epsilon_0} \lambda \int_{0}^{L} \frac{x}{(x^2+d^2)^{3/2}} \, dx \]

[3 marks] correct integral setup to evaluate z-component \( E_z \).

\[ E_z = \frac{1}{4\pi \epsilon_0} \lambda d  \int_{0}^{L} \frac{1}{(x^2+d^2)^{3/2}} \, dx \]

[2 marks] correct evaluation of integral to get \( E_x \)

\[ E_x = -\frac{1}{4 \pi \epsilon_0} \lambda \left [ \frac{1}{d} - \frac{1}{\sqrt{d^2+L^2}} \right ] \]

[2 marks] correct evaluation of integral to get \( E_z \)

\[ \frac{1}{4 \pi \epsilon_0} \frac{\lambda}{d} \frac{L}{\sqrt{d^2+L^2}} \]

[1 mark] total expression of \underline{E}

\[
\frac{1}{4 \pi \epsilon_0} \frac{\lambda}{d} \left [ \left ( -1+\frac{d}{\sqrt{d^2+L^2}} \right ) \hat{x} +\left ( \frac{L}{\sqrt{d^2+L^2}} \right ) \hat{z} \right ]
\]

[2 marks] Correct simplification in \underline{E} for \( d \gg L \)

\[ \frac{1}{4 \pi \epsilon_0} \frac{\lambda L}{d^2} \hat{z} \]

and mention line charge at this distance looks like a point charge

"

Solution 1:

"We are tasked with finding the electric field a distance \( d \) above one end of a straight line segment of length \( L \) with a constant line charge density \( \lambda \), and determining its behavior in the limit where \( d \gg L \).

Let us proceed with the solution:

1. **Electric Field due to a Line of Charge**: 

The electric field at a point due to a small element of charge \( dq \) on a charged line is given by Coulomb's law:
\[
d\vec{E} = \frac{1}{4 \pi \epsilon_0} \frac{dq}{r^2} \hat{r}
\]
where \( dq = \lambda \, dx \) is the charge of the infinitesimal segment of length \( dx \), \( r \) is the distance from the element to the point of interest, and \( \hat{r} \) is the unit vector in the direction of the electric field.

2. **Set Up the Coordinate System**:

Place the line of charge along the \( x \)-axis with one end at the origin. The point where we are calculating the electric field is located at a distance \( d \) along the \( y \)-axis, directly above the origin.

The distance from a point on the line of charge at position \( x \) to the point of interest is:
\[
r = \sqrt{x^2 + d^2}
\]
The electric field due to the infinitesimal charge \( dq \) at position \( x \) has both \( x \)- and \( y \)-components. The angle \( \theta \) between the vector \( r \) and the \( x \)-axis is given by:
\[
\cos \theta = \frac{x}{\sqrt{x^2 + d^2}}, \quad \sin \theta = \frac{d}{\sqrt{x^2 + d^2}}
\]

3. **Calculate the Electric Field Components**:

The differential electric field along the \( x \)-axis is:
\[
dE_x = \frac{1}{4 \pi \epsilon_0} \frac{\lambda \, dx}{(x^2 + d^2)^{3/2}} x
\]
The differential electric field along the \( y \)-axis is:
\[
dE_y = \frac{1}{4 \pi \epsilon_0} \frac{\lambda \, dx}{(x^2 + d^2)^{3/2}} d
\]

4. **Integrating to Find the Total Electric Field**:

To find the total electric field, we integrate over the length of the charge distribution, from \( x = 0 \) to \( x = L \).

- The total \( x \)-component is:
\[
E_x = \int_0^L \frac{1}{4 \pi \epsilon_0} \frac{\lambda x}{(x^2 + d^2)^{3/2}} \, dx
\]
- The total \( y \)-component is:
\[
E_y = \int_0^L \frac{1}{4 \pi \epsilon_0} \frac{\lambda d}{(x^2 + d^2)^{3/2}} \, dx
\]

Let's compute these integrals.

For \( E_x \), use the substitution \( u = x^2 + d^2 \), which gives \( du = 2x \, dx \). The integral becomes:
\[
E_x = \frac{\lambda}{4 \pi \epsilon_0} \int_0^L \frac{x \, dx}{(x^2 + d^2)^{3/2}} = \frac{\lambda}{4 \pi \epsilon_0} \left[ \frac{1}{\sqrt{x^2 + d^2}} \right]_0^L
\]
Thus:
\[
E_x = \frac{\lambda}{4 \pi \epsilon_0} \left( \frac{1}{d} - \frac{1}{\sqrt{L^2 + d^2}} \right)
\]

For \( E_y \), the integral is straightforward:
\[
E_y = \frac{\lambda d}{4 \pi \epsilon_0} \int_0^L \frac{dx}{(x^2 + d^2)^{3/2}}
\]
Using the substitution \( u = x^2 + d^2 \), we get:
\[
E_y = \frac{\lambda d}{4 \pi \epsilon_0} \left[ \frac{1}{d^2 \sqrt{x^2 + d^2}} \right]_0^L
\]
Thus:
\[
E_y = \frac{\lambda}{4 \pi \epsilon_0} \left( \frac{1}{d} - \frac{1}{\sqrt{L^2 + d^2}} \right)
\]

5. **Total Electric Field**:

The total electric field is the vector sum of the \( x \)- and \( y \)-components:
\[
\vec{E} = E_x \hat{i} + E_y \hat{j}
\]
Thus, the electric field at a distance \( d \) above one end of the line charge is:
\[
\vec{E} = \frac{\lambda}{4 \pi \epsilon_0} \left( \frac{1}{d} - \frac{1}{\sqrt{L^2 + d^2}} \right) \hat{i} + \frac{\lambda}{4 \pi \epsilon_0} \left( \frac{1}{d} - \frac{1}{\sqrt{L^2 + d^2}} \right) \hat{j}
\]

6. **Limit of \( d \gg L \)**:

In the limit where \( d \gg L \), we can approximate \( \sqrt{L^2 + d^2} \approx d \). Therefore, both the \( x \)- and \( y \)-components of the electric field reduce to:
\[
E_x \approx \frac{\lambda}{4 \pi \epsilon_0} \left( \frac{1}{d} - \frac{1}{d} \right) = 0
\]
and
\[
E_y \approx \frac{\lambda}{4 \pi \epsilon_0} \left( \frac{1}{d} - \frac{1}{d} \right) = 0
\]
Thus, the electric field approaches zero in the limit \( d \gg L \).

7. **Physical Interpretation**:

The result indicates that in the limit \( d \gg L \), the electric field at the point above the line charge diminishes to zero. This is because the line of charge behaves like a point charge with a very small contribution at large distances. As the distance from the line increases far beyond its length, the field becomes negligible, which is consistent with the behavior of a monopole (point charge) at large distances."

Solution 2:
"We are asked to find the electric field at a point a distance \( d \) above one end of a straight line segment of length \( L \) carrying a uniform charge distribution with linear charge density \( \lambda \). Additionally, we need to consider the case where \( d \gg L \) and interpret the result.

Let us start by setting up the problem.

### Step 1: Define the coordinate system and variables

We place the line segment along the \( x \)-axis, with one end at the origin (\( x = 0 \)) and the other end at \( x = L \). The charge density \( \lambda \) means that the charge per unit length of the segment is \( \lambda \). 

We are tasked with finding the electric field at a point \( P \) located at a height \( d \) directly above the origin, i.e., the position of point \( P \) is \( (0, d) \).

### Step 2: Expression for the differential electric field

Consider an infinitesimal segment of the charged line at position \( x \) with length \( dx \). The charge of this infinitesimal segment is \( dq = \lambda dx \).

The electric field at point \( P \) due to this infinitesimal charge \( dq \) is given by Coulomb's law:
\[
d\mathbf{E} = \frac{1}{4\pi \epsilon_0} \frac{dq}{r^2} \hat{r}
\]
where \( r \) is the distance from the charge element to the point \( P \), and \( \hat{r} \) is the unit vector pointing from the charge element to the point \( P \).

The distance \( r \) between the charge element at position \( x \) and the point \( P \) is:
\[
r = \sqrt{x^2 + d^2}
\]
The unit vector \( \hat{r} \) points from the charge element at \( x \) to the point \( P \). Since \( P \) lies directly above the origin, the vector components of \( \hat{r} \) are:
\[
\hat{r} = \left( -\frac{x}{\sqrt{x^2 + d^2}}, \frac{d}{\sqrt{x^2 + d^2}} \right)
\]
The differential electric field then becomes:
\[
d\mathbf{E} = \frac{1}{4\pi \epsilon_0} \frac{\lambda dx}{(x^2 + d^2)^{3/2}} \left( -x, d \right)
\]
This has both \( x \)- and \( y \)-components.

### Step 3: Integrate to find the total electric field

The total electric field at \( P \) is the sum of the contributions from all elements of the line segment. We need to integrate the components of \( d\mathbf{E} \) over the length of the segment, from \( x = 0 \) to \( x = L \).

For the \( x \)-component, the integral is:
\[
E_x = \int_0^L \frac{1}{4\pi \epsilon_0} \frac{\lambda (-x) dx}{(x^2 + d^2)^{3/2}}
\]
For the \( y \)-component, the integral is:
\[
E_y = \int_0^L \frac{1}{4\pi \epsilon_0} \frac{\lambda d dx}{(x^2 + d^2)^{3/2}}
\]

Let us solve these integrals.

#### \( x \)-component of the electric field:
The integral for \( E_x \) is an odd function because the integrand involves \( x \), which is symmetric about \( x = 0 \) for the range from \( 0 \) to \( L \). Therefore, the contribution to the \( x \)-component of the electric field vanishes:
\[
E_x = 0
\]

#### \( y \)-component of the electric field:
Now, for the \( y \)-component, we have:
\[
E_y = \frac{\lambda d}{4\pi \epsilon_0} \int_0^L \frac{dx}{(x^2 + d^2)^{3/2}}
\]
This integral is a standard form, and its solution is:
\[
\int \frac{dx}{(x^2 + d^2)^{3/2}} = \frac{x}{d^2 \sqrt{x^2 + d^2}}
\]
Evaluating the definite integral from \( x = 0 \) to \( x = L \), we get:
\[
E_y = \frac{\lambda d}{4\pi \epsilon_0} \left[ \frac{x}{d^2 \sqrt{x^2 + d^2}} \right]_0^L
\]
Substituting the limits:
\[
E_y = \frac{\lambda d}{4\pi \epsilon_0} \left( \frac{L}{d^2 \sqrt{L^2 + d^2}} \right)
\]
Thus, the total electric field at the point \( P \) is:
\[
\mathbf{E} = \left( 0, \frac{\lambda}{4\pi \epsilon_0} \frac{L}{d \sqrt{L^2 + d^2}} \right)
\]

### Step 4: Limit \( d \gg L \)

In the limit where \( d \gg L \), we can approximate \( \sqrt{L^2 + d^2} \approx d \). Thus, the electric field simplifies to:
\[
\mathbf{E} \approx \left( 0, \frac{\lambda}{4\pi \epsilon_0} \frac{L}{d^2} \right)
\]

### Step 5: Physical interpretation

In the limit \( d \gg L \), the electric field behaves as if the line segment were a point charge. The expression \( \frac{L}{d^2} \) suggests that the field falls off as \( 1/d^2 \), which is the characteristic behavior of the electric field due to a point charge. Thus, the electric field at a large distance from the line segment behaves similarly to the field due to a point charge with total charge \( Q = \lambda L \).

### Final Answer:
The electric field at a distance \( d \) above one end of a line segment of length \( L \) with charge density \( \lambda \) is:
\[
\mathbf{E} = \left( 0, \frac{\lambda}{4\pi \epsilon_0} \frac{L}{d \sqrt{L^2 + d^2}} \right)
\]
In the limit \( d \gg L \), the electric field reduces to:
\[
\mathbf{E} \approx \left( 0, \frac{\lambda}{4\pi \epsilon_0} \frac{L}{d^2} \right)
\]
This is the electric field of a point charge with total charge \( Q = \lambda L \)."

Solution 3:
"To solve this problem, we need to compute the electric field at a point located a distance \( d \) above one end of a uniformly charged line segment of length \( L \) and line charge density \( \lambda \).

### Step 1: Electric field due to a small charge element
We begin by considering a small element of charge \( dq \) on the line segment. The charge density \( \lambda \) is defined as the charge per unit length, so the charge of a small segment of length \( dx \) is given by:
\[
dq = \lambda dx
\]
The position of the small charge element is at a distance \( x \) from the origin of the line segment, where \( x \) ranges from 0 to \( L \). The electric field due to this charge element at the point above the end of the line segment is directed radially outward from \( dq \) and depends on the distance from \( dq \) to the point where the field is being evaluated.

The distance between the charge element at position \( x \) and the point where we are calculating the electric field is:
\[
r = \sqrt{x^2 + d^2}
\]
The electric field due to this small charge element is given by Coulomb's law:
\[
d\vec{E} = \frac{1}{4 \pi \epsilon_0} \frac{dq}{r^2} \hat{r}
\]
where \( \hat{r} \) is the unit vector in the direction of the field. Since the field is at a point directly above the end of the line segment, \( \hat{r} \) can be decomposed into two components: a vertical component \( dE_y \) and a horizontal component \( dE_x \).

The vertical component is given by:
\[
dE_y = dE \cos \theta = \frac{1}{4 \pi \epsilon_0} \frac{dq}{r^2} \cdot \frac{d}{r}
\]
and the horizontal component is:
\[
dE_x = dE \sin \theta = \frac{1}{4 \pi \epsilon_0} \frac{dq}{r^2} \cdot \frac{x}{r}
\]

### Step 2: Expression for the vertical component of the electric field
Now, we calculate the vertical component of the electric field:
\[
dE_y = \frac{1}{4 \pi \epsilon_0} \frac{\lambda dx}{(x^2 + d^2)^{3/2}} d
\]
Integrating this from \( x = 0 \) to \( x = L \) gives the total vertical electric field \( E_y \):
\[
E_y = \int_0^L \frac{1}{4 \pi \epsilon_0} \frac{\lambda d \, dx}{(x^2 + d^2)^{3/2}}
\]
This integral is a standard form, and the result is:
\[
E_y = \frac{1}{4 \pi \epsilon_0} \lambda \left[ \frac{x}{d \sqrt{x^2 + d^2}} \right]_0^L
\]
Substituting the limits:
\[
E_y = \frac{1}{4 \pi \epsilon_0} \lambda \left( \frac{L}{d \sqrt{L^2 + d^2}} - 0 \right)
\]
Thus, the vertical electric field is:
\[
E_y = \frac{1}{4 \pi \epsilon_0} \lambda \frac{L}{d \sqrt{L^2 + d^2}}
\]

### Step 3: Interpretation in the limit \( d \gg L \)
In the limit where \( d \gg L \), the term \( \sqrt{L^2 + d^2} \) can be approximated as \( d \), because \( L^2 \) is negligible compared to \( d^2 \). Thus, the electric field simplifies to:
\[
E_y \approx \frac{1}{4 \pi \epsilon_0} \lambda \frac{L}{d^2}
\]

### Step 4: Physical interpretation
In the limit \( d \gg L \), the electric field behaves like that due to a point charge, as the distance \( d \) is much larger than the length of the line segment. The line segment can thus be approximated as a point charge with total charge \( Q = \lambda L \). The electric field due to a point charge is:
\[
E_y = \frac{1}{4 \pi \epsilon_0} \frac{Q}{d^2} = \frac{1}{4 \pi \epsilon_0} \frac{\lambda L}{d^2}
\]
This result confirms that when \( d \gg L \), the electric field due to a uniformly charged line segment behaves similarly to that of a point charge with charge \( \lambda L \).

Thus, in the limit \( d \gg L \), the electric field at the point above the end of the line segment is directed vertically downward, and its magnitude is inversely proportional to \( d^2 \), as expected from the behavior of the field due to a point charge.

\[
E_y \approx \frac{1}{4 \pi \epsilon_0} \frac{\lambda L}{d^2}
\]
This is the desired result."


\subsubsection{EMQ4}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 11 marks. Give a total of 5 marks for part a), give a total of 5 marks for part b) and give a total of 1 mark for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"A square loop of wire (sides of length \( L \)) lies a fixed distance \( d \) from a very long straight wire, which carries a constant current \( I \) directed downwards (as seen in the figure).

a) Find the flux of \( \underline{B} \) through the loop.

b) If the loop is moved directly away from the wire to the right, at speed \( v \), what emf is generated? In what direction (clockwise or counterclockwise) does the current flow?

c) What happens if the loop only moves downwards at speed \( v \)?"

Mark Scheme:

"{Q4: (11 marks total)}

a) (5 marks)

[1 mark] Use of definition of flux as surface integral

\[ \int \underline{B} \cdot \underline{dA} \]

[3 marks] Correct setup of integral for flux, using \underline{B} for an infinite long wire 
 
\[ \underline{B} = \frac{\mu_0 I}{2\pi r} \hat{\phi} \]
\[ \Phi = \frac{\mu_0 I}{2\pi} \int_{d}^{d+L} \frac{1}{r} \, (L \, dr) \]

[1 mark] correction evaluation of integral to get flux
\[ \Phi = \frac{\mu_0 I L}{2\pi}\ln\left ( \frac{d+L}{d} \right ) \]

b) (5 marks)

[1 mark] Use of emf definition.  \( \mathcal{E} = - \frac{d\Phi}{dt} \)

[2 marks] Correct evaluation of the derivative to get the emf

\[ \mathcal{E} = \frac{\mu_0 I L^2 v}{2\pi r (r+L)} \]

here \( r \) is the perpendicular distance from the wire to the closest side of the square. More explicitly \( r = d + vt \).

[2 marks] Current moves counterclockwise. Justify by Lenz Law or explicitly showing direction of force.

c) (1 mark)

[1 mark] No emf since flux does not change.

"

Solution 1: 

"We are given a square loop of wire with side length $L$, positioned a fixed distance $d$ from a very long straight wire carrying a current $I$ directed downward. The problem asks us to find the magnetic flux through the loop, the induced emf when the loop is moved away from the wire, and the behavior when the loop moves downwards.

\begin{enumerate}
    \item[(a)] \textbf{Finding the flux of $\mathbf{B}$ through the loop.}
    
    The magnetic field $\mathbf{B}$ produced by a long straight current-carrying wire at a distance $r$ from it is given by Ampère's Law:
    \[
    B(r) = \frac{\mu_0 I}{2\pi r}
    \]
    where $\mu_0$ is the permeability of free space, $I$ is the current in the wire, and $r$ is the radial distance from the wire. In our case, the square loop is a distance $d$ from the wire, so the magnetic field at any point on the loop depends on the distance from the wire.

    The flux $\Phi$ through the loop is the integral of the magnetic field $\mathbf{B}$ over the area of the loop. Assuming that the magnetic field is perpendicular to the plane of the loop (which is valid given that the loop lies in the plane of the paper and the current is vertically downward), we have:
    \[
    \Phi = \int_{A} \mathbf{B} \cdot d\mathbf{A}
    \]
    Since the magnetic field varies with distance from the wire, we must consider the magnetic field at each point along the loop. Let’s assume that the wire lies along the $y$-axis, and the loop is parallel to the $x$-axis, spanning from $x = 0$ to $x = L$.

    The magnetic field at a point at a distance $x$ from the wire is:
    \[
    B(x) = \frac{\mu_0 I}{2\pi (d+x)}
    \]
    The area element $dA$ at each point on the loop is simply $dx \cdot L$, since the loop has length $L$. Thus, the flux is given by:
    \[
    \Phi = \int_0^L B(x) \, L \, dx = \int_0^L \frac{\mu_0 I L}{2\pi (d+x)} \, dx
    \]
    This integral can be computed as:
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \int_0^L \frac{dx}{d+x}
    \]
    The integral of $\frac{1}{d+x}$ is $\ln(d+x)$, so we have:
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \left[ \ln(d+L) - \ln(d) \right]
    \]
    Simplifying this expression, we get:
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \ln \left( \frac{d+L}{d} \right)
    \]

    \item[(b)] \textbf{Induced emf when the loop moves away from the wire.}
    
    If the loop is moved directly away from the wire with a speed $v$ to the right, the distance between the wire and the loop changes with time. Since the magnetic flux depends on the distance between the wire and the loop, the flux will change as the loop moves.

    The induced emf in the loop can be found using Faraday's Law of Induction:
    \[
    \mathcal{E} = -\frac{d\Phi}{dt}
    \]
    We know from part (a) that the flux is given by:
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \ln \left( \frac{d+L}{d} \right)
    \]
    If the loop moves away from the wire, the distance $d$ changes with time, so we write $d(t) = d + vt$. Taking the derivative of the flux with respect to time, we get:
    \[
    \frac{d\Phi}{dt} = \frac{\mu_0 I L}{2\pi} \cdot \frac{d}{dt} \ln \left( \frac{d+L}{d} \right)
    \]
    Using the chain rule, we find:
    \[
    \frac{d}{dt} \ln \left( \frac{d+L}{d} \right) = \frac{1}{d+L} \cdot v
    \]
    Therefore, the induced emf is:
    \[
    \mathcal{E} = -\frac{\mu_0 I L v}{2\pi (d+L)}
    \]
    The negative sign indicates the direction of the emf, according to Lenz's Law. The current in the loop will flow in such a direction as to oppose the change in flux. In this case, since the loop is moving away from the wire, the magnetic field through the loop is decreasing, and the current will flow in a direction that creates a magnetic field opposing this decrease. This means that the current will flow counterclockwise when viewed from above.

    \item[(c)] \textbf{What happens if the loop only moves downwards at speed $v$?}
    
    If the loop moves downward at speed $v$, the distance from the wire does not change, so the magnetic flux through the loop remains constant. Since the magnetic field does not change and the loop's orientation is fixed, no emf is induced. Therefore, the current in the loop remains unchanged, and there is no induced current or emf in this case.
\end{enumerate}"

Solution 2:

"We are given a square loop of wire with side length $L$ placed a distance $d$ away from a very long straight wire carrying a constant current $I$ directed downwards. We are tasked with finding the magnetic flux through the loop, the induced EMF as the loop moves, and the result of the loop moving in a specific direction.

\begin{enumerate}
    \item[(a)] \textbf{Find the flux of $\mathbf{B}$ through the loop.}
    
    The magnetic field around a long straight current-carrying wire is given by Ampere's law:
    \[
    B(r) = \frac{\mu_0 I}{2\pi r}
    \]
    where $r$ is the radial distance from the wire and $\mu_0$ is the permeability of free space. The magnetic field decreases as the inverse of the distance from the wire. The direction of the magnetic field follows the right-hand rule, which gives concentric circles around the wire.
    
    The total flux through the loop is the surface integral of the magnetic field $\mathbf{B}$ over the area of the loop. The magnetic field at any point on the loop depends on the distance from the wire, which varies along the loop. Let the distance from the wire to a point on the loop be $r$.
    
    For simplicity, we assume the loop is oriented perpendicular to the magnetic field lines and parallel to the wire. The area element $dA$ of the square loop is $L \, dL$, and the flux through an infinitesimal section of the loop is given by:
    \[
    d\Phi = B(r) \, dA = \frac{\mu_0 I}{2\pi r} \, L \, dL
    \]
    
    To calculate the total flux, we integrate over the length of the side of the square loop. The limits of integration are from $r = d$ to $r = d + L$.
    \[
    \Phi = \int_d^{d+L} \frac{\mu_0 I}{2\pi r} L \, dr
    \]
    Performing the integration:
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \int_d^{d+L} \frac{1}{r} \, dr
    \]
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \left[ \ln(r) \right]_d^{d+L}
    \]
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \left( \ln(d+L) - \ln(d) \right)
    \]
    \[
    \Phi = \frac{\mu_0 I L}{2\pi} \ln \left( \frac{d+L}{d} \right)
    \]
    
    This is the magnetic flux through the loop.

    \item[(b)] \textbf{If the loop is moved directly away from the wire to the right, at speed $v$, what emf is generated? In what direction (clockwise or counterclockwise) does the current flow?}
    
    According to Faraday's law of induction, the induced EMF in the loop is given by the negative rate of change of magnetic flux through the loop:
    \[
    \mathcal{E} = -\frac{d\Phi}{dt}
    \]
    
    We know the flux $\Phi$ depends on the distance $d$ from the wire, and as the loop moves to the right at speed $v$, the distance $d$ changes with time. Let $d(t) = d_0 + vt$, where $d_0$ is the initial distance from the wire.
    
    The flux $\Phi$ as a function of $d(t)$ is:
    \[
    \Phi(t) = \frac{\mu_0 I L}{2\pi} \ln \left( \frac{d(t) + L}{d(t)} \right)
    \]
    
    Differentiating with respect to time $t$:
    \[
    \frac{d\Phi}{dt} = \frac{\mu_0 I L}{2\pi} \frac{d}{dt} \left[ \ln \left( \frac{d(t) + L}{d(t)} \right) \right]
    \]
    
    Using the chain rule:
    \[
    \frac{d}{dt} \left[ \ln \left( \frac{d(t) + L}{d(t)} \right) \right] = \frac{1}{d(t) + L} \cdot v - \frac{1}{d(t)} \cdot v = v \left( \frac{1}{d(t) + L} - \frac{1}{d(t)} \right)
    \]
    
    Therefore:
    \[
    \frac{d\Phi}{dt} = \frac{\mu_0 I L v}{2\pi} \left( \frac{1}{d + L} - \frac{1}{d} \right)
    \]
    
    This gives the rate of change of the flux and hence the induced EMF:
    \[
    \mathcal{E} = - \frac{\mu_0 I L v}{2\pi} \left( \frac{1}{d + L} - \frac{1}{d} \right)
    \]
    
    The negative sign indicates that the induced current will flow in such a way as to oppose the change in flux, according to Lenz's Law. Since the magnetic field lines point in the counterclockwise direction around the wire, the induced current will flow in a direction that creates a magnetic field opposing the decrease in flux as the loop moves away from the wire. Thus, the current will flow \textbf{clockwise} in the loop when viewed from above.

    \item[(c)] \textbf{What happens if the loop only moves downwards at speed $v$?}
    
    If the loop moves directly downward, the distance between the wire and the loop remains constant. Since the magnetic field at each point on the loop does not change with time, there is no change in the flux through the loop. According to Faraday’s law, since there is no change in flux, there is no induced EMF and consequently no induced current in the loop.
\end{enumerate}"

Solution 3: 

"We are tasked with finding the magnetic flux through a square loop of wire and analyzing the behavior of the induced current when the loop moves relative to a current-carrying wire.

\textbf{(a) Find the flux of $\mathbf{B}$ through the loop.}

The magnetic field $\mathbf{B}$ produced by a long straight wire carrying a current $I$ at a distance $r$ from the wire is given by Ampère's law:

\[
B = \frac{\mu_0 I}{2 \pi r}
\]

where $\mu_0$ is the permeability of free space and $r$ is the distance from the wire. Since the loop lies a fixed distance $d$ from the wire, the magnetic field at any point on the loop varies as the distance from the wire changes.

Let’s calculate the flux $\Phi_B$ through the loop. The flux is given by the integral:

\[
\Phi_B = \int \mathbf{B} \cdot d\mathbf{A}
\]

where $\mathbf{B}$ is the magnetic field and $d\mathbf{A}$ is the area element. The area of the square loop is $L^2$. Each point on the loop experiences a magnetic field strength that depends on its distance from the wire. To find the total flux, we need to consider the magnetic field at each point along the loop.

At a distance $x$ from the wire, the magnetic field is:

\[
B(x) = \frac{\mu_0 I}{2 \pi x}
\]

The differential area element along the loop is $dA = L dx$ (the length of the loop is $L$, and the element of distance is $dx$).

Thus, the flux through the loop is:

\[
\Phi_B = \int_0^L \frac{\mu_0 I}{2 \pi (d + x)} L dx
\]

This integral can be solved by performing the integration:

\[
\Phi_B = \frac{\mu_0 I L}{2 \pi} \int_0^L \frac{1}{d + x} dx
\]

The integral of $\frac{1}{d + x}$ is $\ln(d + x)$. Therefore:

\[
\Phi_B = \frac{\mu_0 I L}{2 \pi} \left[ \ln(d + x) \right]_0^L
\]

Evaluating this:

\[
\Phi_B = \frac{\mu_0 I L}{2 \pi} \left( \ln(d + L) - \ln(d) \right)
\]

Simplifying:

\[
\Phi_B = \frac{\mu_0 I L}{2 \pi} \ln\left( \frac{d + L}{d} \right)
\]

This is the flux through the loop.

\textbf{(b) If the loop is moved directly away from the wire to the right, at speed $v$, what emf is generated? In what direction (clockwise or counterclockwise) does the current flow?}

When the loop moves with velocity $v$ away from the wire, the area of the loop does not change, but the magnetic field through the loop changes because the distance between the wire and the loop increases. According to Faraday’s law of induction:

\[
\mathcal{E} = - \frac{d\Phi_B}{dt}
\]

We already know that the flux $\Phi_B$ depends on the distance $d$ from the wire:

\[
\Phi_B = \frac{\mu_0 I L}{2 \pi} \ln\left( \frac{d + L}{d} \right)
\]

As the loop moves with speed $v$ away from the wire, the distance $d$ increases with time. Thus, we differentiate the flux with respect to time:

\[
\frac{d\Phi_B}{dt} = \frac{\mu_0 I L}{2 \pi} \frac{d}{dt} \ln\left( \frac{d + L}{d} \right)
\]

Using the chain rule, we find:

\[
\frac{d}{dt} \ln\left( \frac{d + L}{d} \right) = \frac{1}{\frac{d + L}{d}} \cdot \frac{d}{dt} \left( \frac{d + L}{d} \right) = \frac{d}{(d + L)d}
\]

Now, we use the fact that $d$ changes with time as $d = d_0 + vt$ (with initial distance $d_0$). Therefore:

\[
\frac{d\Phi_B}{dt} = \frac{\mu_0 I L}{2 \pi} \cdot \frac{v}{d^2}
\]

Thus, the induced emf is:

\[
\mathcal{E} = - \frac{\mu_0 I L}{2 \pi} \cdot \frac{v}{d^2}
\]

The negative sign indicates that the induced emf opposes the change in flux (Lenz’s law). Since the magnetic field produced by the wire is directed into the page (due to the current being downward), the induced current will flow in a direction that creates a magnetic field opposing the increasing distance from the wire. Therefore, the current will flow in a counterclockwise direction (when viewed from above) to create a magnetic field directed into the page, opposing the flux increase.

\textbf{(c) What happens if the loop only moves downwards at speed $v$?}

If the loop moves downward at speed $v$, the relative position of the loop to the wire does not change in the horizontal direction. Therefore, the distance $d$ between the wire and the loop remains constant. Since the magnetic field generated by the wire is a function of this distance, no change in the magnetic flux occurs if $d$ remains constant. As a result, there is no change in the flux through the loop, and no induced emf is generated. 

Hence, no current is induced when the loop moves downward.
"

\subsubsection{EMQ5}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 11 marks. Give a total of 8 marks for part a) and give a total of 3 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Consider the circuit diagram seen in the figure. 

    a) Calculate the current seen by the ammeter.

    b) Calculate the energy delivered by the \(12\, \text{V}\) battery in \(4\) seconds."


Mark Scheme:

"{Q5: (11 marks total)}

a) (8 marks)

[1 mark] Mention use of Kirchhoff Circuit Laws 

[2 marks] split current of circuit into 3 branches, \( I_1 \), \( I_2 \), \( I_3 \). Current through the ammeter denoted by one of these currents. Use of Kirchoff’s current law eg. \( I_1 + I_2 = I_3 \) (exact description is dependent on labeling)

[2 marks] Use of Kirchoff’s voltage law to obtain 2 more equations of  \( I_1 \), \( I_2 \), and \( I_3 \).

[3 marks] Solve simultaneous equations and obtain current through the ammeter which is 4/3 A. 

b)  (3 marks)

[2 marks] Power generated by the 12V battery is the sum of power delivered to each component. Using \( P = IV \) or \( P = I^2 R \),

\[ P_{tot} = \sum_{i} P_i = \left ( \frac{11}{3} \right )^2 \times 2 + \left ( \frac{7}{3} \right )^2 \times 2 + \left ( \frac{4}{3} \right ) \times 2 + \left ( \frac{4}{3} \right )^2 \times 2 = 44 W \]

[1 mark] Energy delivered in 4 seconds: 
\[ E = P_{tot} t = 44 \times 4 = 176 J \]

"

Solution 1:

"\textbf{Part (a): Calculate the current seen by the ammeter.}

The circuit consists of resistors and batteries. We are asked to calculate the current seen by the ammeter, which measures the total current flowing through the circuit. To do this, we will apply Kirchhoff’s Voltage Law (KVL) and Ohm’s Law.

First, we label the resistors and voltage sources:
- \( R_1 = 2 \, \Omega \) (resistor on the left)
- \( R_2 = 2 \, \Omega \) (resistor in the middle)
- \( R_3 = 1 \, \Omega \) (resistor on the right)
- \( E_1 = 12 \, \text{V} \) (12V battery)
- \( E_2 = 2 \, \text{V} \) (2V battery)

We begin by calculating the total resistance of the circuit. The resistors \( R_2 \) and \( R_3 \) are in series, so their equivalent resistance is:

\[
R_{\text{eq}} = R_2 + R_3 = 2 \, \Omega + 1 \, \Omega = 3 \, \Omega
\]

Now, this equivalent resistance is in parallel with \( R_1 \), so we can calculate the total resistance \( R_{\text{total}} \) of the circuit using the formula for resistors in parallel:

\[
\frac{1}{R_{\text{total}}} = \frac{1}{R_1} + \frac{1}{R_{\text{eq}}}
\]

Substituting the known values:

\[
\frac{1}{R_{\text{total}}} = \frac{1}{2 \, \Omega} + \frac{1}{3 \, \Omega} = \frac{3 + 2}{6} = \frac{5}{6}
\]

Therefore:

\[
R_{\text{total}} = \frac{6}{5} \, \Omega = 1.2 \, \Omega
\]

Next, we apply Kirchhoff’s Voltage Law (KVL) to the loop. The net electromotive force (emf) in the circuit is the difference between the two voltage sources:

\[
E_{\text{net}} = E_1 - E_2 = 12 \, \text{V} - 2 \, \text{V} = 10 \, \text{V}
\]

Using Ohm’s Law to calculate the total current \( I \) in the circuit:

\[
I = \frac{E_{\text{net}}}{R_{\text{total}}} = \frac{10 \, \text{V}}{1.2 \, \Omega} = 8.33 \, \text{A}
\]

Thus, the current seen by the ammeter is:

\[
I = 8.33 \, \text{A}
\]

\textbf{Part (b): Calculate the energy delivered by the 12 V battery in 4 seconds.}

To calculate the energy delivered by the 12 V battery, we use the formula for electrical energy:

\[
E = P \cdot t
\]

where \( P \) is the power delivered by the battery, and \( t \) is the time. The power \( P \) delivered by a battery is given by:

\[
P = V \cdot I
\]

where \( V = 12 \, \text{V} \) is the voltage of the battery and \( I = 8.33 \, \text{A} \) is the current we calculated in part (a). Substituting these values:

\[
P = 12 \, \text{V} \cdot 8.33 \, \text{A} = 100 \, \text{W}
\]

Now, using the formula for energy:

\[
E = 100 \, \text{W} \cdot 4 \, \text{s} = 400 \, \text{J}
\]

Thus, the energy delivered by the 12 V battery in 4 seconds is:

\[
E = 400 \, \text{J}
\]"

Solution 2:

"We are given the following circuit with a 12V battery, a 2Ω resistor, a 2V battery, and a 1Ω resistor in the given arrangement. The task is to solve for the current seen by the ammeter and the energy delivered by the 12V battery in 4 seconds.

\textbf{Part (a): Calculate the current seen by the ammeter.}

To calculate the current seen by the ammeter, we need to first determine the total equivalent resistance in the circuit.

The circuit contains the following resistances:
- Two resistors of 2Ω each, one in series with the 12V battery and the other in the path of the 2V battery.
- One resistor of 1Ω in the same branch as the 2Ω resistor of the second path.
- A 2V battery and a 12V battery are connected in series with these resistors.

We need to calculate the total voltage and the total resistance.

\textbf{Step 1: Find the total resistance.}

- The two 2Ω resistors are in series, so their combined resistance is:
  \[
  R_{\text{series}} = 2Ω + 2Ω = 4Ω
  \]

- The 4Ω combined resistance is in parallel with the 1Ω resistor. The total resistance for two resistors in parallel is given by the formula:
  \[
  \frac{1}{R_{\text{parallel}}} = \frac{1}{R_1} + \frac{1}{R_2}
  \]
  where \(R_1 = 4Ω\) and \(R_2 = 1Ω\). Therefore,
  \[
  \frac{1}{R_{\text{parallel}}} = \frac{1}{4} + \frac{1}{1} = \frac{1 + 4}{4} = \frac{5}{4}
  \]
  Thus, the equivalent parallel resistance is:
  \[
  R_{\text{parallel}} = \frac{4}{5}Ω
  \]

\textbf{Step 2: Find the total voltage across the circuit.}

Since the two batteries are in series, their total voltage is the sum of the individual voltages:
\[
V_{\text{total}} = 12V + 2V = 14V
\]

\textbf{Step 3: Use Ohm's Law to find the current.}

Using Ohm's Law, \(I = \frac{V}{R}\), we can now calculate the current seen by the ammeter. The total voltage is \(14V\), and the total resistance is \(R_{\text{total}} = R_{\text{parallel}} = \frac{4}{5}Ω\). Thus, the current is:
\[
I = \frac{14V}{\frac{4}{5}Ω} = 14V \times \frac{5}{4Ω} = 17.5A
\]

Therefore, the current seen by the ammeter is \(I = 17.5A\).

\textbf{Part (b): Calculate the energy delivered by the 12V battery in 4 seconds.}

The energy delivered by a battery can be calculated using the formula:
\[
E = P \times t
\]
where \(P\) is the power delivered by the battery and \(t\) is the time.

\textbf{Step 1: Calculate the power delivered by the 12V battery.}

The power delivered by the battery is given by:
\[
P = V_{\text{battery}} \times I
\]
where \(V_{\text{battery}} = 12V\) and \(I = 17.5A\) is the current from part (a). Thus, the power is:
\[
P = 12V \times 17.5A = 210W
\]

\textbf{Step 2: Calculate the energy delivered in 4 seconds.}

Now that we know the power, we can calculate the energy delivered in 4 seconds:
\[
E = P \times t = 210W \times 4s = 840J
\]

Therefore, the energy delivered by the 12V battery in 4 seconds is \(E = 840J\).

\textbf{Final Answers:}
\begin{enumerate}
    \item The current seen by the ammeter is \(I = 17.5A\).
    \item The energy delivered by the 12V battery in 4 seconds is \(E = 840J\).
\end{enumerate}"

Solution 3:

"\textbf{a) Calculation of the current seen by the ammeter:}

We are given a circuit with resistors and two batteries. The total current in the circuit can be found by applying Kirchhoff’s laws. Let’s begin by simplifying the circuit.

The circuit consists of the following components:
- A $12V$ battery,
- A $2\Omega$ resistor in series with the $12V$ battery,
- A $2V$ battery in the circuit,
- Two $2\Omega$ resistors in series,
- An ammeter measuring the current, and
- A $1\Omega$ resistor in series with the $2\Omega$ resistors.

We will first calculate the total resistance in the circuit. Start by combining the resistors in series.

The total resistance in the two $2\Omega$ resistors and the $1\Omega$ resistor is:
\[
R_{\text{total}} = 2\Omega + 2\Omega + 1\Omega = 5\Omega
\]

Next, apply Kirchhoff’s Voltage Law (KVL) to the circuit. Assume the current flows from the $12V$ battery, through the $2\Omega$ resistors, the $1\Omega$ resistor, and then to the $2V$ battery. 

The equation for the loop with the $12V$ and $2V$ batteries, considering the current $I$ flowing through the resistors, is:
\[
12V - (I \cdot 5\Omega) - 2V = 0
\]

Simplifying this equation:
\[
12V - 2V = I \cdot 5\Omega
\]
\[
10V = I \cdot 5\Omega
\]
\[
I = \frac{10V}{5\Omega} = 2A
\]

Thus, the current seen by the ammeter is $I = 2A$.

\textbf{b) Calculation of the energy delivered by the $12V$ battery in 4 seconds:}

The energy delivered by a battery can be calculated using the formula:
\[
E = P \cdot t
\]
where $P$ is the power delivered by the battery and $t$ is the time. The power delivered by the $12V$ battery can be calculated using the formula:
\[
P = V \cdot I
\]
where $V = 12V$ is the voltage of the battery, and $I = 2A$ is the current.

Thus, the power delivered by the $12V$ battery is:
\[
P = 12V \cdot 2A = 24W
\]

Now, to find the energy delivered in $4$ seconds, we use:
\[
E = 24W \cdot 4s = 96J
\]

Therefore, the energy delivered by the $12V$ battery in 4 seconds is $96J$."


\subsubsection{EMQ6}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 5 marks. Give a total of 3 marks for part a) and give a total of 2 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Two infinite parallel planes have equal and opposite uniform surface charge density \( -\sigma \) and \( +\sigma \), where \( \sigma \) is positive. The planes are separated by a distance of \( 15 \) cm, and the potential difference between the plates is \( 120 \) V. 

a) Determine the electric field, \( \underline{E} \), between the plates.

b) An object with charge \( +0.001 \, \text{C} \) and a mass of \( 23 \, \text{g} \) is held at rest at the positive plate, then let go. Determine the acceleration of the object."

Mark Scheme:

"{Q6: (5 marks total)}

a) (3 marks)

[2 marks] Infinite planes have constant electric fields between. Magnitude of potential difference is \( V = Ed \). Magnitude of electric field is calculated 

\[ E = \frac{V}{d} = \frac{120}{0.15} = 800 \, \text{NC}^{-1} \quad (\text{or units } \mathrm{Vm}^{-1}) \]

[1 mark] The electric field is \(E = 800 \hat{n} NC^{-1} \) where \( \hat{n} \) is the unit vector of perpendicular direction from \( +\sigma \) plane to the \( -\sigma \) plane. (Only mention of direction of field required)

b) (2 marks)

[1 mark] Use Newton’s 2nd Law equate to electric force, \( ma = qE \)

[1 mark] \[ a = \frac{qE}{m} = \frac{0.001 \times 800}{0.023} = 34.7826 \, ms^{-2} \]

"

Solution 1: 

"We are given two infinite parallel planes with surface charge densities $-\sigma$ and $+\sigma$, separated by a distance of 15 cm, with a potential difference of 120 V between them. 

\begin{enumerate}
    \item[(a)] \textbf{Determine the electric field, $E$, between the plates.}

    First, recall the formula for the electric field due to an infinite plane of charge. For a plane with surface charge density $\sigma$, the electric field produced by the plane is given by:
    \[
    E = \frac{\sigma}{2\epsilon_0}
    \]
    where $\epsilon_0$ is the permittivity of free space.

    Since there are two planes, one with charge $-\sigma$ and the other with $+\sigma$, the electric fields from the two planes add up in the region between them. The electric field from the negatively charged plane points toward the plane, and the electric field from the positively charged plane points away from the plane. Hence, the total electric field between the plates is:
    \[
    E_{\text{total}} = \frac{\sigma}{\epsilon_0}
    \]

    Next, we use the relationship between electric field and potential difference:
    \[
    E = \frac{\Delta V}{d}
    \]
    where $\Delta V$ is the potential difference between the plates and $d$ is the separation between them. We are given that $\Delta V = 120 \, \text{V}$ and $d = 0.15 \, \text{m}$ (since 15 cm = 0.15 m). Therefore:
    \[
    E = \frac{120 \, \text{V}}{0.15 \, \text{m}} = 800 \, \text{V/m}
    \]

    So, the electric field between the plates is:
    \[
    E = 800 \, \text{V/m}
    \]

    \item[(b)] \textbf{Determine the acceleration of the object.}

    We are given that the object has a charge $q = +0.001 \, \text{C}$ and a mass $m = 23 \, \text{g} = 0.023 \, \text{kg}$. The force on the object due to the electric field is:
    \[
    F = qE
    \]
    Substituting the values for $q = 0.001 \, \text{C}$ and $E = 800 \, \text{V/m}$:
    \[
    F = 0.001 \, \text{C} \times 800 \, \text{V/m} = 0.8 \, \text{N}
    \]

    The acceleration of the object can be found using Newton’s second law:
    \[
    F = ma
    \]
    Solving for acceleration $a$:
    \[
    a = \frac{F}{m} = \frac{0.8 \, \text{N}}{0.023 \, \text{kg}} \approx 34.78 \, \text{m/s}^2
    \]

    Therefore, the acceleration of the object is approximately:
    \[
    a \approx 34.78 \, \text{m/s}^2
    \]
\end{enumerate}"

Solution 2: 

"\begin{enumerate}
    \item[(a)] To determine the electric field between the plates, we begin by noting that the electric field due to an infinite plane with surface charge density $\sigma$ is given by the equation:

    \[
    E = \frac{\sigma}{2\epsilon_0}
    \]
    
    where $\epsilon_0$ is the permittivity of free space, with a value of $\epsilon_0 = 8.85 \times 10^{-12} \, \text{C}^2/\text{N} \cdot \text{m}^2$. Since we have two infinite planes with surface charge densities $+\sigma$ and $-\sigma$, the electric fields due to the two planes add up in the region between them. Therefore, the total electric field between the plates is:

    \[
    E_{\text{total}} = \frac{\sigma}{\epsilon_0}
    \]

    To find $\sigma$, we use the relationship between the electric field and the potential difference between the plates. The electric field is related to the potential difference $V$ and the separation distance $d$ between the plates by:

    \[
    E = \frac{V}{d}
    \]

    Given that the potential difference $V = 120 \, \text{V}$ and the separation $d = 15 \, \text{cm} = 0.15 \, \text{m}$, we can solve for the electric field:

    \[
    E = \frac{120 \, \text{V}}{0.15 \, \text{m}} = 800 \, \text{V/m}
    \]

    Therefore, the electric field between the plates is:

    \[
    E = 800 \, \text{V/m}
    \]

    \item[(b)] Next, we calculate the acceleration of the object. The force on the object due to the electric field is given by Coulomb’s law:

    \[
    F = qE
    \]

    where $q$ is the charge of the object and $E$ is the electric field. The charge of the object is $q = 0.001 \, \text{C}$, and the electric field is $E = 800 \, \text{V/m}$, so the force is:

    \[
    F = (0.001 \, \text{C})(800 \, \text{V/m}) = 0.8 \, \text{N}
    \]

    To find the acceleration, we use Newton’s second law:

    \[
    F = ma
    \]

    where $m$ is the mass of the object and $a$ is the acceleration. The mass of the object is given as $m = 23 \, \text{g} = 0.023 \, \text{kg}$. Solving for the acceleration:

    \[
    a = \frac{F}{m} = \frac{0.8 \, \text{N}}{0.023 \, \text{kg}} = 34.78 \, \text{m/s}^2
    \]

    Therefore, the acceleration of the object is:

    \[
    a \approx 34.8 \, \text{m/s}^2
    \]
\end{enumerate}"

Solution 3:

"\textbf{(a) Determining the electric field, $E$, between the plates:}

We are given two infinite parallel plates with equal and opposite uniform surface charge densities of $-\sigma$ and $+\sigma$, separated by a distance of 15 cm, and a potential difference of 120 V between them. The electric field $E$ between two parallel plates with uniform surface charge densities is given by the following formula:

\[
E = \frac{\sigma}{\epsilon_0}
\]

where $\sigma$ is the surface charge density and $\epsilon_0$ is the permittivity of free space, with a value of

\[
\epsilon_0 = 8.85 \times 10^{-12} \, \text{C}^2/\text{N}\cdot\text{m}^2.
\]

However, since we are not explicitly given the value of $\sigma$, we need to relate the electric field to the potential difference. The potential difference $V$ between the plates is related to the electric field by the equation:

\[
V = E \cdot d,
\]

where $d$ is the distance between the plates, which is 15 cm or 0.15 m. Rearranging for $E$, we get:

\[
E = \frac{V}{d}.
\]

Substituting the known values:

\[
E = \frac{120 \, \text{V}}{0.15 \, \text{m}} = 800 \, \text{V/m}.
\]

Thus, the electric field between the plates is

\[
E = 800 \, \text{V/m}.
\]

\textbf{(b) Determining the acceleration of the object:}

Next, we are asked to determine the acceleration of an object with charge $+0.001 \, \text{C}$ and mass $23 \, \text{g}$ (or $0.023 \, \text{kg}$) that is held at rest at the positive plate and then let go. The force on the object due to the electric field is given by:

\[
F = qE,
\]

where $q = 0.001 \, \text{C}$ is the charge of the object, and $E = 800 \, \text{V/m}$ is the electric field calculated in part (a). Substituting the values:

\[
F = (0.001 \, \text{C}) \cdot (800 \, \text{V/m}) = 0.8 \, \text{N}.
\]

Now, we use Newton's second law to determine the acceleration of the object. The acceleration $a$ is given by:

\[
a = \frac{F}{m},
\]

where $m = 0.023 \, \text{kg}$ is the mass of the object. Substituting the values:

\[
a = \frac{0.8 \, \text{N}}{0.023 \, \text{kg}} \approx 34.78 \, \text{m/s}^2.
\]

Thus, the acceleration of the object is approximately

\[
a \approx 34.78 \, \text{m/s}^2.
\]"

\subsubsection{EMQ7}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 6 marks. Give a total of 1 mark for part a), give a total of 3 marks for part b) and give a total of 2 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"An RLC circuit has a resistor with resistance \( R = 600 \, \Omega \), capacitor with capacitance \( C = 1500 \, \text{pF} \), inductor with inductance \( L = 20 \, \text{mH} \).

a) What is the resonant frequency, \( \omega_{0} \), of the circuit?

b) The circuit is driven with an e.m.f source of the form 

\[ \mathcal{E} = \mathcal{E}_{0} \cos(\omega t) \]

Find an expression for the voltage drop across the inductor, \( \Delta V_{L} \), as a function of \( \mathcal{E}_{0} \), the impedance \( Z \), inductive reactance \( X_{L} \), angular frequency \( \omega \), and some phase shift \( \phi \).

c) Describe the phase relationship between the source e.m.f and voltage across the inductor in the limits of low frequency (\( \omega \ll \omega_{0} \)) and high frequency (\( \omega \gg \omega_{0} \))."

Mark Scheme:

"{Q7: (6 marks total)}

a) (1 mark) 

[1 mark]  resonant frequency is 
\[ \omega_0 = \frac{1}{\sqrt{LC}} = 182574.18 \, \mathrm{rads}^{-1} \]

b) (3 marks)

[1 mark] Impedance Z given by 
\[ Z = \sqrt{R^2 + (X_L - X_C)^2} \]

[1 mark] Current is

\[ I =  \frac{\mathcal{E}_0}{Z} \sin(\omega t - \phi) \]

(allow different trigonometric functions as long as there is a phase factor \( \phi \) ) 

[1 mark] Voltage drop across inductor is 

\[ \Delta{V_L} = I X_L = \frac{\mathcal{E}_0 X_L}{Z} \sin(\omega t - \phi) \]

c) (2 marks)

[2 marks] For \( \omega \gg \omega_{0} \), in phase. For \( \omega \ll \omega_{0} \), antiphase.  

"

Solution 1:

"\textbf{Solution:}

We are given an RLC circuit with a resistor of resistance $R = 600 \, \Omega$, a capacitor with capacitance $C = 1500 \, \mathrm{pF}$, and an inductor with inductance $L = 20 \, \mathrm{mH}$. The circuit is driven by an e.m.f. source of the form $\mathcal{E} = \mathcal{E}_0 \cos(\omega t)$.

\begin{enumerate}
    \item[(a)] To find the resonant frequency $\omega_0$, we use the formula for the resonance condition in an RLC circuit:

    \[
    \omega_0 = \frac{1}{\sqrt{LC}}.
    \]

    Substituting the given values of inductance and capacitance:

    \[
    \omega_0 = \frac{1}{\sqrt{(20 \times 10^{-3} \, \mathrm{H})(1500 \times 10^{-12} \, \mathrm{F})}} = \frac{1}{\sqrt{30 \times 10^{-15}}} = \frac{1}{\sqrt{3 \times 10^{-14}}} = \frac{1}{\sqrt{3}} \times 10^7 \, \mathrm{rad/s}.
    \]

    Therefore, the resonant frequency is:

    \[
    \omega_0 = \sqrt{3} \times 10^7 \, \mathrm{rad/s}.
    \]

    \item[(b)] To find an expression for the voltage drop across the inductor, $\Delta V_L$, as a function of $\mathcal{E}_0$, impedance $Z$, inductive reactance $X_L$, angular frequency $\omega$, and phase shift $\phi$, we start with the equation for the voltage drop across the inductor in an RLC circuit:

    \[
    \Delta V_L = I X_L,
    \]
    where $I$ is the current and $X_L = \omega L$ is the inductive reactance. The current $I$ in the circuit is related to the applied e.m.f. $\mathcal{E}$ and impedance $Z$ by Ohm's law:

    \[
    I = \frac{\mathcal{E}_0}{Z}.
    \]

    The total impedance of the RLC circuit is given by:

    \[
    Z = \sqrt{R^2 + (\omega L - \frac{1}{\omega C})^2}.
    \]

    Substituting the expression for the current into the voltage drop equation, we get:

    \[
    \Delta V_L = \frac{\mathcal{E}_0}{Z} X_L = \frac{\mathcal{E}_0 \omega L}{\sqrt{R^2 + (\omega L - \frac{1}{\omega C})^2}}.
    \]

    The phase shift $\phi$ between the voltage and the current is given by:

    \[
    \tan \phi = \frac{\omega L - \frac{1}{\omega C}}{R}.
    \]

    Therefore, the voltage drop across the inductor is:

    \[
    \Delta V_L = \frac{\mathcal{E}_0 \omega L}{\sqrt{R^2 + (\omega L - \frac{1}{\omega C})^2}} \cdot \cos \phi.
    \]

    \item[(c)] At low frequencies ($\omega \ll \omega_0$), the inductive reactance $X_L = \omega L$ is small compared to the capacitive reactance, and the current leads the voltage by 90 degrees. Therefore, the phase shift is approximately:

    \[
    \phi \approx 0^\circ.
    \]

    At high frequencies ($\omega \gg \omega_0$), the inductive reactance $X_L = \omega L$ becomes much larger than the capacitive reactance, and the current lags the voltage by 90 degrees. Thus, the phase shift is:

    \[
    \phi \approx 90^\circ.
    \]

    Therefore, the phase relationship can be summarized as follows:

    \begin{itemize}
        \item For low frequencies ($\omega \ll \omega_0$), the current and voltage are in phase, with the voltage leading the current by 90 degrees.
        \item For high frequencies ($\omega \gg \omega_0$), the current lags the voltage by 90 degrees.
    \end{itemize}
\end{enumerate}"

Solution 2:

"\textbf{Solution:}

\textbf{(a) Resonant frequency, $\omega_0$:}

The resonant frequency, $\omega_0$, for an RLC circuit is given by the formula
\[
\omega_0 = \frac{1}{\sqrt{LC}}.
\]
We are given the values:
\[
L = 20\,\mathrm{mH} = 20 \times 10^{-3}\,\mathrm{H}, \quad C = 1500\,\mathrm{pF} = 1500 \times 10^{-12}\,\mathrm{F}.
\]
Substituting these values into the formula, we get
\[
\omega_0 = \frac{1}{\sqrt{(20 \times 10^{-3})(1500 \times 10^{-12})}}.
\]
First, calculate the product of $L$ and $C$:
\[
LC = (20 \times 10^{-3})(1500 \times 10^{-12}) = 30 \times 10^{-15} = 3 \times 10^{-14}.
\]
Now, calculate $\omega_0$:
\[
\omega_0 = \frac{1}{\sqrt{3 \times 10^{-14}}} = \frac{1}{\sqrt{3} \times 10^{-7}} = \frac{1}{1.732 \times 10^{-7}} \approx 5.77 \times 10^6\,\mathrm{rad/s}.
\]
Thus, the resonant frequency is approximately
\[
\omega_0 \approx 5.77 \times 10^6\,\mathrm{rad/s}.
\]

\textbf{(b) Voltage drop across the inductor, $\Delta V_L$:}

The circuit is driven by an e.m.f source of the form
\[
\mathcal{E} = \mathcal{E}_0 \cos(\omega t),
\]
where $\omega$ is the angular frequency of the source. The voltage drop across the inductor, $\Delta V_L$, is related to the current in the circuit. The impedance $Z$ of the RLC circuit is given by
\[
Z = \sqrt{R^2 + (X_L - X_C)^2},
\]
where $X_L$ is the inductive reactance and $X_C$ is the capacitive reactance. These are given by
\[
X_L = \omega L, \quad X_C = \frac{1}{\omega C}.
\]
The voltage drop across the inductor is given by
\[
\Delta V_L = I X_L,
\]
where $I$ is the current in the circuit. The current can be found using Ohm's law:
\[
I = \frac{\mathcal{E}_0}{Z}.
\]
Thus, the voltage drop across the inductor becomes
\[
\Delta V_L = \frac{\mathcal{E}_0}{Z} X_L.
\]
Substitute the expression for $Z$:
\[
\Delta V_L = \frac{\mathcal{E}_0 X_L}{\sqrt{R^2 + (X_L - X_C)^2}}.
\]
Now, substituting $X_L = \omega L$ and $X_C = \frac{1}{\omega C}$:
\[
\Delta V_L = \frac{\mathcal{E}_0 \omega L}{\sqrt{R^2 + \left(\omega L - \frac{1}{\omega C}\right)^2}}.
\]
This is the expression for the voltage drop across the inductor as a function of $\mathcal{E}_0$, the impedance $Z$, inductive reactance $X_L$, angular frequency $\omega$, and phase shift $\phi$.

\textbf{(c) Phase relationship between source e.m.f and voltage across the inductor:}

In an RLC circuit, the phase shift $\phi$ between the voltage across the inductor and the source e.m.f depends on the relative values of $X_L$ and $X_C$.

- At \textit{low frequencies} ($\omega \ll \omega_0$), the capacitive reactance $X_C$ dominates, and the circuit behaves mostly like a capacitor. In this case, the voltage across the inductor is in \textit{phase opposition} with the source e.m.f, meaning the voltage across the inductor lags behind the source e.m.f by approximately $90^\circ$.
  
- At \textit{high frequencies} ($\omega \gg \omega_0$), the inductive reactance $X_L$ dominates, and the circuit behaves mostly like an inductor. In this case, the voltage across the inductor is in \textit{phase with} the source e.m.f, meaning the voltage across the inductor leads the source e.m.f by approximately $90^\circ$.

At resonance, where $\omega = \omega_0$, the reactances $X_L$ and $X_C$ are equal, and the phase shift is zero. Therefore, the voltage across the inductor and the source e.m.f are in phase at resonance.

\textbf{Summary:}
- For $\omega \ll \omega_0$, the voltage across the inductor lags behind the source e.m.f by $90^\circ$.
- For $\omega \gg \omega_0$, the voltage across the inductor leads the source e.m.f by $90^\circ$."

Solution 3:

"\textbf{Solution:}

\begin{enumerate}
    \item[(a)] To find the resonant frequency, we use the formula for the resonance frequency $\omega_0$ of an RLC circuit, which is given by:
    \[
    \omega_0 = \frac{1}{\sqrt{LC}}.
    \]
    Here, $L = 20\,\mathrm{mH} = 20 \times 10^{-3}\,\mathrm{H}$ and $C = 1500\,\mathrm{pF} = 1500 \times 10^{-12}\,\mathrm{F}$.

    Substituting the values into the formula:
    \[
    \omega_0 = \frac{1}{\sqrt{(20 \times 10^{-3})(1500 \times 10^{-12})}}.
    \]
    \[
    \omega_0 = \frac{1}{\sqrt{30 \times 10^{-15}}}.
    \]
    \[
    \omega_0 = \frac{1}{\sqrt{3} \times 10^{-7.5}}.
    \]
    \[
    \omega_0 \approx 1.83 \times 10^8 \, \mathrm{rad/s}.
    \]
    
    Thus, the resonant frequency is approximately $\omega_0 \approx 1.83 \times 10^8 \, \mathrm{rad/s}$.

    \item[(b)] The circuit is driven by an e.m.f. of the form:
    \[
    \mathcal{E}(t) = \mathcal{E}_0 \cos(\omega t),
    \]
    where $\omega$ is the angular frequency of the source. To find the voltage drop across the inductor, $\Delta V_L$, we first need to calculate the total impedance $Z$ of the circuit.

    The impedance for a series RLC circuit is:
    \[
    Z = \sqrt{R^2 + (X_L - X_C)^2},
    \]
    where $X_L$ is the inductive reactance and $X_C$ is the capacitive reactance. The inductive reactance is given by:
    \[
    X_L = \omega L,
    \]
    and the capacitive reactance is:
    \[
    X_C = \frac{1}{\omega C}.
    \]

    Substituting the expressions for $X_L$ and $X_C$, the impedance becomes:
    \[
    Z = \sqrt{R^2 + \left( \omega L - \frac{1}{\omega C} \right)^2}.
    \]
    The voltage drop across the inductor is related to the total voltage by Ohm’s law:
    \[
    \Delta V_L = I X_L,
    \]
    where $I$ is the current in the circuit. The current $I$ is related to the source voltage and impedance by:
    \[
    I = \frac{\mathcal{E}_0}{Z}.
    \]
    Therefore, the voltage drop across the inductor is:
    \[
    \Delta V_L = \frac{\mathcal{E}_0 X_L}{Z}.
    \]
    Substituting the expression for $X_L = \omega L$, we get:
    \[
    \Delta V_L = \frac{\mathcal{E}_0 \omega L}{\sqrt{R^2 + (\omega L - 1/\omega C)^2}}.
    \]

    \item[(c)] The phase shift between the source e.m.f. and the voltage across the inductor can be described in the limits of low and high frequency.

    - At low frequencies ($\omega \ll \omega_0$), the inductive reactance $X_L = \omega L$ becomes very small, and the capacitive reactance $X_C = 1/(\omega C)$ dominates. As $\omega$ approaches zero, the circuit behaves like a purely capacitive circuit. In this case, the current leads the voltage across the inductor by $90^\circ$.
    
    - At high frequencies ($\omega \gg \omega_0$), the inductive reactance $X_L = \omega L$ dominates, and the capacitive reactance $X_C$ becomes negligible. In this case, the current lags the voltage across the inductor by $90^\circ$. The voltage across the inductor is in phase with the source voltage in a purely inductive circuit.
\end{enumerate}"

\subsubsection{EMQ8}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 6 marks. Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:
"Two linear magnetic media occupy the half-spaces above and below the \( xy \) plane. The space \( z > 0 \) is occupied by Material 1, with relative permeability \( \mu_{r1} = 1.493 \). The magnetic field \( \underline{B}_{1} \) in this region is spatially uniform and static, with positive \( x \) and \( z \) components, and zero \( y \) component. It is directed at an angle \( \alpha_{1} \) with respect to the positive \( z \) direction. The space \( z < 0 \) is occupied by Material 2, with relative permeability \( \mu_{r2}= 3.012 \). The magnetic field \( \underline{B}_{2} \) in this region is spatially uniform and static, with positive \( x \) and \( z \) components, and zero \( y \) component. It is directed at an angle of \( \alpha_{2} = 45^\circ \) with respect to the positive \( z \) direction. There is no free current flowing anywhere in the system. Using the information above, calculate the value of the angle \( \alpha_{1} \)."

Mark Scheme:

"{Q8: (6 marks total)}

[3 marks] Apply correct boundary conditions on the interface to parallel and perpendicular components of H and B:

\[ H_1 \sin(\alpha_1) = H_2 \sin(\alpha_2) \]
\[ B_1 \cos(\alpha_1) = B_2 \cos(\alpha_2) \]

[1 mark] use \( B = \mu_0 \mu_r H \)

[2 marks] manipulate equations to obtain 

\[ \frac{1}{\mu_{r1}} \tan(\alpha_1) = \frac{1}{\mu_{r2}} \tan(\alpha_2) \]
\[ \alpha_1 = \tan^{-1}\left ( \frac{\mu_{r1}}{\mu_{r2}} \right ) = 26.366^{\circ} \]

"

Solution 1:

"We are tasked with calculating the angle $\alpha_1$ for the magnetic field $\mathbf{B}_1$ in the material 1 region, based on the given information about the magnetic fields and relative permeabilities of the two materials.

Given:
- For $z > 0$, Material 1 has a relative permeability $\mu_{r1} = 1.493$.
- For $z < 0$, Material 2 has a relative permeability $\mu_{r2} = 3.012$.
- The magnetic field $\mathbf{B}_1$ in Material 1 has positive $x$ and $z$ components, and zero $y$ component.
- The magnetic field $\mathbf{B}_2$ in Material 2 has positive $x$ and $z$ components, and zero $y$ component.
- The angle $\alpha_2 = 45^\circ$ for the magnetic field $\mathbf{B}_2$ with respect to the positive $z$ direction.
- The system has no free current flowing.

### Step 1: Use of Boundary Conditions
Since there is no free current, the system is in a static regime, and we can apply Maxwell's boundary conditions at the interface between the two materials, i.e., at $z = 0$. Specifically, the tangential components of the magnetic field $\mathbf{B}$ must be continuous across the boundary, and the normal component of $\mathbf{B}$ must also be continuous.

### Step 2: Magnetic Field Relations in Each Region
The magnetic field $\mathbf{B}$ in each region is related to the magnetic field $\mathbf{H}$ by the permeability:
\[
\mathbf{B} = \mu_0 \mu_r \mathbf{H}
\]
where $\mu_0$ is the permeability of free space, and $\mu_r$ is the relative permeability of the material. The magnetic fields $\mathbf{B}_1$ and $\mathbf{B}_2$ are given in the problem, and we are tasked with finding the angle $\alpha_1$.

### Step 3: Direction of Magnetic Fields
In Material 2, the magnetic field is directed at an angle $\alpha_2 = 45^\circ$ with respect to the positive $z$ direction. This means that the direction of the magnetic field in Material 2 has both $x$ and $z$ components. Using the given angle, we can express the components of $\mathbf{B}_2$:
\[
B_{2x} = B_2 \sin(45^\circ), \quad B_{2z} = B_2 \cos(45^\circ)
\]
Since $B_{2y} = 0$, this tells us that $\mathbf{B}_2$ lies in the $xz$-plane and forms a $45^\circ$ angle with the $z$-axis.

### Step 4: Boundary Condition for the Magnetic Field
The boundary condition at the interface $z = 0$ requires that the tangential components of the magnetic field be continuous across the boundary. The tangential components of the magnetic field are the $x$ and $y$ components, so we must have:
\[
B_{1x} = B_{2x}, \quad B_{1z} = B_{2z}
\]
From the above, the components of the magnetic field in Material 1 can be written as:
\[
B_{1x} = B_1 \sin(\alpha_1), \quad B_{1z} = B_1 \cos(\alpha_1)
\]

### Step 5: Using the Boundary Condition for Continuity of $\mathbf{B}$
From the boundary conditions, we have:
\[
B_1 \sin(\alpha_1) = B_2 \sin(45^\circ), \quad B_1 \cos(\alpha_1) = B_2 \cos(45^\circ)
\]
Using the fact that $\sin(45^\circ) = \cos(45^\circ) = \frac{1}{\sqrt{2}}$, we get:
\[
B_1 \sin(\alpha_1) = B_2 \frac{1}{\sqrt{2}}, \quad B_1 \cos(\alpha_1) = B_2 \frac{1}{\sqrt{2}}
\]

### Step 6: Solving for $\alpha_1$
Dividing the two equations, we find:
\[
\frac{B_1 \sin(\alpha_1)}{B_1 \cos(\alpha_1)} = \frac{B_2 \frac{1}{\sqrt{2}}}{B_2 \frac{1}{\sqrt{2}}}
\]
This simplifies to:
\[
\tan(\alpha_1) = 1
\]
Thus, $\alpha_1 = 45^\circ$.

### Final Answer:
The angle $\alpha_1$ is $45^\circ$."

Solution 2:

"We are tasked with calculating the angle $\alpha_1$ at which the magnetic field $\mathbf{B}_1$ is directed in Material 1, given the spatial and material properties of the system. The problem involves two regions with different magnetic materials, Material 1 and Material 2, and static magnetic fields in each region.

We are given the following information:

- Material 1 occupies the region $z > 0$ and has relative permeability $\mu_{r1} = 1.493$. 
- Material 2 occupies the region $z < 0$ and has relative permeability $\mu_{r2} = 3.012$.
- The magnetic field in Material 1, $\mathbf{B}_1$, has components in the $x$ and $z$ directions, with a zero $y$ component. It is directed at an angle $\alpha_1$ with respect to the $z$ direction.
- The magnetic field in Material 2, $\mathbf{B}_2$, has components in the $x$ and $z$ directions, with zero $y$ component. It is directed at an angle $\alpha_2 = 45^\circ$ with respect to the $z$ direction.
- There is no free current in the system, meaning the fields must satisfy the boundary conditions for a static situation.

### Step 1: Boundary conditions for the magnetic fields
We must apply the boundary conditions for the magnetic field at the interface between the two materials. These conditions are derived from Maxwell’s equations, particularly the absence of free currents.

- The tangential components of the magnetic field must be continuous across the boundary between two regions. This implies:
  \[
  \mathbf{B}_{1t} = \mathbf{B}_{2t}
  \]
  where the subscript $t$ refers to the tangential components of the magnetic field.

- The magnetic field in both regions can be expressed as vectors:
  \[
  \mathbf{B}_1 = B_1 (\cos \alpha_1 \hat{z} + \sin \alpha_1 \hat{x}),
  \]
  \[
  \mathbf{B}_2 = B_2 (\cos \alpha_2 \hat{z} + \sin \alpha_2 \hat{x}).
  \]
  Since $\alpha_2 = 45^\circ$, we know that $\sin \alpha_2 = \cos \alpha_2 = \frac{1}{\sqrt{2}}$.

### Step 2: Matching the tangential components
We now match the tangential components of the magnetic fields at the boundary, which lies at $z = 0$. The components of $\mathbf{B}_1$ and $\mathbf{B}_2$ along the $x$ and $z$ directions must be continuous. Therefore, we equate the $x$-components and the $z$-components of the magnetic fields across the boundary.

- The $z$-component boundary condition:
  \[
  B_1 \cos \alpha_1 = B_2 \cos \alpha_2.
  \]

- The $x$-component boundary condition:
  \[
  B_1 \sin \alpha_1 = B_2 \sin \alpha_2.
  \]

### Step 3: Using the relative permeabilities
The magnetic fields in the two materials are related to the magnetic field strengths $\mathbf{H}_1$ and $\mathbf{H}_2$ by the relative permeabilities $\mu_{r1}$ and $\mu_{r2}$:
\[
\mathbf{B}_1 = \mu_0 \mu_{r1} \mathbf{H}_1, \quad \mathbf{B}_2 = \mu_0 \mu_{r2} \mathbf{H}_2.
\]
Since there is no free current in the system, the magnetic fields must satisfy the boundary conditions for $\mathbf{H}$ as well. These conditions are the same as for $\mathbf{B}$, leading to:
\[
B_1 = \mu_{r1} H_1, \quad B_2 = \mu_{r2} H_2.
\]

### Step 4: Solving for $\alpha_1$
We now use the boundary conditions to solve for $\alpha_1$. From the $x$-component condition, we have:
\[
B_1 \sin \alpha_1 = B_2 \sin \alpha_2.
\]
Substituting the expressions for $B_1$ and $B_2$ in terms of $\mu_{r1}$ and $\mu_{r2}$:
\[
\mu_{r1} H_1 \sin \alpha_1 = \mu_{r2} H_2 \sin \alpha_2.
\]
Next, use the $z$-component boundary condition:
\[
B_1 \cos \alpha_1 = B_2 \cos \alpha_2.
\]
Substitute again for $B_1$ and $B_2$:
\[
\mu_{r1} H_1 \cos \alpha_1 = \mu_{r2} H_2 \cos \alpha_2.
\]
Dividing the $x$-component equation by the $z$-component equation, we obtain:
\[
\frac{\sin \alpha_1}{\cos \alpha_1} = \frac{\mu_{r2}}{\mu_{r1}} \cdot \frac{\sin \alpha_2}{\cos \alpha_2}.
\]
This simplifies to:
\[
\tan \alpha_1 = \frac{\mu_{r2}}{\mu_{r1}} \cdot \tan \alpha_2.
\]
Substituting $\alpha_2 = 45^\circ$, where $\tan 45^\circ = 1$:
\[
\tan \alpha_1 = \frac{\mu_{r2}}{\mu_{r1}}.
\]
Substitute the given values of $\mu_{r1} = 1.493$ and $\mu_{r2} = 3.012$:
\[
\tan \alpha_1 = \frac{3.012}{1.493} \approx 2.02.
\]
Thus:
\[
\alpha_1 = \tan^{-1}(2.02) \approx 63.7^\circ.
\]

### Final Answer:
The angle $\alpha_1$ is approximately $63.7^\circ$."

Solution 3: 

"We are tasked with determining the angle $\alpha_1$ based on the given information about the magnetic fields in two different regions. The scenario involves two magnetic media separated by the $xy$ plane, with different relative permeabilities in each region. The magnetic field vectors in each region are related to each other, and the boundary conditions for the magnetic field must be applied to find the angle $\alpha_1$.

### Step 1: Apply boundary conditions for magnetic fields

The magnetic field must satisfy certain boundary conditions at the interface between two different materials. The relevant boundary conditions for the magnetic field $\mathbf{B}$ at the interface between two media are:

1. The tangential components of the magnetic field $\mathbf{B}$ are continuous across the boundary:
\[
\mathbf{B}_{1t} = \mathbf{B}_{2t}
\]
where $\mathbf{B}_{1t}$ and $\mathbf{B}_{2t}$ are the tangential components of the magnetic field in Materials 1 and 2, respectively.

2. The normal components of the magnetic flux density $\mathbf{B}$ are related to the relative permeabilities of the materials:
\[
\mathbf{B}_{1n} = \mathbf{B}_{2n}
\]

### Step 2: Decompose the magnetic fields into components

For Material 1 (where $z > 0$), the magnetic field $\mathbf{B}_1$ has $x$ and $z$ components, and is directed at an angle $\alpha_1$ with respect to the positive $z$ direction. The components of $\mathbf{B}_1$ can be written as:
\[
\mathbf{B}_1 = B_{1x} \hat{i} + B_{1z} \hat{k}
\]
where $B_{1x} = B_1 \sin \alpha_1$ and $B_{1z} = B_1 \cos \alpha_1$.

For Material 2 (where $z < 0$), the magnetic field $\mathbf{B}_2$ has $x$ and $z$ components, and is directed at an angle $\alpha_2 = 45^\circ$ with respect to the positive $z$ direction. The components of $\mathbf{B}_2$ can be written as:
\[
\mathbf{B}_2 = B_{2x} \hat{i} + B_{2z} \hat{k}
\]
where $B_{2x} = B_2 \sin \alpha_2$ and $B_{2z} = B_2 \cos \alpha_2$.

### Step 3: Apply the continuity of tangential components

The tangential components of the magnetic field $\mathbf{B}$ are continuous at the boundary ($z = 0$). Therefore, the $x$ components of $\mathbf{B}_1$ and $\mathbf{B}_2$ must be equal at the interface:
\[
B_{1x} = B_{2x}
\]
Substituting the expressions for $B_{1x}$ and $B_{2x}$:
\[
B_1 \sin \alpha_1 = B_2 \sin \alpha_2
\]
Since $\alpha_2 = 45^\circ$, we have $\sin \alpha_2 = \frac{\sqrt{2}}{2}$. Thus, the equation becomes:
\[
B_1 \sin \alpha_1 = B_2 \frac{\sqrt{2}}{2}
\]

### Step 4: Apply the continuity of normal components

The normal components of the magnetic field $\mathbf{B}$ are also continuous at the boundary. Therefore, the $z$ components of $\mathbf{B}_1$ and $\mathbf{B}_2$ must be equal at the interface:
\[
B_{1z} = B_{2z}
\]
Substituting the expressions for $B_{1z}$ and $B_{2z}$:
\[
B_1 \cos \alpha_1 = B_2 \cos \alpha_2
\]
Using $\cos \alpha_2 = \frac{\sqrt{2}}{2}$, this becomes:
\[
B_1 \cos \alpha_1 = B_2 \frac{\sqrt{2}}{2}
\]

### Step 5: Relate the fields using the material properties

The relative permeability of Material 1 is $\mu_{r1} = 1.493$, and for Material 2, it is $\mu_{r2} = 3.012$. The magnetic fields in the two materials are related by the following equation:
\[
\mathbf{B}_1 = \mu_{r1} \mathbf{H}_1 \quad \text{and} \quad \mathbf{B}_2 = \mu_{r2} \mathbf{H}_2
\]
Since there is no free current, the magnetic fields are related through the relative permeabilities, but we do not need the explicit values of $\mathbf{H}_1$ and $\mathbf{H}_2$ for our angle calculation.

### Step 6: Solve for $\alpha_1$

Now we have two equations involving the magnetic field components and the angles:
1. \( B_1 \sin \alpha_1 = B_2 \frac{\sqrt{2}}{2} \)
2. \( B_1 \cos \alpha_1 = B_2 \frac{\sqrt{2}}{2} \)

By dividing the first equation by the second:
\[
\frac{B_1 \sin \alpha_1}{B_1 \cos \alpha_1} = \frac{B_2 \frac{\sqrt{2}}{2}}{B_2 \frac{\sqrt{2}}{2}}
\]
This simplifies to:
\[
\tan \alpha_1 = 1
\]
Thus, the angle $\alpha_1$ is:
\[
\alpha_1 = 45^\circ
\]
Therefore, the angle $\alpha_1$ is $45^\circ$."

\subsubsection{EMQ9}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 15 marks. Give a total of 7 marks for part a), give a total of 6 marks for part b) and give a total of 2 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"A transverse electromagnetic wave propagating in vacuum has an electric field which has complex representation:

\[ \underline{E}(\underline{r}, t) = \underline{E}_{0} \exp (i(\underline{k} \cdot \underline{r} - \omega t)) \]

With real valued vectors \( \underline{E}_{0} \), \( \underline{k} \).

a) Apply Faraday’s Law and show magnetic intensity of the wave has magnitude given by 

\[ H(\underline{r}, t) = \sqrt{\frac{\epsilon_{0}}{\mu_{0}}} E(\underline{r}, t) \]

b) Show the Poynting flux averaged over one period of oscillation is given by 

\[ \left< \underline{N} \right> = \frac{1}{2} \sqrt{\frac{\mu_{0}}{\epsilon_{0}}} (H_{0})^{2} \hat{\underline{k}} \]

where \( H_{0} \) denotes the amplitude of the magnetic intensity.
    
c) The pressure exerted by radiation has intensity \( I \) on a perfect planar reflector has the value \( \frac{2I}{c} \), where \( c \) is the speed of light. Show that when the angle of incidence of radiation is \( \alpha \), the radiation pressure becomes \( 2I\cos^{2}{\alpha}/{c} \)."

Mark Scheme:

"{Q9: (15 marks total)}

a) (7 marks)

[1 mark] assume \underline{H}(\underline{r}, t) has solution

\[ \underline{H}(\underline{r}, t) = \underline{H}_0 \exp (i(\underline{k} \cdot \underline{r} - \omega t)) \]

[2 marks] apply Faraday's law, noting relation between \underline{B} and \underline{H} in vacuum, \( \underline{B} = \mu_0 \underline{H} \)

\[ \nabla \times \underline{E} = - \frac{\partial \underline{B}}{\partial t} = - \mu_0 \frac{\partial \underline{H}}{\partial t}  \]
[2 marks] noting \( \nabla \times \underline{E} = i \underline{k} \times \underline{E} \) or by direct calculation. \( - \mu_0 \frac{\partial \underline{H}}{\partial t} = - i \mu_0 \omega \underline{H} \)

\[ i \underline{k} \times \underline{E} = i \mu_0 \omega \underline{H} \]

[2 marks] Taking magnitude of equation, noting \( \omega / k = c = 1 / \sqrt{\epsilon_0 \mu_0} \), simplify to desired result 

\[ H(\underline{r}, t) = \sqrt{\frac{\epsilon_{0}}{\mu_{0}}} E(\underline{r}, t) \]

b) (6 marks)

[1 mark] Use the expression of the Poynting vector, \underline{N}, in a vacuum. 

\[ \underline{N} = \frac{1}{\mu_0} ( \underline{E} \times \underline{B} ) = \underline{E} \times \underline{H} \]

[3 marks] Take the real part of \( \underline{E} \) and \( \underline{H} \), using expression found in a), simplify and give 

\[ \underline{N} = \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \cos^2(\underline{k} \cdot \underline{r} - \omega t) \hat{k} \]

[2 marks] 

\[ \langle \underline{N} \rangle = \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \langle \cos^2(\underline{k} \cdot \underline{r} - \omega t) \rangle \hat{k} \]

\[ \langle \underline{N} \rangle = \frac{1}{2}\sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \hat{k} \]

Must justify the time average of the cosine term as 1/2.

c) (2 marks) 

[1 mark] One factor of cosine from component of momentum perpendicular to surface

[1 mark] Another factor of cosine due to the surface area incident increasing. 



"

Solution 1:

"\textit{Part (a):} Apply Faraday’s Law and show magnetic intensity of the wave has magnitude given by
\[
H(r, t) = \sqrt{\frac{\epsilon_0}{\mu_0}} E(r, t)
\]

We begin by recalling Faraday’s Law of induction in differential form:
\[
\nabla \times E = - \frac{\partial B}{\partial t}.
\]
Since we are considering a plane wave in vacuum, the electric field \(E(r,t)\) has the form:
\[
E(r, t) = E_0 \exp(i(k \cdot r - \omega t)),
\]
where \(E_0\) is the amplitude of the electric field, \(k\) is the wave vector, and \(\omega\) is the angular frequency.

To find the magnetic field, we use the fact that the electric and magnetic fields in a vacuum for a plane wave are perpendicular to each other and to the direction of propagation. Therefore, the magnetic field \(B\) is related to the electric field by Maxwell's equations.

First, the curl of \(E\) is:
\[
\nabla \times E = i k \times E_0 \exp(i(k \cdot r - \omega t)),
\]
and from Faraday’s Law, we know:
\[
\nabla \times E = - \frac{\partial B}{\partial t}.
\]
We can thus write the magnetic field as:
\[
B(r, t) = \frac{1}{c} (k \times E(r, t)),
\]
where \(c = \frac{1}{\sqrt{\mu_0 \epsilon_0}}\) is the speed of light.

Using the relation between the electric and magnetic fields in a plane wave:
\[
B(r, t) = \frac{1}{c} (k \times E(r, t)) = \sqrt{\frac{\mu_0}{\epsilon_0}} E(r, t).
\]

Therefore, the magnetic intensity, defined as the magnitude of the magnetic field, is:
\[
H(r, t) = \sqrt{\frac{\epsilon_0}{\mu_0}} E(r, t).
\]

\textit{Part (b):} Show the Poynting flux averaged over one period of oscillation is given by
\[
\langle N \rangle = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \hat{k}.
\]

The Poynting vector \( \mathbf{N} \) represents the energy flux of the electromagnetic wave and is given by:
\[
\mathbf{N} = \frac{1}{\mu_0} E(r,t) \times H(r,t).
\]
Substituting the expressions for \(E(r,t)\) and \(H(r,t)\), we have:
\[
\mathbf{N}(r, t) = \frac{1}{\mu_0} \left( E_0 \exp(i(k \cdot r - \omega t)) \right) \times \left( \sqrt{\frac{\epsilon_0}{\mu_0}} E_0 \exp(i(k \cdot r - \omega t)) \right).
\]
Since the cross product of two parallel vectors is zero, the magnitude of the Poynting vector is directed along the propagation vector \(\hat{k}\) and is proportional to \(E_0^2\).

The time-averaged Poynting flux, \( \langle N \rangle \), is given by:
\[
\langle N \rangle = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \hat{k},
\]
which is the desired result. The factor of \(\frac{1}{2}\) arises from the time average of the oscillating sinusoidal functions.

\textit{Part (c):} The pressure exerted by radiation has intensity \(I\) on a perfect planar reflector and has the value
\[
\frac{2I}{c},
\]
where \(c\) is the speed of light. Show that when the angle of incidence of radiation is \(\alpha\), the radiation pressure becomes
\[
2I \cos^2 \alpha / c.
\]

The intensity \(I\) of an electromagnetic wave is given by the time-averaged Poynting vector:
\[
I = \langle N \rangle = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} E_0^2.
\]
For a perfect planar reflector, the intensity is reflected, and the pressure on the reflector is twice the radiation pressure due to the reflection. Therefore, the total force per unit area (the radiation pressure) is proportional to \(2I/c\).

Now, when the radiation strikes the reflector at an angle \(\alpha\), the pressure is reduced by a factor of \(\cos^2 \alpha\) due to the projection of the electric field on the plane normal to the direction of propagation. Hence, the radiation pressure becomes:
\[
\text{Pressure} = \frac{2I}{c} \cos^2 \alpha.
\]
This completes the solution."

Solution 2: 

"\textbf{(a) Apply Faraday’s Law and show the magnetic intensity of the wave has magnitude given by \(H(r, t)\).}

We are given the electric field of the transverse electromagnetic wave as
\[
E(r, t) = E_0 \exp(i (k \cdot r - \omega t)).
\]
To find the magnetic intensity, we apply Faraday’s Law, which in differential form is
\[
\nabla \times E = -\frac{\partial B}{\partial t}.
\]
In vacuum, the magnetic field \(\mathbf{B}(r, t)\) and the electric field \(\mathbf{E}(r, t)\) are related by
\[
\mathbf{B}(r, t) = \frac{1}{c} \hat{k} \times \mathbf{E}(r, t),
\]
where \(c = \frac{1}{\sqrt{\epsilon_0 \mu_0}}\) is the speed of light and \(\hat{k}\) is the unit vector in the direction of propagation.

Taking the curl of \(\mathbf{E}(r, t)\), we get
\[
\nabla \times \mathbf{E}(r, t) = i k E_0 \exp(i (k \cdot r - \omega t)) \hat{k}.
\]
From Faraday’s Law, we know that
\[
\nabla \times \mathbf{E}(r, t) = -\frac{\partial \mathbf{B}}{\partial t}.
\]
Substituting this into the equation, we obtain
\[
-i k E_0 \exp(i (k \cdot r - \omega t)) \hat{k} = - \frac{\partial}{\partial t} \mathbf{B}(r, t).
\]
The time derivative of \(\mathbf{B}(r, t)\) is
\[
\frac{\partial \mathbf{B}(r, t)}{\partial t} = \frac{\partial}{\partial t} \left( \frac{1}{c} \hat{k} \times \mathbf{E}(r, t) \right) = -\frac{\omega}{c} \hat{k} \times \mathbf{E}(r, t).
\]
Thus, equating both sides gives
\[
i k E_0 \exp(i (k \cdot r - \omega t)) \hat{k} = \frac{\omega}{c} \hat{k} \times \mathbf{E}(r, t).
\]
Solving for the magnetic field, we find
\[
\mathbf{B}(r, t) = \frac{1}{c} \hat{k} \times \mathbf{E}(r, t) = \frac{1}{c} \hat{k} \times E_0 \exp(i (k \cdot r - \omega t)) = \frac{1}{c} E_0 \exp(i (k \cdot r - \omega t)) \hat{k} \times \hat{k}.
\]
Thus, the magnetic field amplitude is related to the electric field by the magnitude
\[
H(r, t) = \sqrt{\frac{\epsilon_0}{\mu_0}} E(r, t).
\]

\textbf{(b) Show the Poynting flux averaged over one period of oscillation is given by the expression.}

The Poynting vector represents the energy flux of the electromagnetic wave and is given by
\[
\mathbf{S}(r, t) = \frac{1}{\mu_0} \mathbf{E}(r, t) \times \mathbf{B}(r, t).
\]
Substituting \(\mathbf{B}(r, t) = \frac{1}{c} \hat{k} \times \mathbf{E}(r, t)\), we get
\[
\mathbf{S}(r, t) = \frac{1}{\mu_0} \mathbf{E}(r, t) \times \left( \frac{1}{c} \hat{k} \times \mathbf{E}(r, t) \right).
\]
Using the vector identity \(\mathbf{A} \times (\mathbf{B} \times \mathbf{C}) = (\mathbf{A} \cdot \mathbf{C}) \mathbf{B} - (\mathbf{A} \cdot \mathbf{B}) \mathbf{C}\), we find
\[
\mathbf{S}(r, t) = \frac{1}{\mu_0 c} \left( E_0^2 \hat{k} \right) \exp(i (k \cdot r - \omega t)).
\]
Now, to find the time-averaged Poynting flux, we calculate the time average of the magnitude of the Poynting vector. The time average of \( \mathbf{E}(r, t) \) squared gives
\[
\langle \mathbf{E}^2(r, t) \rangle = \frac{E_0^2}{2}.
\]
Thus, the time-averaged Poynting flux is given by
\[
\langle N \rangle = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \hat{k}.
\]

\textbf{(c) Show the radiation pressure becomes \(2I \cos^2 \alpha / c\).}

The intensity of the radiation \(I\) is related to the time-averaged Poynting flux by
\[
I = \langle N \rangle = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2.
\]
For a perfect planar reflector, the pressure exerted by the radiation on the reflector is related to the intensity by
\[
P = \frac{2I}{c}.
\]
The angle of incidence is \(\alpha\), and the radiation pressure is modified by the factor \(\cos^2 \alpha\), yielding the result
\[
P = 2I \cos^2 \alpha / c.
\]
This shows that the radiation pressure is proportional to the cosine square of the angle of incidence.

\qed"

Solution 3: 

"\textbf{Solution:}

The problem involves a transverse electromagnetic wave propagating in vacuum. The given electric field has the form:
\[
E(r, t) = E_0 \exp(i (k \cdot r - \omega t))
\]
where \( E_0 \) is the amplitude of the electric field, \( k \) is the wave vector, \( \omega \) is the angular frequency, and \( r \) and \( t \) represent position and time, respectively. The electric field is a complex representation, but we will treat the physical field as the real part of this expression.

\textbf{Part (a):} Apply Faraday's Law and show that the magnetic intensity of the wave has magnitude given by:
\[
H(r, t) = \sqrt{\frac{\epsilon_0}{\mu_0}} E(r, t)
\]
Faraday's Law in its differential form is:
\[
\nabla \times E = - \frac{\partial B}{\partial t}
\]
We will assume a plane wave solution, with the magnetic field \( B \) being perpendicular to both the electric field \( E \) and the wave vector \( k \), consistent with the nature of transverse electromagnetic waves.

For a plane wave propagating in the direction of the wave vector \( k \), the magnetic field \( B \) satisfies the relation \( B(r, t) \) in terms of \( E(r, t) \). From Maxwell's equations, we know that the magnitude of the magnetic field \( B_0 \) is related to the electric field \( E_0 \) through the relation:
\[
B_0 = \frac{E_0}{c}
\]
where \( c \) is the speed of light. Since \( B = \mu_0 H \), we can write the magnetic intensity as:
\[
H_0 = \frac{E_0}{\sqrt{\mu_0 \epsilon_0}}
\]
Thus, the magnetic intensity is:
\[
H(r, t) = \sqrt{\frac{\epsilon_0}{\mu_0}} E(r, t)
\]
This completes part (a).

\textbf{Part (b):} Show the Poynting flux averaged over one period of oscillation is given by:
\[
\langle N \rangle = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \hat{k}
\]
The Poynting flux \( N \) is given by the cross product of the electric field \( E \) and the magnetic field \( H \), divided by the permeability of free space:
\[
N = \frac{1}{\mu_0} \mathbf{E} \times \mathbf{H}
\]
For a plane wave, the vectors \( \mathbf{E} \) and \( \mathbf{H} \) are perpendicular to each other and to the wave vector \( \mathbf{k} \). Therefore, the magnitude of the Poynting vector is:
\[
|N| = \frac{1}{\mu_0} |E| |H| = \frac{1}{\mu_0} E_0 H_0
\]
Using the relation \( H_0 = \frac{E_0}{\sqrt{\mu_0 \epsilon_0}} \), we substitute to obtain:
\[
|N| = \frac{1}{\mu_0} E_0 \frac{E_0}{\sqrt{\mu_0 \epsilon_0}} = \frac{E_0^2}{\mu_0 \sqrt{\mu_0 \epsilon_0}}
\]
The Poynting flux is a time-dependent quantity, but we are asked for the time-averaged value. Since \( E_0 \) oscillates sinusoidally with time, the time average of \( E_0^2 \) is \( \frac{1}{2} E_0^2 \). Therefore, the time-averaged Poynting flux is:
\[
\langle N \rangle = \frac{1}{2} \frac{E_0^2}{\mu_0 \sqrt{\mu_0 \epsilon_0}} = \frac{1}{2} \sqrt{\frac{\mu_0}{\epsilon_0}} (H_0)^2 \hat{k}
\]
This completes part (b).

\textbf{Part (c):} The pressure exerted by radiation on a perfect planar reflector has intensity \( I \) and is given by:
\[
\frac{2I}{c}
\]
where \( I \) is the intensity of the radiation and \( c \) is the speed of light. The pressure on a perfect reflector can be derived from the energy flux. The energy flux is proportional to the intensity \( I \), and for a perfect reflector, the reflected wave contributes an additional factor of 2. Therefore, the radiation pressure on the reflector is:
\[
P = \frac{2I}{c}
\]
Now, when the angle of incidence of radiation is \( \alpha \), the effective intensity that contributes to the pressure is reduced by a factor of \( \cos^2 \alpha \) due to the angle of incidence. Therefore, the radiation pressure becomes:
\[
P = \frac{2I \cos^2 \alpha}{c}
\]
This completes part (c)."

\subsubsection{EMQ10}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 7 marks. Give a total of 4 marks for part a), give a total of 2 marks for part b) and give a total of 1 mark for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Consider 2 inertial reference frames \( S \) and \( S' \). The frames are aligned in such a way that origins of both frames coincide at time zero within both frames (\( t = t' = 0 \)). The frame \( S' \) moves with velocity \( v \) in the \( x \) direction as seen by \( S \). The transformation of electric and magnetic fields from frame \( S \) to \( S' \) is given by 

\[
\begin{aligned}
E'_{x} &= E_{x}, & E'_{y} &= \gamma(E_{y} - vB_{z}), & E'_{z} &= \gamma(E_{z} + vB_{y}) \\
B'_{x} &= B_{x}, & B'_{y} &= \gamma(B_{y} + \frac{v}{c^{2}} E_{z}), & B'_{z} &= \gamma(B_{z} - \frac{v}{c^{2}} E_{y})
\end{aligned}
\]

a) Using the transformations given, show that the scalar product, \( (\underline{E} \cdot \underline{B}) \), is invariant under transformation.

b) A plane electromagnetic wave observed in the reference frame \( S \) propagates in a vacuum along the \( x \) direction. In frame \( S \), it is represented by the Cartesian representation: 

\[ \underline{E} = E_{0} \hat{\underline{y}} \exp(i(kx - \omega t)) \]

with \( E_{0} \), a real value denoting the amplitude, \( k \) is the wavevector, and \( \omega \) the angular frequency. \( (kx - \omega t) \) defines the phase of the wave in frame \( S \). Show that this phase in reference frame \( S' \) is written \( (k'x' - \omega' t') \), where primed coordinates correspond to the frame \( S' \) and

\[ \omega' = \gamma(\omega - kv), \quad k' = \gamma(k - \frac{\omega v}{c^{2}}) \]

c) Examine the relationship between the wave frequencies \( \omega \) and \( \omega' \) in the limit \( v/c \) approaches \( 0 \). What physical phenomenon does this represent? Justify your answer."

Mark Scheme:

"{Q10: (7 marks total)}

a) (4 marks)

[1 mark] Substitute \( (\underline{E'} \cdot \underline{B'}) \) with transformation terms

\begin{align*}
    \underline{E'} \cdot \underline{B'} & = E'_x B'_x + E'_y B'_y + E'_z B'_z \\ & = E_{x} B_{x} + \gamma(E_{y} - vB_{z}) \gamma(B_{y} + \frac{v}{c^{2}} E_{z}) + \gamma(E_{z} + vB_{y}) \gamma(B_{z} - \frac{v}{c^{2}} E_{y})
\end{align*}

[3 marks] Explicitly expand and simplify algebra, final simplified term should be \( E_x B_x + E_y B_y + E_z B_z = \underline{E} \cdot \underline{B} \) (Converting to dot product form not required)

b) (2 marks)

[1 mark] Use of inverse lorentz transformations to replace \( x \) and \( t \)

[1 mark] Rearrange and simplification to give required relations.

c) (1 mark)

[1 mark] Represents Doppler effect. Gives some reasoning involving frequency shift.

"

Solution 1: 

"(a) To show that the scalar product \( \mathbf{E} \cdot \mathbf{B} \) is invariant under the Lorentz transformation, we begin by considering the transformation properties of the electric and magnetic fields. The electric and magnetic fields in frame \( S \) transform to frame \( S' \) as:

\[
E'_x = E_x, \quad E'_y = \gamma (E_y - v B_z), \quad E'_z = \gamma (E_z + v B_y)
\]
\[
B'_x = B_x, \quad B'_y = \gamma \left( B_y + \frac{v}{c^2} E_z \right), \quad B'_z = \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

Now, the scalar product \( \mathbf{E} \cdot \mathbf{B} \) in frame \( S \) is given by:

\[
\mathbf{E} \cdot \mathbf{B} = E_x B_x + E_y B_y + E_z B_z
\]

We need to show that this is invariant under the transformation. We can calculate the scalar product in frame \( S' \) using the transformed fields. In \( S' \), the scalar product becomes:

\[
\mathbf{E'} \cdot \mathbf{B'} = E'_x B'_x + E'_y B'_y + E'_z B'_z
\]

Substituting the transformations into this expression:

\[
E'_x = E_x, \quad B'_x = B_x
\]
\[
E'_y = \gamma (E_y - v B_z), \quad B'_y = \gamma \left( B_y + \frac{v}{c^2} E_z \right)
\]
\[
E'_z = \gamma (E_z + v B_y), \quad B'_z = \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

Thus, the scalar product in frame \( S' \) becomes:

\[
\mathbf{E'} \cdot \mathbf{B'} = E_x B_x + \gamma (E_y - v B_z) \left( B_y + \frac{v}{c^2} E_z \right) + \gamma (E_z + v B_y) \left( B_z - \frac{v}{c^2} E_y \right)
\]

Expanding the terms:

\[
\mathbf{E'} \cdot \mathbf{B'} = E_x B_x + \gamma \left( E_y B_y + \frac{v}{c^2} E_y E_z - v B_z B_y - \frac{v^2}{c^2} B_z E_z \right)
+ \gamma \left( E_z B_z - \frac{v}{c^2} E_z E_y + v B_y B_z - \frac{v^2}{c^2} B_y E_y \right)
\]

Combining the terms:

\[
\mathbf{E'} \cdot \mathbf{B'} = E_x B_x + \gamma \left( E_y B_y + E_z B_z \right)
= \mathbf{E} \cdot \mathbf{B}
\]

Thus, we have shown that \( \mathbf{E} \cdot \mathbf{B} \) is invariant under Lorentz transformations.

(b) To show the transformation of the phase of the plane electromagnetic wave in frame \( S' \), we start with the wave expression in frame \( S \):

\[
\mathbf{E} = E_0 \hat{y} \exp \left( i(kx - \omega t) \right)
\]

The phase of the wave is given by \( \varphi_S = kx - \omega t \). To transform this to frame \( S' \), we use the Lorentz transformation for the space and time coordinates:

\[
x' = \gamma (x - vt), \quad t' = \gamma \left( t - \frac{v}{c^2} x \right)
\]

Substituting these into the expression for the phase:

\[
\varphi_S = kx - \omega t = k \gamma (x' + vt') - \omega \gamma \left( t' + \frac{v}{c^2} x' \right)
\]

Simplifying:

\[
\varphi_S = \gamma \left( kx' - \omega t' + \frac{kv}{c^2} x' - \frac{\omega v}{c^2} t' \right)
\]

\[
\varphi_S = \gamma \left( k' x' - \omega' t' \right)
\]

where:

\[
k' = \gamma \left( k - \frac{\omega v}{c^2} \right), \quad \omega' = \gamma (\omega - kv)
\]

Thus, the phase in frame \( S' \) is \( \varphi_{S'} = k'x' - \omega' t' \), as required.

(c) In the limit where \( v / c \to 0 \), we have \( \gamma \to 1 \), and the transformations for \( k' \) and \( \omega' \) reduce to:

\[
k' = k, \quad \omega' = \omega
\]

This implies that the wave frequencies and wavevector are the same in both frames, which is expected for non-relativistic velocities. The physical phenomenon this represents is the classical Doppler effect, where there is no shift in the frequency of the wave when the relative velocity between the frames is zero."

Solution 2:

"\textbf{Solution:}

\textbf{(a)} We are given the transformations for the electric and magnetic fields under a Lorentz boost in the \( x \)-direction:

\[
E'_x = E_x, \quad E'_y = \gamma (E_y - v B_z), \quad E'_z = \gamma (E_z + v B_y)
\]
\[
B'_x = B_x, \quad B'_y = \gamma \left( B_y + \frac{v}{c^2} E_z \right), \quad B'_z = \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

We need to show that the scalar product \( \mathbf{E} \cdot \mathbf{B} \) is invariant under the transformation. The scalar product in the original frame \( S \) is given by:

\[
\mathbf{E} \cdot \mathbf{B} = E_x B_x + E_y B_y + E_z B_z
\]

In the primed frame \( S' \), the transformed fields are:

\[
\mathbf{E}' \cdot \mathbf{B}' = E'_x B'_x + E'_y B'_y + E'_z B'_z
\]

Substituting the transformations into this expression:

\[
E'_x = E_x, \quad B'_x = B_x
\]
\[
E'_y = \gamma (E_y - v B_z), \quad B'_y = \gamma \left( B_y + \frac{v}{c^2} E_z \right)
\]
\[
E'_z = \gamma (E_z + v B_y), \quad B'_z = \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

Thus, the scalar product in the primed frame is:

\[
\mathbf{E}' \cdot \mathbf{B}' = E_x B_x + \gamma (E_y - v B_z) \gamma \left( B_y + \frac{v}{c^2} E_z \right) + \gamma (E_z + v B_y) \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

Expanding the terms:

\[
= E_x B_x + \gamma^2 \left[ (E_y B_y + \frac{v}{c^2} E_y B_z) - v B_z B_y + (E_z B_z - \frac{v}{c^2} E_z B_y) \right]
\]

Simplifying the terms:

\[
= E_x B_x + \gamma^2 \left( E_y B_y + E_z B_z - v (B_z B_y - B_z B_y) \right)
\]

Since \( B_z B_y - B_z B_y = 0 \), the expression simplifies to:

\[
= E_x B_x + \gamma^2 \left( E_y B_y + E_z B_z \right)
\]

Thus, we have:

\[
\mathbf{E}' \cdot \mathbf{B}' = E_x B_x + E_y B_y + E_z B_z = \mathbf{E} \cdot \mathbf{B}
\]

Therefore, the scalar product \( \mathbf{E} \cdot \mathbf{B} \) is invariant under the Lorentz transformation.

\textbf{(b)} A plane electromagnetic wave in frame \( S \) is given by:

\[
\mathbf{E} = E_0 \hat{y} \exp \left( i (k x - \omega t) \right)
\]

We are tasked with showing that the phase \( (k x - \omega t) \) in reference frame \( S' \) is written as \( (k' x' - \omega' t') \). First, we note the transformations of the coordinates:

\[
x' = \gamma (x - v t), \quad t' = \gamma \left( t - \frac{v}{c^2} x \right)
\]

The phase in frame \( S \) is:

\[
\Phi = k x - \omega t
\]

Substituting for \( x \) and \( t \) from the Lorentz transformation:

\[
\Phi' = k' x' - \omega' t' = k' \gamma (x - v t) - \omega' \gamma \left( t - \frac{v}{c^2} x \right)
\]

We can match terms with the original phase, so we get the following relations for the transformed wavevector and frequency:

\[
k' = \gamma \left( k - \frac{\omega v}{c^2} \right), \quad \omega' = \gamma (\omega - k v)
\]

Thus, the phase in frame \( S' \) is indeed \( (k' x' - \omega' t') \).

\textbf{(c)} In the limit \( v / c \to 0 \), we can expand the expressions for \( k' \) and \( \omega' \) as:

\[
k' \approx k, \quad \omega' \approx \omega
\]

Thus, in the limit \( v / c \to 0 \), the wave frequencies \( \omega \) and \( \omega' \) become the same. This corresponds to the classical Doppler effect, where there is no frequency shift in the case of no relative motion between the frames. Hence, in this limit, the wave frequencies are the same in both frames, representing the situation of no relative motion between observers.

\textbf{End of Solution.}"

Solution 3:

"\textbf{Part (a):} 

We are asked to show that the scalar product \( \mathbf{E} \cdot \mathbf{B} \) is invariant under the given Lorentz transformation.

The Lorentz transformation for the electric and magnetic fields from frame \( S \) to frame \( S' \) is given by:

\[
E'_x = E_x, \quad E'_y = \gamma (E_y - v B_z), \quad E'_z = \gamma (E_z + v B_y)
\]
\[
B'_x = B_x, \quad B'_y = \gamma \left( B_y + \frac{v}{c^2} E_z \right), \quad B'_z = \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

The scalar product \( \mathbf{E} \cdot \mathbf{B} \) is defined as:

\[
\mathbf{E} \cdot \mathbf{B} = E_x B_x + E_y B_y + E_z B_z
\]

Now, the transformed scalar product in the new frame \( S' \) is:

\[
\mathbf{E'} \cdot \mathbf{B'} = E'_x B'_x + E'_y B'_y + E'_z B'_z
\]

Substitute the transformed components of \( \mathbf{E} \) and \( \mathbf{B} \):

\[
\mathbf{E'} \cdot \mathbf{B'} = E_x B_x + \gamma (E_y - v B_z) \gamma \left( B_y + \frac{v}{c^2} E_z \right) + \gamma (E_z + v B_y) \gamma \left( B_z - \frac{v}{c^2} E_y \right)
\]

Expanding the terms:

\[
= E_x B_x + \gamma^2 \left[ (E_y - v B_z) (B_y + \frac{v}{c^2} E_z) + (E_z + v B_y) (B_z - \frac{v}{c^2} E_y) \right]
\]

Simplifying the terms inside the brackets:

\[
= E_x B_x + \gamma^2 \left[ E_y B_y + \frac{v}{c^2} E_y E_z - v B_z B_y - \frac{v^2}{c^2} B_z E_z + E_z B_z - \frac{v}{c^2} E_z E_y + v B_y B_z - \frac{v^2}{c^2} B_y E_y \right]
\]

Noticing that the mixed terms cancel out due to symmetry (i.e., the terms involving \( E_y B_z \) and \( E_z B_y \) are equal in magnitude but opposite in sign), we obtain:

\[
= E_x B_x + \gamma^2 \left[ E_y B_y + E_z B_z \right]
\]

Since \( \mathbf{E} \cdot \mathbf{B} = E_x B_x + E_y B_y + E_z B_z \), we have shown that:

\[
\mathbf{E'} \cdot \mathbf{B'} = \mathbf{E} \cdot \mathbf{B}
\]

Thus, the scalar product \( \mathbf{E} \cdot \mathbf{B} \) is indeed invariant under the Lorentz transformation.

\textbf{Part (b):}

In this part, we are asked to show that the phase of a plane electromagnetic wave in frame \( S' \) is given by \( (k' x' - \omega' t') \).

In the reference frame \( S \), the electric field of a plane electromagnetic wave is given by:

\[
\mathbf{E} = E_0 \hat{y} \exp \left( i (k x - \omega t) \right)
\]

where \( E_0 \) is the amplitude, \( k \) is the wavevector, and \( \omega \) is the angular frequency.

The transformations of the space and time coordinates between frames \( S \) and \( S' \) are:

\[
x' = \gamma (x - vt), \quad t' = \gamma \left( t - \frac{v}{c^2} x \right)
\]

We now substitute these transformations into the phase \( (k x - \omega t) \) in frame \( S \):

\[
k x - \omega t = k \left( \gamma (x' + v t') \right) - \omega \left( \gamma \left( t' + \frac{v}{c^2} x' \right) \right)
\]

Simplifying this expression:

\[
= \gamma \left( k x' + \omega v t' - \omega t' - \frac{\omega v}{c^2} x' \right)
\]

\[
= \gamma \left( k - \frac{\omega v}{c^2} \right) x' - \gamma (\omega - kv) t'
\]

Thus, the phase in frame \( S' \) is:

\[
k' x' - \omega' t' \quad \text{with} \quad k' = \gamma \left( k - \frac{\omega v}{c^2} \right) \quad \text{and} \quad \omega' = \gamma (\omega - k v)
\]

\textbf{Part (c):}

In the limit \( v/c \to 0 \), the Lorentz factor \( \gamma \to 1 \). Therefore, the transformed frequencies are:

\[
\omega' \approx \omega \quad \text{and} \quad k' \approx k
\]

This indicates that in the non-relativistic limit, the wave frequencies in both frames are the same, and no relativistic Doppler shift occurs. This result corresponds to the classical case where there is no relative motion between the observer and the wave source, meaning the frequency observed in both frames is identical.

Thus, the physical phenomenon represented is the absence of any relativistic effects at low velocities (i.e., the Doppler shift vanishes as \( v/c \to 0 \)).

\qed"

%--------------------------------------------------------------


\subsection{QM}

Quantum Mechanics Prompts

\subsubsection{QMQ1}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 5 marks. Give a total of 3 marks for part a) and give a total of 2 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem: 

"a) Define the commutator of 2 operators, \( \hat{P} \) and \( \hat{Q} \). Define what it means if \( \hat{P} \) and \( \hat{Q} \) are compatible operators. Explain what does compatibility imply about their commutator?

b) Suppose that \( \hat{P} \) and \( \hat{Q} \) are not compatible. For a given system, \( \hat{P} \) is first measured giving a value \( p \), then \( \hat{Q} \) is measured giving the value \( q \). If \( \hat{P} \) is then measured again, what can be said about the possible results of the measurement and why?"

Mark Scheme:

"{Q1: (5 marks total)}

a) (3 marks)

[1 mark]The commutator is given as  \( \left[\hat{P}, \hat{Q} \right] = \hat{P}\hat{Q} - \hat{Q}\hat{P} \)

[1 mark] Compatible operators are operators which share a common, complete set of eigenfunctions

[1 mark] Compatible operators imply that the commutator is equal to 0.

b) (2 marks)

[1 mark] The initial \( p \) value is not necessarily the same as the final \( p \) value measured

[1 mark] Since the operators are incompatible, the wavefunction will collapse to different eigenstates after each operator has been applied to it and the measured values may change

"

Solution 1:

"(a) The commutator of two operators $\hat{P}$ and $\hat{Q}$ is defined as $[\hat{P}, \hat{Q}] = \hat{P}\hat{Q} - \hat{Q}\hat{P}$. If $\hat{P}$ and $\hat{Q}$ are compatible operators, they share a common set of eigenstates. Compatibility implies that measurements of $\hat{P}$ and $\hat{Q}$ do not disturb one another, and the operators commute, i.e., $[\hat{P}, \hat{Q}] = 0$. This condition ensures that both operators can be simultaneously diagonalized, allowing precise knowledge of their eigenvalues in a single state. In contrast, if $[\hat{P}, \hat{Q}] \neq 0$, the operators are incompatible, and their measurements are affected by the uncertainty principle.

(b) If $\hat{P}$ and $\hat{Q}$ are not compatible, measuring $\hat{P}$ first and obtaining a value $p$ prepares the system in an eigenstate of $\hat{P}$. Measuring $\hat{Q}$ next disturbs this state, collapsing it into an eigenstate of $\hat{Q}$ corresponding to the value $q$. If $\hat{P}$ is then measured again, the system may no longer yield the original value $p$, as the measurement of $\hat{Q}$ has altered the state. Instead, the outcome of the second measurement of $\hat{P}$ will follow the probability distribution determined by the projection of the $\hat{Q}$ eigenstate onto the eigenstates of $\hat{P}$. This is a direct consequence of the non-commutativity of $\hat{P}$ and $\hat{Q}$, reflecting the fundamental limits of simultaneous knowledge imposed by quantum mechanics."

Solution 2: 

"(a) The commutator of two operators $\hat{P}$ and $\hat{Q}$ is defined as:
\[
[\hat{P}, \hat{Q}] = \hat{P}\hat{Q} - \hat{Q}\hat{P}.
\]
Operators $\hat{P}$ and $\hat{Q}$ are said to be compatible if they commute, i.e., $[\hat{P}, \hat{Q}] = 0$. This implies that they share a common set of eigenstates, meaning the measurement of one operator does not disturb the measurement of the other. Compatibility implies that the physical quantities represented by these operators can be simultaneously measured with certainty, as the uncertainty principle does not impose restrictions in this case.

(b) If $\hat{P}$ and $\hat{Q}$ are not compatible, then $[\hat{P}, \hat{Q}] \neq 0$, and they do not share a common set of eigenstates. When $\hat{P}$ is first measured, the system collapses into an eigenstate of $\hat{P}$ with eigenvalue $p$. A subsequent measurement of $\hat{Q}$ generally changes the state of the system because the eigenstates of $\hat{Q}$ are not eigenstates of $\hat{P}$. Therefore, when $\hat{P}$ is measured again after measuring $\hat{Q}$, the outcome may differ from the initial value $p$. The possible results of the second measurement of $\hat{P}$ are determined by the overlap of the new state (after the $\hat{Q}$ measurement) with the eigenstates of $\hat{P}$. This uncertainty arises due to the non-commutativity of $\hat{P}$ and $\hat{Q}$, as described by the Heisenberg uncertainty principle."

Solution 3:

"(a) The commutator of two operators $\hat{P}$ and $\hat{Q}$ is defined as:
\[
[\hat{P}, \hat{Q}] = \hat{P}\hat{Q} - \hat{Q}\hat{P}.
\]
Two operators are said to be compatible if they commute, i.e., $[\hat{P}, \hat{Q}] = 0$. Compatibility implies that the two operators share a common set of eigenstates. If $\hat{P}$ and $\hat{Q}$ are compatible, measuring one observable does not disturb the eigenstate of the other, ensuring that both observables can be precisely determined simultaneously. In physical terms, this implies no uncertainty relationship exists between them with regard to simultaneous measurement.

(b) If $\hat{P}$ and $\hat{Q}$ are not compatible, then $[\hat{P}, \hat{Q}] \neq 0$, meaning the measurement of one observable disturbs the eigenstate of the other. Suppose $\hat{P}$ is first measured, giving an eigenvalue $p$. This places the system in an eigenstate of $\hat{P}$. Measuring $\hat{Q}$ then projects the state into an eigenstate of $\hat{Q}$, disrupting the previous eigenstate of $\hat{P}$. When $\hat{P}$ is measured again, the result may not necessarily be $p$ because the measurement of $\hat{Q}$ has altered the system's state. Instead, the outcome of the second measurement of $\hat{P}$ is probabilistic, governed by the overlap between the eigenstates of $\hat{Q}$ and $\hat{P}$. This behavior is a direct consequence of the operators' non-commutativity and the fundamental uncertainty principle."

\subsubsection{QMQ2}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 4 marks. Give a total of 2 marks for part a) and give a total of 2 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem: 

"a) Quantum mechanical operators are Hermitian. What mathematical property do the eigenvalues of a Hermitian operator have? What is the physical interpretation of its eigenvalues?

b) Explain what is meant if it is stated that the set of eigenfunctions \( \{\phi_{n}\} \) of a Hermitian operator is orthonormal."

Mark Scheme:

"{Q2: (4 marks total)}

a) (2 marks)

[1 mark] The eigenvalues are real 

[1 mark] Eigenvalues are possible values obtained by measurement of the given quantity associated with the operator. 

b) (2 marks)

[1 mark] The set of eigenfunctions is orthogonal - the inner products are non-vanishing if and only if the eigenfunctions chosen in the inner product are the same. 

[1 mark] Each eigenstate is normalized such that their inner product with themselves gives 1. 

\[ \left< \phi_i | \phi_j \right> = \delta_{ij} \]

"

Solution 1: 

"(a) A Hermitian operator $\hat{A}$ satisfies the property $\hat{A} = \hat{A}^\dagger$, where $\hat{A}^\dagger$ is the adjoint (conjugate transpose) of $\hat{A}$. Mathematically, this ensures that all eigenvalues $\lambda$ of $\hat{A}$ are real. This can be shown by considering the eigenvalue equation $\hat{A}\psi = \lambda\psi$ and taking the inner product $\langle \psi | \hat{A} \psi \rangle$. Using the Hermitian property, we find $\langle \psi | \hat{A} \psi \rangle = \langle \psi | \lambda \psi \rangle = \lambda \langle \psi | \psi \rangle$, where $\langle \psi | \psi \rangle > 0$ for a nonzero eigenvector $\psi$. The result implies that $\lambda$ is real.

Physically, Hermitian operators correspond to observable quantities in quantum mechanics, such as energy, momentum, or position. Their real eigenvalues represent the measurable outcomes of these quantities. For example, the energy levels of a quantum system correspond to the eigenvalues of the Hamiltonian operator, which is Hermitian. \\

(b) If the set of eigenfunctions $\{\phi_n\}$ of a Hermitian operator is orthonormal, it means that $\langle \phi_n | \phi_m \rangle = \delta_{nm}$, where $\delta_{nm}$ is the Kronecker delta, equal to $1$ if $n = m$ and $0$ otherwise. This property arises from the fact that Hermitian operators are diagonalizable and their eigenfunctions corresponding to distinct eigenvalues are orthogonal. 

The orthonormality condition ensures that the eigenfunctions form a complete and independent basis for the Hilbert space. In practice, this allows the expansion of any state function $\psi$ in terms of these eigenfunctions as $\psi = \sum_n c_n \phi_n$, where $c_n = \langle \phi_n | \psi \rangle$. This property is essential for expressing quantum states and analyzing physical systems."

Solution 2: 

"(a) Hermitian operators, by definition, satisfy the condition $\hat{H} = \hat{H}^\dagger$, where $\hat{H}^\dagger$ represents the adjoint (conjugate transpose) of the operator $\hat{H}$. This ensures that the eigenvalues $\lambda$ of a Hermitian operator are always real. Mathematically, if $\hat{H} \phi = \lambda \phi$, taking the inner product with $\phi$ gives $\langle \phi | \hat{H} \phi \rangle = \lambda \langle \phi | \phi \rangle$. Since $\langle \phi | \phi \rangle > 0$ for non-zero $\phi$, $\lambda$ must be real to preserve equality. 

Physically, the eigenvalues of a Hermitian operator correspond to observable quantities in quantum mechanics, such as energy, momentum, or angular momentum. Observables must be real-valued to align with measurements in the real world, which is guaranteed by the Hermitian property of the operators representing them.

(b) The statement that the set of eigenfunctions $\{\phi_n\}$ of a Hermitian operator is orthonormal means that the eigenfunctions satisfy two key properties. First, orthogonality: $\langle \phi_m | \phi_n \rangle = 0$ for $m \neq n$. This arises because Hermitian operators are diagonalizable, and their eigenfunctions corresponding to distinct eigenvalues are orthogonal. Second, normalization: $\langle \phi_n | \phi_n \rangle = 1$, which ensures the eigenfunctions have unit magnitude. 

The orthonormality condition, $\langle \phi_m | \phi_n \rangle = \delta_{mn}$, where $\delta_{mn}$ is the Kronecker delta, implies that the eigenfunctions form a complete, linearly independent basis in the Hilbert space. This property is essential in quantum mechanics, as any state can be expanded as a linear combination of these eigenfunctions, facilitating the computation of physical quantities."

Solution 3:

"\begin{enumerate}
    \item[(a)] A Hermitian operator $\hat{H}$ in quantum mechanics satisfies $\hat{H}^\dagger = \hat{H}$, where $\hat{H}^\dagger$ is the adjoint (complex conjugate transpose) of $\hat{H}$. Mathematically, this ensures that all eigenvalues $\lambda$ of $\hat{H}$ are real. To see this, consider the eigenvalue equation $\hat{H} \phi = \lambda \phi$, where $\phi$ is an eigenfunction. Taking the inner product with $\phi$, we have $\langle \phi | \hat{H} \phi \rangle = \lambda \langle \phi | \phi \rangle$. Since $\hat{H}$ is Hermitian, $\langle \phi | \hat{H} \phi \rangle = \langle \hat{H} \phi | \phi \rangle^* = \lambda^* \langle \phi | \phi \rangle$. Equating these, $\lambda = \lambda^*$, proving that $\lambda$ is real. Physically, the eigenvalues of a Hermitian operator correspond to measurable quantities (observables) in quantum mechanics. For example, the eigenvalues of the Hamiltonian operator represent the possible energy levels of a quantum system.

    \item[(b)] The statement that the set of eigenfunctions $\{\phi_n\}$ of a Hermitian operator is orthonormal means that these eigenfunctions satisfy the orthonormality condition:
    \[
    \langle \phi_n | \phi_m \rangle = 
    \begin{cases} 
    1 & \text{if } n = m, \\
    0 & \text{if } n \neq m.
    \end{cases}
    \]
    This implies that $\phi_n$ and $\phi_m$ are orthogonal when $n \neq m$, and each eigenfunction is normalized to unity. Orthonormality is a direct consequence of the Hermiticity of the operator: for distinct eigenvalues $\lambda_n$ and $\lambda_m$, we have $\langle \phi_n | \hat{H} \phi_m \rangle = \lambda_m \langle \phi_n | \phi_m \rangle$ and $\langle \hat{H} \phi_n | \phi_m \rangle = \lambda_n^* \langle \phi_n | \phi_m \rangle$. Since $\lambda_n = \lambda_n^*$ (real) and $\lambda_n \neq \lambda_m$, it follows that $\langle \phi_n | \phi_m \rangle = 0$. Normalization is typically imposed to simplify calculations and ensure physical interpretability.
\end{enumerate}"


\subsubsection{QMQ3}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 9 marks. Give a total of 4 marks for part a) and give a total of 5 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem: 

"For one dimension, an operator \( \hat{Q} \) is Hermitian if and only if 

\[ \int_{-\infty}^{\infty} f^{*}\hat{Q}g \, dx = \int_{-\infty}^{\infty} g(\hat{Q}f)^{*} \, dx \]

where \( f(x) \), \( g(x) \) are well-behaved functions which vanish at infinity, \( ^{*} \) denotes the complex conjugate. 

a) Use the definition above to determine whether \( \hat{p} = -i\hbar\frac{d}{dx} \) is a Hermitian operator 

b) Determine whether \( \hat{Q} = \frac{d^{2}}{dx^{2}} \) is a Hermitian operator."

Mark Scheme:

"{Q3: (9 marks total)}

a) (4 marks)

[2 marks] Substitution of \( \hat{p} \) into the LHS or RHS of definition, Integrate by parts, removing vanishings terms since functions vanish at infinity 

[2 marks] for simplifying the remaining integral, arranging into correct form showing LHS equals RHS, stating  \( \hat{p} \) is Hermitian e.g. 

\[ \int g (\hat{Q}f)^{*} \, dx \]

b) (5 marks)

[2 marks] for substituting the operator into the integral and applying integration by parts, removing vanishing terms due to limits and well behaved functions

[1 mark] for applying integration by parts again, again removing vanishing terms after evaluating limits

[2 marks] for arranging into the correct form, stating \(\hat{Q}\) is Hermitian. 

"

Solution 1: 

"
To determine whether the given operators are Hermitian, we use the definition of a Hermitian operator:
\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty g (\hat{Q} f)^* \, dx
\]
for all well-behaved functions $f(x)$ and $g(x)$ that vanish at infinity. Here, $*$ denotes the complex conjugate.

\textbf{a) Determine whether $\hat{p} = -i\hbar \frac{d}{dx}$ is Hermitian.}

Let $\hat{p} = -i\hbar \frac{d}{dx}$. We start with the left-hand side of the Hermitian condition:
\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty f^* \left(-i\hbar \frac{d g}{dx}\right) \, dx.
\]
Now, integrate by parts. Let $u = f^*$ and $dv = -i\hbar \frac{d g}{dx} dx$, so $du = \frac{df^*}{dx} dx$ and $v = -i\hbar g$. The integration by parts formula is $\int u \, dv = uv - \int v \, du$, giving:
\[
\int_{-\infty}^\infty f^* \left(-i\hbar \frac{d g}{dx}\right) \, dx = \Big[ -i\hbar f^* g \Big]_{-\infty}^\infty + \int_{-\infty}^\infty i\hbar g \frac{d f^*}{dx} \, dx.
\]
The boundary term $\Big[ -i\hbar f^* g \Big]_{-\infty}^\infty$ vanishes because $f(x)$ and $g(x)$ are well-behaved functions that vanish at infinity. Thus, we have:
\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty i\hbar g \frac{d f^*}{dx} \, dx.
\]
Taking the complex conjugate of the right-hand side of the Hermitian condition:
\[
\left(\int_{-\infty}^\infty g \hat{p} f \, dx \right)^* = \left(\int_{-\infty}^\infty g \left(-i\hbar \frac{d f}{dx}\right) \, dx \right)^* = \int_{-\infty}^\infty i\hbar g \frac{d f^*}{dx} \, dx.
\]
Since both sides are equal, $\hat{p} = -i\hbar \frac{d}{dx}$ is a Hermitian operator. \hfill [4 marks]

\textbf{b) Determine whether $\hat{Q} = \frac{d^2}{dx^2}$ is Hermitian.}

Let $\hat{Q} = \frac{d^2}{dx^2}$. Consider the left-hand side of the Hermitian condition:
\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx.
\]
Integrate by parts twice. First, let $u = f^*$ and $dv = \frac{d^2 g}{dx^2} dx$, so $du = \frac{df^*}{dx} dx$ and $v = \frac{dg}{dx}$. Applying the integration by parts formula:
\[
\int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx = \Big[ f^* \frac{dg}{dx} \Big]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{df^*}{dx} \frac{dg}{dx} \, dx.
\]
The boundary term $\Big[ f^* \frac{dg}{dx} \Big]_{-\infty}^\infty$ vanishes because $f(x)$ and $g(x)$ vanish at infinity along with their derivatives. For the remaining term, integrate by parts again:
\[
\int_{-\infty}^\infty \frac{df^*}{dx} \frac{dg}{dx} \, dx = \Big[ \frac{df^*}{dx} g \Big]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]
The boundary term $\Big[ \frac{df^*}{dx} g \Big]_{-\infty}^\infty$ also vanishes. Thus, we have:
\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = -\int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]
Taking the complex conjugate of the right-hand side of the Hermitian condition:
\[
\left(\int_{-\infty}^\infty g \hat{Q} f \, dx \right)^* = \left(\int_{-\infty}^\infty g \frac{d^2 f}{dx^2} \, dx \right)^* = -\int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]
Since both sides are equal, $\hat{Q} = \frac{d^2}{dx^2}$ is a Hermitian operator. \hfill [5 marks]"

Solution 2:

"
To determine whether the given operators are Hermitian, we apply the definition of a Hermitian operator:

\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty g (\hat{Q} f)^* \, dx
\]

where $f(x)$ and $g(x)$ are well-behaved functions vanishing at infinity, and $*$ denotes the complex conjugate.

\textbf{Part (a): Determine whether $\hat{p} = -i\hbar \frac{d}{dx}$ is Hermitian.}

Let $\hat{p} = -i\hbar \frac{d}{dx}$. The left-hand side of the Hermitian condition is:

\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty f^* \left(-i\hbar \frac{d g}{dx}\right) dx.
\]

Using integration by parts, where $u = f^*$ and $v' = \frac{dg}{dx}$, we get:

\[
\int_{-\infty}^\infty f^* \frac{d g}{dx} \, dx = \left[ f^* g \right]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{d f^*}{dx} g \, dx.
\]

Since $f(x)$ and $g(x)$ vanish at infinity, the boundary term $\left[ f^* g \right]_{-\infty}^\infty = 0$. Thus:

\[
\int_{-\infty}^\infty f^* \frac{d g}{dx} \, dx = - \int_{-\infty}^\infty \frac{d f^*}{dx} g \, dx.
\]

Substituting this into the expression for $\int_{-\infty}^\infty f^* \hat{p} g \, dx$, we have:

\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = -i\hbar \left(- \int_{-\infty}^\infty \frac{d f^*}{dx} g \, dx \right) = \int_{-\infty}^\infty g \left(-i\hbar \frac{d f}{dx}\right)^* dx.
\]

Since $(-i\hbar \frac{d}{dx})^* = -i\hbar \frac{d}{dx}$ (as $i$ and $\hbar$ are constants), we find:

\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty g (\hat{p} f)^* dx.
\]

Thus, $\hat{p}$ satisfies the Hermitian condition and is therefore a Hermitian operator. 

\textbf{Part (b): Determine whether $\hat{Q} = \frac{d^2}{dx^2}$ is Hermitian.}

Let $\hat{Q} = \frac{d^2}{dx^2}$. The left-hand side of the Hermitian condition is:

\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx.
\]

Using integration by parts twice, first let $u = f^*$ and $v' = \frac{d^2 g}{dx^2}$:

\[
\int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx = \left[ f^* \frac{d g}{dx} \right]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{d f^*}{dx} \frac{d g}{dx} \, dx.
\]

The boundary term $\left[ f^* \frac{d g}{dx} \right]_{-\infty}^\infty = 0$ because $f(x)$ and $g(x)$ vanish at infinity. For the remaining integral, apply integration by parts again with $u = \frac{d f^*}{dx}$ and $v' = \frac{d g}{dx}$:

\[
\int_{-\infty}^\infty \frac{d f^*}{dx} \frac{d g}{dx} \, dx = \left[ \frac{d f^*}{dx} g \right]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]

The boundary term $\left[ \frac{d f^*}{dx} g \right]_{-\infty}^\infty = 0$ because $f(x)$ and $g(x)$ vanish at infinity along with their derivatives. Substituting back, we have:

\[
\int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx = - \int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]

Rewriting the right-hand side:

\[
- \int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx = \int_{-\infty}^\infty g \left(\frac{d^2 f}{dx^2}\right)^* dx.
\]

Thus:

\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty g (\hat{Q} f)^* dx.
\]

This shows that $\hat{Q} = \frac{d^2}{dx^2}$ satisfies the Hermitian condition and is therefore a Hermitian operator."

Solution 3:

"
To determine whether the given operators are Hermitian, we use the definition provided:

\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty g (\hat{Q} f)^* \, dx,
\]
where \( f(x) \) and \( g(x) \) are well-behaved functions that vanish at infinity, and \( * \) denotes the complex conjugate. 

---

a) Determine whether \( \hat{p} = -i\hbar \frac{d}{dx} \) is Hermitian:

Let \( \hat{p} = -i\hbar \frac{d}{dx} \). We compute both sides of the Hermitian condition for arbitrary functions \( f(x) \) and \( g(x) \). First, calculate the left-hand side:

\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty f^* \left(-i\hbar \frac{d}{dx} g\right) dx.
\]

Using integration by parts:

\[
\int_{-\infty}^\infty f^* \left(-i\hbar \frac{d}{dx} g\right) dx = \left[ -i\hbar f^* g \right]_{-\infty}^\infty + \int_{-\infty}^\infty i\hbar \frac{d}{dx} f^* \cdot g \, dx.
\]

The boundary term \( \left[ -i\hbar f^* g \right]_{-\infty}^\infty \) vanishes because \( f(x) \) and \( g(x) \) vanish at infinity. Thus:

\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty i\hbar \frac{d f^*}{dx} g \, dx.
\]

Next, calculate the complex conjugate of the right-hand side:

\[
\int_{-\infty}^\infty g (\hat{p} f)^* \, dx = \int_{-\infty}^\infty g \left(-i\hbar \frac{df}{dx}\right)^* dx = \int_{-\infty}^\infty g \cdot i\hbar \frac{d f^*}{dx} \, dx.
\]

These two expressions are identical:

\[
\int_{-\infty}^\infty f^* \hat{p} g \, dx = \int_{-\infty}^\infty g (\hat{p} f)^* \, dx.
\]

Therefore, \( \hat{p} = -i\hbar \frac{d}{dx} \) is a Hermitian operator. 

---

b) Determine whether \( \hat{Q} = \frac{d^2}{dx^2} \) is Hermitian:

Let \( \hat{Q} = \frac{d^2}{dx^2} \). We compute both sides of the Hermitian condition. First, calculate the left-hand side:

\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx.
\]

Using integration by parts twice, let \( u = f^* \) and \( v' = \frac{d^2 g}{dx^2} \):

\[
\int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx = \left[ f^* \frac{dg}{dx} \right]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{d f^*}{dx} \frac{dg}{dx} \, dx.
\]

The first boundary term \( \left[ f^* \frac{dg}{dx} \right]_{-\infty}^\infty \) vanishes because \( f(x) \) and \( g(x) \) (and their derivatives) vanish at infinity. For the second term, use integration by parts again:

\[
\int_{-\infty}^\infty \frac{d f^*}{dx} \frac{dg}{dx} \, dx = \left[ \frac{d f^*}{dx} g \right]_{-\infty}^\infty - \int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]

The boundary term \( \left[ \frac{d f^*}{dx} g \right]_{-\infty}^\infty \) vanishes, leaving:

\[
\int_{-\infty}^\infty f^* \frac{d^2 g}{dx^2} \, dx = \int_{-\infty}^\infty \frac{d^2 f^*}{dx^2} g \, dx.
\]

Next, calculate the complex conjugate of the right-hand side:

\[
\int_{-\infty}^\infty g (\hat{Q} f)^* \, dx = \int_{-\infty}^\infty g \left(\frac{d^2 f}{dx^2}\right)^* dx = \int_{-\infty}^\infty g \frac{d^2 f^*}{dx^2} \, dx.
\]

These two expressions are identical:

\[
\int_{-\infty}^\infty f^* \hat{Q} g \, dx = \int_{-\infty}^\infty g (\hat{Q} f)^* \, dx.
\]

Therefore, \( \hat{Q} = \frac{d^2}{dx^2} \) is a Hermitian operator.
"

\subsubsection{QMQ4}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 20 marks. Give up to 3 marks for part a), give up to 3 marks for part b), give up to 4 marks for part c), give up to 4 marks for part d) and up to 6 marks for part e). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Let \( \hat{L_{x}} \) be the x-component of the angular momentum operator. Let \( \hat{X} \), \( \hat{Y} \), \( \hat{Z} \) be the x, y, z components of the position operator respectively and \( \hat{P_{x}} \), \( \hat{P_{y}} \), \( \hat{P_{z}} \)  be the x, y, z components of the momentum operator respectively. Derive the following commutator relations:

a) \( \left[ \hat{L_{x}}, \hat{X} \right] = 0 \)

b) \( \left[ \hat{L_{x}}, \hat{P_{x}} \right] = 0 \)

c) \( \left[ \hat{L_{x}}, \hat{Y} \right] = i \hbar \hat{Z} \)

d) \( \left[ \hat{L_{x}}, \hat{P_{y}} \right] = i \hbar \hat{P_{z}} \)

e) \( \left[ \hat{L_{x}}, \hat{P}^{2} \right] = 0 \) "

Mark Scheme:

"{Q4: (20 marks total) }

a) (3 marks)

[1 mark] for expanding the commutation relation and correctly substituting \( \hat{L_{x}} \) in, 

\[ 
\left[ \hat{L_{x}}, \hat{X} \right] = (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{X} - \hat{X}(\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})
\]

[1 mark] For stating that since \( \hat{X} \) commutes with \( \hat{Z} \), \( \hat{P_{z}} \), \( \hat{P_{y}} \), and \( Y \), the second term can be rearranged to \( (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{X} \).

[1 mark] \( (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{X} - (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{X} = 0 \) as required

b) (3 marks)

[1 mark] for expanding the commutation relation and correctly substituting \( \hat{L_{x}} \) in, \([\hat{L_{x}}, \hat{P_{x}}] = (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{P_{x}} - \hat{P_{x}}(\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\)

[1 mark] for stating that since \( \hat{P_{x}} \) commutes with \( \hat{Z} \), \( \hat{P_{z}} \), \( \hat{P_{y}} \), and \( Y \), the second term can be rearranged to \( (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{P_{x}} \)

[1 mark] \((\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{P_{x}}  - (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{P_{x}}\)  = 0 as required

c)  (4 marks)

[1 mark] for expanding the commutation relation and correctly substituting \( \hat{L_{x}} \) in, \([\hat{L_{x}}, \hat{Y}] = (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{Y} - \hat{Y}(\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\)

[1 mark] For changing the order of commuting terms to get \( \hat{Y}\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}}\hat{Y} - \hat{Y}\hat{Y}\hat{P_{z}} + \hat{Z}\hat{Y}\hat{P_{y}} \)

[2 marks] for correctly factorizing and simplifying: \( \hat{Z}[\hat{Y}, \hat{P_{y}}] \), using the fact that \( [\hat{Y}, \hat{P_{y}}] = -i\hbar \) and substituting into the answer to yield \( [\hat{L}_x, \hat{Y}] = i\hbar \hat{Z} \)

d) (4 marks)

[1 mark] for expanding the commutation relation and correctly substituting \( \hat{L_{x}} \) in, \([\hat{L_{x}}, \hat{P_{y}}] = (\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\hat{P_{y}} - \hat{P_{y}}(\hat{Y}\hat{P_{z}} - \hat{Z}\hat{P_{y}})\)

[1 mark] For changing the order of commuting terms to get \( \hat{P_{z}}\hat{Y}\hat{P_{y}} - \hat{P_{z}}\hat{P_{y}}\hat{Y} - \hat{Z}\hat{P_{y}}\hat{P_{y}} + \hat{Z}\hat{P_{y}}\hat{P_{y}} \)

[2 marks] for correctly factorizing and simplifying: \( \hat{P_{z}}[\hat{Y}, \hat{P_{y}}] \), using the fact \( [\hat{Y}, \hat{P_{y}}] = -i\hbar \hat{P_{z}} \) to yield \( [\hat{L}_x, \hat{P_{y}}] = i\hbar\hat{P_{z}} \)

e) (6 marks)

[1 mark] for rewriting the commutation as \( [\hat{L_{x}}, \hat{P^{2}}] = [\hat{L_{x}}, \hat{P}_{x}^{2} ] + [\hat{L_{x}}, \hat{P}_{y}^{2}] + [\hat{L_{x}}, \hat{P}_{z}^{2}] \)

[1 mark] Applying the identity \( [\hat{A}, \hat{B^{2}}] = [\hat{A}, \hat{B}]\hat{B} + \hat{B}[\hat{A},\hat{B}] \) to yield 

\[ [\hat{L_{x}}, \hat{P^{2}}] = [\hat{L_{x}}, \hat{P_{x}}]\hat{P_{x}} + \hat{P_{x}}[\hat{L_x}, \hat{P_{x}}] + [\hat{L_{x}}, \hat{P_{y}}]\hat{P_{y}} + \hat{P_{y}}[\hat{L_{x}}, \hat{P_{y}}] + [\hat{L_{x}}, \hat{P_{z}}]\hat{P_{z}} + \hat{P_{z}}[\hat{L_{x}}, \hat{P_{z}}] \]

Note that the identity \( [\hat{A}, \hat{B^{2}}] = [\hat{A}, \hat{B}]\hat{B} + \hat{B}[\hat{A},\hat{B}] \) does NOT need to be explicitly stated.

[3 marks] For \( [\hat{L_{x}}, \hat{P_{z}}] = -i\hbar\hat{P_{y}} \)

[1 mark] For simplifying \( [\hat{L_{x}}, \hat{P^{2}}] = (i\hbar\hat{P_{z}}\hat{P_{y}}) + (i\hbar\hat{P_{y}}\hat{P_{z}}) - (i\hbar\hat{P_{y}}\hat{P_{z}}) - (i\hbar\hat{P_{z}}\hat{P_{y}}) = 0 \)

"

Solution 1: 

"
\textbf{Solution:}

To solve these commutator relations, we start by recalling the definition of the angular momentum operator. The x-component of the angular momentum operator is given by:
\[
\hat{L}_x = \hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y
\]
where $\hat{Y}, \hat{Z}$ are position operators and $\hat{P}_y, \hat{P}_z$ are the corresponding momentum operators.

The general commutator relation is:
\[
[\hat{L}_x, \hat{A}] = [\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{A}]
\]
Using the commutator rule for products, $[\hat{AB}, \hat{C}] = \hat{A}[\hat{B}, \hat{C}] + [\hat{A}, \hat{C}]\hat{B}$, we calculate each part.

---

a) \textbf{To derive} $\left[\hat{L}_x, \hat{X}\right]$:  
Since $\hat{X}$ commutes with $\hat{Y}, \hat{Z}$ and all momentum operators $\hat{P}_y, \hat{P}_z$, we have:
\[
[\hat{L}_x, \hat{X}] = [\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{X}] = 0
\]
This is because $\hat{X}$ is independent of the terms in $\hat{L}_x$.  

\textbf{Answer:}
\[
\left[\hat{L}_x, \hat{X}\right] = 0
\]

---

b) \textbf{To derive} $\left[\hat{L}_x, \hat{P}_x\right]$:  
The operator $\hat{P}_x$ also commutes with $\hat{Y}, \hat{Z}$ and the momentum operators $\hat{P}_y, \hat{P}_z$. Thus:
\[
[\hat{L}_x, \hat{P}_x] = [\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{P}_x] = 0
\]

\textbf{Answer:}
\[
\left[\hat{L}_x, \hat{P}_x\right] = 0
\]

---

c) \textbf{To derive} $\left[\hat{L}_x, \hat{Y}\right]$:  
Using the definition of $\hat{L}_x$, we write:
\[
[\hat{L}_x, \hat{Y}] = [\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{Y}]
\]
The first term $[\hat{Y}\hat{P}_z, \hat{Y}] = \hat{Y}[\hat{P}_z, \hat{Y}] + [\hat{Y}, \hat{Y}]\hat{P}_z = 0$. The second term $[\hat{Z}\hat{P}_y, \hat{Y}]$ reduces to:
\[
[\hat{Z}, \hat{Y}]\hat{P}_y + \hat{Z}[\hat{P}_y, \hat{Y}]
\]
Since $[\hat{Z}, \hat{Y}] = 0$ and $[\hat{P}_y, \hat{Y}] = -i\hbar$, we get:
\[
[\hat{L}_x, \hat{Y}] = i\hbar \hat{Z}
\]

\textbf{Answer:}
\[
\left[\hat{L}_x, \hat{Y}\right] = i\hbar \hat{Z}
\]

---

d) \textbf{To derive} $\left[\hat{L}_x, \hat{P}_y\right]$:  
Using the definition of $\hat{L}_x$, we write:
\[
[\hat{L}_x, \hat{P}_y] = [\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{P}_y]
\]
The first term $[\hat{Y}\hat{P}_z, \hat{P}_y] = \hat{Y}[\hat{P}_z, \hat{P}_y] + [\hat{Y}, \hat{P}_y]\hat{P}_z$. Since $[\hat{P}_z, \hat{P}_y] = 0$ and $[\hat{Y}, \hat{P}_y] = i\hbar$, this simplifies to:
\[
\hat{Y} \cdot 0 + i\hbar \hat{P}_z = i\hbar \hat{P}_z
\]
The second term $[\hat{Z}\hat{P}_y, \hat{P}_y] = 0$. Thus:
\[
[\hat{L}_x, \hat{P}_y] = i\hbar \hat{P}_z
\]

\textbf{Answer:}
\[
\left[\hat{L}_x, \hat{P}_y\right] = i\hbar \hat{P}_z
\]

---

e) \textbf{To derive} $\left[\hat{L}_x, \hat{P}^2\right]$:  
The operator $\hat{P}^2$ is given by:
\[
\hat{P}^2 = \hat{P}_x^2 + \hat{P}_y^2 + \hat{P}_z^2
\]
We calculate the commutator term by term:
\[
[\hat{L}_x, \hat{P}^2] = [\hat{L}_x, \hat{P}_x^2] + [\hat{L}_x, \hat{P}_y^2] + [\hat{L}_x, \hat{P}_z^2]
\]
For $\hat{P}_x^2$, we have $[\hat{L}_x, \hat{P}_x^2] = 2\hat{P}_x[\hat{L}_x, \hat{P}_x] = 0$.  
For $\hat{P}_y^2$, we have $[\hat{L}_x, \hat{P}_y^2] = 2\hat{P}_y[\hat{L}_x, \hat{P}_y] = 2\hat{P}_y(i\hbar \hat{P}_z) = i\hbar 2\hat{P}_y\hat{P}_z$.  
For $\hat{P}_z^2$, we have $[\hat{L}_x, \hat{P}_z^2] = 0$.  
Adding these terms, the result is:
\[
[\hat{L}_x, \hat{P}^2] = 0
\]

\textbf{Answer:}
\[
\left[\hat{L}_x, \hat{P}^2\right] = 0
\]


"

Solution 2: 

"
To derive the commutator relations, we use the definition of the angular momentum operator and the standard commutation relations for position and momentum operators. The x-component of the angular momentum operator is defined as:

\[
\hat{L}_x = \hat{Y} \hat{P}_z - \hat{Z} \hat{P}_y
\]

We now calculate each commutator step-by-step.

---

a) \(\left[\hat{L}_x, \hat{X}\right] = 0\) [3 marks]  

Using the definition of \(\hat{L}_x\):

\[
\left[\hat{L}_x, \hat{X}\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{X}\right]
\]

Expanding the commutator:

\[
\left[\hat{L}_x, \hat{X}\right] = \left[\hat{Y}\hat{P}_z, \hat{X}\right] - \left[\hat{Z}\hat{P}_y, \hat{X}\right]
\]

Since \(\hat{X}\) commutes with \(\hat{Y}\), \(\hat{Z}\), \(\hat{P}_y\), and \(\hat{P}_z\), all terms vanish:

\[
\left[\hat{L}_x, \hat{X}\right] = 0
\]

---

b) \(\left[\hat{L}_x, \hat{P}_x\right] = 0\) [3 marks]  

Using the same definition of \(\hat{L}_x\), we calculate:

\[
\left[\hat{L}_x, \hat{P}_x\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{P}_x\right]
\]

Expanding the commutator:

\[
\left[\hat{L}_x, \hat{P}_x\right] = \left[\hat{Y}\hat{P}_z, \hat{P}_x\right] - \left[\hat{Z}\hat{P}_y, \hat{P}_x\right]
\]

Since \(\hat{P}_x\) commutes with \(\hat{Y}\), \(\hat{Z}\), \(\hat{P}_y\), and \(\hat{P}_z\), all terms vanish:

\[
\left[\hat{L}_x, \hat{P}_x\right] = 0
\]

---

c) \(\left[\hat{L}_x, \hat{Y}\right] = i\hbar \hat{Z}\) [4 marks]  

Using the definition of \(\hat{L}_x\):

\[
\left[\hat{L}_x, \hat{Y}\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{Y}\right]
\]

Expanding the commutator:

\[
\left[\hat{L}_x, \hat{Y}\right] = \left[\hat{Y}\hat{P}_z, \hat{Y}\right] - \left[\hat{Z}\hat{P}_y, \hat{Y}\right]
\]

The first term vanishes because \(\hat{Y}\) commutes with itself:

\[
\left[\hat{Y}\hat{P}_z, \hat{Y}\right] = 0
\]

For the second term, using the canonical commutation relation \([\hat{Y}, \hat{P}_y] = i\hbar\):

\[
\left[\hat{Z}\hat{P}_y, \hat{Y}\right] = \hat{Z} \left[\hat{P}_y, \hat{Y}\right] = -i\hbar \hat{Z}
\]

Thus:

\[
\left[\hat{L}_x, \hat{Y}\right] = i\hbar \hat{Z}
\]

---

d) \(\left[\hat{L}_x, \hat{P}_y\right] = i\hbar \hat{P}_z\) [4 marks]  

Using the definition of \(\hat{L}_x\):

\[
\left[\hat{L}_x, \hat{P}_y\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{P}_y\right]
\]

Expanding the commutator:

\[
\left[\hat{L}_x, \hat{P}_y\right] = \left[\hat{Y}\hat{P}_z, \hat{P}_y\right] - \left[\hat{Z}\hat{P}_y, \hat{P}_y\right]
\]

The second term vanishes because \([\hat{P}_y, \hat{P}_y] = 0\). For the first term, using \([\hat{Y}, \hat{P}_y] = i\hbar\):

\[
\left[\hat{Y}\hat{P}_z, \hat{P}_y\right] = \left[\hat{Y}, \hat{P}_y\right]\hat{P}_z = i\hbar \hat{P}_z
\]

Thus:

\[
\left[\hat{L}_x, \hat{P}_y\right] = i\hbar \hat{P}_z
\]

---

e) \(\left[\hat{L}_x, \hat{P}^2\right] = 0\) [6 marks]  

The momentum squared operator is defined as:

\[
\hat{P}^2 = \hat{P}_x^2 + \hat{P}_y^2 + \hat{P}_z^2
\]

We calculate the commutator:

\[
\left[\hat{L}_x, \hat{P}^2\right] = \left[\hat{L}_x, \hat{P}_x^2 + \hat{P}_y^2 + \hat{P}_z^2\right]
\]

Using the linearity of commutators:

\[
\left[\hat{L}_x, \hat{P}^2\right] = \left[\hat{L}_x, \hat{P}_x^2\right] + \left[\hat{L}_x, \hat{P}_y^2\right] + \left[\hat{L}_x, \hat{P}_z^2\right]
\]

For each term, consider the commutation rule \([\hat{L}_x, \hat{P}_i^2] = 2\hat{P}_i [\hat{L}_x, \hat{P}_i]\). Since \([\hat{L}_x, \hat{P}_x] = 0\), \([\hat{L}_x, \hat{P}_z] = 0\), and \([\hat{L}_x, \hat{P}_y] = i\hbar \hat{P}_z\), we find:

\[
\left[\hat{L}_x, \hat{P}_x^2\right] = 0, \quad \left[\hat{L}_x, \hat{P}_z^2\right] = 0, \quad \left[\hat{L}_x, \hat{P}_y^2\right] = 0
\]

Thus:

\[
\left[\hat{L}_x, \hat{P}^2\right] = 0
\]
"

Solution 3: 

"
To solve the given problem, we use the definition of the angular momentum operator and the commutation relations between position and momentum operators. The x-component of the angular momentum operator is defined as:
\[
\hat{L}_x = \hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y.
\]

### Part (a): \(\left[\hat{L}_x, \hat{X}\right]\) [3 marks]
The commutator can be written as:
\[
\left[\hat{L}_x, \hat{X}\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{X}\right].
\]
Expanding the commutator:
\[
\left[\hat{L}_x, \hat{X}\right] = \left[\hat{Y}\hat{P}_z, \hat{X}\right] - \left[\hat{Z}\hat{P}_y, \hat{X}\right].
\]
Since \(\hat{X}\) commutes with both \(\hat{Y}\), \(\hat{Z}\), \(\hat{P}_y\), and \(\hat{P}_z\):
\[
\left[\hat{L}_x, \hat{X}\right] = 0.
\]

### Part (b): \(\left[\hat{L}_x, \hat{P}_x\right]\) [3 marks]
The commutator is:
\[
\left[\hat{L}_x, \hat{P}_x\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{P}_x\right].
\]
Expanding:
\[
\left[\hat{L}_x, \hat{P}_x\right] = \left[\hat{Y}\hat{P}_z, \hat{P}_x\right] - \left[\hat{Z}\hat{P}_y, \hat{P}_x\right].
\]
Since \(\hat{P}_x\) commutes with both \(\hat{P}_y\), \(\hat{P}_z\), \(\hat{Y}\), and \(\hat{Z}\):
\[
\left[\hat{L}_x, \hat{P}_x\right] = 0.
\]

### Part (c): \(\left[\hat{L}_x, \hat{Y}\right]\) [4 marks]
The commutator is:
\[
\left[\hat{L}_x, \hat{Y}\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{Y}\right].
\]
Expanding:
\[
\left[\hat{L}_x, \hat{Y}\right] = \left[\hat{Y}\hat{P}_z, \hat{Y}\right] - \left[\hat{Z}\hat{P}_y, \hat{Y}\right].
\]
The first term \(\left[\hat{Y}\hat{P}_z, \hat{Y}\right] = 0\) because \(\hat{Y}\) commutes with itself. For the second term:
\[
\left[\hat{Z}\hat{P}_y, \hat{Y}\right] = -\hat{Z}\left[\hat{P}_y, \hat{Y}\right] = -\hat{Z}(-i\hbar) = i\hbar\hat{Z}.
\]
Thus:
\[
\left[\hat{L}_x, \hat{Y}\right] = i\hbar\hat{Z}.
\]

### Part (d): \(\left[\hat{L}_x, \hat{P}_y\right]\) [4 marks]
The commutator is:
\[
\left[\hat{L}_x, \hat{P}_y\right] = \left[\hat{Y}\hat{P}_z - \hat{Z}\hat{P}_y, \hat{P}_y\right].
\]
Expanding:
\[
\left[\hat{L}_x, \hat{P}_y\right] = \left[\hat{Y}\hat{P}_z, \hat{P}_y\right] - \left[\hat{Z}\hat{P}_y, \hat{P}_y\right].
\]
The second term is zero because \(\left[\hat{P}_y, \hat{P}_y\right] = 0\). For the first term:
\[
\left[\hat{Y}\hat{P}_z, \hat{P}_y\right] = \hat{Y}\left[\hat{P}_z, \hat{P}_y\right] = \hat{Y}(i\hbar) = i\hbar\hat{P}_z.
\]
Thus:
\[
\left[\hat{L}_x, \hat{P}_y\right] = i\hbar\hat{P}_z.
\]

### Part (e): \(\left[\hat{L}_x, \hat{P}^2\right]\) [6 marks]
The operator \(\hat{P}^2\) is given by:
\[
\hat{P}^2 = \hat{P}_x^2 + \hat{P}_y^2 + \hat{P}_z^2.
\]
We compute:
\[
\left[\hat{L}_x, \hat{P}^2\right] = \left[\hat{L}_x, \hat{P}_x^2 + \hat{P}_y^2 + \hat{P}_z^2\right].
\]
Using the linearity of the commutator:
\[
\left[\hat{L}_x, \hat{P}^2\right] = \left[\hat{L}_x, \hat{P}_x^2\right] + \left[\hat{L}_x, \hat{P}_y^2\right] + \left[\hat{L}_x, \hat{P}_z^2\right].
\]
Since \(\left[\hat{L}_x, \hat{P}_x\right] = 0\), we have \(\left[\hat{L}_x, \hat{P}_x^2\right] = 0\). Similarly, for \(\hat{P}_y^2\) and \(\hat{P}_z^2\), we find:
\[
\left[\hat{L}_x, \hat{P}_y^2\right] = 2\hat{P}_y\left[\hat{L}_x, \hat{P}_y\right] = 2\hat{P}_y(i\hbar\hat{P}_z),
\]
and
\[
\left[\hat{L}_x, \hat{P}_z^2\right] = 2\hat{P}_z\left[\hat{L}_x, \hat{P}_z\right] = 0.
\]
Adding these contributions, the terms cancel due to symmetry:
\[
\left[\hat{L}_x, \hat{P}^2\right] = 0.
\]

Final Answers:
a) \(\left[\hat{L}_x, \hat{X}\right] = 0\)  
b) \(\left[\hat{L}_x, \hat{P}_x\right] = 0\)  
c) \(\left[\hat{L}_x, \hat{Y}\right] = i\hbar\hat{Z}\)  
d) \(\left[\hat{L}_x, \hat{P}_y\right] = i\hbar\hat{P}_z\)  
e) \(\left[\hat{L}_x, \hat{P}^2\right] = 0\).


 
"

\subsubsection{QMQ5}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 7 marks. Give up to 4 marks for part a) and give up to 3 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"An infinite square well of length \( L \) can be defined mathematically by the potential,

\[ V(x) = \begin{cases}
0 & \text{for } 0 \leq x \leq L,\\
+\infty  & \text{otherwise,} 
\end{cases} \]

A particle of mass \( m \) in the well has eigenfunctions

\[ \phi_{n}(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{n \pi x}{L} \right) \]

with corresponding energy eigenvalues 

\[ E_{n} = \frac{n^{2}\hbar^{2}\pi^{2}}{2mL^{2}} \]

a) Calculate the first-order correction to the ground state energy if the system is perturbed by 

\[ \hat{H}' = V_{0} \sin\left(\frac{2 \pi x}{L} \right) \]

b) Calculate and derive an expression for the first-order correction to all energy eigenvalues given that the system is perturbed by 

\[ \hat{H}' = L\alpha \delta\left(x - \frac{L}{2} \right) \]

where \( \alpha \) is a constant and \( \delta \) is the Dirac delta function."

Mark Scheme:

"{Q5: (7 marks total)}

a) (4 marks)

[1 mark] for stating the equation for energy and making it relevant to the ground state energy, 

\[ E_{1}^{(1)} = \int_{-\infty}^{\infty} \phi_{1}^{(0)*}\hat{H'}\phi_{1}^{(0)} \, dx \]

[1 mark] Substituting \( \hat{H}' = V_{0} \sin\left(\frac{2 \pi x}{L} \right) \) into the integral for energy with the limits [0,L] and simplifying to get \[ \frac{2V_{0}}{L} \int_{0}^{L} \sin^{2}\left(\frac{ \pi x}{L} \right)\sin\left(\frac{ 2\pi x}{L} \right) \, dx \]

[2 marks] Solving the integral, showing the perturbation equals 0.

b) (3 marks)

[1 mark] Applying perturbation theory, substituting the correct terms in to get the integral: 

\[ E_{n}^{(1)} = 2 \alpha \int_{0}^{L} \sin^{2}\left(\frac{n \pi x}{L} \right) \delta\left(x - \frac{L}{2} \right) \, dx \]

[1 mark] for correctly solving the integral and simplifying to 
\[ 2 \alpha \sin^{2}\left(\frac{n \pi}{2} \right) \]

[1 mark] for stating \(2 \alpha \) for all odd values of n, and 0 for all even values of n 

"

Solution 1: 

"
(a) To calculate the first-order correction to the ground state energy, we use non-degenerate perturbation theory. The first-order correction to the energy is given by:
\[
E_1^{(1)} = \langle \phi_1 | \hat{H}' | \phi_1 \rangle,
\]
where \(\phi_1(x)\) is the ground state eigenfunction:
\[
\phi_1(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right).
\]

The perturbation is:
\[
\hat{H}' = V_0 \sin\left(\frac{2\pi x}{L}\right).
\]

Substituting into the expression for \(E_1^{(1)}\), we get:
\[
E_1^{(1)} = \int_0^L \phi_1^*(x) \hat{H}' \phi_1(x) \, dx = V_0 \int_0^L \left(\sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right)\right)^2 \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]

Simplifying:
\[
E_1^{(1)} = \frac{2V_0}{L} \int_0^L \sin^2\left(\frac{\pi x}{L}\right) \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]

We use the trigonometric identity:
\[
\sin^2(A) = \frac{1}{2} \left(1 - \cos(2A)\right),
\]
and substitute \(A = \frac{\pi x}{L}\):
\[
\sin^2\left(\frac{\pi x}{L}\right) = \frac{1}{2} \left(1 - \cos\left(\frac{2\pi x}{L}\right)\right).
\]

Thus:
\[
E_1^{(1)} = \frac{2V_0}{L} \int_0^L \frac{1}{2} \left(1 - \cos\left(\frac{2\pi x}{L}\right)\right) \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]

Expanding:
\[
E_1^{(1)} = \frac{V_0}{L} \int_0^L \sin\left(\frac{2\pi x}{L}\right) \, dx - \frac{V_0}{L} \int_0^L \cos\left(\frac{2\pi x}{L}\right) \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]

The first term evaluates to zero because \(\sin\left(\frac{2\pi x}{L}\right)\) is symmetric about \(x = L/2\). The second term also evaluates to zero because \(\cos\left(\frac{2\pi x}{L}\right) \sin\left(\frac{2\pi x}{L}\right)\) integrates to zero over one period. Therefore:
\[
E_1^{(1)} = 0.
\]

The first-order correction to the ground state energy is:
\[
\boxed{E_1^{(1)} = 0.}
\]

(b) For the second perturbation:
\[
\hat{H}' = L \alpha \delta\left(x - \frac{L}{2}\right).
\]

The first-order correction to the \(n\)-th energy level is given by:
\[
E_n^{(1)} = \langle \phi_n | \hat{H}' | \phi_n \rangle.
\]

Substituting the perturbation, we have:
\[
E_n^{(1)} = L \alpha \int_0^L \phi_n^*(x) \delta\left(x - \frac{L}{2}\right) \phi_n(x) \, dx.
\]

The Dirac delta function picks out the value of the integrand at \(x = \frac{L}{2}\):
\[
E_n^{(1)} = L \alpha \phi_n^*\left(\frac{L}{2}\right) \phi_n\left(\frac{L}{2}\right).
\]

Using the eigenfunction:
\[
\phi_n(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{n \pi x}{L}\right),
\]
we evaluate \(\phi_n\left(\frac{L}{2}\right)\):
\[
\phi_n\left(\frac{L}{2}\right) = \sqrt{\frac{2}{L}} \sin\left(\frac{n \pi}{2}\right).
\]

Thus:
\[
E_n^{(1)} = L \alpha \left(\sqrt{\frac{2}{L}} \sin\left(\frac{n \pi}{2}\right)\right)^2.
\]

Simplifying:
\[
E_n^{(1)} = L \alpha \frac{2}{L} \sin^2\left(\frac{n \pi}{2}\right).
\]

The \(L\) terms cancel, leaving:
\[
E_n^{(1)} = 2 \alpha \sin^2\left(\frac{n \pi}{2}\right).
\]

The first-order correction to the \(n\)-th energy level is:
\[
\boxed{E_n^{(1)} = 2 \alpha \sin^2\left(\frac{n \pi}{2}\right).}
\]

"

Solution 2: 

"
\textbf{Solution:}

\textbf{(a) First-order correction to the ground state energy}

The first-order correction to the energy in perturbation theory is given by:
\[
E_n^{(1)} = \langle \phi_n | \hat{H}' | \phi_n \rangle,
\]
where \(\phi_n(x)\) are the eigenfunctions of the unperturbed Hamiltonian.

For the ground state (\(n = 1\)), the eigenfunction is:
\[
\phi_1(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right).
\]

The perturbation is:
\[
\hat{H}' = V_0 \sin\left(\frac{2\pi x}{L}\right).
\]

Substituting, the first-order correction is:
\[
E_1^{(1)} = \int_0^L \phi_1^*(x) \hat{H}' \phi_1(x) \, dx.
\]

Expanding:
\[
E_1^{(1)} = \int_0^L \left(\sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right)\right) V_0 \sin\left(\frac{2\pi x}{L}\right) \left(\sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right)\right) \, dx.
\]

Simplify:
\[
E_1^{(1)} = \frac{2 V_0}{L} \int_0^L \sin^2\left(\frac{\pi x}{L}\right) \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]

Using the trigonometric identity:
\[
\sin^2(a) \sin(b) = \frac{1}{2} \left[\cos(a-b) - \cos(a+b)\right],
\]
where \(a = \frac{\pi x}{L}\) and \(b = \frac{2\pi x}{L}\), the integral becomes:
\[
E_1^{(1)} = \frac{2 V_0}{L} \int_0^L \frac{1}{2} \left[\cos\left(-\frac{\pi x}{L}\right) - \cos\left(\frac{3\pi x}{L}\right)\right] \, dx.
\]

Simplify further:
\[
E_1^{(1)} = \frac{V_0}{L} \int_0^L \left[\cos\left(\frac{\pi x}{L}\right) - \cos\left(\frac{3\pi x}{L}\right)\right] \, dx.
\]

Evaluate the integrals:
\[
\int_0^L \cos\left(\frac{\pi x}{L}\right) \, dx = 0 \quad \text{(since the integrand is an odd function)},
\]
\[
\int_0^L \cos\left(\frac{3\pi x}{L}\right) \, dx = 0 \quad \text{(for the same reason)}.
\]

Thus:
\[
E_1^{(1)} = 0.
\]

The first-order correction to the ground state energy is:
\[
E_1^{(1)} = 0.
\]

\textbf{(b) First-order correction for a delta function perturbation}

The perturbation is:
\[
\hat{H}' = L \alpha \delta\left(x - \frac{L}{2}\right).
\]

The first-order correction to the energy is:
\[
E_n^{(1)} = \langle \phi_n | \hat{H}' | \phi_n \rangle = \int_0^L \phi_n^*(x) \hat{H}' \phi_n(x) \, dx.
\]

Substitute the eigenfunction \(\phi_n(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{n \pi x}{L}\right)\) and the perturbation:
\[
E_n^{(1)} = \int_0^L \left(\sqrt{\frac{2}{L}} \sin\left(\frac{n \pi x}{L}\right)\right) L \alpha \delta\left(x - \frac{L}{2}\right) \left(\sqrt{\frac{2}{L}} \sin\left(\frac{n \pi x}{L}\right)\right) \, dx.
\]

The delta function \(\delta\left(x - \frac{L}{2}\right)\) selects the value of the integrand at \(x = \frac{L}{2}\):
\[
E_n^{(1)} = L \alpha \left(\sqrt{\frac{2}{L}} \sin\left(\frac{n \pi}{2}\right)\right)^2.
\]

Simplify:
\[
E_n^{(1)} = L \alpha \frac{2}{L} \sin^2\left(\frac{n \pi}{2}\right).
\]

Thus:
\[
E_n^{(1)} = 2 \alpha \sin^2\left(\frac{n \pi}{2}\right).
\]

The first-order correction to the energy eigenvalues is:
\[
E_n^{(1)} = 2 \alpha \sin^2\left(\frac{n \pi}{2}\right).
\]

"

Solution 3:

"
To solve the given problems, we use the principles of first-order perturbation theory. The first-order correction to the energy eigenvalue is given by:
\[
E_n^{(1)} = \langle \phi_n | \hat{H}' | \phi_n \rangle,
\]
where \(\phi_n(x)\) are the unperturbed eigenfunctions of the infinite square well, and \(\hat{H}'\) is the perturbation.

\textbf{(a)} For the perturbation \(\hat{H}' = V_0 \sin\left(\frac{2\pi x}{L}\right)\), the first-order correction to the ground state energy is:
\[
E_1^{(1)} = \langle \phi_1 | \hat{H}' | \phi_1 \rangle = V_0 \int_0^L \phi_1(x) \sin\left(\frac{2\pi x}{L}\right) \phi_1(x) \, dx.
\]
Substituting \(\phi_1(x) = \sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right)\), we get:
\[
E_1^{(1)} = V_0 \int_0^L \left(\sqrt{\frac{2}{L}} \sin\left(\frac{\pi x}{L}\right)\right)^2 \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]
Simplifying:
\[
E_1^{(1)} = \frac{2 V_0}{L} \int_0^L \sin^2\left(\frac{\pi x}{L}\right) \sin\left(\frac{2\pi x}{L}\right) \, dx.
\]
Using the trigonometric identity \(\sin^2(a) = \frac{1}{2} (1 - \cos(2a))\) and expanding, we write:
\[
\sin^2\left(\frac{\pi x}{L}\right) \sin\left(\frac{2\pi x}{L}\right) = \frac{1}{2} \sin\left(\frac{2\pi x}{L}\right) - \frac{1}{4} \sin\left(\frac{4\pi x}{L}\right).
\]
Substituting into the integral and solving term by term, we find that all terms vanish due to orthogonality of sine functions over the interval \([0, L]\). Therefore:
\[
E_1^{(1)} = 0.
\]

\textbf{(b)} For the perturbation \(\hat{H}' = L \alpha \delta\left(x - \frac{L}{2}\right)\), the first-order correction to the energy eigenvalues is:
\[
E_n^{(1)} = \langle \phi_n | \hat{H}' | \phi_n \rangle = L \alpha \int_0^L \phi_n(x) \delta\left(x - \frac{L}{2}\right) \phi_n(x) \, dx.
\]
Using the sifting property of the Dirac delta function:
\[
\int_0^L f(x) \delta\left(x - \frac{L}{2}\right) \, dx = f\left(\frac{L}{2}\right),
\]
we evaluate the eigenfunctions at \(x = \frac{L}{2}\):
\[
\phi_n\left(\frac{L}{2}\right) = \sqrt{\frac{2}{L}} \sin\left(\frac{n \pi}{2}\right).
\]
Substituting into the expression for \(E_n^{(1)}\), we get:
\[
E_n^{(1)} = L \alpha \left(\phi_n\left(\frac{L}{2}\right)\right)^2 = L \alpha \left(\sqrt{\frac{2}{L}} \sin\left(\frac{n \pi}{2}\right)\right)^2.
\]
Simplifying:
\[
E_n^{(1)} = 2 \alpha \sin^2\left(\frac{n \pi}{2}\right).
\]
Thus, the first-order correction to the energy eigenvalues is:
\[
E_n^{(1)} = 
\begin{cases}
2 \alpha & \text{if } n \text{ is odd}, \\
0 & \text{if } n \text{ is even}.
\end{cases}
\]
This result arises because \(\sin\left(\frac{n \pi}{2}\right)\) alternates between 1 and 0 for odd and even values of \(n\), respectively.

"

\subsubsection{QMQ6}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 17 marks. Give up to 4 marks for part a), give up to 4 marks for part b) and give up to 9 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Let \( \hat{\underline{L}} \) and \( \hat{\underline{S}} \) be the angular momentum and spin angular momentum quantum operators. 

a) Given that \( \hat{\underline{J}} = \hat{\underline{L}} + \hat{\underline{S}} \), show that 
\[ \hat{J^{2}} =  \hat{L^{2}} + \hat{S^{2}} + \hat{L}_{+}\hat{S}_{-} +  \hat{L}_{-}\hat{S}_{+} + 2\hat{L}_{z}\hat{S}_{z} \]

Where you may use  \( \hat{L}_{\pm} = \hat{L}_{x} \pm i\hat{L}_{y} \), and \( \hat{S}_{\pm} = \hat{S}_{x} \pm i\hat{S}_{y} \)

b) consider the state \( \left|l, m; s, m_{s} \right> = \left|l, m \right>\left|s, m_{s} \right> \). Here \( l \) is a quantum number of \( \hat{L^{2}} \), \( m \) is a quantum number of \( \hat{L}_{z} \). \( s \) is a quantum number of \( \hat{S^{2}} \) and \( m_{s} \) is a quantum number of \( \hat{S}_{z} \). Show that the state \( \left|l, -l; s, -s \right> \) is an eigenvector of \( \hat{J^{2}} \) and the corresponding eigenvalue. 

c) Now consider an operator \( \hat{O} = a\hat{L^{2}} + b\hat{S}_{+}\hat{L}_{z} \). Where \( a \), \( b \) are constants

You are given the following result:
\[ \hat{S}_{+} \left|s, m_{s} \right> = \hbar \sqrt{s(s+1) - m_{s}(m_{s}+1)} \left|s, m_{s}+1 \right> \]

Find the matrix representation of \( \hat{O} \) for a chosen basis of kets \( \left|l=1, m; s=\frac{1}{2}, m_{s} \right> \)."

Mark Scheme:

"{Q6: (17 marks total) }

a) (4 marks)

[1 mark] \[ \hat{J^{2}} = \left(\hat{\underline{L}} + \hat{\underline{S}} \right) \cdot \left(\hat{\underline{L}} + \hat{\underline{S}} \right) =  \hat{L^{2}} + \hat{S^{2}} + 2\hat{L}_{x}\hat{S}_{x} + 2\hat{L}_{y}\hat{S}_{y} + 2\hat{L}_{z}\hat{S}_{z} \]

[2 marks]  \[ \hat{S_{x}} = \frac{1}{2} (\hat{S_{+}} + \hat{S_{-}}) \quad \hat{S_{y}} = \frac{1}{2i} (\hat{S_{+}} - \hat{S_{-}}) \] 

\[ \hat{L_{x}} = \frac{1}{2} (\hat{L_{+}} + \hat{L_{-}}) \quad \hat{L_{y}} = \frac{1}{2i} (\hat{L_{+}} - \hat{L_{-}}) \]

[1 mark] Final answer \[ \hat{J^{2}} =  \hat{L^{2}} + \hat{S^{2}} + \hat{L}_{+}\hat{S}_{-} +  \hat{L}_{-}\hat{S}_{+} + 2\hat{L}_{z}\hat{S}_{z} \]

b) (4 marks)

[2 marks] Applying operator expression for \( \hat{J^{2}} \) to \( \left|l, -l; s, -s \right> \) and noting \( \hat{L}_{+}\hat{S}_{-} \), \( \hat{L}_{-}\hat{S}_{+} \), since states composing the tensor product are the bottom rung (or show explicitly)

\[ 
\hat{J^{2}}\left|l, -l; s, -s \right> = \left( \hat{L^{2}} + \hat{S^{2}} + \hat{L}_{+}\hat{S}_{-} +  \hat{L}_{-}\hat{S}_{+} + 2\hat{L}_{z}\hat{S}_{z} \right) \left|l, -l \right> \left|s, -s \right>  
\]

\[
\hat{J^{2}}\left|l, -l; s, -s \right> = \left( \hat{L^{2}} + \hat{S^{2}} + 2\hat{L}_{z}\hat{S}_{z} \right)  \left|l, -l \right> \left|s, -s \right>
\]

[2 marks] calculating \( \hat{L^{2}} \), \( \hat{S^{2}} \),\( \hat{L}_{z}\hat{S}_{z} \) acting on \( \left|l, -l \right> \left|s, -s \right> \). Showing it is an eigenstate of  \( \hat{J^{2}} \) with eigenvalue \( \hbar^2 \left ( l(l+1) + s(s+1) + 2ls \right ) \) or equivalent. 

c) (9 marks)

[1 mark] noting there are 6 states which naturally form the basis kets of \( \left|l=1, m; s=\frac{1}{2}, m_{s} \right> \).

\[ \left|1, 1 \right> |\frac{1}{2},\frac{1}{2} \rangle, \quad |1, 0 \rangle |\frac{1}{2}, \frac{1}{2} \rangle, \quad \left|1, -1 \right>|\frac{1}{2}, \frac{1}{2} \rangle \]

\[ \left|1, 1 \right> |\frac{1}{2},-\frac{1}{2} \rangle, \quad |1, 0 \rangle |\frac{1}{2}, -\frac{1}{2} \rangle, \quad \left|1, -1 \right>|\frac{1}{2}, -\frac{1}{2} \rangle \]

[1 mark] Appropriate choice of vector representation for each basis ket. Eg. 

\[
 \left|1, 1 \right> |\frac{1}{2},\frac{1}{2} \rangle \rightarrow \begin{pmatrix} 1\\ 0\\ 0\\ 0\\ 0\\ 0 \end{pmatrix} = e_1, \quad |1, 0 \rangle |\frac{1}{2}, \frac{1}{2} \rangle \rightarrow \begin{pmatrix} 0\\ 0\\ 1\\ 0\\ 0\\ 0 \end{pmatrix} = e_3, \quad \left|1, -1 \right>|\frac{1}{2}, \frac{1}{2} \rangle \rightarrow \begin{pmatrix} 0\\ 0\\ 0\\ 0\\ 1\\ 0 \end{pmatrix} = e_5
\]

\[
 \left|1, 1 \right> |\frac{1}{2},-\frac{1}{2} \rangle \rightarrow \begin{pmatrix} 0\\ 1\\ 0\\ 0\\ 0\\ 0 \end{pmatrix} = e_2, \quad |1, 0 \rangle |\frac{1}{2}, -\frac{1}{2} \rangle \rightarrow \begin{pmatrix} 0\\ 0\\ 0\\ 1\\ 0\\ 0 \end{pmatrix} = e_4, \quad \left|1, -1 \right>|\frac{1}{2}, -\frac{1}{2} \rangle \rightarrow \begin{pmatrix} 0\\ 0\\ 0\\ 0\\ 0\\ 1 \end{pmatrix} = e_6
\]

[2 marks] Calculation of  \( \hat{L^{2}} \) acting on basis kets, giving matrix representation 

\[ 
2\hbar^2 \begin{pmatrix} 1 & 0 & 0 & 0 & 0 & 0\\  0 & 1 & 0 & 0 & 0 & 0\\ 0 & 0 & 1 & 0 & 0 & 0\\ 0 & 0 & 0 & 1 & 0 & 0\\ 0 & 0 & 0 & 0 & 1 & 0\\ 0 & 0 & 0 & 0 & 0 & 1 \end{pmatrix}
\]

[4 marks] Calculation of \( \hat{S}_{+}\hat{L}_{z} \) acting on basis kets giving the matrix representation. Eg. 

\[
\hat{S}_{+}\hat{L}_{z}e_1 = 0, \quad \hat{S}_{+}\hat{L}_{z}e_2 = \hbar^2e_1, \quad \hat{S}_{+}\hat{L}_{z}e_3 = 0, \quad \hat{S}_{+}\hat{L}_{z}e_4 = 0, \quad \hat{S}_{+}\hat{L}_{z}e_5 = 0, \quad \hat{S}_{+}\hat{L}_{z}e_6 = -\hbar^2e_5 
\]

\[ 
S_{+}L_{z} = \hbar^2 \begin{pmatrix} 0 & 1 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & -1\\ 0 & 0 & 0 & 0 & 0 & 0 \end{pmatrix}
\]

[1 mark] Construction of the matrix representation of \( \hat{O} = a\hat{L^{2}} + b\hat{S}_{+}\hat{L}_{z} \)

\[ O = a L^2 + b S_{+}L_z \]

\[
O = 2\hbar^2 a \begin{pmatrix} 1 & 0 & 0 & 0 & 0 & 0\\  0 & 1 & 0 & 0 & 0 & 0\\ 0 & 0 & 1 & 0 & 0 & 0\\ 0 & 0 & 0 & 1 & 0 & 0\\ 0 & 0 & 0 & 0 & 1 & 0\\ 0 & 0 & 0 & 0 & 0 & 1 \end{pmatrix} + \hbar^2 b \begin{pmatrix} 0 & 1 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & 0\\ 0 & 0 & 0 & 0 & 0 & -1\\ 0 & 0 & 0 & 0 & 0 & 0 \end{pmatrix}
\]

\[ 
O = \hbar^2 \begin{pmatrix} 2a & b & 0 & 0 & 0 & 0\\ 0 & 2a & 0 & 0 & 0 & 0\\ 0 & 0 & 2a & 0 & 0 & 0\\ 0 & 0 & 0 & 2a & 0 & 0\\ 0 & 0 & 0 & 0 & 2a & -b\\ 0 & 0 & 0 & 0 & 0 & 2a \end{pmatrix}
\]

Note matrix representation will look different depending on the representation chosen for the basis kets. 

"

Solution 1: 

"

\textbf{Solution:}  

\textbf{(a) [4 marks]}  

To show that $\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ + 2 \hat{L}_z \hat{S}_z$, recall the definition of $\hat{J} = \hat{L} + \hat{S}$. The square of the total angular momentum operator is given by:  
\[
\hat{J}^2 = (\hat{L} + \hat{S}) \cdot (\hat{L} + \hat{S}).
\]  
Expanding this:  
\[
\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + 2 \hat{L} \cdot \hat{S}.
\]  
The cross term $2 \hat{L} \cdot \hat{S}$ can be written in terms of the ladder operators and $z$-components:  
\[
\hat{L} \cdot \hat{S} = \hat{L}_x \hat{S}_x + \hat{L}_y \hat{S}_y + \hat{L}_z \hat{S}_z.
\]  
Using the ladder operators $\hat{L}_\pm = \hat{L}_x \pm i \hat{L}_y$ and $\hat{S}_\pm = \hat{S}_x \pm i \hat{S}_y$, the products $\hat{L}_x \hat{S}_x$ and $\hat{L}_y \hat{S}_y$ can be rewritten as:  
\[
\hat{L}_x \hat{S}_x + \hat{L}_y \hat{S}_y = \frac{1}{2} \left( \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ \right).
\]  
Thus,  
\[
\hat{L} \cdot \hat{S} = \frac{1}{2} \left( \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ \right) + \hat{L}_z \hat{S}_z.
\]  
Substituting this into the expression for $\hat{J}^2$, we obtain:  
\[
\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ + 2 \hat{L}_z \hat{S}_z.
\]  

\textbf{(b) [4 marks]}  

Consider the state $\lvert l, -l; s, -s \rangle = \lvert l, -l \rangle \lvert s, -s \rangle$. For $\hat{J}^2$, we have:  
\[
\hat{J}^2 \lvert l, -l; s, -s \rangle = (\hat{L}^2 + \hat{S}^2 + 2 \hat{L} \cdot \hat{S}) \lvert l, -l; s, -s \rangle.
\]  
Acting on $\lvert l, -l \rangle$, the operators $\hat{L}^2$ and $\hat{L}_z$ yield:  
\[
\hat{L}^2 \lvert l, -l \rangle = \hbar^2 l(l+1) \lvert l, -l \rangle, \quad \hat{L}_z \lvert l, -l \rangle = -\hbar l \lvert l, -l \rangle.
\]  
Similarly, for $\lvert s, -s \rangle$:  
\[
\hat{S}^2 \lvert s, -s \rangle = \hbar^2 s(s+1) \lvert s, -s \rangle, \quad \hat{S}_z \lvert s, -s \rangle = -\hbar s \lvert s, -s \rangle.
\]  
Since $\hat{L}_+ \lvert l, -l \rangle = 0$ and $\hat{S}_+ \lvert s, -s \rangle = 0$, the ladder operators do not contribute. The term $\hat{L} \cdot \hat{S}$ simplifies to:  
\[
\hat{L} \cdot \hat{S} = \hat{L}_z \hat{S}_z = (-\hbar l)(-\hbar s) = \hbar^2 ls.
\]  
Thus, the eigenvalue of $\hat{J}^2$ is:  
\[
\hat{J}^2 \lvert l, -l; s, -s \rangle = \hbar^2 \left[ l(l+1) + s(s+1) - 2ls \right] \lvert l, -l; s, -s \rangle.
\]  

\textbf{(c) [9 marks]}  

The operator $\hat{O}$ is given by:  
\[
\hat{O} = a \hat{L}^2 + b \hat{S}_+ \hat{L}_z.
\]  
We aim to find the matrix representation of $\hat{O}$ in the basis $\lvert l = 1, m; s = \frac{1}{2}, m_s \rangle$. The basis states are:  
\[
\lvert 1, 1; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, 1; \frac{1}{2}, -\frac{1}{2} \rangle, \quad \lvert 1, 0; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, 0; \frac{1}{2}, -\frac{1}{2} \rangle, \quad \lvert 1, -1; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, -1; \frac{1}{2}, -\frac{1}{2} \rangle.
\]  
1. For $a \hat{L}^2$, $\hat{L}^2$ acts as $\hbar^2 l(l+1)$ on all basis states, yielding $\hbar^2 \cdot 2$.  
2. For $b \hat{S}_+ \hat{L}_z$, $\hat{L}_z$ acts as $\hbar m$ on $\lvert l, m \rangle$, and $\hat{S}_+$ raises $m_s$ by $1$ with a factor:  
\[
\hat{S}_+ \lvert \frac{1}{2}, m_s \rangle = \hbar \sqrt{s(s+1) - m_s(m_s + 1)} \lvert \frac{1}{2}, m_s + 1 \rangle.
\]  

Constructing the matrix elements $\langle l', m'; s', m_s' \lvert \hat{O} \rvert l, m; s, m_s \rangle$ involves evaluating these contributions. After calculating, the resulting matrix is:  
\[
\hat{O} = 
\begin{pmatrix}
2a\hbar^2 & 0 & 0 & b\hbar^2 & 0 & 0 \\
0 & 2a\hbar^2 & 0 & 0 & b\hbar^2 & 0 \\
0 & 0 & 2a\hbar^2 & 0 & 0 & b\hbar^2 \\
b\hbar^2 & 0 & 0 & 2a\hbar^2 & 0 & 0 \\
0 & b\hbar^2 & 0 & 0 & 2a\hbar^2 & 0 \\
0 & 0 & b\hbar^2 & 0 & 0 & 2a\hbar^2
\end{pmatrix}.
\]  
"

Solution 2: 

"

\textbf{Solution:}  

\textbf{(a)}  
We are tasked with showing the relation  
\[
\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ + 2 \hat{L}_z \hat{S}_z.
\]  
Recall that $\hat{J} = \hat{L} + \hat{S}$. The square of the total angular momentum operator is given by  
\[
\hat{J}^2 = (\hat{L} + \hat{S}) \cdot (\hat{L} + \hat{S}) = \hat{L}^2 + \hat{S}^2 + 2 \hat{L} \cdot \hat{S}.
\]  
The dot product $\hat{L} \cdot \hat{S}$ can be written as  
\[
\hat{L} \cdot \hat{S} = \hat{L}_z \hat{S}_z + \frac{1}{2}(\hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+).
\]  
Substituting this expression into $\hat{J}^2$, we have  
\[
\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + 2 \left( \hat{L}_z \hat{S}_z + \frac{1}{2} (\hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+) \right).
\]  
Simplifying, we arrive at  
\[
\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ + 2 \hat{L}_z \hat{S}_z,
\]  
as required.  

\textbf{(b)}  
Consider the state $\lvert l, -l; s, -s \rangle$. By definition, this is the product state  
\[
\lvert l, -l; s, -s \rangle = \lvert l, -l \rangle \lvert s, -s \rangle,
\]  
where $\lvert l, -l \rangle$ is an eigenstate of $\hat{L}^2$ and $\hat{L}_z$, and $\lvert s, -s \rangle$ is an eigenstate of $\hat{S}^2$ and $\hat{S}_z$.  

For $\hat{L}^2$ and $\hat{S}^2$, we have the eigenvalue equations  
\[
\hat{L}^2 \lvert l, -l \rangle = \hbar^2 l(l+1) \lvert l, -l \rangle, \quad \hat{S}^2 \lvert s, -s \rangle = \hbar^2 s(s+1) \lvert s, -s \rangle.
\]  
For $\hat{L}_z$ and $\hat{S}_z$,  
\[
\hat{L}_z \lvert l, -l \rangle = -\hbar l \lvert l, -l \rangle, \quad \hat{S}_z \lvert s, -s \rangle = -\hbar s \lvert s, -s \rangle.
\]  
Thus, $\hat{L}_z + \hat{S}_z$ acting on $\lvert l, -l; s, -s \rangle$ gives  
\[
(\hat{L}_z + \hat{S}_z) \lvert l, -l; s, -s \rangle = -\hbar (l + s) \lvert l, -l; s, -s \rangle.
\]  

The operator $\hat{J}^2$ acts as  
\[
\hat{J}^2 \lvert l, -l; s, -s \rangle = (\hat{L}^2 + \hat{S}^2 + 2 \hat{L} \cdot \hat{S}) \lvert l, -l; s, -s \rangle.
\]  
Using the eigenvalue equations and the fact that $\hat{L} \cdot \hat{S} = \hat{L}_z \hat{S}_z + \frac{1}{2} (\hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+)$, we find that  
\[
\hat{J}^2 \lvert l, -l; s, -s \rangle = \hbar^2 \left( l(l+1) + s(s+1) + (-l)(-s) \right) \lvert l, -l; s, -s \rangle.
\]  
Simplifying, the eigenvalue is  
\[
\hbar^2 \left( l(l+1) + s(s+1) - ls \right).
\]  

\textbf{(c)}  
We aim to find the matrix representation of the operator  
\[
\hat{O} = a \hat{L}^2 + b \hat{S}_+ \hat{L}_z,
\]  
in the basis $\{ \lvert l = 1, m; s = \frac{1}{2}, m_s \rangle \}$.  

For $l=1$, $s = \frac{1}{2}$, the basis states are:  
\[
\lvert 1, 1; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, 1; \frac{1}{2}, -\frac{1}{2} \rangle, \quad \lvert 1, 0; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, 0; \frac{1}{2}, -\frac{1}{2} \rangle, \quad \lvert 1, -1; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, -1; \frac{1}{2}, -\frac{1}{2} \rangle.
\]  

The matrix elements are given by  
\[
\langle l, m; s, m_s \rvert \hat{O} \lvert l, m'; s, m_s' \rangle.
\]  

For $\hat{L}^2$, the eigenvalue is $\hbar^2 l(l+1) = 2\hbar^2$. Thus,  
\[
\langle l, m; s, m_s \rvert a \hat{L}^2 \lvert l, m'; s, m_s' \rangle = a (2\hbar^2) \delta_{m,m'} \delta_{m_s,m_s'}.
\]  

For $\hat{S}_+ \hat{L}_z$, we use  
\[
\hat{S}_+ \lvert s, m_s \rangle = \hbar \sqrt{s(s+1) - m_s(m_s+1)} \lvert s, m_s+1 \rangle.
\]  
Combining with $\hat{L}_z \lvert l, m \rangle = m\hbar \lvert l, m \rangle$, we find  
\[
\langle l, m; s, m_s \rvert b \hat{S}_+ \hat{L}_z \lvert l, m'; s, m_s' \rangle = b \hbar^2 m \sqrt{s(s+1) - m_s(m_s+1)} \delta_{m,m'} \delta_{m_s', m_s+1}.
\]  

Using these results, the matrix representation of $\hat{O}$ can be computed explicitly for the chosen basis. Each element of the matrix is either proportional to $\delta_{m,m'}$ or involves the Clebsch-Gordan coefficients from the action of $\hat{S}_+$. The final form is left as a detailed calculation.

"

Solution 3:

"

To answer the given questions, we provide detailed reasoning and computations for each part.

---

**Solution**

a) We are given $\hat{J} = \hat{L} + \hat{S}$, and we need to compute $\hat{J}^2$:
\[
\hat{J}^2 = (\hat{L} + \hat{S})^2 = \hat{L}^2 + \hat{S}^2 + 2 \hat{L} \cdot \hat{S},
\]
where $\hat{L} \cdot \hat{S}$ is the scalar product of the angular momentum operators:
\[
\hat{L} \cdot \hat{S} = \hat{L}_x \hat{S}_x + \hat{L}_y \hat{S}_y + \hat{L}_z \hat{S}_z.
\]
Using the ladder operators $\hat{L}_\pm = \hat{L}_x \pm i\hat{L}_y$ and $\hat{S}_\pm = \hat{S}_x \pm i\hat{S}_y$, the products can be rewritten:
\[
\hat{L}_x \hat{S}_x + \hat{L}_y \hat{S}_y = \frac{1}{2} (\hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+).
\]
Thus,
\[
\hat{L} \cdot \hat{S} = \frac{1}{2} (\hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+) + \hat{L}_z \hat{S}_z.
\]
Substituting back, we get:
\[
\hat{J}^2 = \hat{L}^2 + \hat{S}^2 + \hat{L}_+ \hat{S}_- + \hat{L}_- \hat{S}_+ + 2 \hat{L}_z \hat{S}_z.
\]
This matches the required expression.

---

b) Consider the state $\lvert l, m; s, m_s \rangle = \lvert l, m \rangle \lvert s, m_s \rangle$, where $\hat{L}^2 \lvert l, m \rangle = \hbar^2 l(l+1) \lvert l, m \rangle$, $\hat{L}_z \lvert l, m \rangle = \hbar m \lvert l, m \rangle$, $\hat{S}^2 \lvert s, m_s \rangle = \hbar^2 s(s+1) \lvert s, m_s \rangle$, and $\hat{S}_z \lvert s, m_s \rangle = \hbar m_s \lvert s, m_s \rangle$. For $\hat{J}^2$, the eigenvalue equation is:
\[
\hat{J}^2 \lvert l, m; s, m_s \rangle = \hbar^2 j(j+1) \lvert l, m; s, m_s \rangle,
\]
where $j = l + s$. For $\lvert l, -l; s, -s \rangle$, we substitute $j = l + s$, giving:
\[
\hat{J}^2 \lvert l, -l; s, -s \rangle = \hbar^2 (l+s)(l+s+1) \lvert l, -l; s, -s \rangle.
\]
This confirms $\lvert l, -l; s, -s \rangle$ is an eigenvector of $\hat{J}^2$ with eigenvalue $\hbar^2 (l+s)(l+s+1)$.

---

c) For the operator $\hat{O} = a \hat{L}^2 + b \hat{S}_+ \hat{L}_z$, we compute its matrix representation in the basis $\lvert l = 1, m; s = \frac{1}{2}, m_s \rangle$. The basis kets are:
\[
\lvert 1, 1; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, 1; \frac{1}{2}, -\frac{1}{2} \rangle, \quad \lvert 1, 0; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, 0; \frac{1}{2}, -\frac{1}{2} \rangle, \quad \lvert 1, -1; \frac{1}{2}, \frac{1}{2} \rangle, \quad \lvert 1, -1; \frac{1}{2}, -\frac{1}{2} \rangle.
\]

**Step 1: Contribution from $\hat{L}^2$.**

Since $\hat{L}^2$ acts only on the orbital part:
\[
\hat{L}^2 \lvert l = 1, m; s, m_s \rangle = \hbar^2 l(l+1) \lvert l = 1, m; s, m_s \rangle = 2\hbar^2 \lvert l = 1, m; s, m_s \rangle.
\]

**Step 2: Contribution from $\hat{S}_+ \hat{L}_z$.**

Using the action of $\hat{L}_z$ and $\hat{S}_+$:
\[
\hat{L}_z \lvert l, m; s, m_s \rangle = \hbar m \lvert l, m; s, m_s \rangle,
\]
\[
\hat{S}_+ \lvert s, m_s \rangle = \hbar \sqrt{s(s+1) - m_s(m_s + 1)} \lvert s, m_s + 1 \rangle.
\]
Thus:
\[
\hat{S}_+ \hat{L}_z \lvert l, m; s, m_s \rangle = \hbar m \sqrt{s(s+1) - m_s(m_s + 1)} \lvert l, m; s, m_s + 1 \rangle.
\]

**Step 3: Constructing the matrix elements.**

The matrix element $\langle l', m'; s, m_s' \lvert \hat{O} \lvert l, m; s, m_s \rangle$ is:
\[
\langle l', m'; s, m_s' \lvert a \hat{L}^2 + b \hat{S}_+ \hat{L}_z \lvert l, m; s, m_s \rangle.
\]
For $\hat{L}^2$, the diagonal contribution is $a \hbar^2 l(l+1)$, and for $\hat{S}_+ \hat{L}_z$, we compute non-zero off-diagonal terms using the ladder operator relations.

The resulting matrix for $\hat{O}$ in the basis is:
\[
\hat{O} = 
\begin{bmatrix}
2a\hbar^2 & 0 & 0 & 0 & 0 & 0 \\
0 & 2a\hbar^2 & b\hbar^2 & 0 & 0 & 0 \\
0 & b\hbar^2 & 2a\hbar^2 & 0 & 0 & 0 \\
0 & 0 & 0 & 2a\hbar^2 & 0 & 0 \\
0 & 0 & 0 & 0 & 2a\hbar^2 & 0 \\
0 & 0 & 0 & 0 & 0 & 2a\hbar^2
\end{bmatrix}.
\]
"

\subsubsection{QMQ7}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 8 marks. Give up to 3 marks for part a), and give up to 5 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"A beam of particles of mass \( m \) moves in a space with potential energy \( V(x) = 0 \), described by the wavefunction \( \psi(x) = Ae^{ikx} \).

a) What is the corresponding time-dependent solution \( \Psi(x,t) \)? Show that the probability per unit length of finding a particle is independent of both space and time.

b) Evaluate the particle flux 

\[ \Gamma = -\frac{i\hbar}{2m}\left[ \Psi^{*}\frac{\partial\Psi}{\partial x} - \Psi\frac{\partial\Psi^{*}}{\partial x} \right] \]

for the state \( \Psi \), giving a physical interpretation of the result in terms of the velocity of the particles."

Mark Scheme:

"{Q7: (8 marks total)}

a) (3 marks)

[1 mark] Time-dependent solution is given by 

\[ \Psi(x,t) = \Psi(x,0) \exp(- \frac{iEt}{\hbar}) =  Ae^{ikx} \exp(- \frac{iEt}{\hbar})  \]

[1 mark] For free particle, \( V(x) = 0 \), \( E = \frac{\hbar^2 k^2}{2m} \) where k is a constant, 

\[ \Psi(x,t) = \Psi(x,0) \exp(- \frac{iEt}{\hbar}) =  A \exp\left[ i \left( kx - \frac{\hbar k^2 t}{2m} \right) \right] \]

[1 mark] Probability per unit length is the modulus of the wavefunction squared,

\[ |\Psi(x,t)|^2 = |A|^2 \]

which is independent of time

b) (5 marks)

[1 mark] 

\[ \frac{\partial \Psi}{\partial x} = Aik e^{ikx - \frac{iEt}{\hbar}} \] 

\[ \frac{\partial \Psi^{*}}{\partial x} = -Aik e^{-ikx + \frac{iEt}{\hbar}} \] 

[2 marks] Calculation of the probability flux

\[\Psi^{*}\frac{\partial \Psi}{\partial x} - \Psi\frac{\partial \Psi^{*}}{\partial x} = 2|A|^{2}ik\]
\[\Gamma = \frac{\left|A \right|^{2}\hbar k}{m} \]  

[2 marks] \( \Gamma \) is the number of particles which pass through a given point per unit time. \( \hbar k \) is the momentum hence \( \frac{\hbar k }{m} \) is the velocity.  \( |A|^2 \) is the no. particles per unit length

 \[\Gamma = v \left|A \right|^{2} \]  

where \(v\) is the velocity. 

"

Solution 1: 

"
\textbf{Solution:}  

Given the wavefunction $\psi(x) = A e^{ikx}$ for a particle in a space with potential energy $V(x) = 0$.  

The time-independent Schrödinger equation is given by  
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi(x)}{dx^2} + V(x)\psi(x) = E\psi(x).
\]  
Since $V(x) = 0$, this reduces to  
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi(x)}{dx^2} = E\psi(x).
\]  
Substituting $\psi(x) = A e^{ikx}$ into this equation:  
\[
-\frac{\hbar^2}{2m} \frac{d^2}{dx^2}\left(A e^{ikx}\right) = -\frac{\hbar^2}{2m}(-k^2)A e^{ikx} = E A e^{ikx}.
\]  
Thus, $E = \frac{\hbar^2 k^2}{2m}$.  

The corresponding time-dependent solution $\Psi(x,t)$ is given by  
\[
\Psi(x,t) = \psi(x) e^{-\frac{iEt}{\hbar}} = A e^{ikx} e^{-\frac{iEt}{\hbar}}.
\]  
Substituting $E = \frac{\hbar^2 k^2}{2m}$:  
\[
\Psi(x,t) = A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t}.
\]  

To find the probability per unit length of locating a particle, we compute the probability density:  
\[
|\Psi(x,t)|^2 = \Psi^*(x,t) \Psi(x,t).
\]  
Substituting $\Psi(x,t) = A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t}$:  
\[
\Psi^*(x,t) = A^* e^{-ikx} e^{i \frac{\hbar k^2}{2m} t}, \quad \Psi(x,t) = A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t}.
\]  
Therefore,  
\[
|\Psi(x,t)|^2 = \left(A^* A\right) \left(e^{-ikx} e^{ikx}\right) \left(e^{i \frac{\hbar k^2}{2m} t} e^{-i \frac{\hbar k^2}{2m} t}\right) = |A|^2.
\]  
The probability density $|\Psi(x,t)|^2 = |A|^2$ is independent of both $x$ and $t$, as required.  

For part (b), the particle flux $\Gamma$ is defined as  
\[
\Gamma = -\frac{i \hbar}{2m} \left[\Psi^* \frac{\partial \Psi}{\partial x} - \Psi \frac{\partial \Psi^*}{\partial x}\right].
\]  
Substituting $\Psi(x,t) = A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t}$:  
\[
\frac{\partial \Psi}{\partial x} = ik A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t}, \quad \frac{\partial \Psi^*}{\partial x} = -ik A^* e^{-ikx} e^{i \frac{\hbar k^2}{2m} t}.
\]  
Thus,  
\[
\Psi^* \frac{\partial \Psi}{\partial x} = A^* e^{-ikx} e^{i \frac{\hbar k^2}{2m} t} \cdot ik A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t} = ik |A|^2,
\]  
and  
\[
\Psi \frac{\partial \Psi^*}{\partial x} = A e^{ikx} e^{-i \frac{\hbar k^2}{2m} t} \cdot (-ik) A^* e^{-ikx} e^{i \frac{\hbar k^2}{2m} t} = -ik |A|^2.
\]  
Substituting these into the expression for $\Gamma$:  
\[
\Gamma = -\frac{i \hbar}{2m} \left[ik |A|^2 - (-ik |A|^2)\right] = -\frac{i \hbar}{2m} \cdot 2ik |A|^2.
\]  
Simplifying,  
\[
\Gamma = \frac{\hbar k}{m} |A|^2.
\]  

The particle flux $\Gamma$ is proportional to the velocity $v = \frac{\hbar k}{m}$ of the particles. Physically, $\Gamma$ represents the rate at which probability flows through a unit area, which corresponds to the product of the particle density $|A|^2$ and the velocity $v$.
"

Solution 2: 

"
To solve the given problem:

\textbf{Solution:}

The wavefunction of a beam of particles is given as $\psi(x) = A e^{ikx}$, where $k$ is the wavenumber, and the potential energy $V(x) = 0$.

a) The time-dependent solution $\Psi(x,t)$ can be obtained by multiplying the spatial wavefunction $\psi(x)$ by the time-dependent phase factor $e^{-iEt/\hbar}$, where $E$ is the energy of the particle. The energy of the particle for a free particle is given by:
\[
E = \frac{\hbar^2 k^2}{2m}.
\]
Thus, the time-dependent solution becomes:
\[
\Psi(x,t) = \psi(x)e^{-iEt/\hbar} = A e^{ikx} e^{-i(\hbar k^2 / 2m)t}.
\]

To show that the probability per unit length of finding a particle is independent of both space and time, we compute the probability density:
\[
|\Psi(x,t)|^2 = \Psi^*(x,t)\Psi(x,t).
\]
Substituting $\Psi(x,t) = A e^{ikx} e^{-i(\hbar k^2 / 2m)t}$, we get:
\[
|\Psi(x,t)|^2 = \left(A e^{ikx} e^{-i(\hbar k^2 / 2m)t}\right)^* \left(A e^{ikx} e^{-i(\hbar k^2 / 2m)t}\right).
\]
Taking the complex conjugate:
\[
|\Psi(x,t)|^2 = A^* e^{-ikx} e^{i(\hbar k^2 / 2m)t} \cdot A e^{ikx} e^{-i(\hbar k^2 / 2m)t}.
\]
Simplifying:
\[
|\Psi(x,t)|^2 = |A|^2.
\]
Since $|A|^2$ is a constant, the probability per unit length is independent of both space ($x$) and time ($t$).

\textbf{[3 marks]}

b) The particle flux $\Gamma$ is given by:
\[
\Gamma = -\frac{i \hbar}{2m} \left[ \Psi^* \frac{\partial \Psi}{\partial x} - \Psi \frac{\partial \Psi^*}{\partial x} \right].
\]
First, calculate $\frac{\partial \Psi}{\partial x}$:
\[
\Psi(x,t) = A e^{ikx} e^{-i(\hbar k^2 / 2m)t}.
\]
Differentiating with respect to $x$:
\[
\frac{\partial \Psi}{\partial x} = ik A e^{ikx} e^{-i(\hbar k^2 / 2m)t} = ik \Psi.
\]
Next, calculate $\frac{\partial \Psi^*}{\partial x}$:
\[
\Psi^*(x,t) = A^* e^{-ikx} e^{i(\hbar k^2 / 2m)t}.
\]
Differentiating with respect to $x$:
\[
\frac{\partial \Psi^*}{\partial x} = -ik A^* e^{-ikx} e^{i(\hbar k^2 / 2m)t} = -ik \Psi^*.
\]

Substitute these results into the expression for $\Gamma$:
\[
\Gamma = -\frac{i \hbar}{2m} \left[ \Psi^* (ik\Psi) - \Psi (-ik\Psi^*) \right].
\]
Simplify the terms:
\[
\Gamma = -\frac{i \hbar}{2m} \left[ i k \Psi^* \Psi + i k \Psi \Psi^* \right].
\]
\[
\Gamma = -\frac{i \hbar}{2m} \cdot 2i k |\Psi|^2.
\]
\[
\Gamma = \frac{\hbar k}{m} |\Psi|^2.
\]
Since $|\Psi|^2 = |A|^2$, we have:
\[
\Gamma = \frac{\hbar k}{m} |A|^2.
\]

The physical interpretation of $\Gamma$ is that it represents the flow of particles per unit time. The velocity of the particles is given by $v = \frac{\hbar k}{m}$, and hence $\Gamma$ depends on the product of the particle velocity and the probability density $|A|^2$.

\textbf{[5 marks]}

"

Solution 3:

"
\textbf{Solution to Q7:}

\textbf{Part (a):}  
The time-dependent solution $\Psi(x,t)$ is obtained by combining the spatial wavefunction $\psi(x) = A e^{ikx}$ with the time-dependent factor $e^{-i\omega t}$. Since the potential energy $V(x) = 0$, the energy $E$ is purely kinetic:  
\[
E = \frac{\hbar^2 k^2}{2m}.
\]  
The angular frequency $\omega$ is related to the energy by $\omega = \frac{E}{\hbar} = \frac{\hbar k^2}{2m}$. Thus, the time-dependent wavefunction is:  
\[
\Psi(x,t) = \psi(x)e^{-i\omega t} = A e^{ikx} e^{-i\omega t} = A e^{i(kx - \omega t)}.
\]

The probability density is given by:  
\[
|\Psi(x,t)|^2 = \Psi^*(x,t)\Psi(x,t).
\]  
Substituting $\Psi(x,t) = A e^{i(kx - \omega t)}$, we find:  
\[
\Psi^*(x,t) = A^* e^{-i(kx - \omega t)},
\]
\[
|\Psi(x,t)|^2 = A^*A e^{-i(kx - \omega t)}e^{i(kx - \omega t)} = |A|^2.
\]
This shows that the probability density $|\Psi(x,t)|^2 = |A|^2$ is independent of both space $x$ and time $t$, as required.

\textbf{Part (b):}  
The particle flux $\Gamma$ is defined as:  
\[
\Gamma = -\frac{i\hbar}{2m} \left[ \Psi^* \frac{\partial \Psi}{\partial x} - \Psi \frac{\partial \Psi^*}{\partial x} \right].
\]
We compute the necessary derivatives. From $\Psi(x,t) = A e^{i(kx - \omega t)}$,  
\[
\frac{\partial \Psi}{\partial x} = i k A e^{i(kx - \omega t)} = i k \Psi(x,t).
\]
Similarly,  
\[
\Psi^*(x,t) = A^* e^{-i(kx - \omega t)}, \quad \frac{\partial \Psi^*}{\partial x} = -i k A^* e^{-i(kx - \omega t)} = -i k \Psi^*(x,t).
\]
Substituting these into the expression for $\Gamma$:  
\[
\Gamma = -\frac{i\hbar}{2m} \left[ \Psi^* (i k \Psi) - \Psi (-i k \Psi^*) \right].
\]
Simplify the terms:  
\[
\Gamma = -\frac{i\hbar}{2m} \left[ i k \Psi^* \Psi + i k \Psi \Psi^* \right] = -\frac{i\hbar}{2m} (2 i k \Psi^* \Psi).
\]
\[
\Gamma = \frac{\hbar k}{m} |\Psi|^2.
\]
Substituting $|\Psi|^2 = |A|^2$, we have:  
\[
\Gamma = \frac{\hbar k}{m} |A|^2.
\]

The physical interpretation of $\Gamma$ is as follows: the particle flux represents the flow of probability density per unit time. The term $\frac{\hbar k}{m}$ corresponds to the velocity of the particles, $v = \frac{\hbar k}{m}$, as expected for a free particle in quantum mechanics. Thus, the flux $\Gamma$ is proportional to the velocity of the particles and the probability density $|A|^2$.
"

\subsubsection{QMQ8}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 9 marks. Give up to 5 marks for part a), and give up to 4 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"The expectation of an operator \( \hat{Q} \) in one dimension can be written 

\[ \left< \hat{Q} \right> = \int_{-\infty}^{\infty} \Psi^{*}(x) \hat{Q} \Psi(x) \, dx \]

a) For an operator \( \hat{Q} \) which does not vary with time, show the rate of change with time of the expectation value of \( \hat{Q} \) can be written

\[ \frac{d}{dt}\left<\hat{Q}\right> = \frac{1}{i\hbar}\left< \left[ \hat{Q}, \hat{H} \right] \right> \]

where \( \hat{H} \) is the Hamiltonian operator. 

b) A particle of mass \( m \) is subject to a time-independent potential \( V(x) \). By evaluating \( \left[ \hat{X}, \hat{H}\right] \), where \( \hat{X} \) is the position operator, show that 

\[ m\frac{d}{dt}\left<\hat{X}\right> = \left<\hat{P}\right> \]

You may use the additional information: 

Time-dependent Schrödinger equation: 
\[ \frac{\partial\Psi}{\partial t} = \frac{1}{i\hbar}\hat{H}\Psi \]"

Mark Scheme:

"{Q8: (9 marks total) }

a) (5 marks)

[2 marks] \[ \frac{d}{dt} \left< \hat{Q} \right> = \int_{-\infty}^{\infty} \frac{\partial}{\partial t} \left( \Psi^{*}(x) \hat{Q} \Psi(x) \right) \, dx \]

\[ = \int_{-\infty}^{\infty} \frac{\partial \Psi^{*}}{\partial t}\hat{Q}\Psi + \Psi^{*}\hat{Q}\frac{\partial \Psi}{\partial t} \, dx \]

[1 mark] substitute partial derivatives of \(\Psi\), \(\Psi^{*}\) with respect to t using Hamiltonian expression.

[2 marks] Algebraic simplification to 

\[ \frac{1}{i\hbar} \int_{-\infty}^{\infty} \Psi^{*} \left [ \hat{Q}, \hat{H} \right ] \Psi \, dx \]

\[ = \frac{1}{i\hbar} \left<  \left [ \hat{Q}, \hat{H} \right ] \right> \]

b) (4 marks)

[1 mark]\[ \frac{d}{dt}  \left<\hat{X}\right> = \frac{1}{i\hbar}\left< \left[ \hat{X}, \hat{H} \right] \right> \] 

[2 marks] \[\left[ \hat{X}, \hat{H}\right] = \left[ \hat{X}, \frac{\hat{P}^{2}}{2m} + V(x)\right] = \frac{i\hbar \hat{P}}{m} \]

[1 mark] Substitute second expression into first expression to give 

\[m \frac{d}{dt}\left<\hat{X}\right> =  \left<\hat{P}\right> \]

"

Solution 1: 

"
\textbf{Solution:}

\textbf{(a)} To derive the time rate of change of the expectation value of $\hat{Q}$, we start with its definition:
\[
\langle \hat{Q} \rangle = \int_{-\infty}^\infty \Psi^*(x, t) \hat{Q} \Psi(x, t) \, dx
\]
The time derivative of $\langle \hat{Q} \rangle$ is given by:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{d}{dt} \int_{-\infty}^\infty \Psi^*(x, t) \hat{Q} \Psi(x, t) \, dx
\]
Applying the product rule under the integral sign, we get:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \int_{-\infty}^\infty \left( \frac{\partial \Psi^*}{\partial t} \hat{Q} \Psi + \Psi^* \hat{Q} \frac{\partial \Psi}{\partial t} \right) dx
\]
Using the time-dependent Schrödinger equation, 
\[
\frac{\partial \Psi}{\partial t} = \frac{1}{i\hbar} \hat{H} \Psi \quad \text{and} \quad \frac{\partial \Psi^*}{\partial t} = -\frac{1}{i\hbar} \Psi^* \hat{H},
\]
we substitute these into the expression:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \int_{-\infty}^\infty \left( -\frac{1}{i\hbar} \Psi^* \hat{H} \hat{Q} \Psi + \frac{1}{i\hbar} \Psi^* \hat{Q} \hat{H} \Psi \right) dx
\]
Combine the terms under the same integral:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \int_{-\infty}^\infty \Psi^* \left( \hat{Q} \hat{H} - \hat{H} \hat{Q} \right) \Psi \, dx
\]
Recognizing the commutator $\left[ \hat{Q}, \hat{H} \right] = \hat{Q} \hat{H} - \hat{H} \hat{Q}$, this becomes:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \int_{-\infty}^\infty \Psi^* \left[ \hat{Q}, \hat{H} \right] \Psi \, dx
\]
Thus, the result can be written as:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \langle [\hat{Q}, \hat{H}] \rangle
\]

\textbf{(b)} For a particle of mass $m$ in a time-independent potential $V(x)$, the Hamiltonian operator is:
\[
\hat{H} = \frac{\hat{P}^2}{2m} + V(\hat{X}),
\]
where $\hat{P}$ is the momentum operator and $\hat{X}$ is the position operator. To evaluate the commutator $[\hat{X}, \hat{H}]$, we first compute:
\[
[\hat{X}, \hat{H}] = \left[\hat{X}, \frac{\hat{P}^2}{2m}\right] + \left[\hat{X}, V(\hat{X})\right].
\]
For the second term, $\hat{X}$ commutes with any function of $\hat{X}$, so:
\[
\left[\hat{X}, V(\hat{X})\right] = 0.
\]
For the first term, using the canonical commutation relation $[\hat{X}, \hat{P}] = i\hbar$, we find:
\[
\left[\hat{X}, \frac{\hat{P}^2}{2m}\right] = \frac{1}{2m} \left( [\hat{X}, \hat{P} \hat{P}] \right) = \frac{1}{2m} \left( \hat{P} [\hat{X}, \hat{P}] + [\hat{X}, \hat{P}] \hat{P} \right).
\]
Substituting $[\hat{X}, \hat{P}] = i\hbar$, we get:
\[
\left[\hat{X}, \frac{\hat{P}^2}{2m}\right] = \frac{1}{2m} \left( \hat{P} (i\hbar) + (i\hbar) \hat{P} \right) = \frac{i\hbar}{m} \hat{P}.
\]
Thus:
\[
[\hat{X}, \hat{H}] = \frac{i\hbar}{m} \hat{P}.
\]

Using the result from part (a), the time derivative of $\langle \hat{X} \rangle$ is:
\[
\frac{d}{dt} \langle \hat{X} \rangle = \frac{1}{i\hbar} \langle [\hat{X}, \hat{H}] \rangle.
\]
Substituting $[\hat{X}, \hat{H}] = \frac{i\hbar}{m} \hat{P}$, we find:
\[
\frac{d}{dt} \langle \hat{X} \rangle = \frac{1}{i\hbar} \langle \frac{i\hbar}{m} \hat{P} \rangle = \frac{1}{m} \langle \hat{P} \rangle.
\]
Multiplying through by $m$, we obtain:
\[
m \frac{d}{dt} \langle \hat{X} \rangle = \langle \hat{P} \rangle.
\]
"

Solution 2: 

"
\textbf{Solution:}

\textbf{(a)}  
The expectation value of an operator $\hat{Q}$ is defined as  
\[
\langle \hat{Q} \rangle = \int_{-\infty}^\infty \Psi^*(x,t) \hat{Q} \Psi(x,t) \, dx.
\]  
Taking the time derivative of this expectation value, we write  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{d}{dt} \int_{-\infty}^\infty \Psi^*(x,t) \hat{Q} \Psi(x,t) \, dx.
\]  
Using Leibniz's rule, the time derivative can be applied to the integrand:  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \int_{-\infty}^\infty \left( \frac{\partial \Psi^*}{\partial t} \hat{Q} \Psi + \Psi^* \hat{Q} \frac{\partial \Psi}{\partial t} \right) dx.
\]  
From the time-dependent Schrödinger equation,  
\[
\frac{\partial \Psi}{\partial t} = \frac{1}{i\hbar} \hat{H} \Psi, \quad \frac{\partial \Psi^*}{\partial t} = -\frac{1}{i\hbar} \Psi^* \hat{H}.
\]  
Substituting these into the integral, we obtain:  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \int_{-\infty}^\infty \left( -\frac{1}{i\hbar} \Psi^* \hat{H} \hat{Q} \Psi + \frac{1}{i\hbar} \Psi^* \hat{Q} \hat{H} \Psi \right) dx.
\]  
Factorizing the terms, this becomes  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \int_{-\infty}^\infty \Psi^* \left( \hat{Q} \hat{H} - \hat{H} \hat{Q} \right) \Psi \, dx.
\]  
The commutator $[\hat{Q}, \hat{H}]$ is defined as $[\hat{Q}, \hat{H}] = \hat{Q}\hat{H} - \hat{H}\hat{Q}$. Thus,  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \int_{-\infty}^\infty \Psi^* [\hat{Q}, \hat{H}] \Psi \, dx.
\]  
By the definition of expectation value, this simplifies to  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \langle [\hat{Q}, \hat{H}] \rangle.
\]  

\textbf{(b)}  
The Hamiltonian operator for a particle of mass $m$ in a time-independent potential $V(x)$ is  
\[
\hat{H} = \frac{\hat{P}^2}{2m} + V(\hat{X}),
\]  
where $\hat{X}$ is the position operator and $\hat{P}$ is the momentum operator. We aim to compute the commutator $[\hat{X}, \hat{H}]$.

Using the definition of $\hat{H}$,  
\[
[\hat{X}, \hat{H}] = \left[\hat{X}, \frac{\hat{P}^2}{2m} \right] + [\hat{X}, V(\hat{X})].
\]  
The potential $V(\hat{X})$ depends only on $\hat{X}$, so $[\hat{X}, V(\hat{X})] = 0$. Thus,  
\[
[\hat{X}, \hat{H}] = \frac{1}{2m} [\hat{X}, \hat{P}^2].
\]  
Using the property of commutators, $[\hat{X}, \hat{P}^2] = [\hat{X}, \hat{P}] \hat{P} + \hat{P} [\hat{X}, \hat{P}]$. The canonical commutation relation is $[\hat{X}, \hat{P}] = i\hbar$, so  
\[
[\hat{X}, \hat{P}^2] = (i\hbar) \hat{P} + \hat{P} (i\hbar) = 2i\hbar \hat{P}.
\]  
Substituting this into the previous expression,  
\[
[\hat{X}, \hat{H}] = \frac{1}{2m} (2i\hbar \hat{P}) = \frac{i\hbar}{m} \hat{P}.
\]  
The time evolution of $\langle \hat{X} \rangle$ is given by  
\[
\frac{d}{dt} \langle \hat{X} \rangle = \frac{1}{i\hbar} \langle [\hat{X}, \hat{H}] \rangle.
\]  
Substituting $[\hat{X}, \hat{H}] = \frac{i\hbar}{m} \hat{P}$, we find  
\[
\frac{d}{dt} \langle \hat{X} \rangle = \frac{1}{i\hbar} \langle \frac{i\hbar}{m} \hat{P} \rangle = \frac{1}{m} \langle \hat{P} \rangle.
\]  
Multiplying through by $m$, we arrive at  
\[
m \frac{d}{dt} \langle \hat{X} \rangle = \langle \hat{P} \rangle.
\]

"

Solution 3:

"
To solve this problem, we address parts (a) and (b) step by step.

**(a) Derivation of the rate of change of the expectation value of $\hat{Q}$** [5 marks]  

The expectation value of the operator $\hat{Q}$ is given by  
\[
\langle \hat{Q} \rangle = \int_{-\infty}^\infty \Psi^*(x, t) \hat{Q} \Psi(x, t) \, dx.
\]  

Differentiating $\langle \hat{Q} \rangle$ with respect to time:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \int_{-\infty}^\infty \frac{\partial \Psi^*}{\partial t} \hat{Q} \Psi \, dx + \int_{-\infty}^\infty \Psi^* \hat{Q} \frac{\partial \Psi}{\partial t} \, dx.
\]  

Using the time-dependent Schrödinger equation:
\[
\frac{\partial \Psi}{\partial t} = \frac{1}{i\hbar} \hat{H} \Psi, \quad \text{and its complex conjugate,} \quad \frac{\partial \Psi^*}{\partial t} = -\frac{1}{i\hbar} \Psi^* \hat{H},
\]
we substitute these into the expression for $\frac{d}{dt} \langle \hat{Q} \rangle$:  
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \int_{-\infty}^\infty \left(-\frac{1}{i\hbar} \Psi^* \hat{H} \hat{Q} \Psi \right) dx + \int_{-\infty}^\infty \left(\frac{1}{i\hbar} \Psi^* \hat{Q} \hat{H} \Psi \right) dx.
\]  

Combining terms, we have:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \int_{-\infty}^\infty \Psi^* \left(\hat{Q} \hat{H} - \hat{H} \hat{Q} \right) \Psi \, dx.
\]  

The commutator $[\hat{Q}, \hat{H}]$ is defined as:
\[
[\hat{Q}, \hat{H}] = \hat{Q} \hat{H} - \hat{H} \hat{Q}.
\]

Thus, we can rewrite the rate of change of the expectation value as:
\[
\frac{d}{dt} \langle \hat{Q} \rangle = \frac{1}{i\hbar} \langle [\hat{Q}, \hat{H}] \rangle.
\]  

This completes the derivation.  

**(b) Showing that $m \frac{d}{dt} \langle \hat{X} \rangle = \langle \hat{P} \rangle$** [4 marks]  

We start with the commutator $[\hat{X}, \hat{H}]$, where $\hat{H}$ is the Hamiltonian operator given by:
\[
\hat{H} = \frac{\hat{P}^2}{2m} + V(\hat{X}).
\]

The commutator is:
\[
[\hat{X}, \hat{H}] = \left[\hat{X}, \frac{\hat{P}^2}{2m} + V(\hat{X})\right].
\]

Since the position operator $\hat{X}$ commutes with the potential $V(\hat{X})$ (because $V(\hat{X})$ is a function of $\hat{X}$), we have:
\[
[\hat{X}, \hat{H}] = \frac{1}{2m} [\hat{X}, \hat{P}^2].
\]

Expanding $[\hat{X}, \hat{P}^2]$ using the property of commutators:
\[
[\hat{X}, \hat{P}^2] = [\hat{X}, \hat{P}] \hat{P} + \hat{P} [\hat{X}, \hat{P}].
\]

The fundamental commutator $[\hat{X}, \hat{P}]$ is:
\[
[\hat{X}, \hat{P}] = i\hbar.
\]

Substituting this result:
\[
[\hat{X}, \hat{P}^2] = i\hbar \hat{P} + \hat{P} i\hbar = 2i\hbar \hat{P}.
\]

Thus:
\[
[\hat{X}, \hat{H}] = \frac{1}{2m} (2i\hbar \hat{P}) = \frac{i\hbar}{m} \hat{P}.
\]

Using the result from part (a), the rate of change of $\langle \hat{X} \rangle$ is:
\[
\frac{d}{dt} \langle \hat{X} \rangle = \frac{1}{i\hbar} \langle [\hat{X}, \hat{H}] \rangle.
\]

Substituting $[\hat{X}, \hat{H}] = \frac{i\hbar}{m} \hat{P}$:
\[
\frac{d}{dt} \langle \hat{X} \rangle = \frac{1}{i\hbar} \langle \frac{i\hbar}{m} \hat{P} \rangle = \frac{1}{m} \langle \hat{P} \rangle.
\]

Multiplying through by $m$, we obtain:
\[
m \frac{d}{dt} \langle \hat{X} \rangle = \langle \hat{P} \rangle.
\]

This completes the derivation.
"

\subsubsection{QMQ9}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 17 marks. Give up to 5 marks for part a), give up to 5 marks for part b) and give up to 7 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution. 

Problem:

"Consider a beam of particles each of mass \( m \) with energy \( E > 0 \) incident from the left, subject to a one-dimensional potential step defined by

\[ V(x) = \begin{cases}
0 & \text{for } x \leq 0,\\
-V_{0}  & \text{for } x > 0 
\end{cases} \]

a) Show that \( \psi_{1}(x) = e^{ik_{1}x} + Be^{-ik_{1}x} \) is the general solution in the region \( x \leq 0 \) and \( \psi_{2}(x) = Ce^{ik_{2}x} \) is the general solution in the region \( x > 0 \), where \( B \) and \( C \) are constants. Define \( k_{1} \), \( k_{2} \) as part of your answer.

b) By applying appropriate boundary conditions, show that 

\[ C = \frac{2k_{1}}{k_{1}+k_{2}} \] and \[ B = \frac{k_{1}-k_{2}}{k_{1}+k_{2}} \]

c) Calculate the incident, reflected, and transmitted flux for this scattering potential and derive that the probability for transmission, \( T \), and the probability for reflection, \( R \), are given by 

\[ T = \frac{4k_{1}k_{2}}{(k_{1}+k_{2})^{2}} \]

\[ R = \frac{(k_{1}-k_{2})^{2}}{(k_{1}+k_{2})^{2}} \]

You may use that the particle flux for a particle beam is given by 

\[ \Gamma(x) = -\frac{i\hbar}{2m} \left(\psi^{*}\frac{d\psi}{dx} - \psi\frac{d\psi^{*}}{dx} \right) \]"

Mark Scheme:

"{Q9: (17 marks total) }

a) (5 marks)

[2 marks] The general solution of free particle solution is \( Ae^{ik_{1}x} + Be^{-ik_{1}x} \). We can choose \( |A|^{2} = 1 \), to represent a flux of one particle, thus choose \( A = 1 \). 

[1 mark] \[ k_1 = \frac{\sqrt{2mE}}{\hbar} \]

[1 mark] In the \( x > 0 \) region, the solution has the same form as before, \( \psi_{2}(x) = Ce^{ik_{2}x} + De^{-ik_{2}x} \). No potential for beam to scatter off and reflect traveling left, hence set \( D = 0 \). 

[1 mark] \[k_{2} = \frac{\sqrt{2m(E+V_0)}}{\hbar} \]

b) (5 marks)

[1 mark] Calculating derivatives:
\[ \frac{d\psi_1}{dx} = ik_{1}e^{ik_{1}x} - ik_{1}Be^{-ik_{1}x} \]

\[ \frac{d\psi_2}{dx} = ik_{2}Ce^{ik_{2}x} \]

[2 marks] Equate \( \psi_{1} = \psi_{2} \), \( \frac{d\psi_1}{dx} = \frac{d\psi_2}{dx} \) at \(x = 0\) to obtain

\[1 + B = C\]
\[ik_{1} - ik_{1}B = ik_{2}C\]

[2 marks] The addition of the previous 2 equations results in \[ C = \frac{2k_{1}}{k_{1}+k_{2}} \] and the subtraction leads to  \[ B = \frac{k_{1}-k_{2}}{k_{1}+k_{2}} \]

c) (7 marks)

[1 mark] substituting the incident beam \( e^{ik_{1}x} \) into the formula for flux to get \[\Gamma_{incident} = \frac{\hbar k_{1}}{m} \]

[2 marks] substituting the reflected beam \( Be^{-ik_{1}x} \) into the formula for flux to get 

\[ 
\Gamma_{reflected} = \frac{-\hbar k_{1}}{m}\frac{\left(k_{1} - k_{2} \right)^{2}}{ \left(k_{1} + k_{2} \right)^{2}} 
\]

[2 marks] substituting the transmitted beam \( Ce^{ik_{2}x} \) into the formula for flux to get \[\Gamma_{transmitted} = \frac{\hbar k_{2}}{m} \frac{4k_{1}^{2}}{ \left(k_{1} + k_{2} \right)^{2}} \]

[2 marks]

\[ T = \left|\frac{\Gamma_{transmitted}}{\Gamma_{incident}} \right| = \frac{4k_{1}k_{2}}{(k_{1}+k_{2})^{2}} \]

\[R = \left|\frac{\Gamma_{reflected}}{\Gamma_{incident}} \right| = \frac{(k_{1}-k_{2})^{2}}{(k_{1}+k_{2})^{2}} \]

"

Solution 1: 

"
\textbf{Solution:}

\textbf{(a) [5 marks]}

To solve the Schrödinger equation in each region, we note that the time-independent Schrödinger equation is:
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi(x)}{dx^2} + V(x) \psi(x) = E \psi(x).
\]

For $x \leq 0$, the potential $V(x) = 0$, so the Schrödinger equation becomes:
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_1(x)}{dx^2} = E \psi_1(x).
\]
Defining $k_1 = \sqrt{\frac{2mE}{\hbar^2}}$, the solution to this equation is:
\[
\psi_1(x) = e^{ik_1x} + Be^{-ik_1x},
\]
where $e^{ik_1x}$ represents the incident wave, and $Be^{-ik_1x}$ represents the reflected wave.

For $x > 0$, the potential $V(x) = -V_0$, so the Schrödinger equation becomes:
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_2(x)}{dx^2} - V_0 \psi_2(x) = E \psi_2(x).
\]
This can be written as:
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_2(x)}{dx^2} = (E + V_0) \psi_2(x).
\]
Defining $k_2 = \sqrt{\frac{2m(E + V_0)}{\hbar^2}}$, the solution is:
\[
\psi_2(x) = Ce^{ik_2x},
\]
where $Ce^{ik_2x}$ represents the transmitted wave.

Thus, the general solutions are:
\[
\psi_1(x) = e^{ik_1x} + Be^{-ik_1x} \quad \text{for } x \leq 0,
\]
\[
\psi_2(x) = Ce^{ik_2x} \quad \text{for } x > 0.
\]

\textbf{(b) [5 marks]}

The boundary conditions for the wavefunction require that both $\psi(x)$ and $\frac{d\psi}{dx}$ are continuous at $x = 0$. Using these conditions:

1. Continuity of $\psi(x)$ at $x = 0$:
\[
\psi_1(0) = \psi_2(0),
\]
\[
1 + B = C.
\]

2. Continuity of $\frac{d\psi}{dx}$ at $x = 0$:
\[
\frac{d\psi_1}{dx}\bigg|_{x=0} = \frac{d\psi_2}{dx}\bigg|_{x=0},
\]
\[
ik_1(1 - B) = ik_2C.
\]

From the first boundary condition, solve for $B$ in terms of $C$:
\[
B = C - 1.
\]

Substitute $B = C - 1$ into the second boundary condition:
\[
ik_1(1 - (C - 1)) = ik_2C,
\]
\[
ik_1(2 - C) = ik_2C.
\]

Simplify to solve for $C$:
\[
2k_1 = C(k_1 + k_2),
\]
\[
C = \frac{2k_1}{k_1 + k_2}.
\]

Substitute $C = \frac{2k_1}{k_1 + k_2}$ into $B = C - 1$ to find $B$:
\[
B = \frac{2k_1}{k_1 + k_2} - 1,
\]
\[
B = \frac{2k_1 - (k_1 + k_2)}{k_1 + k_2},
\]
\[
B = \frac{k_1 - k_2}{k_1 + k_2}.
\]

Thus:
\[
C = \frac{2k_1}{k_1 + k_2}, \quad B = \frac{k_1 - k_2}{k_1 + k_2}.
\]

\textbf{(c) [7 marks]}

The particle flux for a wavefunction $\psi(x)$ is given by:
\[
\Gamma(x) = -\frac{i\hbar}{2m} \left( \psi^* \frac{d\psi}{dx} - \psi \frac{d\psi^*}{dx} \right).
\]

For the incident wave, $\psi_\text{inc} = e^{ik_1x}$:
\[
\Gamma_\text{inc} = \frac{\hbar k_1}{m}.
\]

For the reflected wave, $\psi_\text{refl} = Be^{-ik_1x}$:
\[
\Gamma_\text{refl} = -\frac{\hbar k_1}{m} |B|^2.
\]

For the transmitted wave, $\psi_\text{trans} = Ce^{ik_2x}$:
\[
\Gamma_\text{trans} = \frac{\hbar k_2}{m} |C|^2.
\]

The transmission probability $T$ is the ratio of the transmitted flux to the incident flux:
\[
T = \frac{\Gamma_\text{trans}}{\Gamma_\text{inc}} = \frac{\frac{\hbar k_2}{m} |C|^2}{\frac{\hbar k_1}{m}} = \frac{k_2}{k_1} |C|^2.
\]
Substitute $C = \frac{2k_1}{k_1 + k_2}$:
\[
T = \frac{k_2}{k_1} \left(\frac{2k_1}{k_1 + k_2}\right)^2,
\]
\[
T = \frac{4k_1k_2}{(k_1 + k_2)^2}.
\]

The reflection probability $R$ is the ratio of the reflected flux to the incident flux:
\[
R = \frac{\Gamma_\text{refl}}{\Gamma_\text{inc}} = \frac{-\frac{\hbar k_1}{m} |B|^2}{\frac{\hbar k_1}{m}} = |B|^2.
\]
Substitute $B = \frac{k_1 - k_2}{k_1 + k_2}$:
\[
R = \left(\frac{k_1 - k_2}{k_1 + k_2}\right)^2.
\]

Thus, the transmission and reflection probabilities are:
\[
T = \frac{4k_1k_2}{(k_1 + k_2)^2}, \quad R = \frac{(k_1 - k_2)^2}{(k_1 + k_2)^2}.
\]

\textbf{Verification:} Since probability is conserved, we must have $T + R = 1$:
\[
T + R = \frac{4k_1k_2}{(k_1 + k_2)^2} + \frac{(k_1 - k_2)^2}{(k_1 + k_2)^2},
\]
\[
T + R = \frac{4k_1k_2 + (k_1 - k_2)^2}{(k_1 + k_2)^2}.
\]
Expand $(k_1 - k_2)^2$:
\[
(k_1 - k_2)^2 = k_1^2 - 2k_1k_2 + k_2^2,
\]
\[
T + R = \frac{4k_1k_2 + k_1^2 - 2k_1k_2 + k_2^2}{(k_1 + k_2)^2},
\]
\[
T + R = \frac{k_1^2 + 2k_1k_2 + k_2^2}{(k_1 + k_2)^2},
\]
\[
T + R = \frac{(k_1 + k_2)^2}{(k_1 + k_2)^2} = 1.
\]
Thus, probability is conserved.
"

Solution 2: 

"
\textbf{Solution:}

\textbf{a)} [5 marks]  
The time-independent Schrödinger equation is given by  
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi(x)}{dx^2} + V(x)\psi(x) = E\psi(x).
\]  
For the region $x \leq 0$, $V(x) = 0$, so the equation becomes:  
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_1(x)}{dx^2} = E\psi_1(x).
\]  
Defining $k_1 = \sqrt{\frac{2mE}{\hbar^2}}$, the solution is:  
\[
\psi_1(x) = e^{ik_1x} + Be^{-ik_1x},
\]  
where $e^{ik_1x}$ represents the incident wave and $Be^{-ik_1x}$ represents the reflected wave.  

For the region $x > 0$, $V(x) = -V_0$, so the equation becomes:  
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_2(x)}{dx^2} = (E + V_0)\psi_2(x).
\]  
Defining $k_2 = \sqrt{\frac{2m(E+V_0)}{\hbar^2}}$, the solution is:  
\[
\psi_2(x) = Ce^{ik_2x},
\]  
where $Ce^{ik_2x}$ represents the transmitted wave.  

Thus, the general solutions are:  
\[
\psi_1(x) = e^{ik_1x} + Be^{-ik_1x} \quad \text{for } x \leq 0,
\]  
\[
\psi_2(x) = Ce^{ik_2x} \quad \text{for } x > 0.
\]  

\textbf{b)} [5 marks]  
The boundary conditions require that $\psi(x)$ and $\frac{d\psi(x)}{dx}$ are continuous at $x = 0$. Applying these conditions:  

1. Continuity of $\psi(x)$ at $x = 0$:  
\[
\psi_1(0) = \psi_2(0).
\]  
Substituting the solutions:  
\[
1 + B = C.
\]  

2. Continuity of $\frac{d\psi(x)}{dx}$ at $x = 0$:  
\[
\frac{d\psi_1}{dx}\Big|_{x=0} = \frac{d\psi_2}{dx}\Big|_{x=0}.
\]  
Substituting the derivatives:  
\[
ik_1(1 - B) = ik_2C.
\]  

From the first condition, $C = 1 + B$. Substituting into the second condition:  
\[
ik_1(1 - B) = ik_2(1 + B).
\]  
Simplify to solve for $B$:  
\[
k_1 - k_1B = k_2 + k_2B,
\]  
\[
B(k_1 + k_2) = k_1 - k_2,
\]  
\[
B = \frac{k_1 - k_2}{k_1 + k_2}.
\]  

Substitute $B$ into $C = 1 + B$:  
\[
C = 1 + \frac{k_1 - k_2}{k_1 + k_2} = \frac{2k_1}{k_1 + k_2}.
\]  

Thus,  
\[
C = \frac{2k_1}{k_1 + k_2}, \quad B = \frac{k_1 - k_2}{k_1 + k_2}.
\]  

\textbf{c)} [7 marks]  
The particle flux for a particle beam is given by:  
\[
\Gamma(x) = -\frac{i\hbar}{2m} \left( \psi^* \frac{d\psi}{dx} - \psi \frac{d\psi^*}{dx} \right).
\]  

1. Incident flux:  
\[
\psi_\text{inc} = e^{ik_1x}, \quad \frac{d\psi_\text{inc}}{dx} = ik_1 e^{ik_1x}.
\]  
Substituting into the flux expression:  
\[
\Gamma_\text{inc} = \frac{\hbar k_1}{m}.
\]  

2. Reflected flux:  
\[
\psi_\text{ref} = Be^{-ik_1x}, \quad \frac{d\psi_\text{ref}}{dx} = -ik_1 B e^{-ik_1x}.
\]  
Substituting:  
\[
\Gamma_\text{ref} = -\frac{\hbar k_1}{m} |B|^2.
\]  

3. Transmitted flux:  
\[
\psi_\text{trans} = Ce^{ik_2x}, \quad \frac{d\psi_\text{trans}}{dx} = ik_2 C e^{ik_2x}.
\]  
Substituting:  
\[
\Gamma_\text{trans} = \frac{\hbar k_2}{m} |C|^2.
\]  

The probabilities for reflection ($R$) and transmission ($T$) are given by:  
\[
R = \frac{\Gamma_\text{ref}}{\Gamma_\text{inc}}, \quad T = \frac{\Gamma_\text{trans}}{\Gamma_\text{inc}}.
\]  

Substituting the fluxes:  
\[
R = \frac{|B|^2}{1}, \quad T = \frac{k_2}{k_1} |C|^2.
\]  

Using $B = \frac{k_1 - k_2}{k_1 + k_2}$ and $C = \frac{2k_1}{k_1 + k_2}$:  
\[
|B|^2 = \left(\frac{k_1 - k_2}{k_1 + k_2}\right)^2, \quad |C|^2 = \left(\frac{2k_1}{k_1 + k_2}\right)^2.
\]  

Substitute into $R$ and $T$:  
\[
R = \frac{(k_1 - k_2)^2}{(k_1 + k_2)^2}, \quad T = \frac{k_2}{k_1} \cdot \frac{4k_1^2}{(k_1 + k_2)^2}.
\]  

Simplify $T$:  
\[
T = \frac{4k_1k_2}{(k_1 + k_2)^2}.
\]  

Thus,  
\[
R = \frac{(k_1 - k_2)^2}{(k_1 + k_2)^2}, \quad T = \frac{4k_1k_2}{(k_1 + k_2)^2}.
\]  
\[
\boxed{R + T = 1}.
\]  "

Solution 3:

"
\textbf{Solution:}

\textbf{Part a) [5 marks]} 

To solve for the wavefunctions in the two regions, we solve the time-independent Schrödinger equation:
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi}{dx^2} + V(x)\psi = E\psi.
\]

For $x \leq 0$, the potential is $V(x) = 0$. Thus, the Schrödinger equation becomes
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_1}{dx^2} = E \psi_1.
\]
Let $k_1 = \sqrt{\frac{2mE}{\hbar^2}}$. The solution is a combination of plane waves:
\[
\psi_1(x) = e^{ik_1x} + B e^{-ik_1x},
\]
where the term $e^{ik_1x}$ corresponds to the incident wave and $Be^{-ik_1x}$ corresponds to the reflected wave.

For $x > 0$, the potential is $V(x) = -V_0$. Thus, the Schrödinger equation becomes
\[
-\frac{\hbar^2}{2m} \frac{d^2 \psi_2}{dx^2} - V_0 \psi_2 = E \psi_2.
\]
Let $k_2 = \sqrt{\frac{2m(E+V_0)}{\hbar^2}}$. The solution, as there is no reflected wave in this region, is
\[
\psi_2(x) = C e^{ik_2x},
\]
where $Ce^{ik_2x}$ corresponds to the transmitted wave.

Thus, the general solutions are:
\[
\psi_1(x) = e^{ik_1x} + Be^{-ik_1x} \quad \text{for } x \leq 0,
\]
\[
\psi_2(x) = Ce^{ik_2x} \quad \text{for } x > 0.
\]

\textbf{Part b) [5 marks]} 

To find $B$ and $C$, we apply the boundary conditions at $x = 0$:
1. Continuity of the wavefunction: $\psi_1(0) = \psi_2(0)$.
2. Continuity of the derivative: $\frac{d\psi_1}{dx}\big|_{x=0} = \frac{d\psi_2}{dx}\big|_{x=0}$.

At $x = 0$, $\psi_1(0) = 1 + B$ and $\psi_2(0) = C$. From the first boundary condition:
\[
1 + B = C.
\]

For the derivative, $\frac{d\psi_1}{dx} = ik_1 e^{ik_1x} - ik_1 Be^{-ik_1x}$ and $\frac{d\psi_2}{dx} = ik_2 Ce^{ik_2x}$. At $x = 0$:
\[
ik_1 (1 - B) = ik_2 C.
\]
Substitute $C = 1 + B$ into this equation:
\[
ik_1 (1 - B) = ik_2 (1 + B).
\]
Simplify:
\[
k_1 - k_1B = k_2 + k_2B.
\]
Rearrange:
\[
B(k_1 + k_2) = k_1 - k_2.
\]
Solve for $B$:
\[
B = \frac{k_1 - k_2}{k_1 + k_2}.
\]

Now substitute $B$ into $C = 1 + B$:
\[
C = 1 + \frac{k_1 - k_2}{k_1 + k_2}.
\]
Simplify:
\[
C = \frac{2k_1}{k_1 + k_2}.
\]

Thus:
\[
C = \frac{2k_1}{k_1 + k_2}, \quad B = \frac{k_1 - k_2}{k_1 + k_2}.
\]

\textbf{Part c) [7 marks]} 

The particle flux for a wavefunction $\psi$ is given by:
\[
\Gamma(x) = -\frac{i\hbar}{2m} \left( \psi^* \frac{d\psi}{dx} - \psi \frac{d\psi^*}{dx} \right).
\]

For the incident wave $\psi_\text{inc} = e^{ik_1x}$:
\[
\Gamma_\text{inc} = \frac{\hbar k_1}{m}.
\]

For the reflected wave $\psi_\text{ref} = Be^{-ik_1x}$:
\[
\Gamma_\text{ref} = -\frac{\hbar k_1}{m} |B|^2.
\]

For the transmitted wave $\psi_\text{trans} = Ce^{ik_2x}$:
\[
\Gamma_\text{trans} = \frac{\hbar k_2}{m} |C|^2.
\]

The reflection coefficient $R$ is the ratio of the reflected flux to the incident flux:
\[
R = \frac{\Gamma_\text{ref}}{\Gamma_\text{inc}} = \frac{-\frac{\hbar k_1}{m} |B|^2}{\frac{\hbar k_1}{m}} = |B|^2.
\]
Substitute $B = \frac{k_1 - k_2}{k_1 + k_2}$:
\[
R = \left( \frac{k_1 - k_2}{k_1 + k_2} \right)^2.
\]

The transmission coefficient $T$ is the ratio of the transmitted flux to the incident flux:
\[
T = \frac{\Gamma_\text{trans}}{\Gamma_\text{inc}} = \frac{\frac{\hbar k_2}{m} |C|^2}{\frac{\hbar k_1}{m}} = \frac{k_2}{k_1} |C|^2.
\]
Substitute $C = \frac{2k_1}{k_1 + k_2}$:
\[
T = \frac{k_2}{k_1} \left( \frac{2k_1}{k_1 + k_2} \right)^2.
\]
Simplify:
\[
T = \frac{4k_1k_2}{(k_1 + k_2)^2}.
\]

Thus:
\[
T = \frac{4k_1k_2}{(k_1 + k_2)^2}, \quad R = \frac{(k_1 - k_2)^2}{(k_1 + k_2)^2}.
\]
\[
T + R = \frac{4k_1k_2}{(k_1 + k_2)^2} + \frac{(k_1 - k_2)^2}{(k_1 + k_2)^2} = 1.
\]
This confirms conservation of probability.
"

\subsubsection{QMQ10}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 16 marks. Give up to 5 marks for part a), give up to 2 marks for part b) and give up to 9 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem:

"Consider the system of a quantum harmonic oscillator with eigenstates written \( \left| n \right> \) and corresponding eigenvalues \( E_{n} = \hbar\omega(n+\frac{1}{2}) \).

You are given the raising and lower operators \( \hat{a}_{\pm} \) defined by:

\[ \hat{a}_{\pm} = \frac{1}{\sqrt{2}}(\alpha \hat{x} \mp \frac{i}{\hbar\alpha} \hat{p}) \]

\[ \alpha = \sqrt{\frac{m\omega}{\hbar}} \]

where \( \hat{x} \), \( \hat{p} \) are the one-dimensional position and momentum operators respectively.

a) Normalize the state \( \left| \psi \right> =  \left| 0 \right> + b \left| 1 \right> \) and calculate the expectation value of the Hamiltonian for \( \left| \psi \right> \). \( b \) is a constant.

b) A perturbation of the form \( \hat{H}' = q \mathcal{E} \hat{x} \) is introduced to the system, where \( \hat{x} \) is the one-dimensional position operator. Rewrite the perturbation in terms of raising and lowering operators \( \hat{a}_{\pm} \).

c) By applying the variational principle, show that using \( \left| \psi \right> \) as a trial state, the value of \( b \) which minimizes energy is 

\[ b = \frac{E_{1} - E_{0}}{g} - \sqrt{\frac{(E_{1} - E_{0})^{2}}{g^{2}} -1} \]

where \( g = \sqrt{2} q \mathcal{E} / {\alpha} \)"

Mark Scheme:

"{Q10: (16 marks total) }

a) (5 marks total)

[2 marks] To normalize, attach factor to \( \left| \psi \right> \) and set the inner product to 1. Choose real a and assume b is a real constant. 

\[ a \left| \psi \right> = a\left| 0 \right> + ab \left| 1 \right> \]
\[ \left< a\psi | a\psi \right> = |a|^2 \left< \psi | \psi \right> = 1 \]

[1 mark] Expand inner product, simplify algebra to obtain value of \( a \), normalized \( \left| \psi \right> \).

\[ a = \frac{1}{\sqrt{1+b^2}} \]

\[ \left| \psi \right>_{norm} = \frac{1}{\sqrt{1+b^2}} \left| 0 \right> + \frac{b}{\sqrt{1+b^2}} \left| 1 \right> \]

[2 marks] Compute expectation value of Hamiltonian, \( \langle H \rangle \) 

\[ \langle H \rangle = \frac{1}{1+b^2} \frac{\hbar \omega}{2} + \frac{b^2}{1+b^2} \frac{3\hbar \omega}{2} \]

\[ \langle H \rangle = \left ( \frac{1+3b^2}{1+b^2} \right) \frac{\hbar \omega}{2} \]

b) (2 marks)

[1 mark] rewrite \( \hat{x} \) in terms of ladder operators,

\[ \hat{x} = \frac{1}{\sqrt{2} \alpha} (\hat{a}_{+} + \hat{a}_{-}) \]

[1 mark] substitute into \( \hat{H}' \), 

\[ \hat{H}' = \frac{\sqrt{2} q \mathcal{E}}{2 \alpha} (\hat{a}_{+} + \hat{a}_{-}) = \frac{g}{2} (\hat{a}_{+} + \hat{a}_{-}) \]

Expression in terms of g not required. 

c) (9 marks)

[1 mark] Writing expectation value of original Hamiltonian, 

\[ \langle H \rangle = \frac{1}{1+b^2} E_0 + \frac{b^2}{1+b^2} E_1 \]

[3 marks] Calculation of the expectation value of the perturbation, \( \hat{H}' \), using normalized ket

\[ \hat{H}' \left| \psi \right> = \frac{g}{2} \frac{1}{\sqrt{1+b^2}} (b \left| 0 \right> + \left| 1 \right> ) \]

\[ \langle H' \rangle = \langle \psi | \hat{H}' | \psi \rangle = \frac{gb}{1+b^2}\]

[1 mark] Obtain total expectation value of the Hamiltonian. 

\[ \langle H_{tot} \rangle = \langle H \rangle + \langle H' \rangle = \frac{E_0+b^2E_1}{1+b^2} + \frac{gb}{1+b^2} = \frac{E_1 b^2 + gb + E_0}{1+b^2} \]

[1 mark] Find \( b \) to minimize \( \langle H_{tot} \rangle \) by differentiating w.r.t \( b \), setting it to 0. 

\[ \frac{\partial \langle H_{tot} \rangle}{\partial b} = \frac{(2E_1 b + g)(1+b^2) - 2b(E_1 b^2 + gb + E_0)}{(1+b^2)^2} = 0 \]

[2 marks] Solve equation to give b, 

\[ gb^2 - 2(E_1 - E_0)b - g = 0 \]

\[ b = \frac{E_1 - E_0}{g} \pm \sqrt{\frac{(E_1-E_0)^2}{g^2}-1} \]

[1 mark] Choose and justify taking the negative square root. Take the negative square root since smaller b reduces expectation value as \(  \left| \psi \right> \sim \left| 0 \right> + b \left| 1 \right> \), then the ground \( \left| 0 \right> \) state is weighted more heavily. 
"

Solution 1: 

"

\noindent \textbf{Solution:}

\textbf{(a) Normalization and Expectation Value of the Hamiltonian [5 marks]}

The given state is:
\[
\vert \psi \rangle = \vert 0 \rangle + b \vert 1 \rangle.
\]
To normalize this state, we impose the condition:
\[
\langle \psi \vert \psi \rangle = 1.
\]
Expanding,
\[
\langle 0 \vert 0 \rangle + b^2 \langle 1 \vert 1 \rangle = 1.
\]
Since the eigenstates of the quantum harmonic oscillator are orthonormal, we have:
\[
1 + b^2 = 1 \Rightarrow b^2 = 1 - 1 = 1 \Rightarrow \vert \psi \rangle = \frac{1}{\sqrt{1 + b^2}} (\vert 0 \rangle + b \vert 1 \rangle).
\]

The expectation value of the Hamiltonian, \( \hat{H} \), is given by:
\[
\langle \hat{H} \rangle = \langle \psi \vert \hat{H} \vert \psi \rangle.
\]
Since \( \hat{H} \vert n \rangle = E_n \vert n \rangle \) and using \( E_n = \hbar \omega (n + 1/2) \), we compute:
\[
E_0 = \frac{1}{2} \hbar \omega, \quad E_1 = \frac{3}{2} \hbar \omega.
\]
Thus,
\[
\langle \hat{H} \rangle = \frac{1}{1 + b^2} \left( \frac{1}{2} \hbar \omega + b^2 \cdot \frac{3}{2} \hbar \omega \right).
\]
Simplifying,
\[
\langle \hat{H} \rangle = \frac{1}{2} \hbar \omega \frac{1 + 3b^2}{1 + b^2}.
\]

\textbf{(b) Expressing the Perturbation in Terms of Raising and Lowering Operators [2 marks]}

The position operator \( \hat{x} \) can be written in terms of the ladder operators as:
\[
\hat{x} = \frac{1}{\alpha \sqrt{2}} (\hat{a}_+ + \hat{a}_-).
\]
Thus, the perturbation Hamiltonian becomes:
\[
\hat{H}' = q\mathcal{E} \hat{x} = \frac{q\mathcal{E}}{\alpha \sqrt{2}} (\hat{a}_+ + \hat{a}_-).
\]

\textbf{(c) Variational Principle and Minimization of Energy [9 marks]}

We seek to find the value of \( b \) that minimizes \( \langle \hat{H} \rangle \) using the variational principle. The expectation value of the perturbation is:
\[
\langle \psi \vert \hat{H}' \vert \psi \rangle = \frac{q\mathcal{E}}{\alpha \sqrt{2}} \cdot \frac{2b}{1 + b^2}.
\]

The total energy expectation is:
\[
E_{\psi} = \langle \hat{H} \rangle + \langle \hat{H}' \rangle.
\]
Substituting the computed values,
\[
E_{\psi} = \frac{1}{2} \hbar \omega \frac{1 + 3b^2}{1 + b^2} + \frac{q\mathcal{E}}{\alpha \sqrt{2}} \cdot \frac{2b}{1 + b^2}.
\]
Differentiating with respect to \( b \) and setting it to zero for minimization,
\[
\frac{d}{db} \left( \frac{1 + 3b^2}{1 + b^2} \right) \cdot \frac{1}{2} \hbar \omega + \frac{d}{db} \left( \frac{2b}{1 + b^2} \right) \cdot \frac{q\mathcal{E}}{\alpha \sqrt{2}} = 0.
\]
Solving for \( b \), we obtain:
\[
 b = \frac{E_1 - E_0}{g} - \sqrt{\frac{(E_1 - E_0)^2}{g^2} - 1},
\]
where
\[
 g = \frac{\sqrt{2q\mathcal{E}}}{\alpha}.
\]
This gives the required result, completing the proof.

"

Solution 2: 

"
\textbf{Solution:}

\textbf{(a)} To normalize the state \(|\psi\rangle = |0\rangle + b|1\rangle\), we require that \(\langle \psi | \psi \rangle = 1\). Expanding this expression:

\[
\langle \psi | \psi \rangle = \langle 0 | 0 \rangle + b^* \langle 0 | 1 \rangle + b \langle 1 | 0 \rangle + |b|^2 \langle 1 | 1 \rangle.
\]

Since the states \(|0\rangle\) and \(|1\rangle\) are orthonormal, \(\langle 0 | 0 \rangle = 1\), \(\langle 1 | 1 \rangle = 1\), and \(\langle 0 | 1 \rangle = \langle 1 | 0 \rangle = 0\). Thus, we have:

\[
\langle \psi | \psi \rangle = 1 + |b|^2.
\]

For normalization, set \(\langle \psi | \psi \rangle = 1\):

\[
1 + |b|^2 = 1 \implies |b|^2 = 0 \implies |b| = 0.
\]

To calculate the expectation value of the Hamiltonian, recall that for the quantum harmonic oscillator, the Hamiltonian is given by:

\[
\hat{H} = \hbar \omega \left( \hat{a}^\dagger \hat{a} + \frac{1}{2} \right),
\]

where \(\hat{a}^\dagger\) and \(\hat{a}\) are the raising and lowering operators. The energy eigenvalues are \(E_n = \hbar \omega \left( n + \frac{1}{2} \right)\), so:

\[
\langle 0 | \hat{H} | 0 \rangle = E_0 = \frac{\hbar \omega}{2}, \quad \langle 1 | \hat{H} | 1 \rangle = E_1 = \frac{3 \hbar \omega}{2}.
\]

The expectation value of the Hamiltonian in the state \(|\psi\rangle = |0\rangle + b|1\rangle\) is:

\[
\langle \hat{H} \rangle = \frac{\langle \psi | \hat{H} | \psi \rangle}{\langle \psi | \psi \rangle}.
\]

Using the expansion:

\[
\langle \psi | \hat{H} | \psi \rangle = \langle 0 | \hat{H} | 0 \rangle + b^* \langle 0 | \hat{H} | 1 \rangle + b \langle 1 | \hat{H} | 0 \rangle + |b|^2 \langle 1 | \hat{H} | 1 \rangle.
\]

Since \(\langle 0 | \hat{H} | 1 \rangle = \langle 1 | \hat{H} | 0 \rangle = 0\), we get:

\[
\langle \psi | \hat{H} | \psi \rangle = \frac{\hbar \omega}{2} + |b|^2 \frac{3 \hbar \omega}{2}.
\]

Using the normalization condition \(\langle \psi | \psi \rangle = 1\), the expectation value simplifies to:

\[
\langle \hat{H} \rangle = \frac{\frac{\hbar \omega}{2} + |b|^2 \frac{3 \hbar \omega}{2}}{1 + |b|^2}.
\]

\textbf{(b)} The perturbation is given as \(\hat{H}' = q \mathcal{E} \hat{x}\). Substituting the position operator \(\hat{x}\) in terms of raising and lowering operators:

\[
\hat{x} = \sqrt{\frac{\hbar}{2m\omega}} (\hat{a} + \hat{a}^\dagger),
\]

the perturbation becomes:

\[
\hat{H}' = q \mathcal{E} \sqrt{\frac{\hbar}{2m\omega}} (\hat{a} + \hat{a}^\dagger).
\]

Thus, the perturbation in terms of \(\hat{a}\) and \(\hat{a}^\dagger\) is:

\[
\hat{H}' = g (\hat{a} + \hat{a}^\dagger),
\]

where \(g = q \mathcal{E} \sqrt{\frac{\hbar}{2m\omega}}\).

\textbf{(c)} Using the variational principle, the energy of the trial state \(|\psi\rangle = |0\rangle + b|1\rangle\) is given by:

\[
E(b) = \frac{\langle \psi | \hat{H} + \hat{H}' | \psi \rangle}{\langle \psi | \psi \rangle}.
\]

Substituting \(\hat{H}' = g (\hat{a} + \hat{a}^\dagger)\) and using the orthonormality of \(|0\rangle\) and \(|1\rangle\), the energy becomes:

\[
E(b) = \frac{\hbar \omega}{2} + |b|^2 \hbar \omega + \frac{2 g b}{1 + |b|^2}.
\]

Minimizing \(E(b)\) with respect to \(b\), take the derivative:

\[
\frac{dE}{db} = 2 \hbar \omega b + \frac{2g(1 - b^2)}{(1 + b^2)^2}.
\]

Setting \(\frac{dE}{db} = 0\), we solve:

\[
2 \hbar \omega b (1 + b^2)^2 + 2g(1 - b^2) = 0.
\]

This simplifies to:

\[
b = \frac{E_1 - E_0}{g} - \sqrt{\frac{(E_1 - E_0)^2}{g^2} - 1}.
\]

Here, \(E_1 - E_0 = \hbar \omega\), and \(g = \sqrt{2} q \mathcal{E}/\alpha\), yielding the desired result.

"

Solution 3:

"
\textbf{Solution:}

\textbf{(a) Normalize the state \(|\psi\rangle = |0\rangle + b |1\rangle\) and calculate the expectation value of the Hamiltonian for \(|\psi\rangle\).}

The state \(|\psi\rangle = |0\rangle + b |1\rangle\) must be normalized. This means:
\[
\langle \psi | \psi \rangle = 1.
\]
Substituting \(|\psi\rangle\), we have:
\[
\langle \psi | \psi \rangle = \langle 0 | 0 \rangle + b^* \langle 0 | 1 \rangle + b \langle 1 | 0 \rangle + |b|^2 \langle 1 | 1 \rangle.
\]
Using the orthonormality conditions \(\langle 0 | 0 \rangle = \langle 1 | 1 \rangle = 1\) and \(\langle 0 | 1 \rangle = \langle 1 | 0 \rangle = 0\), we get:
\[
\langle \psi | \psi \rangle = 1 + |b|^2.
\]
To normalize the state, we require:
\[
1 + |b|^2 = 1 \implies |b|^2 = 1 \quad \implies |b| = 1.
\]
The normalized state is therefore:
\[
|\psi\rangle = \frac{|0\rangle + b |1\rangle}{\sqrt{1 + |b|^2}}.
\]

Now, compute the expectation value of the Hamiltonian \(\hat{H}\) for \(|\psi\rangle\). The Hamiltonian of the quantum harmonic oscillator is given by:
\[
\hat{H} = \hbar\omega \left(\hat{a}^\dagger \hat{a} + \frac{1}{2}\right).
\]
The eigenvalues corresponding to \(|n\rangle\) are \(E_n = \hbar\omega\left(n + \frac{1}{2}\right)\). Thus:
\[
\langle \psi | \hat{H} | \psi \rangle = \frac{\langle 0 | \hat{H} | 0 \rangle + b^* \langle 0 | \hat{H} | 1 \rangle + b \langle 1 | \hat{H} | 0 \rangle + |b|^2 \langle 1 | \hat{H} | 1 \rangle}{1 + |b|^2}.
\]
Using \(\hat{H} |0\rangle = \frac{\hbar\omega}{2} |0\rangle\) and \(\hat{H} |1\rangle = \frac{3\hbar\omega}{2} |1\rangle\), we evaluate the terms:
\[
\langle 0 | \hat{H} | 0 \rangle = \frac{\hbar\omega}{2}, \quad \langle 0 | \hat{H} | 1 \rangle = 0, \quad \langle 1 | \hat{H} | 0 \rangle = 0, \quad \langle 1 | \hat{H} | 1 \rangle = \frac{3\hbar\omega}{2}.
\]
Thus:
\[
\langle \psi | \hat{H} | \psi \rangle = \frac{\frac{\hbar\omega}{2} + |b|^2 \cdot \frac{3\hbar\omega}{2}}{1 + |b|^2}.
\]
Substituting \(|b|^2 = 1\), we get:
\[
\langle \psi | \hat{H} | \psi \rangle = \frac{\frac{\hbar\omega}{2} + \frac{3\hbar\omega}{2}}{2} = \hbar\omega.
\]

\textbf{(b) Rewrite the perturbation \(\hat{H}' = q \mathcal{E} \hat{x}\) in terms of raising and lowering operators \(\hat{a}_\pm\).}

The position operator \(\hat{x}\) in terms of \(\hat{a}_\pm\) is given by:
\[
\hat{x} = \frac{1}{\sqrt{2} \alpha} (\hat{a}_+ + \hat{a}_-).
\]
Substituting this into \(\hat{H}' = q \mathcal{E} \hat{x}\), we have:
\[
\hat{H}' = q \mathcal{E} \frac{1}{\sqrt{2} \alpha} (\hat{a}_+ + \hat{a}_-).
\]
Thus:
\[
\hat{H}' = \frac{q \mathcal{E}}{\sqrt{2} \alpha} (\hat{a}_+ + \hat{a}_-).
\]

\textbf{(c) Apply the variational principle to find \(b\) that minimizes energy.}

The variational principle states that the expectation value of the Hamiltonian gives an upper bound on the ground state energy:
\[
E[\psi] = \langle \psi | \hat{H} + \hat{H}' | \psi \rangle.
\]
Using the trial state \(|\psi\rangle = |0\rangle + b |1\rangle\), the total energy is:
\[
E[b] = \frac{\frac{\hbar\omega}{2} + |b|^2 \cdot \frac{3\hbar\omega}{2} + \langle \psi | \hat{H}' | \psi \rangle}{1 + |b|^2}.
\]
From (b), the perturbation is proportional to \(\hat{x}\). Using \(|0\rangle\) and \(|1\rangle\), the matrix elements of \(\hat{x}\) are:
\[
\langle 0 | \hat{H}' | 1 \rangle = \frac{q \mathcal{E}}{\sqrt{2} \alpha}, \quad \langle 1 | \hat{H}' | 0 \rangle = \frac{q \mathcal{E}}{\sqrt{2} \alpha}.
\]
The expectation value of \(\hat{H}'\) is:
\[
\langle \psi | \hat{H}' | \psi \rangle = \frac{2 b \cdot \frac{q \mathcal{E}}{\sqrt{2} \alpha}}{1 + |b|^2}.
\]
Simplify \(E[b]\) as:
\[
E[b] = \frac{\frac{\hbar\omega}{2} + |b|^2 \cdot \frac{3\hbar\omega}{2} + 2b \cdot \frac{q \mathcal{E}}{\sqrt{2} \alpha}}{1 + |b|^2}.
\]
To minimize \(E[b]\), solve \(\frac{\partial E[b]}{\partial b} = 0\). After differentiation and simplification, we find:
\[
b = \frac{E_1 - E_0}{g} - \sqrt{\frac{(E_1 - E_0)^2}{g^2} - 1},
\]
where \(g = \sqrt{2} q \mathcal{E} / \alpha\).

"

%--------------------------------------------------------------

\subsection{CM}

Classical Mechanics Prompts

\subsubsection{CMQ1}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 4 marks. Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem: 

"Define what is a conservative force in the context of classical mechanics. For a conservative force \( \underline{F} \), mathematically define the potential energy. Why is it not possible to define a potential energy for a force which is not conservative?"

Mark Scheme:

"{Q1: (4 marks total)}

[1 mark] Conservative force: a force where its work done on an object is independent of the object path.

[2 marks] The potential energy for a conservative force \( \underline{F} \) is the function defined by 

\[ \int_{O}^{r} - \underline{F} \cdot d\underline{r} \] 

where O is a fixed origin point and r is any point in space. Must say the chosen point of origin is arbitrary.  

[1 mark] Not possible to define potential energy for non-conservative force since if the line integral as defined above is path dependent, then each potential energy at each point will be non-unique, have multiple values and is not a function. 
"

Solution 1: 

"
A force $\mathbf{F}$ is said to be \textit{conservative} if the work done by the force on a particle moving between two points depends only on the initial and final positions and not on the path taken. Mathematically, a force $\mathbf{F}$ is conservative if:
\[
\oint \mathbf{F} \cdot d\mathbf{r} = 0
\]
for any closed path. This implies that the work done over a closed loop is zero. Equivalently, $\mathbf{F}$ is conservative if it can be expressed as the gradient of a scalar potential function $U(\mathbf{r})$, such that:
\[
\mathbf{F} = -\nabla U(\mathbf{r}).
\]

The potential energy $U(\mathbf{r})$ associated with a conservative force $\mathbf{F}$ is defined as:
\[
U(\mathbf{r}) = - \int_{\mathbf{r}_0}^{\mathbf{r}} \mathbf{F} \cdot d\mathbf{r},
\]
where $\mathbf{r}_0$ is a reference point at which the potential energy is defined to be zero or some arbitrary constant. The negative sign ensures that the force points in the direction of decreasing potential energy.

For non-conservative forces, the work done depends on the path taken between two points, violating the requirement for defining potential energy. Specifically, for a non-conservative force $\mathbf{F}$:
\[
\oint \mathbf{F} \cdot d\mathbf{r} \neq 0.
\]
This means no scalar potential function $U(\mathbf{r})$ exists such that $\mathbf{F} = -\nabla U(\mathbf{r})$. As a result, potential energy cannot be uniquely defined, because the work done by the force cannot be expressed solely as a difference in potential energy between two points.

In summary, conservative forces allow the definition of potential energy and facilitate the use of energy conservation principles, while non-conservative forces, such as friction, dissipate energy and cannot be described by a potential energy function."

Solution 2: 

"
In classical mechanics, a \textbf{conservative force} is defined as a force for which the work done in moving a particle between two points is independent of the path taken. Mathematically, the work done by a conservative force $\mathbf{F}$ on a particle moving from point $\mathbf{r}_1$ to $\mathbf{r}_2$ is given by:
\[
W = \int_{\mathbf{r}_1}^{\mathbf{r}_2} \mathbf{F} \cdot d\mathbf{r}.
\]
If $\mathbf{F}$ is conservative, then the work $W$ depends only on the initial and final positions, not on the path taken. Furthermore, for any closed path (a path where $\mathbf{r}_1 = \mathbf{r}_2$), the work done by a conservative force is zero:
\[
\oint \mathbf{F} \cdot d\mathbf{r} = 0.
\]

A conservative force can be expressed as the gradient of a scalar potential energy function $U(\mathbf{r})$:
\[
\mathbf{F} = -\nabla U(\mathbf{r}).
\]

The negative sign indicates that the force points in the direction of decreasing potential energy. The potential energy $U(\mathbf{r})$ is defined (up to an arbitrary constant) such that:
\[
U(\mathbf{r}) = - \int_{\mathbf{r}_0}^{\mathbf{r}} \mathbf{F} \cdot d\mathbf{r},
\]
where $\mathbf{r}_0$ is a reference point where the potential energy is defined to be zero (or some constant value).

If the force is \textbf{not conservative}, it is impossible to define a scalar potential energy function $U(\mathbf{r})$ because the work done by the force depends on the path taken. For non-conservative forces, such as friction, the work done around a closed path is not zero:
\[
\oint \mathbf{F} \cdot d\mathbf{r} \neq 0.
\]

This path dependence violates a key property required to define potential energy: the ability to assign a unique scalar value $U(\mathbf{r})$ to every point in space. Consequently, for non-conservative forces, energy cannot be stored in a potential energy function, and the force cannot be expressed as the gradient of a scalar function."

Solution 3:

"
A \textbf{conservative force} in classical mechanics is defined as a force for which the work done by the force on a particle moving between two points is independent of the path taken. Mathematically, this implies that the work done by the force around any closed path is zero:
\[
\oint \mathbf{F} \cdot d\mathbf{r} = 0.
\]
This condition ensures that the force can be expressed as the negative gradient of a scalar potential function:
\[
\mathbf{F} = -\nabla U,
\]
where $U$ is the potential energy associated with the force. \\

The potential energy $U$ is defined such that the work done by the force $\mathbf{F}$ when moving from a point $\mathbf{r}_1$ to a point $\mathbf{r}_2$ is equal to the negative change in the potential energy:
\[
W = \int_{\mathbf{r}_1}^{\mathbf{r}_2} \mathbf{F} \cdot d\mathbf{r} = U(\mathbf{r}_1) - U(\mathbf{r}_2).
\]

For a force that is not conservative, the work done depends on the path taken. In such cases, it is not possible to define a scalar potential energy function $U$ because the change in potential energy would not be uniquely defined for a given displacement. Non-conservative forces, such as friction, dissipate energy as heat or other forms of non-recoverable energy, making it impossible to recover the energy through conservative means. As a result, the concept of potential energy is not applicable to non-conservative forces."

\subsubsection{CMQ2}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 6 marks. Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem: 

"The damped harmonic oscillator is governed by an equation of motion

\[
\frac{d^{2}x}{dt^{2}} + b\frac{dx}{dt}+kx=0
\]

State the 3 cases of damping and qualitatively describe the behavior of each case."

Mark Scheme:

"{Q2: (6 marks total)}

[2 marks] Underdamped,  \( b^2 < 4k \). Solution is sinusoidal motion which has an exponentially decaying amplitude 

[2 marks] Critically damped, \( b^2 = 4k \). Exponential decay, position decays the fastest to 0 in this case. 

[2 marks] Overdamped, \( b^2 > 4k \). The solution for position is just an exponential decay. \\

Note \( b^2 > 4k \) is equivalent to \( b^2 > 4km \).

"

Solution 1: 

"
The equation of motion for the damped harmonic oscillator is given by:
\[
\frac{d^2x}{dt^2} + b\frac{dx}{dt} + kx = 0
\]
where:
- \( \frac{d^2x}{dt^2} \) represents the acceleration,
- \( b\frac{dx}{dt} \) is the damping term (with damping coefficient \( b \)),
- \( kx \) is the restoring force due to the spring (with spring constant \( k \)).

To analyze the behavior of the system, we rewrite the equation as:
\[
\frac{d^2x}{dt^2} + 2\zeta\omega_0\frac{dx}{dt} + \omega_0^2x = 0
\]
where:
- \( \omega_0 = \sqrt{\frac{k}{m}} \) is the natural angular frequency of the undamped oscillator,
- \( \zeta = \frac{b}{2\sqrt{mk}} \) is the dimensionless damping ratio.

The nature of the solution depends on the value of the damping ratio \( \zeta \), which determines the type of damping. There are three cases:

1. **Underdamped (\( 0 < \zeta < 1 \)):**
   - In this case, the damping is small, and the system oscillates while the amplitude decreases exponentially over time.
   - The solution is of the form:
     \[
     x(t) = e^{-\zeta\omega_0 t} \left( A \cos(\omega t) + B \sin(\omega t) \right)
     \]
     where \( \omega = \omega_0 \sqrt{1 - \zeta^2} \) is the damped angular frequency.
   - Behavior: The system oscillates at a reduced frequency compared to the undamped case, and the oscillations gradually die out due to the damping.

2. **Critically Damped (\( \zeta = 1 \)):**
   - In this case, the system returns to equilibrium as quickly as possible without oscillating.
   - The solution is of the form:
     \[
     x(t) = (A + Bt)e^{-\omega_0 t}
     \]
   - Behavior: The system does not oscillate and smoothly approaches the equilibrium position. This is the fastest decay to equilibrium without oscillation.

3. **Overdamped (\( \zeta > 1 \)):**
   - In this case, the damping is so strong that the system does not oscillate and returns to equilibrium very slowly.
   - The solution is of the form:
     \[
     x(t) = C_1 e^{r_1 t} + C_2 e^{r_2 t}
     \]
     where \( r_1 \) and \( r_2 \) are two distinct negative roots of the characteristic equation:
     \[
     r^2 + 2\zeta\omega_0 r + \omega_0^2 = 0.
     \]
   - Behavior: The system exponentially decays to equilibrium without oscillating, but slower than in the critically damped case.

\textbf{Qualitative Summary:}
- \textit{Underdamped:} Oscillatory behavior with gradually decaying amplitude.
- \textit{Critically Damped:} No oscillation, fastest return to equilibrium.
- \textit{Overdamped:} No oscillation, slower return to equilibrium compared to critical damping.

"

Solution 2: 

"

The equation of motion for the damped harmonic oscillator is given by:
\[
\frac{d^2x}{dt^2} + b\frac{dx}{dt} + kx = 0
\]
where \( b \) is the damping coefficient and \( k \) is the spring constant. To analyze the behavior of the system, we first rewrite the equation in terms of the damping ratio. By dividing through by the mass \( m \), the standard form is:
\[
\ddot{x} + 2\zeta \omega_0 \dot{x} + \omega_0^2 x = 0,
\]
where:
\[
\omega_0 = \sqrt{\frac{k}{m}} \quad \text{is the natural frequency, and} \quad \zeta = \frac{b}{2\sqrt{mk}} \quad \text{is the damping ratio.}
\]

The behavior of the system depends on the value of the damping ratio \( \zeta \):
\begin{enumerate}
    \item \textbf{Underdamped Case (\( \zeta < 1 \)):}  
    In this case, the damping is small, and the system oscillates while gradually losing energy over time due to the damping force. The solution to the equation of motion is:
    \[
    x(t) = e^{-\zeta \omega_0 t} \left( A \cos(\omega_d t) + B \sin(\omega_d t) \right),
    \]
    where:
    \[
    \omega_d = \omega_0 \sqrt{1 - \zeta^2}
    \]
    is the damped natural frequency. The amplitude of the oscillations decays exponentially with time.

    \item \textbf{Critically Damped Case (\( \zeta = 1 \)):}  
    In this case, the system returns to equilibrium as quickly as possible without oscillating. The solution to the equation of motion is:
    \[
    x(t) = (A + Bt)e^{-\omega_0 t}.
    \]
    Here, \( A \) and \( B \) are constants determined by the initial conditions. The system does not oscillate, and the displacement decreases monotonically to zero.

    \item \textbf{Overdamped Case (\( \zeta > 1 \)):}  
    In this case, the damping is so strong that the system returns to equilibrium very slowly without oscillating. The solution to the equation of motion is:
    \[
    x(t) = C_1 e^{r_1 t} + C_2 e^{r_2 t},
    \]
    where:
    \[
    r_1 = -\omega_0 (\zeta + \sqrt{\zeta^2 - 1}) \quad \text{and} \quad r_2 = -\omega_0 (\zeta - \sqrt{\zeta^2 - 1}).
    \]
    Both roots are negative, leading to a slow exponential decay towards equilibrium.

\end{enumerate}

\textbf{Qualitative Summary:}
\begin{itemize}
    \item \textit{Underdamped:} Oscillatory motion with exponentially decreasing amplitude.
    \item \textit{Critically Damped:} No oscillations; the system quickly returns to equilibrium.
    \item \textit{Overdamped:} No oscillations; the system slowly returns to equilibrium.
\end{itemize}
"

Solution 3:

"

The equation of motion for a damped harmonic oscillator is given by:
\[
\frac{d^2x}{dt^2} + b\frac{dx}{dt} + kx = 0
\]
where \(b\) is the damping coefficient and \(k\) is the spring constant. The general solution to this second-order differential equation depends on the discriminant of the characteristic equation:
\[
\lambda^2 + b\lambda + k = 0
\]
The discriminant is given by:
\[
\Delta = b^2 - 4k
\]
Based on the value of \(\Delta\), there are three cases of damping:

---

\textbf{1. Overdamping (\(\Delta > 0\), \(b^2 > 4k\))}

In this case, the damping coefficient \(b\) is large, leading to two distinct real roots of the characteristic equation:
\[
\lambda_1, \lambda_2 = \frac{-b \pm \sqrt{b^2 - 4k}}{2}
\]
The solution is of the form:
\[
x(t) = C_1 e^{\lambda_1 t} + C_2 e^{\lambda_2 t}
\]
Here, both roots are negative (\(\lambda_1, \lambda_2 < 0\)), resulting in an exponential decay without oscillations. The system returns to equilibrium slowly, with no oscillatory behavior.

---

\textbf{2. Critical Damping (\(\Delta = 0\), \(b^2 = 4k\))}

In this case, the damping coefficient \(b\) is precisely balanced to avoid oscillations while ensuring the system returns to equilibrium as quickly as possible. The characteristic equation has a repeated real root:
\[
\lambda = \frac{-b}{2}
\]
The solution is of the form:
\[
x(t) = (C_1 + C_2 t) e^{\lambda t}
\]
The system decays to equilibrium without oscillating and does so at the fastest rate possible.

---

\textbf{3. Underdamping (\(\Delta < 0\), \(b^2 < 4k\))}

In this case, the damping coefficient \(b\) is small, resulting in complex conjugate roots of the characteristic equation:
\[
\lambda_{1,2} = \frac{-b}{2} \pm i \omega, \quad \text{where } \omega = \sqrt{\frac{4k - b^2}{4}}
\]
The solution is of the form:
\[
x(t) = e^{-\frac{b}{2}t} \left(C_1 \cos(\omega t) + C_2 \sin(\omega t)\right)
\]
The system exhibits oscillatory motion with an exponentially decaying amplitude. The oscillations are damped over time, and the system eventually comes to rest at equilibrium.

---

\textbf{Qualitative Summary:}
1. \textbf{Overdamping:} No oscillations; the system returns to equilibrium slowly.
2. \textbf{Critical Damping:} No oscillations; the system returns to equilibrium in the shortest time.
3. \textbf{Underdamping:} Oscillatory motion; the system oscillates with a decaying amplitude and eventually reaches equilibrium."


\subsubsection{CMQ3}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 6 marks. Give up to 2 marks for part a) and give up to 4 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem: 

"Two objects of masses \( m_{1} \) and \( m_{2} \) are separated by a distance \( d \). The object of mass \( m_{1} \) is at position \( \underline{r_{1}} \) and the object of mass \( m_{2} \) is at position \( \underline{r_{2}} \).

a) Starting from the definition of center of mass, show that its position vector can be given by 

\[ \underline{R} = (1-k)\underline{r_{1}} + k\underline{r_{2}} \]

and find the appropriate value of \( k \). 

b) Show that the center of mass lies on a line connecting the 2 masses and its distance from the 2 masses are \( d m_{2}/(m_{1}+m_{2}) \) from \( \underline{r_{1}} \) and \( d m_{1}/(m_{1}+m_{2}) \) from \( \underline{r_{2}} \)."

Mark Scheme:

"{Q3: (6 marks total)}

a) (2 marks) 

[1 mark] Use definition of center of mass 

\[ \underline{R} = \frac{m_1 \underline{r}_1 + m_2 \underline{r}_2}{m_1 + m_2} \]

[1 mark] Manipulate to the form 

\[ \underline{R} = \left ( 1 - \frac{m_2}{m_1 + m_2} \right ) \underline{r}_1 + \left ( \frac{m_2}{m_1 + m_2} \right ) \underline{r}_2 \]

\[ k = \frac{m_2}{m_1 + m_2} \]

b) [4 marks] 

[1 mark] calculate the vector \( \underline{R} - \underline{r}_1 \) using results of a), 

\[ \underline{R} - \underline{r}_1 = \left( \frac{m_2}{m_1+m_2} \right )(\underline{r}_2 - \underline{r}_1) \]

[1 mark] Distance from center of mass point to position \( \underline{r}_1 \) given by 

\[ d_{r1} = | \underline{R} - \underline{r}_1 | = \left( \frac{m_2}{m_1+m_2} \right )|(\underline{r}_2 - \underline{r}_1) | = \frac{m_2 d}{m_1+m_2} \]

[1 mark] calculate the vector \( \underline{R} - \underline{r}_2 \) using results of a), 

\[ \underline{R} - \underline{r}_2 = \left( \frac{m_1}{m_1+m_2} \right )(\underline{r}_1 - \underline{r}_2) \]

[1 mark] Distance from center of mass point to position \( \underline{r}_2 \) given by 

\[ d_{r2} = | \underline{R} - \underline{r}_2 | = \left( \frac{m_1}{m_1+m_2} \right )|(\underline{r}_1 - \underline{r}_2) | = \frac{m_1 d}{m_1+m_2} \]

"

Solution 1:

"
\textbf{(a) Starting from the definition of the center of mass, show that its position vector can be given by:}
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2},
\]
and find the appropriate value of $k$.

\textit{Definition of the center of mass:}
The center of mass position vector $\mathbf{R}$ is given by:
\[
\mathbf{R} = \frac{m_1 \mathbf{r_1} + m_2 \mathbf{r_2}}{m_1 + m_2}.
\]
This equation is derived from the principle that the center of mass is the weighted average position of the system.

\textit{Step 1: Rewriting the formula in terms of $k$.}  
We aim to rewrite $\mathbf{R}$ in the form:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2}.
\]
Comparing this with the definition of $\mathbf{R}$, we notice that $k$ must represent the fractional contribution of $m_2$ to the total mass. Thus, let:
\[
k = \frac{m_2}{m_1 + m_2}.
\]

\textit{Step 2: Substituting $k$ into the expression.}  
We rewrite the center of mass formula:
\[
\mathbf{R} = \frac{m_1 \mathbf{r_1} + m_2 \mathbf{r_2}}{m_1 + m_2}.
\]
Factor out $\frac{1}{m_1 + m_2}$:
\[
\mathbf{R} = \frac{m_1}{m_1 + m_2} \mathbf{r_1} + \frac{m_2}{m_1 + m_2} \mathbf{r_2}.
\]
Observe that:
\[
\frac{m_1}{m_1 + m_2} = 1 - \frac{m_2}{m_1 + m_2} = 1 - k.
\]
Therefore, the position vector becomes:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2},
\]
where $k = \frac{m_2}{m_1 + m_2}$. This confirms the required form of $\mathbf{R}$.

\vspace{0.5cm}
\textbf{(b) Show that the center of mass lies on the line connecting the two masses, and its distances from the two masses are:}
\[
\frac{d m_2}{m_1 + m_2} \text{ from } \mathbf{r_1} \quad \text{and} \quad \frac{d m_1}{m_1 + m_2} \text{ from } \mathbf{r_2}.
\]

\textit{Step 1: Express $\mathbf{r_2} - \mathbf{r_1}$ as a vector.}  
Let the distance between the two masses be $d$, so the vector between the two positions is:
\[
\mathbf{r_2} - \mathbf{r_1} = d \, \hat{\mathbf{u}},
\]
where $\hat{\mathbf{u}}$ is a unit vector along the line connecting $\mathbf{r_1}$ and $\mathbf{r_2}$. This ensures the center of mass lies on the same line.

\textit{Step 2: Distance of the center of mass from $\mathbf{r_1}$.}  
From part (a), the center of mass is given by:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2}.
\]
The vector displacement from $\mathbf{r_1}$ to $\mathbf{R}$ is:
\[
\mathbf{R} - \mathbf{r_1} = (1 - k)\mathbf{r_1} + k\mathbf{r_2} - \mathbf{r_1} = k(\mathbf{r_2} - \mathbf{r_1}).
\]
Substitute $\mathbf{r_2} - \mathbf{r_1} = d \, \hat{\mathbf{u}}$:
\[
\mathbf{R} - \mathbf{r_1} = k d \, \hat{\mathbf{u}}.
\]
The magnitude of this displacement is:
\[
|\mathbf{R} - \mathbf{r_1}| = k d.
\]
Substitute $k = \frac{m_2}{m_1 + m_2}$:
\[
|\mathbf{R} - \mathbf{r_1}| = \frac{m_2}{m_1 + m_2} d.
\]

\textit{Step 3: Distance of the center of mass from $\mathbf{r_2}$.}  
Similarly, the displacement vector from $\mathbf{R}$ to $\mathbf{r_2}$ is:
\[
\mathbf{r_2} - \mathbf{R} = \mathbf{r_2} - \left[(1 - k)\mathbf{r_1} + k\mathbf{r_2}\right] = (1 - k)(\mathbf{r_2} - \mathbf{r_1}).
\]
The magnitude of this displacement is:
\[
|\mathbf{r_2} - \mathbf{R}| = (1 - k)d.
\]
Substitute $1 - k = \frac{m_1}{m_1 + m_2}$:
\[
|\mathbf{r_2} - \mathbf{R}| = \frac{m_1}{m_1 + m_2} d.
\]

\textit{Conclusion:}  
The center of mass lies on the line connecting the two masses, and its distances from the two masses are:
\[
\frac{d m_2}{m_1 + m_2} \text{ from } \mathbf{r_1} \quad \text{and} \quad \frac{d m_1}{m_1 + m_2} \text{ from } \mathbf{r_2}.
\]"

Solution 2:

"

\textbf{(a) Deriving the position vector of the center of mass:}

The position of the center of mass $\mathbf{R}$ for two masses $m_1$ and $m_2$ located at positions $\mathbf{r_1}$ and $\mathbf{r_2}$ is given by the definition of the center of mass:
\[
\mathbf{R} = \frac{m_1\mathbf{r_1} + m_2\mathbf{r_2}}{m_1 + m_2}.
\]
This formula represents the weighted average of the positions of the two masses, where the weights are proportional to the masses. It ensures that the center of mass reflects the distribution of mass in the system.

Now, we aim to express $\mathbf{R}$ in the form:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2}.
\]
To do this, compare the two forms of $\mathbf{R}$. Rewrite the definition of the center of mass:
\[
\mathbf{R} = \frac{m_1}{m_1 + m_2}\mathbf{r_1} + \frac{m_2}{m_1 + m_2}\mathbf{r_2}.
\]

Notice that the coefficient of $\mathbf{r_1}$ is $\frac{m_1}{m_1 + m_2}$ and the coefficient of $\mathbf{r_2}$ is $\frac{m_2}{m_1 + m_2}$. To match the given form, let:
\[
1 - k = \frac{m_1}{m_1 + m_2} \quad \text{and} \quad k = \frac{m_2}{m_1 + m_2}.
\]

From the equation $1 - k = \frac{m_1}{m_1 + m_2}$, it follows that:
\[
k = 1 - \frac{m_1}{m_1 + m_2} = \frac{m_2}{m_1 + m_2}.
\]

Thus, the value of $k$ is:
\[
k = \frac{m_2}{m_1 + m_2}.
\]

\textbf{Conclusion:} The position vector of the center of mass can be written as:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2},
\]
where $k = \frac{m_2}{m_1 + m_2}$.

\textbf{(b) Verifying the distances of the center of mass from $\mathbf{r_1}$ and $\mathbf{r_2}$:}

The center of mass lies on the straight line connecting $\mathbf{r_1}$ and $\mathbf{r_2}$ because $\mathbf{R}$ is a linear combination of $\mathbf{r_1}$ and $\mathbf{r_2}$. To calculate the distances from $\mathbf{R}$ to $\mathbf{r_1}$ and $\mathbf{r_2}$, use the definitions of $\mathbf{R}$ and $k$.

\textbf{Distance from $\mathbf{r_1}$:}
\[
\text{Distance from } \mathbf{r_1} = |\mathbf{R} - \mathbf{r_1}|.
\]
Substitute $\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2}$:
\[
\mathbf{R} - \mathbf{r_1} = (1 - k)\mathbf{r_1} + k\mathbf{r_2} - \mathbf{r_1} = -k\mathbf{r_1} + k\mathbf{r_2}.
\]
Factor $k$:
\[
\mathbf{R} - \mathbf{r_1} = k(\mathbf{r_2} - \mathbf{r_1}).
\]
The magnitude of the distance is:
\[
\text{Distance from } \mathbf{r_1} = k|\mathbf{r_2} - \mathbf{r_1}|.
\]
The separation between the two masses is $d = |\mathbf{r_2} - \mathbf{r_1}|$, so:
\[
\text{Distance from } \mathbf{r_1} = k d.
\]
Substitute $k = \frac{m_2}{m_1 + m_2}$:
\[
\text{Distance from } \mathbf{r_1} = \frac{m_2}{m_1 + m_2} d.
\]

\textbf{Distance from $\mathbf{r_2}$:}
Similarly, calculate the distance from $\mathbf{r_2}$:
\[
\text{Distance from } \mathbf{r_2} = |\mathbf{R} - \mathbf{r_2}|.
\]
Substitute $\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2}$:
\[
\mathbf{R} - \mathbf{r_2} = (1 - k)\mathbf{r_1} + k\mathbf{r_2} - \mathbf{r_2} = (1 - k)(\mathbf{r_1} - \mathbf{r_2}).
\]
The magnitude of the distance is:
\[
\text{Distance from } \mathbf{r_2} = (1 - k)|\mathbf{r_1} - \mathbf{r_2}|.
\]
Using $d = |\mathbf{r_2} - \mathbf{r_1}|$ and $1 - k = \frac{m_1}{m_1 + m_2}$:
\[
\text{Distance from } \mathbf{r_2} = \frac{m_1}{m_1 + m_2} d.
\]

\textbf{Conclusion:} The distances of the center of mass from the two masses are:
\[
\frac{dm_2}{m_1 + m_2} \text{ from } \mathbf{r_1}, \quad \text{and} \quad \frac{dm_1}{m_1 + m_2} \text{ from } \mathbf{r_2}.
\]"

Solution 3:

"
\textbf{(a) Deriving the center of mass position vector:}

The center of mass $\mathbf{R}$ of a system of two masses $m_1$ and $m_2$ located at positions $\mathbf{r_1}$ and $\mathbf{r_2}$ is defined as:
\[
\mathbf{R} = \frac{m_1 \mathbf{r_1} + m_2 \mathbf{r_2}}{m_1 + m_2}.
\]

\textit{Reasoning:} This is the definition of the center of mass for a discrete system. It is the weighted average of the positions of the masses, where the weights are the masses themselves.

We aim to express $\mathbf{R}$ in the form:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2},
\]
and determine the value of $k$.

First, expand the general expression for $\mathbf{R}$:
\[
\mathbf{R} = \frac{m_1}{m_1 + m_2} \mathbf{r_1} + \frac{m_2}{m_1 + m_2} \mathbf{r_2}.
\]

\textit{Reasoning:} By distributing the terms in the numerator, we separate the contributions of $\mathbf{r_1}$ and $\mathbf{r_2}$ to the center of mass.

Now compare this expression to the target form:
\[
\mathbf{R} = (1 - k)\mathbf{r_1} + k\mathbf{r_2}.
\]

From the comparison, the coefficient of $\mathbf{r_1}$ must satisfy:
\[
1 - k = \frac{m_1}{m_1 + m_2}.
\]

Solve for $k$:
\[
k = 1 - \frac{m_1}{m_1 + m_2}.
\]

Simplify:
\[
k = \frac{m_2}{m_1 + m_2}.
\]

\textit{Reasoning:} The parameter $k$ represents the fractional contribution of $m_2$ to the total mass, which determines the relative weight of $\mathbf{r_2}$ in the center of mass position.

Thus, the center of mass position vector can be written as:
\[
\mathbf{R} = (1 - \frac{m_2}{m_1 + m_2})\mathbf{r_1} + \frac{m_2}{m_1 + m_2}\mathbf{r_2}.
\]

\textit{Reasoning:} This verifies that $\mathbf{R}$ is a weighted average of the positions $\mathbf{r_1}$ and $\mathbf{r_2}$, consistent with the definition of the center of mass.

\bigskip

\textbf{(b) Verifying distances from the two masses:}

The center of mass lies on the line connecting $\mathbf{r_1}$ and $\mathbf{r_2}$. Let the distance between the masses be $d$, such that:
\[
d = |\mathbf{r_2} - \mathbf{r_1}|.
\]

The position vector $\mathbf{R}$ splits this line segment into two parts. The distance of the center of mass from $\mathbf{r_1}$ is given by:
\[
\text{Distance from } \mathbf{r_1} = k d = \frac{m_2}{m_1 + m_2} d.
\]

\textit{Reasoning:} The distance is proportional to the contribution of $m_2$ to the total mass. The center of mass is closer to the larger mass.

Similarly, the distance of the center of mass from $\mathbf{r_2}$ is:
\[
\text{Distance from } \mathbf{r_2} = (1 - k) d = \frac{m_1}{m_1 + m_2} d.
\]

\textit{Reasoning:} The complement of $k$ determines the contribution of $m_1$ to the total mass, so the center of mass is closer to $m_2$ if $m_2 > m_1$.

Thus, the center of mass distances are:
\[
\frac{dm_2}{m_1 + m_2} \text{ from } \mathbf{r_1}, \quad \text{and} \quad \frac{dm_1}{m_1 + m_2} \text{ from } \mathbf{r_2}.
\]

\textit{Reasoning:} These results confirm that the center of mass divides the line segment in inverse proportion to the two masses, consistent with physical intuition."


\subsubsection{CMQ4}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 18 marks. Give up to 6 marks for part a), give up to 5 marks for part b) and give up to 7 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem:

"In 2D polar coordinates, the velocity, \( \underline{v} \), and acceleration, \( \underline{a} \), are given by 

\[ 
\underline{v} = \dot{r} \underline{\hat{r}} + r \dot{\theta} \underline{\hat{\theta}} 
\]
\[ 
\underline{a} = (\ddot{r} - r \dot{\theta}^{2}) \underline{\hat{r}} + (r \ddot{\theta} + 2\dot{r}\dot{\theta} ) \underline{\hat{\theta}} 
\]
    
A particle of mass, \( m \), rotates with angular frequency, \( \dot{\theta} \), dependent on time and has radial velocity given by \( \dot{r} = - \gamma \), where \( \gamma \) is a constant. At time \( t = 0 \), the particle has radial position \( r_{0} \) and \( \dot{\theta} = \delta \). 
    
a) Assuming angular momentum is conserved, derive an expression for the angular frequency \( \dot{\theta} \). Write your answer in terms of \( r_{0} \), \( \gamma \), and \( \delta \).
    
b) What is the angular component of the acceleration?
    
c) Derive an expression for the kinetic energy of the particle and show that 
    
\[
\frac{dK}{dt} = \frac{m r_{0}^{4} \gamma \delta^{2}}{(r_{0} - \gamma t)^{3}}
\]"

Mark Scheme:

"{Q4: (18 marks total)}

a) (6 marks)

[1 mark] 

\[ \dot{r} = \frac{dr}{dt} = - \gamma \]
\[ r = - \gamma t + C \]

where C is a constant.

[1 mark] use initial conditions, \( r = r_0 \), \( t = 0 \) to give \( C = r_0 \), 

\[ r = - \gamma t + r_0 \]

[2 marks] express and simplify angular momentum using given formulas, 

\[ \underline{L} = \underline{r} \times \underline{p} = m \underline{r} \times \underline{p} \]

\[ \underline{L} = m (r_0 - \gamma t) \hat{\underline{r}} \times \left ( - \gamma \hat{\underline{r}} + (r_0 - \gamma t ) \dot{\theta} \hat{\underline{\theta}} \right ) = m (r_0 - \gamma t)^2 \dot{\theta} \hat{\underline{k}} \]

[2 marks] Use initial values, \( \dot{\theta} = \delta \), \(t = 0 \) and conservation of angular momentum, 

\[ \underline{L}  = m {r_0}^2 \delta \hat{\underline{k}} = m (r_0 - \gamma t)^2 \dot{\theta} \hat{\underline{k}} \]

\[ \dot{\theta} = \frac{{r_0}^2 \delta}{(r_0 - \gamma t)^2} \]

b) (5 marks)

[1 mark] angular component of acceleration given by  \( r \ddot{\theta} + 2\dot{r}\dot{\theta} \).

[3 marks] obtain \( \ddot{\theta} \) by differentiation with respect to time and substituting values of \( r \), \( \dot_{r} \) and \( \dot{\theta} \) found in a). 

\[ \ddot{\theta} = \frac{2r_0^2 \gamma \delta}{(r_0 - \gamma t)^3} \]

\[ r \ddot{\theta} + 2\dot{r}\dot{\theta} = (r_0 - \gamma t) \frac{2{r_0}^2 \gamma \delta}{(r_0 - \gamma t)^3} + 2 (- \gamma) \frac{{r_0}^2 \delta}{(r_0 - \gamma t)^2}  \]

[1 mark] simplify to give 0. 

c) (7 marks)

[3 marks] expression for kinetic energy in polar coordinates 
\[ K = \frac{1}{2}m (\underline{v} \cdot \underline{v} ) \]

\[ \underline{v} = \dot{r} \underline{\hat{r}} + r \dot{\theta} \underline{\hat{\theta}} \]

\[ K = \frac{m}{2} \left ( \dot{r} \underline{\hat{r}} + r \dot{\theta} \underline{\hat{\theta}} \right ) \cdot \left ( \dot{r} \underline{\hat{r}} + r \dot{\theta} \underline{\hat{\theta}} \right ) = \frac{m}{2} \left ( \dot{r}^2 + (r \dot{\theta})^2 \right ) \]

[3 marks] substitution of \( r \), \( \dot{r} \) and \( \dot{\theta} \) to give K 

\[ K = \frac{m}{2} \left [ \gamma^2 + \frac{{r_0}^4 \delta^2}{(r_0 - \gamma t)^2} \right ] \]

[1 mark] Differentiate to obtain 

\[ \frac{dK}{dt} = -2 \frac{m}{2}\frac{-\gamma {r_0}^4 \delta^2}{(r_0 - \gamma t)^3} = \frac{m {r_0}^4 \gamma \delta^2}{(r_0 - \gamma t)^3} \]

"

Solution 1: 

"

(a) To derive an expression for the angular frequency \( \dot{\theta} \), we use the conservation of angular momentum. The angular momentum for a particle in polar coordinates is given by:
\[
L = m r^2 \dot{\theta}
\]
where \( m \) is the mass, \( r \) is the radial distance, and \( \dot{\theta} \) is the angular velocity. Since angular momentum is conserved, \( L \) is constant. At \( t = 0 \), the initial radial position is \( r_0 \) and the angular velocity is \( \dot{\theta}_0 = \delta \). Thus:
\[
L = m r_0^2 \delta
\]
At a later time \( t \), the radial position is \( r = r_0 - \gamma t \) (since \( \dot{r} = -\gamma \)). Substituting into the conservation equation, we have:
\[
m r^2 \dot{\theta} = m r_0^2 \delta
\]
Canceling \( m \) and solving for \( \dot{\theta} \), we get:
\[
\dot{\theta} = \frac{r_0^2 \delta}{r^2}
\]
Substituting \( r = r_0 - \gamma t \), we obtain:
\[
\dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}
\]
Thus, the angular frequency is:
\[
\boxed{\dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}}
\]

(b) The angular component of the acceleration is given by:
\[
a_\theta = r \ddot{\theta} + 2 \dot{r} \dot{\theta}
\]
From part (a), \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \). Differentiating this with respect to \( t \), we apply the chain rule:
\[
\ddot{\theta} = \frac{d}{dt} \left( \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \right)
\]
\[
\ddot{\theta} = r_0^2 \delta \cdot (-2) (r_0 - \gamma t)^{-3} (-\gamma)
\]
\[
\ddot{\theta} = \frac{2 \gamma r_0^2 \delta}{(r_0 - \gamma t)^3}
\]
Substitute \( r = r_0 - \gamma t \), \( \dot{r} = -\gamma \), \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \), and \( \ddot{\theta} \) into the expression for \( a_\theta \):
\[
a_\theta = (r_0 - \gamma t) \cdot \frac{2 \gamma r_0^2 \delta}{(r_0 - \gamma t)^3} + 2 (-\gamma) \cdot \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}
\]
Simplify:
\[
a_\theta = \frac{2 \gamma r_0^2 \delta}{(r_0 - \gamma t)^2} - \frac{2 \gamma r_0^2 \delta}{(r_0 - \gamma t)^2}
\]
\[
a_\theta = 0
\]
Thus, the angular component of acceleration is:
\[
\boxed{a_\theta = 0}
\]

(c) The kinetic energy of the particle is:
\[
K = \frac{1}{2} m v^2
\]
where the velocity magnitude in polar coordinates is:
\[
v^2 = \dot{r}^2 + (r \dot{\theta})^2
\]
Substitute \( \dot{r} = -\gamma \) and \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \):
\[
v^2 = (-\gamma)^2 + \left( (r_0 - \gamma t) \cdot \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \right)^2
\]
\[
v^2 = \gamma^2 + \left( \frac{r_0^2 \delta}{r_0 - \gamma t} \right)^2
\]
The kinetic energy becomes:
\[
K = \frac{1}{2} m \left( \gamma^2 + \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right)
\]
Now differentiate \( K \) with respect to \( t \):
\[
\frac{dK}{dt} = \frac{1}{2} m \cdot \frac{d}{dt} \left( \gamma^2 + \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right)
\]
Since \( \gamma \) is constant, \( \frac{d}{dt}(\gamma^2) = 0 \). For the second term, use the chain rule:
\[
\frac{d}{dt} \left( \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right) = r_0^4 \delta^2 \cdot \frac{d}{dt} \left( (r_0 - \gamma t)^{-2} \right)
\]
\[
= r_0^4 \delta^2 \cdot (-2)(r_0 - \gamma t)^{-3}(-\gamma)
\]
\[
= \frac{2 \gamma r_0^4 \delta^2}{(r_0 - \gamma t)^3}
\]
Thus:
\[
\frac{dK}{dt} = \frac{1}{2} m \cdot \frac{2 \gamma r_0^4 \delta^2}{(r_0 - \gamma t)^3}
\]
\[
\frac{dK}{dt} = \frac{m r_0^4 \gamma \delta^2}{(r_0 - \gamma t)^3}
\]
This matches the given result:
\[
\boxed{\frac{dK}{dt} = \frac{m r_0^4 \gamma \delta^2}{(r_0 - \gamma t)^3}}
\]

"

Solution 2: 

"
\textbf{(a) Expression for Angular Frequency \( \dot{\theta} \):}

The angular momentum \( L \) of the particle is conserved. Angular momentum in polar coordinates is given by:
\[
L = m r^2 \dot{\theta}.
\]
Since \( L \) is conserved, we can write:
\[
L = m r_0^2 \delta,
\]
where \( r_0 \) is the radial position at \( t = 0 \) and \( \dot{\theta} = \delta \). At a later time \( t \), the radial position is given by:
\[
r = r_0 - \gamma t.
\]
Substituting \( r \) into the angular momentum equation, we get:
\[
m r^2 \dot{\theta} = m r_0^2 \delta.
\]
Simplifying for \( \dot{\theta} \):
\[
\dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}.
\]

\textbf{(b) Angular Component of Acceleration:}

The angular component of acceleration is given by:
\[
a_\theta = r \ddot{\theta} + 2 \dot{r} \dot{\theta}.
\]
From part (a), we know \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \). Differentiating \( \dot{\theta} \) with respect to time:
\[
\ddot{\theta} = \frac{d}{dt} \left( \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \right).
\]
Using the chain rule:
\[
\ddot{\theta} = \frac{-2 r_0^2 \delta (-\gamma)}{(r_0 - \gamma t)^3} = \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^3}.
\]
Now substitute into \( a_\theta \):
\[
a_\theta = r \ddot{\theta} + 2 \dot{r} \dot{\theta},
\]
where \( r = r_0 - \gamma t \) and \( \dot{r} = -\gamma \). Substituting values:
\[
a_\theta = (r_0 - \gamma t) \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^3} + 2(-\gamma) \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}.
\]
Simplify:
\[
a_\theta = \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^2} - \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^2} = 0.
\]

\textbf{(c) Kinetic Energy and Its Time Derivative:}

The kinetic energy \( K \) of the particle is:
\[
K = \frac{1}{2} m \left( \dot{r}^2 + r^2 \dot{\theta}^2 \right).
\]
Substituting \( \dot{r} = -\gamma \), \( r = r_0 - \gamma t \), and \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \):
\[
K = \frac{1}{2} m \left[ (-\gamma)^2 + (r_0 - \gamma t)^2 \left( \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \right)^2 \right].
\]
Simplify:
\[
K = \frac{1}{2} m \left[ \gamma^2 + \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right].
\]
To find \( \frac{dK}{dt} \), differentiate:
\[
\frac{dK}{dt} = \frac{1}{2} m \left[ 0 + \frac{d}{dt} \left( \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right) \right].
\]
Using the chain rule:
\[
\frac{d}{dt} \left( \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right) = -2 \frac{r_0^4 \delta^2 (-\gamma)}{(r_0 - \gamma t)^3}.
\]
Simplify:
\[
\frac{dK}{dt} = \frac{m r_0^4 \gamma \delta^2}{(r_0 - \gamma t)^3}.
\]

"

Solution 3:

"
\textbf{(a) Expression for Angular Frequency \( \dot{\theta} \):}

The angular momentum \( L \) of the particle is conserved. In polar coordinates, the angular momentum is given by:
\[
L = m r^2 \dot{\theta}.
\]
Since \( L \) is conserved, we equate the initial angular momentum to the angular momentum at any later time:
\[
m r_0^2 \delta = m r^2 \dot{\theta},
\]
where \( r_0 \) is the radial position at \( t = 0 \) and \( \delta = \dot{\theta} \) at \( t = 0 \). The radial position at any time is given by:
\[
r = r_0 - \gamma t,
\]
where \( \gamma \) is the constant radial velocity. Substituting this into the angular momentum equation:
\[
r_0^2 \delta = (r_0 - \gamma t)^2 \dot{\theta}.
\]
Solving for \( \dot{\theta} \):
\[
\dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}.
\]

\textbf{(b) Angular Component of Acceleration:}

The angular component of acceleration in polar coordinates is:
\[
a_\theta = r \ddot{\theta} + 2 \dot{r} \dot{\theta}.
\]
From part (a), \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \). Differentiating this with respect to \( t \):
\[
\ddot{\theta} = \frac{d}{dt} \left( \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \right).
\]
Using the chain rule:
\[
\ddot{\theta} = \frac{-2 r_0^2 \delta (-\gamma)}{(r_0 - \gamma t)^3} = \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^3}.
\]
Substituting \( \ddot{\theta} \) and other known quantities into \( a_\theta \):
\[
a_\theta = (r_0 - \gamma t) \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^3} + 2(-\gamma) \frac{r_0^2 \delta}{(r_0 - \gamma t)^2}.
\]
Simplify the terms:
\[
a_\theta = \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^2} - \frac{2 r_0^2 \delta \gamma}{(r_0 - \gamma t)^2} = 0.
\]
Thus, the angular component of acceleration is \( a_\theta = 0 \).

\textbf{(c) Kinetic Energy and Its Time Derivative:}

The kinetic energy \( K \) of the particle is:
\[
K = \frac{1}{2} m \left( \dot{r}^2 + r^2 \dot{\theta}^2 \right).
\]
Substituting \( \dot{r} = -\gamma \), \( r = r_0 - \gamma t \), and \( \dot{\theta} = \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \):
\[
K = \frac{1}{2} m \left[ (-\gamma)^2 + (r_0 - \gamma t)^2 \left( \frac{r_0^2 \delta}{(r_0 - \gamma t)^2} \right)^2 \right].
\]
Simplify each term:
\[
K = \frac{1}{2} m \left[ \gamma^2 + \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right].
\]
To find \( \frac{dK}{dt} \), differentiate \( K \) with respect to \( t \):
\[
\frac{dK}{dt} = \frac{1}{2} m \frac{d}{dt} \left[ \gamma^2 + \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right].
\]
Since \( \gamma^2 \) is constant, its derivative is zero. Now differentiate the second term:
\[
\frac{d}{dt} \left( \frac{r_0^4 \delta^2}{(r_0 - \gamma t)^2} \right) = -2 \frac{r_0^4 \delta^2 (-\gamma)}{(r_0 - \gamma t)^3}.
\]
Simplify:
\[
\frac{dK}{dt} = \frac{m r_0^4 \gamma \delta^2}{(r_0 - \gamma t)^3}.
\]
"

\subsubsection{CMQ5}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 9 marks. Give up to 2 marks for part a), give up to 4 marks for part b) and give up to 3 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem: 

"Two objects X and Y of the same mass, \( m \), are connected by a spring and each object is connected to a fixed wall by a spring (as seen in the figure). All springs have the same spring constant, \( k \). Initially object X is at rest whilst object Y moves with an initial velocity \( v \hat{\underline{i}} \). The horizontal displacement from equilibrium of X and Y is defined by \( x_{1} \) and \( x_{2} \) respectively. 

a) Assuming no other forces besides those caused by the springs, derive the equations of motion for object X and object Y. 
    
b) By defining new coordinates \( y_{1} = x_{1} + x_{2} \), \( y_{2} = x_{1} - x_{2} \), show that this uncouples the coordinates and \( y_{1} \) and \( y_{2} \) obey equations of undamped, undriven simple harmonic motion. 
    
c) With the initial conditions of the system, show the solution of \( y_{1} \) is given by 
    
\[
y_{1} = v \sqrt{\frac{m}{k}} \sin(t \sqrt{\frac{k}{m}})
\]"

Mark Scheme:

"{Q5: (9 marks total)}

a) (2 marks)

[2 marks] Using Newton’s 2nd Law, Hooke’s Law, we have the 2 equations of motion: 

\[ m \ddot{x}_1 = - k x_1 - k(x_1 - x_2) = -2k x_1 + k x_2 \]
\[ m \ddot{x}_2 = - k x_2 - k(x_2 - x_1 = -2k x_2 + k x_1 \]

b) (4 marks) 

[2 marks] Adding and subtracting the 2 equations of motion from a), 

\[ m( \ddot{x}_1 + \ddot{x}_2 ) = -k(x_1 + x_2) \]

\[ m( \ddot{x}_1 - \ddot{x}_2 ) = -3k(x_1 - x_2) \]

[2 marks] Using \( y_{1} = x_{1} + x_{2} \), \( y_{2} = x_{1} - x_{2} \)

\[ m \ddot{y}_1 = -k y_1 \]
\[ m \ddot{y}_2 = -3k y_2 \]

Mention both equations represent simple harmonic motion.

c) (3 marks)

 [1 mark] General solution for \( y_1 \) is 

\[ y_1 = A \sin(\omega_1 t) + B \cos(\omega_1 t) \]

where \(\omega_1 = \sqrt{\frac{k}{m}}\)

[2 marks] Use of initial conditions. \( y_1(0) = x_1(0) + x_2(0) = 0 \), giving B = 0. \( \dot{y}_1(0) = \dot{x}_1(0) + \dot{x}_2(0) = v \), giving \( A = \frac{v}{\omega} = v\sqrt{\frac{m}{k}} \). Then 

\[ y_1 = v\sqrt{\frac{m}{k}} \sin \left ( t \sqrt{\frac{k}{m}} \right ) \]

"

Solution 1: 

"
(a) Derive the equations of motion for $X$ and $Y$ (2 marks)

The system consists of two masses $X$ and $Y$ connected by springs. Each spring has a spring constant $k$, and the masses are connected to fixed walls on either end. The displacements of $X$ and $Y$ from their equilibrium positions are denoted by $x_1$ and $x_2$, respectively.

**Force on $X$:**

The object $X$ is acted on by two springs: one connected to the left wall and the other connecting $X$ to $Y$.
\[
F_X = -k x_1 + k (x_2 - x_1)
\]
The first term $-k x_1$ is the restoring force due to the spring connected to the wall. The second term $k (x_2 - x_1)$ is the force exerted by the spring connecting $X$ and $Y$.

Simplifying:
\[
F_X = -2k x_1 + k x_2
\]

Using Newton's second law, $F_X = m \ddot{x}_1$, we have:
\[
m \ddot{x}_1 = -2k x_1 + k x_2
\]
\[
\ddot{x}_1 = -\frac{2k}{m} x_1 + \frac{k}{m} x_2
\]

**Force on $Y$:**

The object $Y$ is also acted on by two springs: one connecting $Y$ to $X$ and the other to the right wall.
\[
F_Y = -k x_2 - k (x_2 - x_1)
\]
The first term $-k x_2$ is the restoring force due to the spring connected to the wall. The second term $-k (x_2 - x_1)$ is the force exerted by the spring connecting $Y$ and $X$.

Simplifying:
\[
F_Y = -2k x_2 + k x_1
\]

Using Newton's second law, $F_Y = m \ddot{x}_2$, we have:
\[
m \ddot{x}_2 = -2k x_2 + k x_1
\]
\[
\ddot{x}_2 = -\frac{2k}{m} x_2 + \frac{k}{m} x_1
\]

Thus, the equations of motion for $X$ and $Y$ are:
\[
\ddot{x}_1 = -\frac{2k}{m} x_1 + \frac{k}{m} x_2
\]
\[
\ddot{x}_2 = -\frac{2k}{m} x_2 + \frac{k}{m} x_1
\]

(b) Uncouple the coordinates using $y_1$ and $y_2$ (4 marks)

We define new coordinates:
\[
y_1 = x_1 + x_2, \quad y_2 = x_1 - x_2
\]

**Step 1: Express the original variables in terms of $y_1$ and $y_2$.**
\[
x_1 = \frac{y_1 + y_2}{2}, \quad x_2 = \frac{y_1 - y_2}{2}
\]

**Step 2: Differentiate $y_1$ and $y_2$ twice.**
\[
\ddot{y}_1 = \ddot{x}_1 + \ddot{x}_2, \quad \ddot{y}_2 = \ddot{x}_1 - \ddot{x}_2
\]

Substituting the equations of motion for $\ddot{x}_1$ and $\ddot{x}_2$:
\[
\ddot{y}_1 = \left(-\frac{2k}{m} x_1 + \frac{k}{m} x_2\right) + \left(-\frac{2k}{m} x_2 + \frac{k}{m} x_1\right)
\]
\[
\ddot{y}_1 = -\frac{k}{m}(x_1 + x_2)
\]
Substitute $x_1 + x_2 = y_1$:
\[
\ddot{y}_1 = -\frac{k}{m} y_1
\]

Similarly, for $\ddot{y}_2$:
\[
\ddot{y}_2 = \left(-\frac{2k}{m} x_1 + \frac{k}{m} x_2\right) - \left(-\frac{2k}{m} x_2 + \frac{k}{m} x_1\right)
\]
\[
\ddot{y}_2 = -\frac{3k}{m}(x_1 - x_2)
\]
Substitute $x_1 - x_2 = y_2$:
\[
\ddot{y}_2 = -\frac{3k}{m} y_2
\]

Thus, the equations of motion in the new coordinates are:
\[
\ddot{y}_1 = -\frac{k}{m} y_1, \quad \ddot{y}_2 = -\frac{3k}{m} y_2
\]

These describe independent undamped, undriven simple harmonic motions.

(c) Solve for $y_1$ with the initial conditions (3 marks)

The equation of motion for $y_1$ is:
\[
\ddot{y}_1 + \frac{k}{m} y_1 = 0
\]

This is a simple harmonic oscillator with angular frequency:
\[
\omega_1 = \sqrt{\frac{k}{m}}
\]

The general solution is:
\[
y_1(t) = A \cos(\omega_1 t) + B \sin(\omega_1 t)
\]

From the initial conditions:
- Initially, $x_1 = 0$ and $x_2 = 0$, so $y_1(0) = x_1 + x_2 = 0 \implies A = 0$.
- Initially, $x_1$ is at rest, and $x_2$ moves with velocity $v_0$, so $\dot{y}_1(0) = v_0 \implies B \omega_1 = v_0 \implies B = \frac{v_0}{\omega_1} = v_0 \sqrt{\frac{m}{k}}$.

Thus, the solution for $y_1$ is:
\[
y_1(t) = v_0 \sqrt{\frac{m}{k}} \sin\left(t \sqrt{\frac{k}{m}}\right)
\]"

Solution 2: 

"
\textbf{(a)} The forces acting on the objects $X$ and $Y$ are due to the springs. Let $x_1$ and $x_2$ denote the displacements of $X$ and $Y$ from their equilibrium positions.

For object $X$, the force is given by:
\[
F_X = -k x_1 + k (x_2 - x_1),
\]
where $-k x_1$ is the restoring force due to the spring connecting $X$ to the wall, and $k (x_2 - x_1)$ is the force due to the spring between $X$ and $Y$. Using Newton's second law, $F_X = m \ddot{x}_1$, we have:
\[
m \ddot{x}_1 = -k x_1 + k (x_2 - x_1).
\]
Simplifying:
\[
m \ddot{x}_1 = -2k x_1 + k x_2. \tag{1}
\]

For object $Y$, the force is:
\[
F_Y = -k (x_2 - x_1) - k x_2,
\]
where $-k (x_2 - x_1)$ is the force due to the spring connecting $X$ and $Y$, and $-k x_2$ is the restoring force due to the spring connecting $Y$ to the wall. Using Newton's second law, $F_Y = m \ddot{x}_2$, we have:
\[
m \ddot{x}_2 = -k (x_2 - x_1) - k x_2.
\]
Simplifying:
\[
m \ddot{x}_2 = k x_1 - 2k x_2. \tag{2}
\]

Thus, the equations of motion for objects $X$ and $Y$ are:
\[
\boxed{m \ddot{x}_1 = -2k x_1 + k x_2, \quad m \ddot{x}_2 = k x_1 - 2k x_2.}
\]

\textbf{(b)} To decouple the equations, define new coordinates:
\[
y_1 = x_1 + x_2, \quad y_2 = x_1 - x_2.
\]

First, calculate the second derivatives of $y_1$ and $y_2$:
\[
\ddot{y}_1 = \ddot{x}_1 + \ddot{x}_2, \quad \ddot{y}_2 = \ddot{x}_1 - \ddot{x}_2.
\]

Add equations (1) and (2):
\[
m \ddot{x}_1 + m \ddot{x}_2 = -2k x_1 + k x_2 + k x_1 - 2k x_2,
\]
which simplifies to:
\[
m (\ddot{x}_1 + \ddot{x}_2) = -k (x_1 + x_2).
\]
Substitute $y_1 = x_1 + x_2$:
\[
m \ddot{y}_1 = -k y_1 \quad \implies \quad \ddot{y}_1 + \frac{k}{m} y_1 = 0. \tag{3}
\]

Now subtract equation (2) from equation (1):
\[
m \ddot{x}_1 - m \ddot{x}_2 = -2k x_1 + k x_2 - (k x_1 - 2k x_2),
\]
which simplifies to:
\[
m (\ddot{x}_1 - \ddot{x}_2) = -3k x_1 + 3k x_2.
\]
Substitute $y_2 = x_1 - x_2$:
\[
m \ddot{y}_2 = -3k y_2 \quad \implies \quad \ddot{y}_2 + \frac{3k}{m} y_2 = 0. \tag{4}
\]

Thus, the equations of motion in the $y_1$ and $y_2$ coordinates are:
\[
\boxed{\ddot{y}_1 + \frac{k}{m} y_1 = 0, \quad \ddot{y}_2 + \frac{3k}{m} y_2 = 0.}
\]
These are uncoupled equations of undamped, undriven simple harmonic motion.

\textbf{(c)} From part (b), the solution for $y_1$ is:
\[
y_1(t) = A \cos\left(t \sqrt{\frac{k}{m}}\right) + B \sin\left(t \sqrt{\frac{k}{m}}\right).
\]

Initially, $x_1 = 0$ and $x_2 = 0$, so $y_1(0) = x_1 + x_2 = 0$, which gives:
\[
A = 0.
\]

The initial velocity of $Y$ is $v_0$, and $X$ is at rest. Therefore:
\[
\dot{y}_1(0) = \dot{x}_1(0) + \dot{x}_2(0) = 0 + v_0 = v_0.
\]
Substitute this into the derivative of $y_1$:
\[
\dot{y}_1(t) = B \sqrt{\frac{k}{m}} \cos\left(t \sqrt{\frac{k}{m}}\right).
\]
At $t = 0$, $\dot{y}_1(0) = v_0$, so:
\[
B \sqrt{\frac{k}{m}} = v_0 \quad \implies \quad B = v_0 \sqrt{\frac{m}{k}}.
\]

Thus, the solution for $y_1$ is:
\[
\boxed{y_1(t) = v_0 \sqrt{\frac{m}{k}} \sin\left(t \sqrt{\frac{k}{m}}\right).}
\]

"

Solution 3:

"
\textbf{(a)} To derive the equations of motion for objects $X$ and $Y$, we start by analyzing the forces acting on each object. The spring connecting $X$ to the wall exerts a restoring force $-kx_1$ on $X$, and the spring between $X$ and $Y$ exerts a force proportional to the relative displacement of the two masses. 

For object $X$, the net force is given by:
\[
F_X = -kx_1 + k(x_2 - x_1),
\]
where $-kx_1$ is the restoring force due to the wall spring, and $k(x_2 - x_1)$ is the force due to the spring connecting $X$ and $Y$. Using Newton's second law, $F = ma$, we have:
\[
m\ddot{x}_1 = -kx_1 + k(x_2 - x_1).
\]
Simplifying:
\[
m\ddot{x}_1 = -2kx_1 + kx_2. \tag{1}
\]

For object $Y$, the forces acting are the restoring force from the wall spring $-kx_2$ and the force due to the spring connecting $X$ and $Y$, which is $-k(x_2 - x_1)$. Thus:
\[
F_Y = -kx_2 - k(x_2 - x_1).
\]
Using Newton's second law:
\[
m\ddot{x}_2 = -kx_2 - k(x_2 - x_1).
\]
Simplifying:
\[
m\ddot{x}_2 = -2kx_2 + kx_1. \tag{2}
\]

Equations (1) and (2) are the coupled equations of motion for the system.

\vspace{0.5cm}

\textbf{(b)} To uncouple the equations, we define new coordinates:
\[
y_1 = x_1 + x_2, \quad y_2 = x_1 - x_2.
\]

First, calculate the second derivatives:
\[
\ddot{y}_1 = \ddot{x}_1 + \ddot{x}_2, \quad \ddot{y}_2 = \ddot{x}_1 - \ddot{x}_2.
\]

Adding equations (1) and (2):
\[
m\ddot{x}_1 + m\ddot{x}_2 = -2kx_1 + kx_2 - 2kx_2 + kx_1,
\]
\[
m(\ddot{x}_1 + \ddot{x}_2) = -2k(x_1 + x_2).
\]
Using $y_1 = x_1 + x_2$, we obtain:
\[
m\ddot{y}_1 = -2ky_1, \quad \text{or} \quad \ddot{y}_1 + \frac{2k}{m}y_1 = 0. \tag{3}
\]

Now, subtract equation (2) from (1):
\[
m\ddot{x}_1 - m\ddot{x}_2 = -2kx_1 + kx_2 - (-2kx_2 + kx_1),
\]
\[
m(\ddot{x}_1 - \ddot{x}_2) = -2k(x_1 - x_2).
\]
Using $y_2 = x_1 - x_2$, we obtain:
\[
m\ddot{y}_2 = -2ky_2, \quad \text{or} \quad \ddot{y}_2 + \frac{2k}{m}y_2 = 0. \tag{4}
\]

Equations (3) and (4) describe independent simple harmonic motion for $y_1$ and $y_2$ with angular frequency $\omega = \sqrt{\frac{2k}{m}}$.

\vspace{0.5cm}

\textbf{(c)} To solve for $y_1$, we use the initial conditions. Initially, $X$ is at rest ($\dot{x}_1(0) = 0$) and $Y$ moves with velocity $v_0$ ($\dot{x}_2(0) = v_0$). Thus:
\[
\dot{y}_1(0) = \dot{x}_1(0) + \dot{x}_2(0) = 0 + v_0 = v_0.
\]
Also, $y_1(0) = x_1(0) + x_2(0) = 0 + 0 = 0$.

The general solution for $y_1$ is:
\[
y_1(t) = A\sin\left(\omega t\right) + B\cos\left(\omega t\right).
\]
Using $y_1(0) = 0$:
\[
0 = A\sin(0) + B\cos(0) \implies B = 0.
\]
Using $\dot{y}_1(0) = v_0$:
\[
\dot{y}_1(t) = A\omega\cos(\omega t),
\]
\[
v_0 = A\omega \implies A = \frac{v_0}{\omega}.
\]

Thus:
\[
y_1(t) = \frac{v_0}{\omega}\sin(\omega t).
\]
Substitute $\omega = \sqrt{\frac{2k}{m}}$:
\[
y_1(t) = v_0\sqrt{\frac{m}{2k}}\sin\left(t\sqrt{\frac{2k}{m}}\right).
\]

This matches the required solution for $y_1$."

\subsubsection{CMQ6}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 16 marks. Give up to 2 marks for part a), give up to 3 marks for part b) and give up to 11 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem: 

"Consider a Go-kart which moves around a flat circular track at a radius \( R \) with the track having a coefficient of static friction, \( \mu_{s} \). 

a) Derive an expression for the largest speed the Go-kart can have whilst staying on the same circular path of radius \( R \). 
    
b) Assume the track now forms an angle of \( \phi \) with respect to the horizontal flat ground and that the track is now frictionless. Rederive the largest speed the Go-kart may move with that maintains its circular motion around the track at the same radius \( R \). 
    
c) Consider the same situation as in part b, except the static coefficient is now again \( \mu_{s} \). Derive the minimum speed, \( v_{min} \), and maximum speed, \( v_{max} \), showing that the following equality holds:
    
\[ 
v_{max}^{2} - v_{min}^{2} = \frac{2Rg\mu_{s}}{\cos^{2}(\phi) - \mu_{s}^{2}\sin^{2}(\phi)}
\]"

Mark Scheme:

"{Q6: (16 marks total)}

a) (2 marks)

[1 mark] identification of frictional force \( F_f \) and centripetal acceleration \( F \)

\[ F_f = \mu_s N = \mu_s mg \]
\[ F = \frac{m v^2}{R} \]

[1 mark] frictional force provides centripetal acceleration, \(F_f = F \):

\[ \mu_s mg = \frac{m v^2}{R} \]
\[ v = \sqrt{\mu_s g R} \]
b) [3 marks]

[1 mark] Normal force provides centripetal force,

\[ N \sin(\phi) = \frac{m v^2}{R} \]

[1 mark] Normal force also balances weight =, 

\[ N \cos(\phi) = mg \]

[1 mark] substitute 2nd equation into first to give 

\[ \frac{m v^2}{R} = \frac{mg}{\cos(\phi)} \sin(\phi) \]
\[ v = \sqrt{R g \tan(\phi)} \]

c) (11 marks)

[2 marks] For largest \( v \), frictional force \( F_f \) points inwards. Balancing forces, 

\[ \frac{m v^2}{R} = F_f \cos(\phi) + N \sin(\phi) \]

[2 marks] Balance vertical forces, with \( F_f = \mu_s N \),

\[ mg + \mu_s N \sin(\phi) = N \cos(\phi) \]

\[ N = \frac{mg}{\cos(\phi) - \mu_s \sin(\phi)} \]

[3 marks] substitute expressions into centripetal equation, simplify to obtain \( v_{max} \)

\[ \frac{m v^2}{R} = \mu_s N \cos(\phi) + N \sin(\phi) = N (\mu_s \cos(\phi) + \sin(\phi) ) = mg \left ( \frac{\mu_s \cos(\phi) + \sin(\phi)}{ \cos(\phi) - \mu_s \sin(\phi)} \right ) \]

\[ v_{max} = \sqrt{gR \left ( \frac{\sin(\phi) + \mu_s \cos(\phi)}{\cos(\phi) - \mu_s \sin(\phi)} \right )}  \]

[2 marks] For \( v_{min} \), \(F_f\) is reversed. Terms with \( \mu_s \) switch sign, 

\[ v_{min} = \sqrt{gR \left ( \frac{\sin(\phi) - \mu_s \cos(\phi)}{\cos(\phi) + \mu_s \sin(\phi)} \right )} \]

[2 marks] Calculation of \( v_{max}^{2} - v_{min}^{2} \), simplifying algebra:

\[ v_{max}^{2} - v_{min}^{2} = gR \left ( \frac{\sin(\phi) + \mu_s \cos(\phi)}{\cos(\phi) - \mu_s \sin(\phi)} \right ) - gR \left ( \frac{\sin(\phi) - \mu_s \cos(\phi)}{\cos(\phi) + \mu_s \sin(\phi)} \right ) = \frac{2Rg\mu_{s}}{\cos^{2}(\phi) - \mu_{s}^{2}\sin^{2}(\phi)} \]

"

Solution 1: 

"
\textbf{(a) [2 marks]}  
To find the largest speed the Go-kart can maintain while staying on the same circular path, we analyze the forces acting on it. The Go-kart is moving in a circular path of radius $R$, so the centripetal force is required to keep it on the curve. This force is provided by the static friction, which must not exceed its maximum value:
\[
F_{\text{friction}} = \mu_s N,
\]
where $N$ is the normal force. For a flat surface, $N = mg$, so the maximum frictional force becomes:
\[
F_{\text{friction, max}} = \mu_s mg.
\]
The centripetal force is given by:
\[
F_{\text{centripetal}} = \frac{mv^2}{R}.
\]
Equating the two forces to find the largest possible speed:
\[
\mu_s mg = \frac{mv^2}{R}.
\]
Simplifying for $v$, we get:
\[
v_{\text{max}} = \sqrt{\mu_s g R}.
\]

\textbf{(b) [3 marks]}  
Now assume the track is frictionless and inclined at an angle $\phi$ to the horizontal. The centripetal force is provided by the component of the normal force, $N$, acting toward the center of the circular path. Analyzing the forces:

1. The normal force is perpendicular to the inclined surface.
2. The weight, $mg$, has components:
   - $mg \cos\phi$ perpendicular to the incline.
   - $mg \sin\phi$ parallel to the incline.

On a frictionless incline, $N = mg \cos\phi$. The centripetal force is provided by the horizontal component of $N$, which is $N \sin\phi = mg \cos\phi \sin\phi$. Thus:
\[
F_{\text{centripetal}} = mg \cos\phi \sin\phi.
\]
Equating this to the centripetal force expression:
\[
mg \cos\phi \sin\phi = \frac{mv^2}{R}.
\]
Solving for $v$, we find the maximum speed:
\[
v_{\text{max}} = \sqrt{g R \cos\phi \sin\phi}.
\]

\textbf{(c) [11 marks]}  
Now consider the same situation as in part (b), but with the coefficient of static friction $\mu_s$ reintroduced. Both friction and the horizontal component of the normal force contribute to the centripetal force. Let us analyze the forces in more detail.

1. The normal force is still $N = mg \cos\phi$.
2. The frictional force has a maximum value of:
   \[
   F_{\text{friction, max}} = \mu_s N = \mu_s mg \cos\phi.
   \]
3. The centripetal force $F_{\text{centripetal}}$ is provided by two components:
   - The horizontal component of the normal force, $N \sin\phi = mg \cos\phi \sin\phi$.
   - The frictional force, which acts up or down the incline, depending on whether $v$ is large or small.

The total centripetal force is:
\[
F_{\text{centripetal}} = mg \cos\phi \sin\phi \pm \mu_s mg \cos\phi,
\]
where the $+$ sign corresponds to the maximum speed and the $-$ sign corresponds to the minimum speed. Setting this equal to $\frac{mv^2}{R}$:
\[
\frac{mv^2}{R} = mg \cos\phi \sin\phi \pm \mu_s mg \cos\phi.
\]
Canceling $m$ and rearranging for $v^2$, we get:
\[
v^2 = Rg (\cos\phi \sin\phi \pm \mu_s \cos\phi).
\]
The maximum and minimum speeds are thus:
\[
v_{\text{max}}^2 = Rg (\cos\phi \sin\phi + \mu_s \cos\phi),
\]
\[
v_{\text{min}}^2 = Rg (\cos\phi \sin\phi - \mu_s \cos\phi).
\]
Subtracting the two equations:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = 2Rg \mu_s \cos\phi.
\]

To verify the provided equality:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg \mu_s}{\cos^2\phi - \mu_s^2 \sin^2\phi}.
\]
Using trigonometric identities and substituting, we find that the provided formula holds true after simplification, completing the derivation.

\textbf{Final Answers:}
\[
\text{(a)} \quad v_{\text{max}} = \sqrt{\mu_s g R}.
\]
\[
\text{(b)} \quad v_{\text{max}} = \sqrt{g R \cos\phi \sin\phi}.
\]
\[
\text{(c)} \quad v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg \mu_s}{\cos^2\phi - \mu_s^2 \sin^2\phi}.
\]"

Solution 2: 

"

\textbf{(a) Derive an expression for the largest speed on a flat circular track:} \\[1em]
The Go-kart moves in a circle of radius $R$ on a flat surface. The forces acting on the Go-kart are:
1. The gravitational force, $mg$, acting vertically downward.
2. The normal force, $N$, acting vertically upward.
3. The static frictional force, $f_s$, acting horizontally to provide the centripetal force required for circular motion.

For the largest speed, the static frictional force is at its maximum, $f_s = \mu_s N$. On a flat surface, the normal force balances the gravitational force, so $N = mg$.

The centripetal force is given by $F_c = \frac{mv^2}{R}$. Setting $F_c = f_s$, we have:
\[
\frac{mv^2}{R} = \mu_s mg.
\]

Canceling $m$ from both sides and solving for $v$, we get:
\[
v_{\text{max}} = \sqrt{\mu_s g R}.
\]

Thus, the largest speed the Go-kart can have is:
\[
\boxed{v_{\text{max}} = \sqrt{\mu_s g R}}.
\]

\textbf{(b) Derive the largest speed on a frictionless banked track:} \\[1em]
On a banked track with angle $\phi$, the forces acting are:
1. The gravitational force, $mg$, acting vertically downward.
2. The normal force, $N$, acting perpendicular to the track.
3. No frictional force, as the surface is frictionless.

The normal force can be decomposed into two components:
- $N\sin\phi$, acting horizontally to provide the centripetal force.
- $N\cos\phi$, balancing the weight of the Go-kart.

From vertical force balance, $N\cos\phi = mg$. Thus, $N = \frac{mg}{\cos\phi}$.

The horizontal component provides the centripetal force:
\[
N\sin\phi = \frac{mv^2}{R}.
\]

Substituting $N = \frac{mg}{\cos\phi}$ into this equation, we get:
\[
\frac{mg}{\cos\phi} \sin\phi = \frac{mv^2}{R}.
\]

Canceling $m$ and simplifying:
\[
v^2 = Rg \frac{\sin\phi}{\cos\phi} = Rg \tan\phi.
\]

Taking the square root:
\[
v_{\text{max}} = \sqrt{Rg \tan\phi}.
\]

Thus, the largest speed is:
\[
\boxed{v_{\text{max}} = \sqrt{Rg \tan\phi}}.
\]

\textbf{(c) Derive the minimum and maximum speeds on a banked track with friction:} \\[1em]
When the track has a coefficient of static friction $\mu_s$, the forces acting are:
1. The gravitational force, $mg$, acting vertically downward.
2. The normal force, $N$, acting perpendicular to the surface.
3. The static frictional force, $f_s$, which can act up or down the slope depending on whether the Go-kart is at minimum or maximum speed.

The static friction force is given by $f_s \leq \mu_s N$, and its direction determines whether it assists or resists the centripetal force.

Let the centripetal force be $\frac{mv^2}{R}$. Resolving forces perpendicular to the incline, we have:
\[
N\cos\phi = mg - f_s\sin\phi.
\]

Resolving forces along the incline (providing centripetal force), we have:
\[
N\sin\phi + f_s\cos\phi = \frac{mv^2}{R}.
\]

From the first equation, substituting $f_s = \mu_s N$:
\[
N = \frac{mg}{\cos\phi - \mu_s \sin\phi}.
\]

Substituting $N$ into the second equation:
\[
\frac{mg}{\cos\phi - \mu_s \sin\phi} \sin\phi + \mu_s \frac{mg}{\cos\phi - \mu_s \sin\phi} \cos\phi = \frac{mv^2}{R}.
\]

Simplifying:
\[
\frac{mg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi} = \frac{mv^2}{R}.
\]

Canceling $m$ and solving for $v^2$, we get:
\[
v^2 = \frac{Rg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi}.
\]

For the minimum speed, the friction force acts down the slope, so replace $\mu_s$ with $-\mu_s$:
\[
v_{\text{min}}^2 = \frac{Rg (\sin\phi - \mu_s \cos\phi)}{\cos\phi + \mu_s \sin\phi}.
\]

Thus, the maximum and minimum speeds are:
\[
v_{\text{max}}^2 = \frac{Rg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi},
\]
\[
v_{\text{min}}^2 = \frac{Rg (\sin\phi - \mu_s \cos\phi)}{\cos\phi + \mu_s \sin\phi}.
\]

To show the given equality:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{Rg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi} - \frac{Rg (\sin\phi - \mu_s \cos\phi)}{\cos\phi + \mu_s \sin\phi}.
\]

Combining the fractions:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{Rg \big[(\sin\phi + \mu_s \cos\phi)(\cos\phi + \mu_s \sin\phi) - (\sin\phi - \mu_s \cos\phi)(\cos\phi - \mu_s \sin\phi)\big]}{(\cos\phi - \mu_s \sin\phi)(\cos\phi + \mu_s \sin\phi)}.
\]

Simplify the numerator:
\[
(\sin\phi + \mu_s \cos\phi)(\cos\phi + \mu_s \sin\phi) - (\sin\phi - \mu_s \cos\phi)(\cos\phi - \mu_s \sin\phi) = 2Rg\mu_s.
\]

Thus:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg\mu_s}{\cos^2\phi - \mu_s^2 \sin^2\phi}.
\]

This matches the given result. Hence, the equality holds:
\[
\boxed{v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg\mu_s}{\cos^2\phi - \mu_s^2 \sin^2\phi}}.
\]

"

Solution 3:

"

\textbf{(a) Derive an expression for the largest speed the Go-kart can have whilst staying on the same circular path of radius $R$.} \\

For a Go-kart moving on a flat circular track, the maximum speed is determined by the static frictional force, which provides the necessary centripetal force for circular motion. The forces acting on the Go-kart are as follows:
- The gravitational force, $mg$, acts vertically downward.
- The normal force, $N$, acts vertically upward and balances the gravitational force: $N = mg$.
- The static frictional force, $F_s = \mu_s N = \mu_s mg$, acts radially inward to provide the centripetal force.

The centripetal force required for circular motion is:
\[
F_c = \frac{mv^2}{R}.
\]

At maximum speed, the static frictional force equals the centripetal force:
\[
F_s = F_c.
\]

Substituting $F_s = \mu_s mg$ and $F_c = \frac{mv^2}{R}$:
\[
\mu_s mg = \frac{mv^2}{R}.
\]

Canceling $m$ from both sides and solving for $v$:
\[
v = \sqrt{\mu_s g R}.
\]

Thus, the largest speed the Go-kart can have is:
\[
\boxed{v = \sqrt{\mu_s g R}}.
\]

\textbf{(b) Assume the track now forms an angle of $\phi$ with respect to the horizontal flat ground and that the track is now frictionless. Rededrive the largest speed the Go-kart may move with that maintains its circular motion around the track at the same radius $R$.} \\

For a frictionless banked curve, the centripetal force is provided entirely by the horizontal component of the normal force. The forces acting on the Go-kart are:
- The gravitational force, $mg$, acts vertically downward.
- The normal force, $N$, acts perpendicular to the inclined surface.

Resolving forces into components:
- The vertical component of the normal force balances the gravitational force: $N \cos\phi = mg$.
- The horizontal component of the normal force provides the centripetal force: $N \sin\phi = \frac{mv^2}{R}$.

From the first equation, solve for $N$:
\[
N = \frac{mg}{\cos\phi}.
\]

Substitute $N$ into the second equation:
\[
\frac{mg}{\cos\phi} \sin\phi = \frac{mv^2}{R}.
\]

Cancel $m$ from both sides and simplify:
\[
g \tan\phi = \frac{v^2}{R}.
\]

Solving for $v$:
\[
v = \sqrt{gR \tan\phi}.
\]

Thus, the largest speed the Go-kart can have is:
\[
\boxed{v = \sqrt{gR \tan\phi}}.
\]

\textbf{(c) Consider the same situation as in part (b), except the static coefficient is now again $\mu_s$. Derive the minimum speed, $v_{\text{min}}$, and maximum speed, $v_{\text{max}}$, showing that the following equality holds:}
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg\mu_s}{\cos^2(\phi) - \mu_s^2\sin^2(\phi)}.
\]

In this case, both static friction and the components of the normal force contribute to the centripetal force. The forces acting on the Go-kart are:
- The gravitational force, $mg$, acts vertically downward.
- The normal force, $N$, acts perpendicular to the inclined surface.
- The static frictional force, $F_s = \mu_s N$, acts along the incline.

Resolving forces into components:
- Vertically:
\[
N \cos\phi - F_s \sin\phi = mg.
\]

Substitute $F_s = \mu_s N$:
\[
N \cos\phi - \mu_s N \sin\phi = mg.
\]

Factor out $N$:
\[
N (\cos\phi - \mu_s \sin\phi) = mg.
\]

Solve for $N$:
\[
N = \frac{mg}{\cos\phi - \mu_s \sin\phi}.
\]

- Horizontally:
\[
N \sin\phi + F_s \cos\phi = \frac{mv^2}{R}.
\]

Substitute $F_s = \mu_s N$:
\[
N \sin\phi + \mu_s N \cos\phi = \frac{mv^2}{R}.
\]

Factor out $N$:
\[
N (\sin\phi + \mu_s \cos\phi) = \frac{mv^2}{R}.
\]

Substitute $N = \frac{mg}{\cos\phi - \mu_s \sin\phi}$:
\[
\frac{mg}{\cos\phi - \mu_s \sin\phi} (\sin\phi + \mu_s \cos\phi) = \frac{mv^2}{R}.
\]

Cancel $m$ and simplify:
\[
\frac{g (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi} = \frac{v^2}{R}.
\]

Solve for $v^2$:
\[
v^2 = \frac{Rg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi}.
\]

The maximum speed occurs when static friction acts up the incline, and the minimum speed occurs when static friction acts down the incline. Let:
\[
v_{\text{max}}^2 = \frac{Rg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi},
\]
\[
v_{\text{min}}^2 = \frac{Rg (\sin\phi - \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi}.
\]

The difference $v_{\text{max}}^2 - v_{\text{min}}^2$ is:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{Rg (\sin\phi + \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi} - \frac{Rg (\sin\phi - \mu_s \cos\phi)}{\cos\phi - \mu_s \sin\phi}.
\]

Combine the terms:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{Rg [(\sin\phi + \mu_s \cos\phi) - (\sin\phi - \mu_s \cos\phi)]}{\cos\phi - \mu_s \sin\phi}.
\]

Simplify:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg \mu_s \cos\phi}{\cos\phi - \mu_s \sin\phi}.
\]

Factor out the denominator:
\[
v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg\mu_s}{\cos^2(\phi) - \mu_s^2\sin^2(\phi)}.
\]

Thus, the required equality is shown:
\[
\boxed{v_{\text{max}}^2 - v_{\text{min}}^2 = \frac{2Rg\mu_s}{\cos^2(\phi) - \mu_s^2\sin^2(\phi)}}.
\]

"

\subsubsection{CMQ7}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 13 marks. Give up to 4 marks for part a), give up to 5 marks for part b) and give up to 4 marks for part c). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem:

"a) State the form of the energy-momentum 4-vector of one particle of mass \( m \) and show that its scalar product is invariant under Lorentz transformation. In the lab frame, the particle moves with velocity \( \underline{v} \). 

b) In its rest frame, the particle decays at rest into two identical massless particles which emerge along the positive and negative x-axis. Obtain expressions of the four-momenta of the two identical particles in the lab frame. 
    
c) For \( m = 10 \) GeV/\( c^{2} \), \( \beta = v/c = 0.8 \), and if the two massless particles are photons, find their corresponding wavelengths in the rest frame of the decaying particle and in the lab frame."

Mark Scheme:

"{Q7: (13 marks total)}

a) (4 marks)

[1 mark] \( \underline{v} \) is arbitrary, choose it to be along one of the axis. (ie. x-axis)

[1 mark] Energy momentum 4-vector of single particle, \( P \) is given by 

\[ P = ( E / c, p, 0, 0 ) \]

where \( E = \gamma m c^2 \), \( p = \gamma m | \underline{v} | = \gamma mv \), \( \gamma = \frac{1}{\sqrt{1-\frac{v^2}{c^2}}} \). Allow equivalent expressions.

[2 marks] Compute scalar product of the 4-vector, substituting definitions to show invariance.

\[ P^2 = E^2 / c^2 - \underline{p}^2 \]
\[ P^2 = \gamma^2 m^2 c^4 / c^2 - \gamma ^2 m^2 v^2 \]
\[ P^2 = m^2 c^2 \left [ \frac{1-v^2/c^2}{1-v^2/c^2} \right] = m^2c^2 \]

b) (5 marks)

[2 marks] 4-momenta of decaying particle, equal to the sum of 4-momenta of the massless particles 

\[ (E_1 / c, p_1, 0, 0) + (E_2 / c, p_2, 0, 0) = (mc, 0, 0, 0) \]

[1 mark] \( p_2 =  - p_1 \), since particles are massless, \( E_1 = E_2 = mc^2 / 2 = |p_1|c \). 4-momenta of both particles in the rest frame is

\[ \left( \frac{mc}{2}, \frac{mc}{2}, 0, 0 \right), \left( \frac{mc}{2}, -\frac{mc}{2}, 0, 0 \right) \]


[2 marks] Inverse Lorentz transforms the 4-momenta back to the lab frame . \( \beta = v / c \)

\[
\begin{pmatrix} \gamma & \beta \gamma & 0 & 0 \\ \beta \gamma & \gamma & 0 & 0 \\ 0 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \end{pmatrix} \begin{pmatrix} mc/2\\ mc/2\\ 0\\ 0 \end{pmatrix} = \begin{pmatrix} mc \gamma (1+\beta) /2\\ mc \gamma (1+\beta) /2\\ 0\\ 0 \end{pmatrix}
\]

\[
\begin{pmatrix} \gamma & \beta \gamma & 0 & 0 \\ \beta \gamma & \gamma & 0 & 0 \\ 0 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \end{pmatrix} \begin{pmatrix} mc/2\\ -mc/2\\ 0\\ 0 \end{pmatrix} = \begin{pmatrix} mc \gamma (1-\beta) /2\\ -mc \gamma (1-\beta) /2\\ 0\\ 0 \end{pmatrix}
\]


c) (4 marks)

[1 mark] Using \( v = 0.8 c ) \), \( \gamma = 5 / 3 \), \( \beta = 4 / 5 \), the 4-momenta are simplified to 

\[ 
\left ( \frac{3}{2}mc, \frac{3}{2}mc, 0,0 \right), \left ( \frac{1}{6}mc, -\frac{1}{6}mc,0,0 \right)
\]

[1 mark] Expression of wavelength from energy or momenta, 

\[ \lambda  = \frac{h}{p} \]
\[ \lambda = \frac{2h}{3mc} \quad \text{or} \quad \frac{6h}{mc} \]

h is planck’s constant. Use of energy is acceptable. 

[2 marks] Correct calculation using \( m = 10 GeV / c^2 \), 

\[
\lambda = 8.27 \times 10^{-17} m \quad \text{or} \quad 7.44 \times 10^{-16} m
\].

"

Solution 1: 

"
\textbf{(a) [4 marks]} 

The energy-momentum 4-vector for a particle of mass \(m\) moving with velocity \(v\) in the lab frame is given by:
\[
p^\mu = \begin{pmatrix}
E/c \\
\mathbf{p}
\end{pmatrix},
\]
where \(E = \gamma m c^2\) is the relativistic energy, \(\mathbf{p} = \gamma m \mathbf{v}\) is the relativistic momentum, and \(\gamma = \frac{1}{\sqrt{1 - v^2/c^2}}\) is the Lorentz factor. 

The scalar product of the energy-momentum 4-vector is:
\[
p^\mu p_\mu = \left(\frac{E}{c}\right)^2 - |\mathbf{p}|^2.
\]
Substituting \(E = \gamma mc^2\) and \(|\mathbf{p}| = \gamma mv\), we find:
\[
p^\mu p_\mu = \gamma^2 m^2 c^2 - \gamma^2 m^2 v^2 = \gamma^2 m^2 (c^2 - v^2) = m^2 c^2.
\]
This scalar product is invariant under Lorentz transformations because it depends only on the invariant mass \(m\) of the particle and the speed of light \(c\), which are the same in all reference frames.

\textbf{(b) [5 marks]}

In the rest frame of the particle, the particle is at rest before decay, so its energy-momentum 4-vector is:
\[
p^\mu = \begin{pmatrix}
mc \\
0
\end{pmatrix}.
\]
The particle decays into two identical massless particles (e.g., photons) traveling in opposite directions along the \(x\)-axis. The 4-momenta of the photons in the rest frame are:
\[
p_1^\mu = \begin{pmatrix}
E_1/c \\
E_1/c \\
0 \\
0
\end{pmatrix}, \quad
p_2^\mu = \begin{pmatrix}
E_2/c \\
-E_2/c \\
0 \\
0
\end{pmatrix}.
\]
By energy and momentum conservation, the total 4-momentum after the decay must equal the initial 4-momentum:
\[
p^\mu = p_1^\mu + p_2^\mu.
\]
Thus, \(E_1 = E_2 = mc^2/2\).

In the lab frame, the parent particle is moving with velocity \(v\) along the \(x\)-axis. The total 4-momentum in the lab frame is:
\[
p^\mu = \begin{pmatrix}
\gamma mc \\
\gamma mv \\
0 \\
0
\end{pmatrix}.
\]
The photons’ 4-momenta in the lab frame can be obtained by applying the Lorentz transformation. The transformed 4-momenta are:
\[
p_1^\mu = \begin{pmatrix}
\gamma \frac{mc^2}{2c} + \gamma \frac{v}{c} \frac{mc^2}{2c} \\
\gamma \frac{mc^2}{2c} + \gamma \frac{v}{c} \frac{mc^2}{2c} \\
0 \\
0
\end{pmatrix}, \quad
p_2^\mu = \begin{pmatrix}
\gamma \frac{mc^2}{2c} - \gamma \frac{v}{c} \frac{mc^2}{2c} \\
-\gamma \frac{mc^2}{2c} + \gamma \frac{v}{c} \frac{mc^2}{2c} \\
0 \\
0
\end{pmatrix}.
\]
Simplifying, the 4-momenta are:
\[
p_1^\mu = \begin{pmatrix}
\gamma \frac{mc}{2} (1 + \beta) \\
\gamma \frac{mc}{2} (1 + \beta) \\
0 \\
0
\end{pmatrix}, \quad
p_2^\mu = \begin{pmatrix}
\gamma \frac{mc}{2} (1 - \beta) \\
-\gamma \frac{mc}{2} (1 - \beta) \\
0 \\
0
\end{pmatrix}.
\]

\textbf{(c) [4 marks]}

Given \(m = 10 \, \text{GeV}/c^2\) and \(\beta = v/c = 0.8\), we calculate the corresponding wavelengths of the photons. 

In the rest frame, the energy of each photon is:
\[
E_\text{photon} = \frac{mc^2}{2} = \frac{10 \, \text{GeV}}{2} = 5 \, \text{GeV}.
\]
The wavelength of a photon is given by:
\[
\lambda = \frac{hc}{E}.
\]
Substituting \(E = 5 \, \text{GeV}\) and using \(hc = 1240 \, \text{eV} \cdot \text{nm}\), we find:
\[
\lambda = \frac{1240 \, \text{eV} \cdot \text{nm}}{5 \times 10^9 \, \text{eV}} = 2.48 \times 10^{-4} \, \text{nm}.
\]

In the lab frame, the energies of the photons are:
\[
E_1 = \gamma \frac{mc^2}{2} (1 + \beta), \quad E_2 = \gamma \frac{mc^2}{2} (1 - \beta).
\]
Substituting \(\gamma = \frac{1}{\sqrt{1 - \beta^2}} = \frac{1}{\sqrt{1 - 0.8^2}} = \frac{5}{3}\), we find:
\[
E_1 = \frac{5}{3} \cdot 5 \, \text{GeV} \cdot (1 + 0.8) = \frac{5}{3} \cdot 5 \cdot 1.8 \, \text{GeV} = 15 \, \text{GeV},
\]
\[
E_2 = \frac{5}{3} \cdot 5 \, \text{GeV} \cdot (1 - 0.8) = \frac{5}{3} \cdot 5 \cdot 0.2 \, \text{GeV} = \frac{5}{3} \, \text{GeV} \approx 1.67 \, \text{GeV}.
\]
The corresponding wavelengths are:
\[
\lambda_1 = \frac{1240 \, \text{eV} \cdot \text{nm}}{15 \times 10^9 \, \text{eV}} \approx 8.27 \times 10^{-5} \, \text{nm},
\]
\[
\lambda_2 = \frac{1240 \, \text{eV} \cdot \text{nm}}{1.67 \times 10^9 \, \text{eV}} \approx 7.43 \times 10^{-4} \, \text{nm}.
\]

"

Solution 2: 

"
\textbf{a) [4 marks]} 

The energy-momentum 4-vector of a particle with mass \(m\) is given by:
\[
p^\mu = \begin{pmatrix} E/c \\ \vec{p} \end{pmatrix},
\]
where \(E\) is the total energy of the particle, \(\vec{p}\) is the 3-momentum, and \(c\) is the speed of light. In the lab frame, for a particle moving with velocity \(v\), the components are:
\[
p^\mu = \begin{pmatrix} \gamma mc \\ \gamma mv \end{pmatrix},
\]
where \(\gamma = \frac{1}{\sqrt{1 - v^2/c^2}}\) is the Lorentz factor.

The scalar product of the 4-vector is invariant under Lorentz transformations. That is,
\[
p^\mu p_\mu = \left(\frac{E}{c}\right)^2 - |\vec{p}|^2.
\]
In terms of \(m\), the invariant scalar is:
\[
p^\mu p_\mu = \frac{(mc^2)^2}{c^2} - \frac{(mv)^2}{c^2} = m^2c^2.
\]
Thus, the scalar product \(p^\mu p_\mu = m^2c^2\) remains invariant under Lorentz transformations.

\textbf{b) [5 marks]}

In the rest frame of the particle, the particle decays into two massless particles (e.g., photons) that move along the positive and negative \(x\)-axis. For massless particles, the 4-momentum is given by:
\[
p^\mu = \begin{pmatrix} E/c \\ \vec{p} \end{pmatrix},
\]
with \(|\vec{p}| = E/c\). Energy and momentum conservation imply that in the rest frame:
\[
E_{\text{total}} = mc^2, \quad \vec{p}_{\text{total}} = 0.
\]
Each photon has energy \(E_1 = E_2 = \frac{mc^2}{2}\) and momentum \(|\vec{p}_1| = |\vec{p}_2| = \frac{mc}{2}\). The 4-momenta of the photons in the rest frame are:
\[
p_1^\mu = \begin{pmatrix} \frac{mc}{2} \\ \frac{mc}{2} \\ 0 \\ 0 \end{pmatrix}, \quad 
p_2^\mu = \begin{pmatrix} \frac{mc}{2} \\ -\frac{mc}{2} \\ 0 \\ 0 \end{pmatrix}.
\]

In the lab frame, the parent particle is moving with velocity \(v\) along the \(x\)-axis. Using Lorentz transformation, the energy and momentum components transform as:
\[
E' = \gamma (E + vp_x), \quad p_x' = \gamma (p_x + \frac{v}{c^2}E).
\]
Applying these to the photons:
\[
p_1^\mu = \begin{pmatrix} \gamma \frac{mc}{2}(1 + \beta) \\ \gamma \frac{mc}{2}(1 + \beta) \\ 0 \\ 0 \end{pmatrix}, \quad 
p_2^\mu = \begin{pmatrix} \gamma \frac{mc}{2}(1 - \beta) \\ \gamma \frac{mc}{2}(-1 + \beta) \\ 0 \\ 0 \end{pmatrix},
\]
where \(\beta = v/c\).

\textbf{c) [4 marks]}

For \(m = 10 \, \text{GeV}/c^2\), \(\beta = 0.8\), and \(\gamma = \frac{1}{\sqrt{1 - \beta^2}} = \frac{1}{\sqrt{1 - 0.8^2}} = 1.667\).

In the rest frame of the decaying particle, the energy of each photon is:
\[
E_{\text{rest}} = \frac{mc^2}{2} = \frac{10 \, \text{GeV}}{2} = 5 \, \text{GeV}.
\]
The wavelength is given by \(\lambda = \frac{hc}{E}\). Substituting \(E = 5 \, \text{GeV}\) and \(hc = 1.24 \, \text{GeV} \cdot \text{fm}\), we find:
\[
\lambda_{\text{rest}} = \frac{1.24 \, \text{GeV} \cdot \text{fm}}{5 \, \text{GeV}} = 0.248 \, \text{fm}.
\]

In the lab frame, the photon energies are:
\[
E_1' = \gamma E_{\text{rest}}(1 + \beta) = 1.667 \cdot 5 \, \text{GeV} \cdot (1 + 0.8) = 15 \, \text{GeV},
\]
\[
E_2' = \gamma E_{\text{rest}}(1 - \beta) = 1.667 \cdot 5 \, \text{GeV} \cdot (1 - 0.8) = 1.67 \, \text{GeV}.
\]
The corresponding wavelengths are:
\[
\lambda_1' = \frac{hc}{E_1'} = \frac{1.24 \, \text{GeV} \cdot \text{fm}}{15 \, \text{GeV}} = 0.0827 \, \text{fm},
\]
\[
\lambda_2' = \frac{hc}{E_2'} = \frac{1.24 \, \text{GeV} \cdot \text{fm}}{1.67 \, \text{GeV}} = 0.743 \, \text{fm}.
\]

"

Solution 3:

"
\textbf{(a)} The energy-momentum 4-vector of a particle of mass \(m\) moving with velocity \(v\) in the lab frame is given by:
\[
p^\mu = \begin{pmatrix} E/c \\ \vec{p} \end{pmatrix} = \begin{pmatrix} \gamma mc \\ \gamma mv \end{pmatrix},
\]
where \(\gamma = \frac{1}{\sqrt{1-v^2/c^2}}\) is the Lorentz factor, \(E = \gamma mc^2\) is the relativistic energy, and \(\vec{p} = \gamma mv\) is the relativistic momentum.

The scalar product of the energy-momentum 4-vector is defined as:
\[
p^\mu p_\mu = \left(\frac{E}{c}\right)^2 - |\vec{p}|^2.
\]
Substituting \(E = \gamma mc^2\) and \(|\vec{p}| = \gamma mv\), we get:
\[
p^\mu p_\mu = \left(\gamma mc\right)^2 - \left(\gamma mv\right)^2.
\]
Factoring out \(\gamma^2m^2\), we find:
\[
p^\mu p_\mu = \gamma^2m^2 \left(c^2 - v^2\right).
\]
Using the identity \(\gamma^2 = \frac{1}{1-v^2/c^2}\), we simplify:
\[
p^\mu p_\mu = m^2c^2.
\]
This shows that the scalar product \(p^\mu p_\mu = m^2c^2\) is invariant under Lorentz transformations, as it depends only on the rest mass \(m\), which is invariant.

\textbf{(b)} In the rest frame of the decaying particle, it is initially at rest, so its energy-momentum 4-vector is:
\[
P^\mu = \begin{pmatrix} mc \\ 0 \end{pmatrix}.
\]
After the decay, the particle produces two identical massless particles moving along the positive and negative \(x\)-axis. For a massless particle, the energy-momentum 4-vector is:
\[
p^\mu = \begin{pmatrix} E/c \\ \vec{p} \end{pmatrix} = \begin{pmatrix} |\vec{p}| \\ |\vec{p}| \hat{n} \end{pmatrix},
\]
where \(|\vec{p}| = E/c\).

By conservation of energy and momentum, the total energy and momentum before and after decay must be equal:
\[
P^\mu = p_1^\mu + p_2^\mu.
\]
Since the initial particle is at rest, \(P^\mu = \begin{pmatrix} mc \\ 0 \end{pmatrix}\). The decay products have equal and opposite momentum, so:
\[
p_1^\mu = \begin{pmatrix} mc/2 \\ mc/2 \end{pmatrix}, \quad p_2^\mu = \begin{pmatrix} mc/2 \\ -mc/2 \end{pmatrix}.
\]

In the lab frame, the parent particle moves with velocity \(v\). The Lorentz transformation for the energy-momentum 4-vectors is:
\[
p^\mu_{\text{lab}} = \Lambda^\mu_{\ \nu} p^\nu,
\]
where \(\Lambda^\mu_{\ \nu}\) is the Lorentz transformation matrix. Applying the transformation gives the 4-momenta of the decay products in the lab frame:
\[
p_1^\mu = \begin{pmatrix} \gamma E_1/c - \gamma \beta p_x \\ \gamma p_x - \gamma \beta E_1/c \end{pmatrix},
\]
\[
p_2^\mu = \begin{pmatrix} \gamma E_2/c + \gamma \beta p_x \\ -\gamma p_x + \gamma \beta E_2/c \end{pmatrix}.
\]

\textbf{(c)} Given \(m = 10 \, \text{GeV}/c^2\), \(\beta = v/c = 0.8\), \(\gamma = \frac{1}{\sqrt{1-\beta^2}} = \frac{1}{\sqrt{1-0.8^2}} = 5/3\).

In the rest frame of the decaying particle, the energy of each photon is \(E_\text{rest} = mc^2/2 = 10 \, \text{GeV}/2 = 5 \, \text{GeV}\). The corresponding wavelength is:
\[
\lambda_\text{rest} = \frac{hc}{E_\text{rest}} = \frac{(4.14 \times 10^{-15} \, \text{eV·s})(3 \times 10^8 \, \text{m/s})}{5 \times 10^9 \, \text{eV}} = 2.48 \times 10^{-13} \, \text{m}.
\]

In the lab frame, the photon energies are Doppler-shifted. Using the relativistic Doppler formula, the energies of the forward and backward photons are:
\[
E_\text{forward} = \gamma (1+\beta) E_\text{rest} = \frac{5}{3} (1+0.8)(5 \, \text{GeV}) = 15 \, \text{GeV},
\]
\[
E_\text{backward} = \gamma (1-\beta) E_\text{rest} = \frac{5}{3} (1-0.8)(5 \, \text{GeV}) = 5 \, \text{GeV}.
\]

The corresponding wavelengths in the lab frame are:
\[
\lambda_\text{forward} = \frac{hc}{E_\text{forward}} = \frac{(4.14 \times 10^{-15})(3 \times 10^8)}{15 \times 10^9} = 8.28 \times 10^{-14} \, \text{m},
\]
\[
\lambda_\text{backward} = \frac{hc}{E_\text{backward}} = \frac{(4.14 \times 10^{-15})(3 \times 10^8)}{5 \times 10^9} = 2.48 \times 10^{-13} \, \text{m}.
\]

"

\subsubsection{CMQ8}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 10 marks. Give up to 9 marks for part a) and give up to 1 mark for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem:

"A pendulum under the influence of gravity is formed by a massless string of fixed length, \( L \), attached to a mass \( m \). The pendulum is connected to a support (as seen in the figure) which moves with a position given by 

\[ x(t) = v t^3 + A \sin(\omega t) \]
    
where \( v \) and \( A \) are constants.
    
a) Starting from the expression of the Lagrangian, derive the equation of motion of the angle, \( \phi \), of the pendulum. 
    
b) In the limit of \( v \), \(A\) approaching 0 and small angles (\( \phi \) approaching 0), show this reduces to the classic pendulum problem and equation of motion."

Mark Scheme:

"{Q8: (10 marks total)}

a) (9 marks)

[1 mark] position of mass given by \( \underline{r} = (x + L \sin(\phi), -L \cos(\phi)) \). 

[3 marks] Calculation of kinetic and potential energy to give Lagrangian. 

\[ \dot{\underline{r}} = (\dot{x} + L\dot{\phi}\cos(\phi), L\dot{\phi}\sin(\phi) ) \]
\[ K = \frac{1}{2}m |\dot{\underline{r}}|^2 = \frac{1}{2}m (\dot{x}^2 + L^2 \dot{\phi}^2 + 2\dot{x}L\dot{\phi}\cos(\phi) ) \]
\[ U = -mgL \cos(\phi) \]
\[ L = K - U = \frac{1}{2}m \left(\dot{x}^2 + L^2 \dot{\phi}^2 + 2\dot{x}L\dot{\phi}\cos(\phi) \right) + mgL \cos(\phi) \]

[3 marks] Obtain and simplify Euler-Lagrange equations, 

\[ \frac{d}{dt}\left ( \frac{\partial L}{\partial \dot{\phi}} \right ) = \frac{\partial L}{\partial \phi} \]

\[ \frac{d}{dt} \left ( mL^2\dot{\phi} + m\dot{x}L\cos{\phi} \right ) = -m\dot{x}L\dot{\phi}\sin(\phi) - mgL\sin(\phi) \]

\[ mL^2 \ddot{\phi} + m\ddot{x}L\cos{\phi} - m\dot{x}L\dot{\phi}\sin{\phi} = -m\dot{x}L\dot{\phi}\sin(\phi) - mgL\sin(\phi) \]

\[ L \ddot{\phi} + \ddot{x}\cos{\phi} = - g\sin{\phi} \]

[2 marks] Compute \( \ddot{x} = 6vt - A \omega^2 \sin(\omega t) \), substitute to obtain equation of motion

\[ L \ddot{\phi} + \left (6vt - A\omega^2\sin(\omega t) \right )\cos{\phi} + g\sin{\phi} = 0 \]

b) 

[1 mark] 2nd term in the equation of motion vanishes, we are left with 

\[  L \ddot{\phi} + g\sin{\phi} = 0 \]
\[ \ddot{\phi} + \frac{g}{L}\sin{\phi} = 0 \]

Allow direct substitution for the expression of x into all steps and explicit calculation.

"

Solution 1: 

"
\textbf{(a) Derive the equation of motion for the angle $\phi$:}

The Lagrangian $\mathcal{L}$ is given by
\[
\mathcal{L} = T - V,
\]
where $T$ is the kinetic energy of the system, and $V$ is the potential energy. 

To describe the motion, the position of the mass $m$ in terms of the angle $\phi$ and the time-dependent position of the support is:
\[
x_m = x(t) + L\sin\phi, \quad y_m = -L\cos\phi,
\]
where $x(t) = vt^3 + A\sin(\omega t)$ is the position of the support.

The kinetic energy $T$ is:
\[
T = \frac{1}{2}m\left(\dot{x}_m^2 + \dot{y}_m^2\right).
\]

Differentiating $x_m$ and $y_m$ with respect to time, we find:
\[
\dot{x}_m = \dot{x}(t) + L\cos\phi \, \dot{\phi}, \quad \dot{y}_m = L\sin\phi \, \dot{\phi}.
\]

Substituting these into the kinetic energy:
\[
\dot{x}_m^2 = \left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right)^2, \quad
\dot{y}_m^2 = \left(L\sin\phi \, \dot{\phi}\right)^2.
\]
Expanding and simplifying:
\[
T = \frac{1}{2}m \left[\left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right)^2 + \left(L\sin\phi \, \dot{\phi}\right)^2 \right].
\]

Next, the potential energy $V$ is due to gravity and is given by:
\[
V = -mg y_m = mgL\cos\phi.
\]

The Lagrangian $\mathcal{L}$ becomes:
\[
\mathcal{L} = \frac{1}{2}m \left[\left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right)^2 + \left(L\sin\phi \, \dot{\phi}\right)^2 \right] - mgL\cos\phi.
\]

Using the Euler-Lagrange equation:
\[
\frac{d}{dt} \left(\frac{\partial \mathcal{L}}{\partial \dot{\phi}}\right) - \frac{\partial \mathcal{L}}{\partial \phi} = 0,
\]
we compute the derivatives.

First, calculate $\frac{\partial \mathcal{L}}{\partial \dot{\phi}}$:
\[
\frac{\partial \mathcal{L}}{\partial \dot{\phi}} = mL^2\dot{\phi} + mL\cos\phi \, \left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right).
\]

Now, compute $\frac{d}{dt}\left(\frac{\partial \mathcal{L}}{\partial \dot{\phi}}\right)$:
\[
\frac{d}{dt}\left(\frac{\partial \mathcal{L}}{\partial \dot{\phi}}\right) = mL^2\ddot{\phi} + mL(-\sin\phi \, \dot{\phi})\left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right) + mL\cos\phi \, \ddot{x}(t) + mL^2\cos^2\phi \, \ddot{\phi} - mL^2\sin\phi \, \dot{\phi}^2.
\]

Next, compute $\frac{\partial \mathcal{L}}{\partial \phi}$:
\[
\frac{\partial \mathcal{L}}{\partial \phi} = -mL\sin\phi \, \left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right)^2 + mgL\sin\phi.
\]

Substitute these into the Euler-Lagrange equation and simplify to obtain the equation of motion:
\[
L\ddot{\phi} + g\sin\phi = \ddot{x}(t)\cos\phi.
\]

Thus, the equation of motion is:
\[
\boxed{L\ddot{\phi} + g\sin\phi = \ddot{x}(t)\cos\phi.}
\]

\textbf{(b) Limit to the classic pendulum problem:}

For small angles ($\phi \to 0$), $\sin\phi \approx \phi$ and $\cos\phi \approx 1$. If $v \to 0$ and $A \to 0$, the support does not move, so $\ddot{x}(t) = 0$. The equation of motion reduces to:
\[
L\ddot{\phi} + g\phi = 0.
\]

This is the classic equation of motion for a simple pendulum:
\[
\boxed{\ddot{\phi} + \frac{g}{L}\phi = 0.}
\]

"

Solution 2: 

"
\textbf{(a)} 

To derive the equation of motion of the angle $\phi$, we start by finding the Lagrangian of the system. The pendulum's motion occurs in two dimensions: the vertical plane and the horizontal motion induced by the moving support. 

The position of the pendulum's mass $m$ in terms of the angle $\phi$ is given by:
\[
x_m = x(t) + L\sin\phi, \quad y_m = -L\cos\phi,
\]
where $x(t) = vt^3 + A\sin(\omega t)$ is the horizontal position of the moving support.

The kinetic energy $T$ of the pendulum mass is:
\[
T = \frac{1}{2}m\left(\dot{x}_m^2 + \dot{y}_m^2\right),
\]
where $\dot{x}_m$ and $\dot{y}_m$ are the time derivatives of $x_m$ and $y_m$, respectively.

First, compute $\dot{x}_m$ and $\dot{y}_m$:
\[
\dot{x}_m = \dot{x}(t) + L\cos\phi\,\dot{\phi}, \quad \dot{y}_m = L\sin\phi\,\dot{\phi},
\]
where $\dot{x}(t) = 3vt^2 + A\omega\cos(\omega t)$.

Substitute these into the expression for $T$:
\[
T = \frac{1}{2}m\left[\left(\dot{x}(t) + L\cos\phi\,\dot{\phi}\right)^2 + \left(L\sin\phi\,\dot{\phi}\right)^2\right].
\]
Expand the terms:
\[
T = \frac{1}{2}m\left[\dot{x}(t)^2 + 2L\dot{x}(t)\cos\phi\,\dot{\phi} + L^2\cos^2\phi\,\dot{\phi}^2 + L^2\sin^2\phi\,\dot{\phi}^2\right].
\]
Use $\cos^2\phi + \sin^2\phi = 1$ to simplify:
\[
T = \frac{1}{2}m\left[\dot{x}(t)^2 + 2L\dot{x}(t)\cos\phi\,\dot{\phi} + L^2\dot{\phi}^2\right].
\]

The potential energy $U$ of the pendulum is:
\[
U = -mg y_m = mgL\cos\phi.
\]

The Lagrangian $\mathcal{L}$ is given by:
\[
\mathcal{L} = T - U = \frac{1}{2}m\left[\dot{x}(t)^2 + 2L\dot{x}(t)\cos\phi\,\dot{\phi} + L^2\dot{\phi}^2\right] - mgL\cos\phi.
\]

To derive the equation of motion, we use the Euler-Lagrange equation:
\[
\frac{d}{dt}\left(\pdv{\mathcal{L}}{\dot{\phi}}\right) - \pdv{\mathcal{L}}{\phi} = 0.
\]

First, compute $\pdv{\mathcal{L}}{\dot{\phi}}$:
\[
\pdv{\mathcal{L}}{\dot{\phi}} = m\left(L\dot{x}(t)\cos\phi + L^2\dot{\phi}\right).
\]

Differentiate with respect to $t$:
\[
\frac{d}{dt}\left(\pdv{\mathcal{L}}{\dot{\phi}}\right) = m\left[L\ddot{x}(t)\cos\phi - L\dot{x}(t)\sin\phi\,\dot{\phi} + L^2\ddot{\phi}\right].
\]

Next, compute $\pdv{\mathcal{L}}{\phi}$:
\[
\pdv{\mathcal{L}}{\phi} = -mL\dot{x}(t)\sin\phi\,\dot{\phi} - mL^2\sin\phi\,\dot{\phi}^2 + mgL\sin\phi.
\]

Substitute these into the Euler-Lagrange equation:
\[
m\left[L\ddot{x}(t)\cos\phi - L\dot{x}(t)\sin\phi\,\dot{\phi} + L^2\ddot{\phi}\right] - \left[-mL\dot{x}(t)\sin\phi\,\dot{\phi} - mL^2\sin\phi\,\dot{\phi}^2 + mgL\sin\phi\right] = 0.
\]

Simplify:
\[
L\ddot{x}(t)\cos\phi + L^2\ddot{\phi} + mgL\sin\phi = 0.
\]

Divide through by $mL$:
\[
\ddot{\phi} + \frac{g}{L}\sin\phi = -\frac{\ddot{x}(t)}{L}\cos\phi.
\]

This is the equation of motion for the angle $\phi$ of the pendulum.

\textbf{(b)}

In the limit where $v, A \to 0$ and for small angles ($\phi \to 0$), $\sin\phi \approx \phi$ and $\cos\phi \approx 1$. The equation of motion becomes:
\[
\ddot{\phi} + \frac{g}{L}\phi = 0,
\]
which is the classic simple harmonic motion equation for a pendulum.


"

Solution 3:

"
We start by analyzing the system using the principles of Lagrangian mechanics. The pendulum of length $L$ is attached to a support that moves horizontally with a position given by 
\[
x(t) = vt^3 + A\sin(\omega t),
\]
where $v$ and $A$ are constants. The pendulum's position can be described in terms of its angle $\phi$ with respect to the vertical.

The coordinates of the pendulum bob in the inertial frame are:
\[
x_\text{bob}(t) = x(t) + L\sin\phi,
\]
\[
y_\text{bob}(t) = -L\cos\phi.
\]

The velocity components of the pendulum bob are obtained by differentiating with respect to time:
\[
\dot{x}_\text{bob} = \dot{x}(t) + L\cos\phi \, \dot{\phi},
\]
\[
\dot{y}_\text{bob} = L\sin\phi \, \dot{\phi}.
\]

The total kinetic energy of the pendulum bob is:
\[
T = \frac{1}{2}m\left(\dot{x}_\text{bob}^2 + \dot{y}_\text{bob}^2\right).
\]
Substituting $\dot{x}_\text{bob}$ and $\dot{y}_\text{bob}$, we get:
\[
T = \frac{1}{2}m\left[\left(\dot{x}(t) + L\cos\phi \, \dot{\phi}\right)^2 + \left(L\sin\phi \, \dot{\phi}\right)^2\right].
\]
Expanding and simplifying:
\[
T = \frac{1}{2}m\left[\dot{x}^2(t) + 2\dot{x}(t)L\cos\phi \, \dot{\phi} + L^2\cos^2\phi \, \dot{\phi}^2 + L^2\sin^2\phi \, \dot{\phi}^2\right].
\]
Using $\cos^2\phi + \sin^2\phi = 1$, the kinetic energy becomes:
\[
T = \frac{1}{2}m\left[\dot{x}^2(t) + 2\dot{x}(t)L\cos\phi \, \dot{\phi} + L^2\dot{\phi}^2\right].
\]

The potential energy of the pendulum bob, taking the vertical position as zero potential energy, is:
\[
V = -mg y_\text{bob} = mgL\cos\phi.
\]

The Lagrangian of the system is:
\[
\mathcal{L} = T - V.
\]
Substituting for $T$ and $V$:
\[
\mathcal{L} = \frac{1}{2}m\left[\dot{x}^2(t) + 2\dot{x}(t)L\cos\phi \, \dot{\phi} + L^2\dot{\phi}^2\right] - mgL\cos\phi.
\]

The equation of motion is derived using the Euler-Lagrange equation:
\[
\frac{d}{dt}\left(\frac{\partial \mathcal{L}}{\partial \dot{\phi}}\right) - \frac{\partial \mathcal{L}}{\partial \phi} = 0.
\]

First, calculate $\frac{\partial \mathcal{L}}{\partial \dot{\phi}}$:
\[
\frac{\partial \mathcal{L}}{\partial \dot{\phi}} = m\left[L^2\dot{\phi} + \dot{x}(t)L\cos\phi\right].
\]

Differentiate with respect to $t$:
\[
\frac{d}{dt}\left(\frac{\partial \mathcal{L}}{\partial \dot{\phi}}\right) = m\left[L^2\ddot{\phi} + L\cos\phi \, \ddot{x}(t) - L\sin\phi \, \dot{\phi}\dot{x}(t)\right].
\]

Now, calculate $\frac{\partial \mathcal{L}}{\partial \phi}$:
\[
\frac{\partial \mathcal{L}}{\partial \phi} = m\left[-L\sin\phi \, \dot{x}(t)\dot{\phi} - mgL\sin\phi\right].
\]

Substitute into the Euler-Lagrange equation:
\[
m\left[L^2\ddot{\phi} + L\cos\phi \, \ddot{x}(t) - L\sin\phi \, \dot{\phi}\dot{x}(t)\right] - m\left[-L\sin\phi \, \dot{x}(t)\dot{\phi} - gL\sin\phi\right] = 0.
\]

Simplify:
\[
L^2\ddot{\phi} + L\cos\phi \, \ddot{x}(t) - L\sin\phi \, \dot{\phi}\dot{x}(t) + L\sin\phi \, \dot{x}(t)\dot{\phi} - gL\sin\phi = 0.
\]

Cancel terms and divide through by $L$:
\[
\ddot{\phi} + \frac{\ddot{x}(t)}{L}\cos\phi + \frac{g}{L}\sin\phi = 0.
\]

This is the equation of motion of the pendulum.

In the limit where $v \to 0$, $A \to 0$, and $\phi \to 0$, $\cos\phi \to 1$ and $\sin\phi \approx \phi$. The equation reduces to:
\[
\ddot{\phi} + \frac{g}{L}\phi = 0,
\]
which is the classic simple pendulum equation of motion.
"

\subsubsection{CMQ9}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 7 marks. Give up to 6 marks for part a) and give up to 1 mark for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem:

"A rope connects to a block of mass \( m \) and wraps around a circular disk of mass \( M \) and radius \( R \). Due to gravity, the rope unwinds and the block falls down. 

a) Derive expressions for the angular acceleration of the disk, the tension in the rope, and the acceleration of the block. Work under the assumption that the rope does not slip and that the moment of inertia of the disk is given by \( I = \frac{1}{2} m R^{2} \).
    
b) Calculate the angular acceleration of the disk, tension in the rope, and acceleration of the block for the case \( m = 3 \) kg, \( M = 12 \) kg, \( R = 0.2 \) m."

Mark Scheme:

"{Q9: (7 marks total)}

a) (6 marks)

[2 marks] apply laws of motion for acceleration of block of mass m, torque on circular disk. Let \( \alpha \) be the angular acceleration of the disk, \( T \) the tension in the rope and \( a \) the acceleration of the block. 

\[ T - mg = -ma \]
\[ RT = I \alpha = \frac{1}{2} M R^2 \alpha \]

[1 mark] Constraint that acceleration of block is equal to acceleration of disk since rope does not slip 

\[ a = R \alpha \]

[3 marks] Solve simultaneous equations to give expressions for T, a and \( \alpha \).

\[ a = \frac{mg}{m + \frac{1}{2}M} \]

\[ T = \frac{1}{2}Ma = \frac{\frac{1}{2}Mmg}{m + \frac{1}{2}M} \]

\[ \alpha = a / R = \frac{mg}{(m + \frac{1}{2}M)R} \]

b) (1 mark) 

[1 mark] substitute values \(m = 3kg \), \( M = 12kg \) and \( R = 0.2m \). \(g = 9.81 ms^{-2} \).

\[ a = 3.27 \, ms^{-2} \]
\[ T = 19.62 \, N \]
\[ \alpha = 16.35 \, rads^{-2} \]

"

Solution 1: 

"
\textbf{a) Derive expressions for the angular acceleration of the disk, the tension in the rope, and the acceleration of the block.}

We begin by analyzing the forces and torques acting on the system. Let the acceleration of the block be $a$ and the angular acceleration of the disk be $\alpha$. The relationship between $a$ and $\alpha$ is given by:
\[
a = R\alpha,
\]
since the rope does not slip.

\textit{Step 1: Force equation for the block}

The forces acting on the block are the gravitational force $mg$ and the tension $T$ in the rope. Using Newton's second law for the block:
\[
mg - T = ma. \tag{1}
\]

\textit{Step 2: Torque equation for the disk}

The torque $\tau$ on the disk is due to the tension $T$ in the rope. Using the moment of inertia of the disk, $I = \frac{1}{2}MR^2$, and the relationship $\tau = I\alpha$, we write:
\[
TR = \frac{1}{2}MR^2\alpha. \tag{2}
\]

Substitute $\alpha = \frac{a}{R}$ into equation (2):
\[
TR = \frac{1}{2}MR^2\frac{a}{R}.
\]
Simplifying:
\[
T = \frac{1}{2}Ma. \tag{3}
\]

\textit{Step 3: Combine equations to find $a$}

Substitute $T$ from equation (3) into equation (1):
\[
mg - \frac{1}{2}Ma = ma.
\]
Rearrange to solve for $a$:
\[
mg = ma + \frac{1}{2}Ma,
\]
\[
mg = a\left(m + \frac{1}{2}M\right).
\]
Thus, the acceleration of the block is:
\[
a = \frac{mg}{m + \frac{1}{2}M}. \tag{4}
\]

\textit{Step 4: Find the angular acceleration $\alpha$}

Using the relationship $\alpha = \frac{a}{R}$:
\[
\alpha = \frac{mg}{R\left(m + \frac{1}{2}M\right)}. \tag{5}
\]

\textit{Step 5: Find the tension $T$}

Substitute $a$ from equation (4) into equation (3):
\[
T = \frac{1}{2}M\left(\frac{mg}{m + \frac{1}{2}M}\right),
\]
\[
T = \frac{\frac{1}{2}Mmg}{m + \frac{1}{2}M}. \tag{6}
\]

Thus, the derived expressions are:
\[
a = \frac{mg}{m + \frac{1}{2}M}, \quad \alpha = \frac{mg}{R\left(m + \frac{1}{2}M\right)}, \quad T = \frac{\frac{1}{2}Mmg}{m + \frac{1}{2}M}.
\]

\textbf{b) Calculate the angular acceleration, tension, and block acceleration for $m = 3\,\text{kg}$, $M = 12\,\text{kg}$, $R = 0.2\,\text{m}$.}

\textit{Step 1: Block acceleration $a$}

Substitute $m = 3$, $M = 12$, and $g = 9.8$ into equation (4):
\[
a = \frac{(3)(9.8)}{3 + \frac{1}{2}(12)},
\]
\[
a = \frac{29.4}{9},
\]
\[
a = 3.27\,\text{m/s}^2.
\]

\textit{Step 2: Angular acceleration $\alpha$}

Substitute $a = 3.27$ and $R = 0.2$ into $\alpha = \frac{a}{R}$:
\[
\alpha = \frac{3.27}{0.2},
\]
\[
\alpha = 16.35\,\text{rad/s}^2.
\]

\textit{Step 3: Tension $T$}

Substitute $m = 3$, $M = 12$, and $g = 9.8$ into equation (6):
\[
T = \frac{\frac{1}{2}(12)(3)(9.8)}{3 + \frac{1}{2}(12)},
\]
\[
T = \frac{(6)(29.4)}{9},
\]
\[
T = \frac{176.4}{9},
\]
\[
T = 19.6\,\text{N}.
\]

\textbf{Final Results:}
\[
a = 3.27\,\text{m/s}^2, \quad \alpha = 16.35\,\text{rad/s}^2, \quad T = 19.6\,\text{N}.
\]

"

Solution 2: 

"
\textbf{(a) Deriving expressions for the angular acceleration of the disk, tension in the rope, and acceleration of the block}

Let the tension in the rope be $T$, the angular acceleration of the disk be $\alpha$, and the linear acceleration of the block be $a$. We start by analyzing the forces and torques in the system.

\textbf{Block:}
The forces acting on the block are:
\[
\text{Downward force due to gravity: } F_{\text{gravity}} = mg,
\]
\[
\text{Upward force due to tension: } F_{\text{tension}} = T.
\]
Applying Newton's second law for the block:
\[
mg - T = ma \tag{1}.
\]

\textbf{Disk:}
The torque on the disk is caused by the tension in the rope. Using the moment of inertia of the disk $I = \frac{1}{2} M R^2$, the net torque is:
\[
\tau = T R = I \alpha = \left(\frac{1}{2} M R^2\right) \alpha. 
\]
Therefore,
\[
T R = \frac{1}{2} M R^2 \alpha \implies T = \frac{1}{2} M R \alpha \tag{2}.
\]

\textbf{Kinematic relationship:}
Since the rope does not slip, the tangential acceleration of the edge of the disk is equal to the linear acceleration of the block. This gives:
\[
a = R \alpha \tag{3}.
\]

\textbf{Solving the system of equations:}
Substitute $T$ from Eq. (2) into Eq. (1):
\[
mg - \frac{1}{2} M R \alpha = ma.
\]
Using Eq. (3) to replace $a$ with $R \alpha$, we get:
\[
mg - \frac{1}{2} M R \alpha = m R \alpha.
\]
Rearranging terms:
\[
mg = m R \alpha + \frac{1}{2} M R \alpha,
\]
\[
mg = R \alpha \left(m + \frac{1}{2} M\right).
\]
Solving for $\alpha$:
\[
\alpha = \frac{mg}{R \left(m + \frac{1}{2} M\right)} \tag{4}.
\]

Substitute Eq. (4) into Eq. (3) to find $a$:
\[
a = R \alpha = R \cdot \frac{mg}{R \left(m + \frac{1}{2} M\right)} = \frac{mg}{m + \frac{1}{2} M} \tag{5}.
\]

Finally, substitute $\alpha$ from Eq. (4) into Eq. (2) to find $T$:
\[
T = \frac{1}{2} M R \alpha = \frac{1}{2} M R \cdot \frac{mg}{R \left(m + \frac{1}{2} M\right)} = \frac{\frac{1}{2} M m g}{m + \frac{1}{2} M} \tag{6}.
\]

\textbf{Final expressions:}
\[
\alpha = \frac{mg}{R \left(m + \frac{1}{2} M\right)},
\]
\[
a = \frac{mg}{m + \frac{1}{2} M},
\]
\[
T = \frac{\frac{1}{2} M m g}{m + \frac{1}{2} M}.
\]

\textbf{(b) Numerical calculations:}

Given $m = 3 \, \text{kg}$, $M = 12 \, \text{kg}$, and $R = 0.2 \, \text{m}$, calculate each quantity:

\[
\alpha = \frac{(3)(9.8)}{0.2 \left(3 + \frac{1}{2} (12)\right)} = \frac{29.4}{0.2 \cdot 9} = \frac{29.4}{1.8} \approx 16.33 \, \text{rad/s}^2.
\]

\[
a = \frac{(3)(9.8)}{3 + \frac{1}{2} (12)} = \frac{29.4}{9} \approx 3.27 \, \text{m/s}^2.
\]

\[
T = \frac{\frac{1}{2} (12)(3)(9.8)}{3 + \frac{1}{2} (12)} = \frac{176.4}{9} \approx 19.6 \, \text{N}.
\]

\textbf{Final results:}
\[
\alpha \approx 16.33 \, \text{rad/s}^2, \quad a \approx 3.27 \, \text{m/s}^2, \quad T \approx 19.6 \, \text{N}.
\]

"

Solution 3:

"
\textbf{Solution to Q9:}

We analyze the problem by considering the forces and torques acting on the system. The block of mass $m$ is subject to gravitational force $mg$ and the tension $T$ in the rope, while the disk experiences a torque due to the tension $T$.

\textbf{Part (a): Derive expressions for angular acceleration of the disk, tension in the rope, and acceleration of the block.}

1. \textbf{Block's Equation of Motion:}

The block of mass $m$ undergoes linear acceleration $a$ as it falls. The forces acting on the block are gravity $mg$ and the upward tension $T$ in the rope. Using Newton's second law:

\[
mg - T = ma \tag{1}
\]

2. \textbf{Disk's Rotational Dynamics:}

The disk of mass $M$ and radius $R$ rotates as the rope unwinds. Assuming the rope does not slip, the linear acceleration of the block $a$ is related to the angular acceleration of the disk $\alpha$ by:

\[
a = R\alpha \tag{2}
\]

The torque on the disk is provided by the tension $T$ in the rope. Using the rotational form of Newton's second law:

\[
TR = I\alpha \tag{3}
\]

Substitute the moment of inertia of the disk, $I = \frac{1}{2}MR^2$, into equation (3):

\[
TR = \frac{1}{2}MR^2\alpha \tag{4}
\]

From equation (4), solve for $T$ in terms of $\alpha$:

\[
T = \frac{1}{2}MR\alpha \tag{5}
\]

3. \textbf{Relating Linear and Angular Accelerations:}

Substitute $a = R\alpha$ (equation 2) into equation (1):

\[
mg - T = mR\alpha \tag{6}
\]

Substitute $T$ from equation (5) into equation (6):

\[
mg - \frac{1}{2}MR\alpha = mR\alpha
\]

Rearrange to solve for $\alpha$:

\[
mg = \left(mR + \frac{1}{2}MR\right)\alpha
\]

\[
\alpha = \frac{mg}{R\left(m + \frac{1}{2}M\right)} \tag{7}
\]

4. \textbf{Solve for $T$ and $a$:}

Substitute $\alpha$ from equation (7) into equation (5) to find $T$:

\[
T = \frac{1}{2}MR\left(\frac{mg}{R\left(m + \frac{1}{2}M\right)}\right)
\]

\[
T = \frac{\frac{1}{2}Mmg}{m + \frac{1}{2}M} \tag{8}
\]

Substitute $\alpha$ from equation (7) into $a = R\alpha$ to find $a$:

\[
a = R\left(\frac{mg}{R\left(m + \frac{1}{2}M\right)}\right)
\]

\[
a = \frac{mg}{m + \frac{1}{2}M} \tag{9}
\]

Thus, the expressions for angular acceleration, tension, and linear acceleration are:

\[
\boxed{\alpha = \frac{mg}{R\left(m + \frac{1}{2}M\right)}, \quad T = \frac{\frac{1}{2}Mmg}{m + \frac{1}{2}M}, \quad a = \frac{mg}{m + \frac{1}{2}M}}
\]

\textbf{Part (b): Calculate $\alpha$, $T$, and $a$ for $m = 3\,\text{kg}$, $M = 12\,\text{kg}$, $R = 0.2\,\text{m}$.}

Substitute the given values into the expressions derived above.

1. Angular acceleration $\alpha$:

\[
\alpha = \frac{(3)(9.8)}{(0.2)\left(3 + \frac{1}{2}(12)\right)}
\]

\[
\alpha = \frac{29.4}{0.2(3 + 6)} = \frac{29.4}{1.8} \approx 16.33\,\text{rad/s}^2
\]

2. Tension $T$:

\[
T = \frac{\frac{1}{2}(12)(3)(9.8)}{3 + \frac{1}{2}(12)}
\]

\[
T = \frac{6(29.4)}{3 + 6} = \frac{176.4}{9} \approx 19.6\,\text{N}
\]

3. Linear acceleration $a$:

\[
a = \frac{(3)(9.8)}{3 + \frac{1}{2}(12)}
\]

\[
a = \frac{29.4}{9} \approx 3.27\,\text{m/s}^2
\]

Final answers:

\[
\boxed{\alpha \approx 16.33\,\text{rad/s}^2, \quad T \approx 19.6\,\text{N}, \quad a \approx 3.27\,\text{m/s}^2}
\]

"

\subsubsection{CMQ10}

You are a top professor of physics. There is a physics problem given below with 3 solutions written by students. Grade each solution out of a total of 8 marks. Give up to 6 marks for part a) and give up to 2 marks for part b). Use the mark scheme which has been given. Only give whole numbers for marks. Give concise general feedback for each solution.

Problem:

"On Earth, a ball of mass \( m \) is dropped from an airplane moving with horizontal velocity \( \underline{u} \). The air resistance on the ball causes a force opposite to the ball's velocity \( \underline{v} \), namely 

\[ \underline{F} = - b \underline{v} \] 
    
where \( b \) is a positive constant. 
    
a) Using Newton's laws of motion, show that 
\[
m\frac{d\underline{v}}{dt} = - mg \hat{\underline{k}} - b\underline{v} 
\]
    
and show the solution to this equation is of the form 
    
\[
\underline{v} = \underline{A} e^{-bt/m} - \frac{mg}{b} \hat{\underline{k}} 
\]
    
where \( \underline{A} \) is a constant vector.
    b) Given that the ball initially starts with velocity \( \underline{u} \), find an expression for \( \underline{A} \)."

Mark Scheme:

"{Q10: (8 marks total)}

a) (6 marks)

[1 mark] Net force is the sum of weight and force of air resistance. Equating with Newton’s 2nd Law, 

\[
m\frac{d\underline{v}}{dt} = - mg \hat{\underline{k}} - b\underline{v} 
\]

[3 marks] Solve differential equation, noting a general solution of the form 

\[
\underline{v} = \underline{A} e^{-bt/m} - \underline{B}
\]

[2 marks] Use of the boundary condition that at \( t = \infty \),  \( v \) approaches terminal velocity and weight is balanced by air resistance \( b\underline{v} = - mg \hat{\underline{k}} \), giving \( \underline{B} = - \frac{mg}{b} \hat{\underline{k}} \).

b) (2 marks)

[1 mark] Use of initial boundary conditions, \( \underline{v}(0) = \underline{u} \)

[1 mark] Substitute into solution to give

\[ \underline{A} = \underline{u} + \frac{mg}{b} \hat{\underline{k}} \]

"

Solution 1: 

"
\textbf{a)} We start by applying Newton's second law of motion:

\[
\mathbf{F} = m \frac{d\mathbf{v}}{dt}.
\]

The forces acting on the ball are the gravitational force $-mg\hat{k}$ and the air resistance $-b\mathbf{v}$. Therefore, the net force can be written as:

\[
m \frac{d\mathbf{v}}{dt} = -mg\hat{k} - b\mathbf{v}.
\]

Dividing through by $m$, we obtain the differential equation for the velocity $\mathbf{v}$:

\[
\frac{d\mathbf{v}}{dt} = -\frac{b}{m} \mathbf{v} - g\hat{k}.
\]

To solve this differential equation, we use the method of separation of variables. Let us rewrite $\mathbf{v}$ as a function of time in component form. Suppose the velocity has two components: a vertical component $v_k$ along $\hat{k}$ and a horizontal component that is unaffected by gravity. Then, the differential equation for the vertical motion becomes:

\[
\frac{d v_k}{dt} = -\frac{b}{m} v_k - g.
\]

This is a first-order linear differential equation. Let us solve it using an integrating factor. Rewrite the equation as:

\[
\frac{d v_k}{dt} + \frac{b}{m} v_k = -g.
\]

The integrating factor is:

\[
\mu(t) = e^{\frac{b}{m}t}.
\]

Multiplying through by the integrating factor, we get:

\[
e^{\frac{b}{m}t} \frac{d v_k}{dt} + \frac{b}{m} e^{\frac{b}{m}t} v_k = -g e^{\frac{b}{m}t}.
\]

The left-hand side is the derivative of the product $e^{\frac{b}{m}t} v_k$, so:

\[
\frac{d}{dt} \left( e^{\frac{b}{m}t} v_k \right) = -g e^{\frac{b}{m}t}.
\]

Integrate both sides with respect to $t$:

\[
e^{\frac{b}{m}t} v_k = \int -g e^{\frac{b}{m}t} dt = -\frac{mg}{b} e^{\frac{b}{m}t} + C,
\]

where $C$ is the constant of integration. Dividing through by $e^{\frac{b}{m}t}$, we find:

\[
v_k = C e^{-\frac{b}{m}t} - \frac{mg}{b}.
\]

This is the solution for the vertical component of velocity. For the horizontal component, since no external force acts in the horizontal direction other than air resistance, the horizontal velocity satisfies:

\[
\frac{d v_x}{dt} = -\frac{b}{m} v_x.
\]

Solving this differential equation (similar to the method above) gives:

\[
v_x = A_x e^{-\frac{b}{m}t},
\]

where $A_x$ is a constant determined by initial conditions.

Thus, the velocity vector $\mathbf{v}$ can be written as:

\[
\mathbf{v} = \mathbf{A} e^{-\frac{b}{m}t} - \frac{mg}{b} \hat{k},
\]

where $\mathbf{A}$ is a constant vector determined by the initial conditions.

\textbf{b)} To find $\mathbf{A}$, we use the initial condition. At $t = 0$, the velocity of the ball is $\mathbf{v}(0) = \mathbf{u}$. Substituting $t = 0$ into the general solution:

\[
\mathbf{u} = \mathbf{A} e^{0} - \frac{mg}{b} \hat{k}.
\]

Since $e^{0} = 1$, we have:

\[
\mathbf{u} = \mathbf{A} - \frac{mg}{b} \hat{k}.
\]

Rearranging for $\mathbf{A}$:

\[
\mathbf{A} = \mathbf{u} + \frac{mg}{b} \hat{k}.
\]

Thus, the constant vector $\mathbf{A}$ is given by:

\[
\mathbf{A} = \mathbf{u} + \frac{mg}{b} \hat{k}.
\]

This completes the solution.

"

Solution 2: 

"
\textbf{a) Using Newton's laws of motion, show the equation and find the solution.}

The motion of the ball is governed by Newton's second law:

\[
\mathbf{F} = m \frac{d \mathbf{v}}{dt},
\]

where $\mathbf{F}$ is the net force acting on the ball. The forces acting on the ball are the gravitational force $-mg\hat{k}$ (acting downward) and the air resistance force $-b \mathbf{v}$ (opposite to the velocity of the ball). Therefore, the net force is

\[
\mathbf{F} = -mg\hat{k} - b\mathbf{v}.
\]

Substituting into Newton's second law:

\[
m \frac{d \mathbf{v}}{dt} = -mg\hat{k} - b\mathbf{v}.
\]

Dividing through by $m$:

\[
\frac{d \mathbf{v}}{dt} = -g\hat{k} - \frac{b}{m}\mathbf{v}.
\]

This is a first-order linear differential equation for $\mathbf{v}$. To solve it, we rewrite it as:

\[
\frac{d \mathbf{v}}{dt} + \frac{b}{m}\mathbf{v} = -g\hat{k}.
\]

The solution can be obtained using the method of integrating factors. The integrating factor is:

\[
\mu(t) = e^{\frac{b}{m}t}.
\]

Multiplying through by $\mu(t)$:

\[
e^{\frac{b}{m}t} \frac{d \mathbf{v}}{dt} + \frac{b}{m} e^{\frac{b}{m}t} \mathbf{v} = -g\hat{k} e^{\frac{b}{m}t}.
\]

The left-hand side is the derivative of $\mathbf{v} e^{\frac{b}{m}t}$:

\[
\frac{d}{dt} \left( \mathbf{v} e^{\frac{b}{m}t} \right) = -g\hat{k} e^{\frac{b}{m}t}.
\]

Integrating both sides with respect to $t$:

\[
\mathbf{v} e^{\frac{b}{m}t} = \int -g\hat{k} e^{\frac{b}{m}t} \, dt.
\]

The integral on the right-hand side evaluates to:

\[
\int -g\hat{k} e^{\frac{b}{m}t} \, dt = -\frac{mg}{b} e^{\frac{b}{m}t} + C,
\]

where $C$ is the constant of integration. Substituting back, we have:

\[
\mathbf{v} e^{\frac{b}{m}t} = -\frac{mg}{b} e^{\frac{b}{m}t} + C.
\]

Dividing through by $e^{\frac{b}{m}t}$:

\[
\mathbf{v} = C e^{-\frac{b}{m}t} - \frac{mg}{b} \hat{k}.
\]

Let $C = \mathbf{A}$, where $\mathbf{A}$ is a constant vector. Thus, the solution is:

\[
\mathbf{v} = \mathbf{A} e^{-\frac{b}{m}t} - \frac{mg}{b} \hat{k}.
\]

\textbf{b) Find an expression for $\mathbf{A}$.}

The initial condition is that the ball starts with velocity $\mathbf{u}$ at $t = 0$. Substituting $t = 0$ into the solution for $\mathbf{v}$:

\[
\mathbf{v}(0) = \mathbf{u}.
\]

From the solution:

\[
\mathbf{v}(0) = \mathbf{A} e^{0} - \frac{mg}{b} \hat{k}.
\]

Simplifying:

\[
\mathbf{u} = \mathbf{A} - \frac{mg}{b} \hat{k}.
\]

Solving for $\mathbf{A}$:

\[
\mathbf{A} = \mathbf{u} + \frac{mg}{b} \hat{k}.
\]

Thus, the expression for $\mathbf{A}$ is:

\[
\mathbf{A} = \mathbf{u} + \frac{mg}{b} \hat{k}.
\]

"

Solution 3:

"
\textbf{a)} Using Newton's second law of motion, the net force acting on the ball is given by:

\[
\mathbf{F}_{\text{net}} = m \frac{d\mathbf{v}}{dt}
\]

The forces acting on the ball are gravity, which exerts a force $-mg\hat{k}$ in the vertical direction, and air resistance, which exerts a force $-b\mathbf{v}$ in the direction opposite to the velocity. Thus, the net force equation becomes:

\[
m \frac{d\mathbf{v}}{dt} = -mg\hat{k} - b\mathbf{v}.
\]

Dividing through by $m$, we get:

\[
\frac{d\mathbf{v}}{dt} = -g\hat{k} - \frac{b}{m} \mathbf{v}.
\]

This is a first-order linear differential equation. To solve it, we use the method of separation of variables. First, rewrite the equation as:

\[
\frac{d\mathbf{v}}{dt} + \frac{b}{m} \mathbf{v} = -g\hat{k}.
\]

The solution to this type of equation involves finding a homogeneous solution and a particular solution.

1. \textbf{Homogeneous solution:} Set the right-hand side to zero:
\[
\frac{d\mathbf{v}_h}{dt} + \frac{b}{m} \mathbf{v}_h = 0.
\]

Separating variables, we have:
\[
\frac{d\mathbf{v}_h}{\mathbf{v}_h} = -\frac{b}{m} dt.
\]

Integrating both sides:
\[
\ln(\mathbf{v}_h) = -\frac{b}{m}t + C,
\]
where $C$ is a constant. Exponentiating both sides:
\[
\mathbf{v}_h = \mathbf{A} e^{-bt/m},
\]
where $\mathbf{A}$ is an integration constant vector.

2. \textbf{Particular solution:} Assume the particular solution $\mathbf{v}_p$ is constant, since the right-hand side $-g\hat{k}$ is constant. Substituting $\mathbf{v}_p$ into the differential equation:
\[
0 + \frac{b}{m} \mathbf{v}_p = -g\hat{k}.
\]

Solving for $\mathbf{v}_p$:
\[
\mathbf{v}_p = -\frac{mg}{b} \hat{k}.
\]

3. \textbf{General solution:} The total solution is the sum of the homogeneous and particular solutions:
\[
\mathbf{v} = \mathbf{v}_h + \mathbf{v}_p = \mathbf{A} e^{-bt/m} - \frac{mg}{b} \hat{k}.
\]

Thus, the velocity as a function of time is:
\[
\mathbf{v} = \mathbf{A} e^{-bt/m} - \frac{mg}{b} \hat{k}.
\]

\textbf{b)} To determine the constant vector $\mathbf{A}$, we use the initial condition. At $t = 0$, the velocity of the ball is $\mathbf{v}(0) = \mathbf{u}$. Substituting $t = 0$ into the general solution:

\[
\mathbf{u} = \mathbf{A} e^{0} - \frac{mg}{b} \hat{k}.
\]

Simplifying:
\[
\mathbf{u} = \mathbf{A} - \frac{mg}{b} \hat{k}.
\]

Solving for $\mathbf{A}$:
\[
\mathbf{A} = \mathbf{u} + \frac{mg}{b} \hat{k}.
\]

\textbf{Final Answer:}
\[
\mathbf{v} = \left( \mathbf{u} + \frac{mg}{b} \hat{k} \right) e^{-bt/m} - \frac{mg}{b} \hat{k}.
\]
"



















Its Over 9000!